### 2017-02-08 doc/manuscript.Rmd
Major manuscript updates.

---
title: Characterizing the Human Virome and Colorectal Cancer Link Using Machine Learning and Network Theory
author: 
- name: Geoffrey D Hannigan
  affiliation: 1
- name: Melissa B Duhaime
  affiliation: 2
- name: Mack T Ruffin IV
  affiliation: 3
- name: Charlie C Koumpouras
  affiliation: 1
- name: Patrick D Schloss
  affiliation: 1,\*
address:
  - code: 1
    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 2
    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48108
  - code: 3
    address: Department of Family Medicine, University of Michigan, Ann Arbor, Michigan, 48108
  - code: \*
    address: To whom correspondence may be addressed.
output:
  md_document:
    variant: markdown
---

&nbsp;

| ***Corresponding Author Information***
| Patrick D Schloss, PhD
| 1150 W Medical Center Dr. 1526 MSRB I
| Ann Arbor, Michigan 48109
| Phone: (734) 647-5801
| Email: pschloss@umich.edu

| ***Journal***: PNAS ([*Preparation Details*](http://www.pnas.org/site/authors/procedures.xhtml))
| ***Major Classification***: Biological Sciences
| ***Minor Classification***: Microbiology

***Keywords***: Colorectal Cancer, Virome, Machine Learning

***Text Length***: 34,182 / ~39,000 Characters

\* *Figures included for internal editing purposes*

\newpage

```{r loadfiles, echo=FALSE}
sigfig = 3
alphadf <- read.delim(file = "../rtables/diversity.tsv", header = TRUE, sep = "\t")
twoauc <- read.delim(file = "../rtables/twoclass-auc.tsv", header = TRUE, sep = "\t")
twoss <- read.delim(file = "../rtables/twoclass-auc-statsig.tsv", header = TRUE, sep = "\t")
threeauc <- read.delim(file = "../rtables/threeclass-auc.tsv", header = TRUE, sep = "\t")
threesig <- read.delim(file = "../rtables/threeclass-sig.tsv", header = TRUE, sep = "\t")
idcount <- read.delim(file = "../rtables/idcount.tsv", header = FALSE, sep = "\t")
phagecount <- read.delim(file = "../rtables/phagecount.tsv", header = FALSE, sep = "\t")
corstats <- read.delim(file = "../rtables/cor-stats.tsv", header = TRUE, sep = "\t")
```

# Abstract

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although the cause of colorectal cancer remains unclear, it has been strongly linked to gut bacterial communities. Viruses are another important component of the gut microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the gut virome (virus community) role in colorectal cancer using a cohort of 90 human subjects with either healthy, precancerous, or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic methods to compare the virome's role to that of the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting an association between the virome and colorectal cancer, but virome-based classification models were highly associated with colorectal cancer. Bacteriophages, not eukaryotic viruses, made up the majority of the CRC-associated virome, suggesting the community was indirectly linked to colorectal cancer, modulating bacterial community structure and functionality. Phages with broader host ranges had a more significant role in cancer development. These results provide foundational evidence that phage communities are associated with colorectal cancer, and that broadly tropic phages play more significant roles. Because of its importance, the virome should be considered in future colorectal cancer studies, as well as other cancer types.

**Word Count**: 229

# Significance Statement
Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. It's progression and severity is linked to gut bacterial communities. Little is known about the cancer-associated gut virus communities and their influence on bacteria. We characterize this influence of gut virus community structure on colorectal cancer in humans. The link between the gut virome and colorectal cancer was established by using virome-based machine learning models to accurately classify patient cancer status. The cancer-virus link was driven primarily by bacteriophages (bacterial viruses). Wider phage host ranges were more strongly linked to colorectal cancer. The results suggest an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.

**Word Count**: 119

\newpage

# Introduction
Cancer remains a devastating and persistent plague on humanity. Although cancer is still a primary cause of morbidity and mortality worldwide, we have made considerable therapeutic progress in recent decades. Perhaps one of our most impactful advances has not been in cancer treatment, but rather in detecting cancers at early stages, thereby improving treatment efficacy. This has been evident in a variety of prominent cancer types, with one of the most notable being colorectal cancer.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans will have died from the disease [@Siegel:2014jo]. Although it remains a major health problem, the impact of colorectal cancer has been reduced by improved screening and prevention efforts [@Zauber:2015cr; @Siegel:2014jo].

Development of colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Screening methods are effective, but their invasiveness has created a lack of compliance, creating a need for accurate, non-invasive screening methods. One such method is screening associated gut microbial communities.

Although the cause of colorectal cancer remains unclear, it has been strongly associated with gut bacterial communities [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. This association has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the gut microbiome. Viruses are another important component of the gut microbial community that have yet to be studied in the context of colorectal cancer.

Due to their mutagenic abilities and their propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Additionally, because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages are potentially indirectly linked to cancer. The gut virome (the virus community of the gut) has the potential to impact health and disease (e.g. cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. We aim to take this field of research further by beginning to assess the role of the virome in colorectal cancer.

Here we present a study of the colorectal cancer virome, highlighting the viruses most associated with the cancer state and the virome's utility for prognosis and diagnosis. We report that, like the association between the bacterial community and colorectal cancer was driven primarily by *Fusobacterium*, the association between the virome and colorectal cancer was driven by broadly infectious, bacteriophage hubs within the phage-bacteria ecosystem. By creating effective classification models using virus community signatures, we were able to accurately classify stool samples as cancerous, precancerous, or healthy. The implications of these findings are threefold. *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative, virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer. We expect this study will provide opportunities for continued research into the role of the virome in human cancer development.

# Results

## The Colorectal Cancer Virome Cohort
The study cohort consisted of 90 human subjects, 30 of which served as healthy controls, 30 of which had adenoma lesions consistent with a precancerous state, and 30 which had carcinoma lesions consistent with colorectal cancer **(Figure \ref{sampleproc})**. Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques. The 16S rRNA gene sequences were reported in a previous publication but re-analyzed here [@Zackular:2014fba]. The other half of the stool samples were purified for virus like particles (VLPs) and consequent genomic DNA extraction, followed by shotgun metagenomic sequencing. The VLP purification allowed us to observe the *active virome* because we only sequence those viruses that are encapsulated.

Virus DNA was purified by re-suspending the stool in buffer and removing contaminating cells (e.g. human, bacteria, etc) by filtration followed by contaminating cell lysis and degradation of the released genomic DNA **(Figure \ref{sampleproc})**. The resulting genomic DNA was used to prepare a shotgun metagenomic sequencing library that was sequenced on the Illumina HiSeq4000 platform. Each run was performed with a blank control to detect any contaminants from reagents. Only one of the nine viral controls detected DNA, which was of a minimal concentration, providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls were sparsely sequenced and were mostly removed while sub-sampling to even depths **(Figure \ref{qualcontrol} B)**.

The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) with the majority containing hundreds of related contigs **(Figure \ref{contigqc} - \ref{clustercontigqc})**. These operational units, which are conceptually similar to the operational taxonomic units (OTUs) used in 16S analysis, allow us to study bacterial and phage entities as highly related genomic entities.

## Diversity is Insufficient for Virome-Based Cancer Classification
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiosis"). We therefore initially evaluated the diversity of virome OGUs and their association with colorectal cancer. We utilized the Bray-Curtis dissimilarity metric to evaluate the differences in communities between disease states. To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced negative controls.

There was no statistically significant clustering of the disease groups, as visualized by NMDS ordination (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu} A**). An Anosim test with a post hoc multivariate Tukey test was used to calculate the statistical significance of the differences between the disease groups based on the variance around the cluster centroids **(Figure \ref{betaogu} B**). There were no significant differences between the disease groups, although there was a strongly significant difference between the blank controls (those few that remained after sub-sampling) and the rest of the study groups, further supporting the quality of our sample set (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**.

In addition to beta diversity, we also calculated the differences in virus alpha diversity associated with colorectal cancer. Again we found no significant alterations in either Shannon entropy or richness of the virus communities **(Figure \ref{betaogu} C-D)**. Overall, standard diversity metrics were insufficient for capturing the differences in the virus communities between disease states. This suggested to us that a more sophisticated approach for understanding the microbial community may be required, such as machine learning classification.

## Virome-Based Machine Learning Provides Accurate Cancer Classification
Previous work has shown that 16S rRNA community signatures are effective for classifying stool samples as originating from healthy, precancerous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja]. This is valuable because it presents a potential alternative screening approach to an invasive and expensive colonoscopy. The exceptional performance of bacterial signatures in these predictive models also supports a role for bacteria in colorectal cancer. Here we built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.

We built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals. These models were based on virus metagenomic community signatures or bacterial 16S rRNA gene signatures. We also included a whole metagenomic sequence set to ensure viral metagenomic observations were not a trait of metagenomics in general. Each operational units' relative abundance was used in the feature set. The same model approach was used for all three datasets, and the only difference was the data used to learn the model. We confirmed that our model using bacterial 16S data replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba].

We compared the bacterial 16S rRNA gene model to a model built using the virome signatures. The viral model performed significantly slightly better than the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively, after fifteen random forest iterations **(Figure \ref{predmodel} A - B)**. To confirm that this observation was due to the virome signature itself, and was not a trait of metagenomic datasets in general, we built a model using whole metagenomic community signatures. This model performed poorly with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`, lending support to our observation that classification performance is a trait unique to the virus metagenome **(Figure \ref{predmodel} A - B)**.

To evaluate the synergistic capabilities of the bacterial and viral signatures within the model, we built a combinatory model using both bacteria community and virome data. The combination model yielded modest but significantly improved performance beyond the virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This suggests the virus and bacterial communities may have synergistic capabilities for classifying stool as belonging to cancerous individuals.

The association between the two communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was iteratively removed. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with it's previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature was also driven primarily by a few operational genomic units, suggesting a role in cancer development **(Figure \ref{predmodel} D)**. The identified important viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified"). Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant viral dark matter (uncharacterized virus genomic space) associated with the human virome. This is common in the virome, and studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz].

When the bacterial and viral community signatures were combined, two bacterial and two viral organisms primarily drove the community association with cancer **(Figure \ref{predmodel} E)**. The most important microbes were two unidentified viruses, followed by *Bacteroides* and *Fusobacterium*.

## The Gut Virome Classifies Cancerous, Pre-Cancerous, and Healthy States
After evaluating our ability to classify samples as cancerous or healthy, we incorporated the precancerous adenoma samples into the model and evaluated our ability to classify all three states in our total sample set. We used three-class random forest models for the bacterial 16S and viral sample sets. The bacterial signature model yielded an AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded an overall AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). Both models were best able to classify cancer samples from healthy or precancerous samples, but struggled to distinguish precancerous from healthy or cancerous **(Figure \ref{threewaymodel} A-B)**. The cancerous signal was the most discriminatory of the three sample types.

Many of the microbes important for the two-class (cancer vs healthy) bacteria and virus models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium was the same *Fusobacterium* between the two and three class models, supporting its significance to the association between cancer and the bacterial communities **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. Unlike the two-class virome model, the viruses most important to the three-class model were identified bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**. A *Podoviridae* was most important, followed by two *Siphoviridae* phages.

The classification model determined cancer state by incorporating the relative abundance profiles of the microbes within each community. The signatures ranged from notably high abundance of some OGUs, low abundance of some OGUs, and an absence of other OGUs **(Figure \ref{threewaymodel} F)**. Not all important OGUs were of increased abundance. The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.

## Bacteriophages Drive Link Between Virome and Colorectal Cancer
The virome-based model was able to accurately classify stool samples as cancerous, precancerous, or healthy. Not only is this important for establishing an alternative diagnostic model, but it also suggests an underlying biological importance for viruses in colorectal cancer. Above we used our classification models to evaluate which virus OGUs were most highly linked to colorectal cancer. We were able to further characterize these important OGUs to better understand the underlying biological link between the virome and colorectal cancer.

The role of the virome in colorectal cancer could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that are important for colorectal cancer, we utilized the longest sequences from the OGUs as the representative sequences to be used for taxonomic classification. These sequences were aligned to a set of all reference virus genomes, including bacteriophages and eukaryotic viruses. A strict e-value threshold of 1e-25 was used to improve our confidence in the matches between the genome sequences. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel} E)**. Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes. Thus the majority of the OGUs are bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities. This is consistent with previous reports suggesting the gut virome is primarily phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb].

## Broadly Infectious Phages Play Greater Role in Colorectal Cancer
Reference-based analyses of the virome perform poorly due to sparse reference databases and genomic modularity. Much of the virome is described as genomic "dark matter" and taxonomic identities can be largely uninformative (e.g. *Siphoviridae* describes the phage morphology under an electron microscope). Sometimes these identities are also used to infer bacterial hosts. Instead of relying on genome alignments to infer the identities and host ranges of the phage OGUs, we employed a network-based technique to understand the roles of the OGUs in the greater microbiome context, as previously described (cite other network preprint here). We implemented a random forest model to predict which bacterial OGUs are infected by which phage OGUs, resulting in an ecological network of the bacteria and phages within the community **(Figure \ref{network} A)**. We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model. We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), indicating that phage network hubs may play more important roles in colorectal cancer **(Figure \ref{network} B)**.

# Discussion
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because some bacteria also have oncogenic properties, bacteriophages may play an indirect role in promoting carcinogenesis by altering bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential, the link between virus gut communities (the human gut virome) and cancer has yet to be evaluated. Here we show that, like gut bacterial communities, the gut virome is associated with colorectal cancer. Our findings support a model for oncogenesis by phage-modulated bacterial community composition.

Previous work has supported a causative role for bacterial community composition and colorectal cancer progression, as well as an indirect role for those antibiotic agents which alter that composition [@Zackular:2016en]. Zackular *et al* showed that administering antibiotics (*A*) and thereby killing subsets of bacterial communities (*B*) was sufficient for altering colorectal cancer progression (*C*) in an inflammation-based murine model for colorectal cancer. These findings established a model of cancer dependence on altered bacterial community composition, which in turn relied on antibiotic administration.

$$ A\rightarrow B\rightarrow C $$

In addition to bacterial communities, our findings highlight a strong link between gut virus communities and colorectal cancer. The signal is dominated by bacteriophages, suggesting the virome may indirectly promote cancer by modulating bacterial community composition within the gut. These results have led us to suggest an alternative model for microbiome-driven colorectal cancer that depends on bacteriophages (*P*), instead of antibiotics, as community modulators.

$$ P\rightarrow B\rightarrow C $$

Our work supports a model in which bacteriophage communities play a role in colorectal cancer progression by altering bacterial community composition, which is already known to impact colorectal cancer. Alpha and beta diversity of the phage communities were not different between cancer states, but a more sophisticated random forest approach suggested that abundance relationships between phage taxa are highly linked to cancer status. Some select viral genomic units (phages) stood out as being highly important to the cancer-phage link, and this importance was correlated with centrality in the bacteria-phage ecological network. This suggests a model in which the phages most associated with colorectal cancer are also hubs within the bacteria-phage ecological network, thus allowing their abundance changes at the center of the interaction network to ripple throughout the rest of the community and promote the cancer-related ecosystem **(Figure \ref{modelsummary})**.

When interpreting these results, it is worth noting that we are studying these communities as operational units, making us unable to draw conclusions with a high taxonomic resolution. This is especially true for bacteriophages, which are poorly annotated and have taxonomy of limited utility. Bacterial operational taxonomic units often represent relatively large categories of bacterial taxa, such as genera, that can represent a variety of species and strains. Likewise, metagenomic operational genomic units represent operationally informative classifications for understanding the community, but should not be interpreted as single phages or bacterial strains.

Further work will be required to support the potential causal relationships outlined in this study. Although we provide evidence for gut bacteriophage communities to alter bacterial communities toward a cancerous state, we must follow up with experimental validation of these observations. The clear pathway toward these answers will be the utilization of murine models as described by Zackular *et al* which showed the link between antibiotic administration and colorectal cancer development [@Zackular:2016en].

In addition to the therapeutic ramifications for understanding the colorectal cancer microbiome, our findings provide a proof-of-principle that viruses, while under-appreciated and understudied in the human microbiome, are an important contributer to human disease that has the potential to provide an abundance of information that supplements that of bacterial communities. Evidence has suggested that the virome is a crucial component to the microbiome and that bacteriophages are important players. Bacteriophage and bacterial communities cannot thrive without each other [@RodriguezValera:2009cr]. Not only is the human virome an important part of human health and disease, but it appears to have a particular significance in cancer research.

# Methods

## Analysis Source Code & Availability
All associated source code and Makefile are available for review at the following GitHub repository: https://github.com/SchlossLab/Hannigan-2016-ColonCancerVirome.

## Study Design and Patient Sampling
This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent. Design and sampling of this sample set have been reported previously [@Zackular:2014fba]. Briefly, whole evacuated stool was collected from patients who were 18 years of age or older, able to provide informed consent, have had colonoscopy and histologically confirmed colonic disease status, had not had surgery, had not had chemotherapy or radiation, and were free of known co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and inflammatory bowel disease. Samples were collected from four locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor (Michigan, USA). Ninety patients were recruited to the study, thirty of which were designated healthy, thirty with detected adenomas, and thirty with detected carcinomas.

## 16S Data Acquisition & Processing
The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba]. Sequence (fastq) and metadata files were downloaded from http://www.mothur.org/MicrobiomeBiomarkerCRC. The 16S rRNA gene sequences were analyzed as described previously, relying on the Mothur analytical toolkit (v1.37.0) [@Sze:2016iw; @Schloss:2009do]. Briefly, the sequences were de-replicated, screened for chimeras using UCHIME [@Edgar:2011gy] and the SILVA database [@Pruesse:2007jc], and binned into operational taxonomic units (OTUS) using a 97% similarity threshold. Abundance was normalized for uneven sequencing depth by randomly sub-sampling to 10,000 sequences, as previously reported [@Baxter:2016dja].

## Whole Metagenomic Library Preparation & Sequencing
DNA was extracted from stool samples using the PowerSoil-htp 96 Well Soil DNA Isolation Kit (Mo Bio Laboratories) using an EPMotion 5075 pipetting system. Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT library preparation kit according to the standard kit protocol. The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution. The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.

## Virus Metagenomic Library Preparation & Sequencing
Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz]. Briefly, an aliquot of stool (~0.1g) was resuspended in SM buffer and vortexed to facilitate resuspension. The resuspended stool was centrifuged to remove major particulate debris, followed by filtering through a 0.22µm filter to remove smaller contaminants. The filtered supernatant was treated with chloroform to lyse contaminating cells including bacteria, human, fungi, etc. The exposed genomic DNA from the lysed cells was degraded by treating the samples with DNase. The DNA was extracted from the purified VLPs using the Wizard PCR Purification Preparation Kit (Promega). Disease classes were staggered across purification runs to prevent run variation as a confounding factor. Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT library preparation kit according to the standard kit protocol. The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution. The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz]. The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.

## Metagenome Quality Control
Both the viral and whole metagenomic sample sets were subjected to the same quality control procedures. The sequences were obtained as de-multiplexed fastq files from the HiSeq platform and subjected to 5' and 3' adapter trimming using the CutAdapt program (v1.9.1) with an error rate of 0.1 and an overlap of 10 [@Martin:2011eu]. The FastX toolkit (v0.0.14) was used to quality trim the reads to a minimum length of 75bp and a minimum quality score of 30 [@FASTXToolkit:wr]. Reads mapping to the human genome were removed using the DeconSeq algorithm (v0.4.3) and default parameters [@Schmieder:2011fo].

## Contig Assembly & Abundance
Contigs were assembled using paired end read files that were purged of sequences without a corresponding pair (e.g. One read removed due to low quality). The Megahit program (v1.0.6) was used to assemble contigs for each sample using a minimum contig length of 1000 bp and iterating assemblies from 21-mers to 101-mers by 20 [@Li:2016kd]. Contigs from the virus and whole metagenomic sample sets were concatenated within their respective groups. Abundance of the contigs within each sample was calculated by aligning sequences back to the concatenated contig files using the bowtie2 global aligner (v2.2.1), with a 25 bp seed length and an allowance of one mismatch [@Langmead:2012jh]. Abundance was corrected for contig reference length and the number of contigs included in each operational genomic unit. Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1e6 and 5e5 reads, respectively, and removing samples with less total samples than the threshold. Thresholds were set for maximizing sequence information while minimizing numbers of lost samples.

## Operational Genomic Unit Classification
Much like operational taxonomic units (OGUs) are used as an operational definition of similar 16S rRNA gene sequences in absence of taxonomic identification, we operationally defined closely related contig sequences as operational genomic units (OGUs) in the absence of taxonomic identity. OGUs were defined with the CONCOCT algorithm (v0.4.0) which bins related contigs by similar tetra-mer and co-abundance profiles within samples using a variational Bayesian approach [@Alneberg:2014fc]. CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria due to computational limitations.

## Diversity
Alpha and beta diversity were calculated using the operational genomic unit abundance profiles for each sample. Sequences were sub-sampled down to 100,000 sequences. Samples with less than the cutoff were removed from the analysis. Alpha diversity was calculated using the Shannon Entropy and Richness metrics. Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (Anosim) with a post-hoc multivariate Tukey test. All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE.

## Classification Modeling
Classification modeling was performed in R using the Caret package [@caretClassificatio:_U2Lit_1]. OTU and OGU abundance data was preprocessed by removing features (OTUs and OGUs) that were present in less than half of the samples. This served both as an effective feature reduction technique and made the calculations computationally feasible. The binary random forest model was trained using the Area Under the ROC Curve (AUC) and the three-class random forest model was trained using the mean AUC. Both were validated using five-fold cross validation. Each training set was repeated five times, and the model was tuned across five iterations of mtry values. For consistency and accurate comparison between feature groups (e.g. bacteria, virus), the sample model parameters were used for each group. The maximum AUC during training was recorded across 10 iterations of each group model creation to test the significance of the differences between feature set performance. Statistical significance was evaluated using a Wilcoxon test between two categories, or a pairwise Wilcoxon test with Bonferroni corrected p-values when comparing more than two categories.

## Taxonomic Identification of Operational Genomic Units
Viral operational genomic units (OGUs) were identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives. The longest contiguous sequence in each operational genomic unit was used as a representative sequence for classification. Each representative sequence was aligned to the reference genome database using the tblastx alignment algorithm (v2.2.27) and a strict similarity threshold (e-value < 1e-25) [@Camacho:2009fc]. Annotation was interpreted as phage, eukaryotic virus, or unknown.

## Ecological Network Analysis & Correlations
The ecological network of the bacterial and phage operational genomic units were constructed and analyzed as previously described (cite network preprint here). Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1). The degree of phage centrality was quantified using the alpha centrality metric in the igraph CRAN package. A Spearman correlation was performed between model importance and phage centrality scores.

# Conflicts of Interest
The authors declare no conflicts of interest.

# Acknowledgments
We thank the members of the Schloss lab for their underlying contributions. GD Hannigan was supported in part by the Michigan Molecular Mechanisms of Microbial Pathogenesis Fellowship.

\newpage

# Figures

![*Cohort and sample processing outline. Thirty subject stool samples were collected from healthy, adenoma (pre-cancer), and carcinoma (cancer) patients. Stool samples were split into two aliquots, the first of which was used for bacterial sequencing and the second which was used for virus sequencing. Bacterial sequencing was done using both 16S rRNA amplicon and whole metagenomic shotgun sequencing techniques. Virus samples were purified for viruses using filtration and a combination of chloroform (bacterial lysis) and DNase (exposed genomic DNA degradation). The resulting encapsulated virus DNA was sequenced using whole metagenomic shotgun sequencing.* \label{sampleproc}](../figures/sample-processing.pdf){ width=100% }

\newpage

![*Diversity calculations comparing cancer states of the colorectal virome, based on relative abundance of operational genomic units in each sample. A) NMDS ordination of community samples, colored for cancerous (green), pre-cancerous (red), and healthy (yellow). B) Differences in means between disease group centroids with 95% confidence intervals based on an Anosim test with a post hoc multivariate Tukey test. Comparisons (indicated on y-axis) in which the intervals cross the zero mean difference line (dashed line) were not significantly different. C) Shannon diversity and D) richness alpha diversity quantification comparing pre-cancerous (grey), cancerous (red), and healthy (tan) states.*\label{betaogu}](../figures/diversity-beta-ogu.pdf){ width=100% }

\newpage

![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S signatures, whole metagenomic signatures, and a combination of virome and 16S signatures. A) ROC curve for visualizing the performance of each of the models for classifying stool as coming from either a cancerous or healthy individual. B) Quantification of the AUC variation for each model, and how it compares to each of the other models based on 15 iterations. A pairwise Wilcoxon test with a Bonferroni multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01). C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S classification model when removed from the classification model. Results based on 25 iterations. OTU features are colored by taxonomic identity. D) Mean decrease in accuracy of each operational genomic unit in the virome classification model. E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S and virome features.*\label{predmodel}](../figures/predmodel-viromebacteria.pdf){ width=100% }

\newpage

![*ROC curves from A) virome and B) bacterial 16S three-class random forest models tuned on mean AUC. Each curve represents the ability of the specified class to be classified against the other two classes. C) Quantification of the mean AUC variation for each model based on 10 model iterations. A pairwise Wilcoxon test with a Bonferroni multiple hypothesis correction demonstrated that the models are significantly different (alpha = 0.01). D) Mean decrease in accuracy when virome operational genomic units and E) bacterial 16S OTUs are removed from the respective three-class classification models. Results based on 25 iterations. F) Relative abundance of the six most important virome OGUs in the model, with the most important on the right. Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=100% }

\newpage

![*Community network analysis utilizing predicted interactions between bacteria and phage operational genomic units. A) Visualization of the community network for our colorectal cancer cohort. B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OGU. A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}](../figures/NetworkAndScatter.pdf){ width=80% }

\newpage

![*Example visualization of the proposed model for bacteriophage modulation of colorectal cancer. A) Example network diagrams illustrating the impact of changes on keystone phages (\* asterisk) within the microbiome. A hypothetical ecological network may being at a healthy equilibrium but an alteration to the abundance of a keystone phage can impact the remainder of the network. Highly connected phage host abundances may be altered by the original disruption, which would ripple out to the remainder of the model. B) Another visualization of the impact a change in keystone phage abundance could have on the remainder of the network.*\label{modelsummary}](../figures/ExampleHypothesis.pdf){ width=75% }

\newpage

# Supplemental Figures

\beginsupplement

![*Basic Quality Control Metrics. A) VLP genomic DNA yield from all sequenced samples. Each bar represents a sample which is grouped and colored by its associated disease group. B) Sequence yield following quality control including quality score filtering and human decontamination.*\label{qualcontrol}](../figures/qualitycontrol.pdf){ width=100% }

\newpage

![*Length and coverage statistics. A) Heated scatter plot demonstrating the distribution of contig coverage (number of sequences mapping to each contig) and contig length for the virus metagenomic sample set. B) Scatter plot illustrating the distribution of operational genomic unit (OGU) length and sequence coverage for the virus metagenomic sample set. C) Heated scatter plot demonstrating the distribution of contig coverage and length for the whole metagenomic sample set. D) Scatter plot illustrating the distribution of operational genomic unit (OGU) length and sequence coverage for the whole metagenomic sample set.*\label{contigqc}](../figures/qc-contiglengthcoverage.pdf){ width=75% }

\newpage

![*Operational genomic unit composition stats. A) Strip chart demonstrating the length and frequency of contigs within each operational genomic unit of the virome sample set. The y-axis is the operational genomic unit identifier, and x-axis is the length of each contig, and each dot represents a contig found within the specified operational genomic unit. B) Density plot (analogous to histogram) of the number of virome operational genomic units containing the specific number of contigs, as indicated by the x-axis. C-D) Sample plots as panels C and D, but for the whole metagenomic sample set.*\label{clustercontigqc}](../figures/contiglengthbycluster.pdf){ width=75% }

\newpage

![*Beta-diversity comparing disease states and the study negative controls. A) NMDS ordination of community samples, colored by disease state. B) Differences in means between disease group centroids with 95% confidence intervals based on an Anosim test with a post hoc multivariate Tukey test. Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}](../figures/diversity-beta-ogu-negative.pdf){ width=100% }

\newpage

# References

### 2017-02-24 doc/manuscript.Rmd
General document push. In this case, more data figures.

@@
 ```
+
+\newpage
 
@@
 
+## Phage Replication Style Identification
+Phage OGU replication sytle was identified using methods described previously [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
+Briefly, we identified lysogenic phage OGUs as representative contigs containing at least one of three genomic markers: 1) phage integrase genes, 2) prophage genes from the ACLAME database, 3) genomic similarity to bacterial reference genomes.
+Integrase genes were identified in phage OGU representative contigs by aligning the contigs to a reference database of all known phage integrase genes from the Uniprot database (Uniprot search term: "organism:phage gene:int NOT putative").
+Prophage genes were identified in the same way, using the ACLAME set of reference prophage genes.
+In both cases, the blastx algorithm was used with an e-value of 10e-5.
+Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes.
+To accomplish this, representative phage contigs were aligned to the European Nucleotide Archive bacterial genome reference set using the blastn algorithm (e-value < 10e-25).
+
 # Conflicts of Interest
@@
 The resulting encapsulated virus DNA was sequenced using whole metagenomic shotgun sequencing.* \label{sampleproc}](../figures/sample-processing.pdf){ width=100% }
-
-\newpage
-
-![*Diversity calculations comparing cancer states of the colorectal virome, based on relative abundance of operational genomic units in each sample.
-A) NMDS ordination of community samples, colored for cancerous (green), pre-cancerous (red), and healthy (yellow).
-B) Differences in means between disease group centroids with 95% confidence intervals based on an Anosim test with a post hoc multivariate Tukey test.
-Comparisons (indicated on y-axis) in which the intervals cross the zero mean difference line (dashed line) were not significantly different.
-C) Shannon diversity and D) richness alpha diversity quantification comparing pre-cancerous (grey), cancerous (red), and healthy (tan) states.*\label{betaogu}](../figures/diversity-beta-ogu.pdf){ width=100% }
 
@@
 
+![*Diversity calculations comparing cancer states of the colorectal virome, based on relative abundance of operational genomic units in each sample.
+A) NMDS ordination of community samples, colored for cancerous (green), pre-cancerous (red), and healthy (yellow).
+B) Differences in means between disease group centroids with 95% confidence intervals based on an Anosim test with a post hoc multivariate Tukey test.
+Comparisons (indicated on y-axis) in which the intervals cross the zero mean difference line (dashed line) were not significantly different.
+C) Shannon diversity and D) richness alpha diversity quantification comparing pre-cancerous (grey), cancerous (red), and healthy (tan) states.*\label{betaogu}](../figures/diversity-beta-ogu.pdf){ width=100% }
+
+\newpage
+
 ![*Beta-diversity comparing disease states and the study negative controls.
@@
 
+![*Lysogenic phage relative abundance in disease states.
+Phage OGUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
+No disease groups were statistically significant.*\label{betaogunegative}](../figures/lysogenic-relabund.pdf){ width=50% }
+
+\newpage
+
+![*Relative abundance correlations between bacterial OTUs and virome OGUs.
+A) Pearson correlation R values between all bacterial OTUs (x-axis) and viral OGUs (y-axis) with blue being positively correlated and red being negatively correlated.
+Operational units are organized by importance in their colorectal cancer classification models, such that the most important units are in the top left corner.
+B) Magnification of the boxed region in pannel (A), highlighting the correlation between the most important bacterial OTUs and virome OGUs.
+The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as pannel (A).
+C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OGUs.*\label{betaogunegative}](../figures/phage-bacteria-cor.pdf){ width=100% }
+
+\newpage
+
 # References

### 2017-03-02 doc/manuscript.Rmd
Have not pushed in a while. Lots of changes, especially to the discussion.

@@
 ---
-title: Characterizing the Human Virome and Colorectal Cancer Link Using Machine Learning and Network Theory
+title: Characterizing the Role of the Human Virome in Colorectal Cancer
 author: 
@@
   - code: 3
-    address: Department of Family Medicine, University of Michigan, Ann Arbor, Michigan, 48108
+    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
   - code: \*
@@
 Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world.
-Although the cause of colorectal cancer remains unclear, it has been strongly linked to gut bacterial communities.
-Viruses are another important component of the gut microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential.
-We evaluated the gut virome (virus community) role in colorectal cancer using a cohort of 90 human subjects with either healthy, precancerous, or cancerous colons.
+Although the cause of colorectal cancer remains unclear, it has been strongly linked to colon bacterial communities.
+Viruses are another important component of the colon microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential.
+We evaluated the colon virome (virus community) role in colorectal cancer using a cohort of 90 human subjects with either healthy, precancerous, or cancerous colons.
 We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic methods to compare the virome's role to that of the bacterial community.
@@
 Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide.
-It's progression and severity is linked to gut bacterial communities.
-Little is known about the cancer-associated gut virus communities and their influence on bacteria.
-We characterize this influence of gut virus community structure on colorectal cancer in humans.
-The link between the gut virome and colorectal cancer was established by using virome-based machine learning models to accurately classify patient cancer status.
+It's progression and severity is linked to colon bacterial communities.
+Little is known about the cancer-associated colon virus communities and their influence on bacteria.
+We characterize this influence of colon virus community structure on colorectal cancer in humans.
+The link between the colon virome and colorectal cancer was established by using virome-based machine learning models to accurately classify patient cancer status.
 The cancer-virus link was driven primarily by bacteriophages (bacterial viruses).
@@
 Screening methods are effective, but their invasiveness has created a lack of compliance, creating a need for accurate, non-invasive screening methods.
-One such method is screening associated gut microbial communities.
+One such method is screening associated colon microbial communities.
 
-Although the cause of colorectal cancer remains unclear, it has been strongly associated with gut bacterial communities [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
+Although the cause of colorectal cancer remains unclear, it has been strongly associated with colon bacterial communities [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
 This association has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja].
-While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the gut microbiome.
-Viruses are another important component of the gut microbial community that have yet to be studied in the context of colorectal cancer.
+While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome.
+Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.
 
@@
 Additionally, because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages are potentially indirectly linked to cancer.
-The gut virome (the virus community of the gut) has the potential to impact health and disease (e.g.
+The colon virome (the virus community of the colon) has the potential to impact health and disease (e.g.
 cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
@@
 
-## Virome-Based Machine Learning Provides Accurate Cancer Classification
+## Machine Learning Links the Virome to Colorectal Cancer
 Previous work has shown that 16S rRNA community signatures are effective for classifying stool samples as originating from healthy, precancerous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja].
 This is valuable because it presents a potential alternative screening approach to an invasive and expensive colonoscopy.
-The exceptional performance of bacterial signatures in these predictive models also supports a role for bacteria in colorectal cancer.
+The exceptional performance of bacteria in these classification models also supports a role for bacteria in colorectal cancer.
 Here we built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.
@@
 We compared the bacterial 16S rRNA gene model to a model built using the virome signatures.
-The viral model performed significantly slightly better than the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively, after fifteen random forest iterations **(Figure \ref{predmodel} A - B)**.
-To confirm that this observation was due to the virome signature itself, and was not a trait of metagenomic datasets in general, we built a model using whole metagenomic community signatures.
+The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively, after fifteen random forest iterations **(Figure \ref{predmodel} A - B)**.
+
+To confirm that this observation was due to the virome community itself, and was not a trait of metagenomic datasets in general, we built a model using whole metagenomic community signatures.
 This model performed poorly with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`, lending support to our observation that classification performance is a trait unique to the virus metagenome **(Figure \ref{predmodel} A - B)**.
+Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs.
+Further filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**.
+The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.
 
@@
 
-## The Gut Virome Classifies Cancerous, Pre-Cancerous, and Healthy States
+## The colon Virome Allows for Cancerous, Adenomatous, and Healthy Classification
 After evaluating our ability to classify samples as cancerous or healthy, we incorporated the precancerous adenoma samples into the model and evaluated our ability to classify all three states in our total sample set.
@@
 
-## Bacteriophages Drive Link Between Virome and Colorectal Cancer
+## Temperate Bacteriophages Drive Link Between Virome and Colorectal Cancer
 The virome-based model was able to accurately classify stool samples as cancerous, precancerous, or healthy.
@@
 Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes.
-Thus the majority of the OGUs are bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities.
-This is consistent with previous reports suggesting the gut virome is primarily phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb].
 
-## Broadly Infectious Phages Play Greater Role in Colorectal Cancer
-Reference-based analyses of the virome perform poorly due to sparse reference databases and genomic modularity.
-Much of the virome is described as genomic "dark matter" and taxonomic identities can be largely uninformative (e.g.
-*Siphoviridae* describes the phage morphology under an electron microscope).
-Sometimes these identities are also used to infer bacterial hosts.
-Instead of relying on genome alignments to infer the identities and host ranges of the phage OGUs, we employed a network-based technique to understand the roles of the OGUs in the greater microbiome context, as previously described (cite other network preprint here).
-We implemented a random forest model to predict which bacterial OGUs are infected by which phage OGUs, resulting in an ecological network of the bacteria and phages within the community **(Figure \ref{network} A)**.
+We evaluated whether the phages in the community were primarily lytic (lyse their hosts) or temperate (lysogenic; can integrate into their host's genome, as well as lyse the cell).
+We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes.
+This approach was done as described in previous work [@Minot:2011ez; @Hannigan:2015fz].
+We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{threewaymodel} E)**.
+Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes.
+These findings are consistent with previous reports suggesting the colon virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
+
+## Influential Bacteria and Phages Associate Indirectly
+Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were predators of the influential bacteria, and thus influenced their relative abundance through predation (e.g.
+One of the important phages targets *Fusobacterium*).
+If this hypothesis was true, we would expect that the relative abundance between many important bacteria and phages to be correlated.
+We correlated the relative abundance values of every phage OGU and bacterial OTU and found strikingly low correlation between the bacterial and phage abundances **(Figure \ref{correlations} A,C)**.
+There was an overall absence of correlations between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**.
+This evidence supported our null hypothesis that the influential phages are not predators of the influential bacterial.
+
+We further investigated the infectious capabilities and potential host ranges of the influential phages by building a machine learning model to predict which phage OGUs infect which bacteria in the overall community.
+The predicted interactions were then used to build a network which was used for subsequent analysis.
+This approach was done as has been previously described (**Cite Network Preprint**).
+This analysis revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**.
 We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model.
-We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), indicating that phage network hubs may play more important roles in colorectal cancer **(Figure \ref{network} B)**.
+We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), indicating that phages important in driving colorectal cancer are also community hubs **(Figure \ref{network} B)**.
+
+## Influential Phage Identities Shift During CRC Progression
+Because our cohort sampled healthy, adenomatous, and cancerous colons, we are able to gain initial insight into virus community shifts during cancer progression.
+We evaluated the community changes at the two disease transitions (healthy to adenomatous and adenomatous to cancerous) by building individual random forest models to compare only those sample classes.
+We found that, while bacterial 16S rRNA models perform equally well for all disease class comparisons, the virome models performed worst when seperating healthy from adenomatous, better when seperating adenomatous from cancerous, and the best when seperating healthy from cancerous **(Figure \ref{transitionmodels} A-B)**.
+Like bacterial communities, the most important phage OGUs provide insight into which organisms are driving the transitions between disease states.
+Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members drove the transitions from healthy to adenomatous and adenomatous to cancerous, with one phage exception **(Figure \ref{transitionmodels} C-E)**.
+A *Myoviridae* OGU (Cluster 188) was the seventh most important phage driving the transition from a healthy to adenomatous state, and was the most important phage driving the progression from an adenomatous to carcingenic state.
+Therefore, like bacteria, there are distinct phages assocaited with the transitions from a healthy to an adenomatous colon, and from an adenomatous to a cancerous colon.
 
@@
 Because some bacteria also have oncogenic properties, bacteriophages may play an indirect role in promoting carcinogenesis by altering bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
-Despite their carcinogenic potential, the link between virus gut communities (the human gut virome) and cancer has yet to be evaluated.
-Here we show that, like gut bacterial communities, the gut virome is associated with colorectal cancer.
-Our findings support a model for oncogenesis by phage-modulated bacterial community composition.
+Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colon communities (the human colon virome) and colorectal cancer has yet to be evaluated.
+Here we show that, like colon bacterial communities, the colon virome is associated with colorectal cancer.
+Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.
 
-Previous work has supported a causative role for bacterial community composition and colorectal cancer progression, as well as an indirect role for those antibiotic agents which alter that composition [@Zackular:2016en].
-Zackular *et al* showed that administering antibiotics (*A*) and thereby killing subsets of bacterial communities (*B*) was sufficient for altering colorectal cancer progression (*C*) in an inflammation-based murine model for colorectal cancer.
-These findings established a model of cancer dependence on altered bacterial community composition, which in turn relied on antibiotic administration.
+We found that the basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states.
+This led us to implement a more sophisticated machine learning approach (random forest classificaiton model) that allowed us to detect strong associations between the colon virus communities and colorectal cancer.
+Unlike other human cancer types that are often assocaited with human viruses, colorectal cancer is primarily assocaited with altered bacteriophage communities.
+These phage communities were not predators of the most influential bacteria (e.g.
+*Fusobacterium* linked to decreases in *Fusobacterium* phages), as demonstrated by the lack of correlation between the abundance of these entities.
+Additionally, we identified a correlation in which those phages strongly linked to colorectal cancer were also identified as influential hubs within the overall phage-bacteria network, linking the influence of a phage on the community as a whole to its importance to cancer progression.
+Our previous work has shown that modifying colon bacterial communities with antibiotics alters colorectal cancer progression and tumor burden in mice, and suggests phage-mediated community alterations may have a similar effect, especially if those phages were broadly infectious [@Zackular:2016en].
+Together this data supports a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
 
-$$ A\rightarrow B\rightarrow C $$
+To better understand the potential mechanisms by which phages modulate the colon bacterial community, we built a working hypothesis that incorporated our findings into the Flynn model for colorectal cancer developement **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
+We hypothesized that the broadly infectious phages in the colon began lysing, and thereby disrupting, the bacterial communities to open a niche in which opportunistic bacteria were able to colonize, such as *Fusobacterium nucleatum*.
+Once the "driver" bacteria had established themselves in the epithelium, other opporunistic "passenger" bacteria were able to adhere to the driver, colonize, and begin establishing a biofilm.
+We feel it is likely that the phages played a role in biofilm dispersal and growth, by lysing bacteria within the biofilm, a process shown to be important for effective biofilm growth.
+The oncogenic bacteria were then able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response.
+As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria important to our cancer classification model.
+As the bacteria entered their oncogenic synergy with the epithelium, we hypothesize the phages could continue mediating biofilm dispersal, as well as support the colonized oncogenic bacteria by lysing competing cells to maintain the niche and provide nutrients to the other bacteria.
+In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide the future, mostly functional studies, by our group and others, into the role the virome plays in colorectal cancer.
 
-In addition to bacterial communities, our findings highlight a strong link between gut virus communities and colorectal cancer.
-The signal is dominated by bacteriophages, suggesting the virome may indirectly promote cancer by modulating bacterial community composition within the gut.
-These results have led us to suggest an alternative model for microbiome-driven colorectal cancer that depends on bacteriophages (*P*), instead of antibiotics, as community modulators.
+Although we have evidence for the interaction between the bacteriophage and bacterial community, we are not able to address whether the phages are directly interacting with the human host **(Figure \ref{modelsummary} A)**.
+Like most foreign entities, bacteriophages are known to be capable of acting as human antigens and thereby initiate or contribute to an immune response.
+In the context of colorectal cancer, some phages may have been induced out of their bacterial host (most phages were temperate) and interacted with the human host to promote the cancer-associated inflammatory response.
+Because we are not able to address this behavior with our dataset, we noted it as a known unknown, and another plausible mechanism by which some phages could be associated with colorectal cancer developement.
 
-$$ P\rightarrow B\rightarrow C $$
-
-Our work supports a model in which bacteriophage communities play a role in colorectal cancer progression by altering bacterial community composition, which is already known to impact colorectal cancer.
-Alpha and beta diversity of the phage communities were not different between cancer states, but a more sophisticated random forest approach suggested that abundance relationships between phage taxa are highly linked to cancer status.
-Some select viral genomic units (phages) stood out as being highly important to the cancer-phage link, and this importance was correlated with centrality in the bacteria-phage ecological network.
-This suggests a model in which the phages most associated with colorectal cancer are also hubs within the bacteria-phage ecological network, thus allowing their abundance changes at the center of the interaction network to ripple throughout the rest of the community and promote the cancer-related ecosystem **(Figure \ref{modelsummary})**.
-
-When interpreting these results, it is worth noting that we are studying these communities as operational units, making us unable to draw conclusions with a high taxonomic resolution.
-This is especially true for bacteriophages, which are poorly annotated and have taxonomy of limited utility.
-Bacterial operational taxonomic units often represent relatively large categories of bacterial taxa, such as genera, that can represent a variety of species and strains.
-Likewise, metagenomic operational genomic units represent operationally informative classifications for understanding the community, but should not be interpreted as single phages or bacterial strains.
-
-Further work will be required to support the potential causal relationships outlined in this study.
-Although we provide evidence for gut bacteriophage communities to alter bacterial communities toward a cancerous state, we must follow up with experimental validation of these observations.
-The clear pathway toward these answers will be the utilization of murine models as described by Zackular *et al* which showed the link between antibiotic administration and colorectal cancer development [@Zackular:2016en].
+A notable observation from our analysis was the lack of performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences.
+This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies, and the advantage of 16S rRNA over whole metagenomic shotgun sequencing.
+We found that there were six bacterial OTUs that drove the performance of the 16S rRNA classiciation model, and these OTUs were all sparsely present and lowly abundant.
+Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model.
+The bacterial metagenomic OGUs represented only the most abundant taxa, which was uninformative for this application.
+There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but these approaches did not utilize OGU clustering like we did here, and the models only performed **as well** as the 16S rRNA model.
+Thus the targeted 16S rRNA sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples.
+Despite a loss of enthusiasm in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
 
@@
 
-![*Example visualization of the proposed model for bacteriophage modulation of colorectal cancer.
-A) Example network diagrams illustrating the impact of changes on keystone phages (\* asterisk) within the microbiome.
-A hypothetical ecological network may being at a healthy equilibrium but an alteration to the abundance of a keystone phage can impact the remainder of the network.
-Highly connected phage host abundances may be altered by the original disruption, which would ripple out to the remainder of the model.
-B) Another visualization of the impact a change in keystone phage abundance could have on the remainder of the network.*\label{modelsummary}](../figures/ExampleHypothesis.pdf){ width=75% }
+![*Working hypothesis of how the bacteriophage community is associated with colorectal cancer and the associated bacterial community.*\label{modelsummary}](../figures/WorkingHypothesis.pdf){ width=100% }
 
@@
 Phage OGUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
-No disease groups were statistically significant.*\label{betaogunegative}](../figures/lysogenic-relabund.pdf){ width=50% }
+No disease groups were statistically significant.*\label{replicationstyles}](../figures/lysogenic-relabund.pdf){ width=50% }
 
@@
 ![*Relative abundance correlations between bacterial OTUs and virome OGUs.
-A) Pearson correlation R values between all bacterial OTUs (x-axis) and viral OGUs (y-axis) with blue being positively correlated and red being negatively correlated.
+A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OGUs (y-axis) with blue being positively correlated and red being negatively correlated.
 Operational units are organized by importance in their colorectal cancer classification models, such that the most important units are in the top left corner.
@@
 The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as pannel (A).
-C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OGUs.*\label{betaogunegative}](../figures/phage-bacteria-cor.pdf){ width=100% }
+C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OGUs.*\label{correlations}](../figures/phage-bacteria-cor.pdf){ width=100% }
+
+\newpage
+
+![*Comparison of bacterial 16S rRNA classification models with and without OTUs whose median relative abundance are greater than zero.
+A) Classification model performance (measured as area under the curve) for bacteria models using 16S rRNA data both with and without filtering of samples whose median was zero.
+Significance was calculated using a Wilcoxon rank sum test, and the resulting p-value is shown.
+The random area under the curve (0.5) is marked with a dashed line.
+B) Relative abundance of the six bacterial OTUs removed when filtered for OTUs with median relative abundance of zero.
+OTU relative abundance is seperated by healthy (red) and cancerous (grey) samples.
+Relative abundance of 1% is marked by the dashed line.*\label{16scompare}](../figures/filtered16S.pdf){ width=100% }
+
+\newpage
+
+![*Transition of colorectal cancer importance through disease progression.*\label{transitionmodels}](../figures/temporalmodels.pdf){ width=80% }
 

### 2017-03-14 doc/manuscript.Rmd
More major revisions.

@@
 Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world.
-Although the cause of colorectal cancer remains unclear, it has been strongly linked to colon bacterial communities.
-Viruses are another important component of the colon microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential.
-We evaluated the colon virome (virus community) role in colorectal cancer using a cohort of 90 human subjects with either healthy, precancerous, or cancerous colons.
-We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic methods to compare the virome's role to that of the bacterial community.
-We found that alpha and beta diversity metrics were insufficient for detecting an association between the virome and colorectal cancer, but virome-based classification models were highly associated with colorectal cancer.
-Bacteriophages, not eukaryotic viruses, made up the majority of the CRC-associated virome, suggesting the community was indirectly linked to colorectal cancer, modulating bacterial community structure and functionality.
-Phages with broader host ranges had a more significant role in cancer development.
-These results provide foundational evidence that phage communities are associated with colorectal cancer, and that broadly tropic phages play more significant roles.
-Because of its importance, the virome should be considered in future colorectal cancer studies, as well as other cancer types.
+Although the majority of colorectal cancer case causes remains unclear, it's progression has been linked to colonic bacterial community composition.
+Viruses are another important component of the colonic microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential.
+We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons.
+We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community.
+We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated virome-based random forest models identified significant changes in the virus community.
+The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting the community was indirectly linked to colorectal cancer by modulating bacterial community structure and functionality.
+These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
+Together our findings add to the existing model for the role the microbiome plays in colorectal cancer development, thus providing us with a more complete understanding of colorectal cancer etiology.
 
-**Word Count**: 229
+**Word Count**: 226
 
@@
 Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide.
-It's progression and severity is linked to colon bacterial communities.
-Little is known about the cancer-associated colon virus communities and their influence on bacteria.
-We characterize this influence of colon virus community structure on colorectal cancer in humans.
-The link between the colon virome and colorectal cancer was established by using virome-based machine learning models to accurately classify patient cancer status.
-The cancer-virus link was driven primarily by bacteriophages (bacterial viruses).
-Wider phage host ranges were more strongly linked to colorectal cancer.
+Its progression and severity have been linked to colonic bacterial community composition.
+Little is known about cancer-associated colon virus communities and their influence on bacteria.
+We began addressing this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients, and how they compare to bacterial community changes.
+The altered colorectal cancer virome was used in virome-based machine learning models to accurately classify patient cancer status.
+The cancer-virus classifier was driven primarily by bacteriophages (bacterial viruses).
+Wider phage host ranges were more strongly associated with colorectal cancer.
 The results suggest an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.
 
-**Word Count**: 119
+**Word Count**: 125
 
@@
 # Introduction
-Cancer remains a devastating and persistent plague on humanity.
-Although cancer is still a primary cause of morbidity and mortality worldwide, we have made considerable therapeutic progress in recent decades.
-Perhaps one of our most impactful advances has not been in cancer treatment, but rather in detecting cancers at early stages, thereby improving treatment efficacy.
-This has been evident in a variety of prominent cancer types, with one of the most notable being colorectal cancer.
+Due to their mutagenic abilities and their propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
+Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer.
+The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g.
+cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
+The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggest that colorectal cancer may be linked to altered virus communities.
 
@@
 The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans will have died from the disease [@Siegel:2014jo].
-Although it remains a major health problem, the impact of colorectal cancer has been reduced by improved screening and prevention efforts [@Zauber:2015cr; @Siegel:2014jo].
+Although the majority of colorectal cancer case causes remains unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
+This work has led to the proposal of disease models in which bacteria colonize the colon, develop biofilms, promote infiltration, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
+This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja].
+While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome.
+Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.
+We evaluated colorectal cancer disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer developement: healthy, adenomatous, and cancerous.
 
-Development of colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e.
+Colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e.
 adenoma) in the large intestine [@Fearon:2011go].
 If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death.
-Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr].
-Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz].
-Screening methods are effective, but their invasiveness has created a lack of compliance, creating a need for accurate, non-invasive screening methods.
-One such method is screening associated colon microbial communities.
+Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr], and survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz].
+Thus work such as this, which aims to power early detection and prevention of progression from early cancer stages, has a great potential to inform therapeutic development.
 
-Although the cause of colorectal cancer remains unclear, it has been strongly associated with colon bacterial communities [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
-This association has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja].
-While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome.
-Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.
-
-Due to their mutagenic abilities and their propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Additionally, because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages are potentially indirectly linked to cancer.
-The colon virome (the virus community of the colon) has the potential to impact health and disease (e.g.
-cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
-We aim to take this field of research further by beginning to assess the role of the virome in colorectal cancer.
-
-Here we present a study of the colorectal cancer virome, highlighting the viruses most associated with the cancer state and the virome's utility for prognosis and diagnosis.
-We report that, like the association between the bacterial community and colorectal cancer was driven primarily by *Fusobacterium*, the association between the virome and colorectal cancer was driven by broadly infectious, bacteriophage hubs within the phage-bacteria ecosystem.
-By creating effective classification models using virus community signatures, we were able to accurately classify stool samples as cancerous, precancerous, or healthy.
+In this study we report that colorectal cancer is associated with variation in colonic virus communities.
+The majority of viruses identified within the virome were temperate bacteriophages.
+Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was also driven by a subset of influential phages.
+Our data suggest that the influential phages do not exclusively infect influential bacterial (e.g.
+*Fusobacterium* phages infecting the *Fusobacterium*), but rather act through the community as a whole.
 The implications of these findings are threefold.
 *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process.
-*Second*, we present a supplementary, or even alternative, virus-based approach for classification modeling of colorectal cancer using stool samples.
+*Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples.
 *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.
-We expect this study will provide opportunities for continued research into the role of the virome in human cancer development.
 
@@
 ## The Colorectal Cancer Virome Cohort
-The study cohort consisted of 90 human subjects, 30 of which served as healthy controls, 30 of which had adenoma lesions consistent with a precancerous state, and 30 which had carcinoma lesions consistent with colorectal cancer **(Figure \ref{sampleproc})**.
+The study cohort consisted of 90 human subjects, 30 of which had healthy colons, 30 of which had adenomas, and 30 which had carcinomas **(Figure \ref{sampleproc})**.
 Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques.
-The 16S rRNA gene sequences were reported in a previous publication but re-analyzed here [@Zackular:2014fba].
 The other half of the stool samples were purified for virus like particles (VLPs) and consequent genomic DNA extraction, followed by shotgun metagenomic sequencing.
-The VLP purification allowed us to observe the *active virome* because we only sequence those viruses that are encapsulated.
+The VLP purification allowed us to observe the *active virome* because we only sequenced those viruses that are encapsulated.
 
-Virus DNA was purified by re-suspending the stool in buffer and removing contaminating cells (e.g.
-human, bacteria, etc) by filtration followed by contaminating cell lysis and degradation of the released genomic DNA **(Figure \ref{sampleproc})**.
-The resulting genomic DNA was used to prepare a shotgun metagenomic sequencing library that was sequenced on the Illumina HiSeq4000 platform.
-Each run was performed with a blank control to detect any contaminants from reagents.
-Only one of the nine viral controls detected DNA, which was of a minimal concentration, providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
+Virus DNA was purified before sequencing, which allowed us to analyze DNA from within virus capsids **(Figure \ref{sampleproc})**.
+Each extraction protocol was performed with a blank control to detect any contaminants from reagents.
+Only one of the nine controls contained detectable DNA, which was of a minimal concentration, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
 As was expected, these controls were sparsely sequenced and were mostly removed while sub-sampling to even depths **(Figure \ref{qualcontrol} B)**.
+The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**.
+Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs; conceptually similar to 16S rRNA gene operational taxonomic units) with the majority containing hundreds of related contigs **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
 
-The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**.
-Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) with the majority containing hundreds of related contigs **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
-These operational units, which are conceptually similar to the operational taxonomic units (OTUs) used in 16S analysis, allow us to study bacterial and phage entities as highly related genomic entities.
+## Virus Diversity and Colorectal Cancer
+Microbiome and disease associations are often described as being of an altered diversity (i.e.
+"dysbiotic").
+We therefore initially evaluated the diversity of virome OGUs and their association with colorectal cancer.
+We utilized the Bray-Curtis dissimilarity metric to evaluate differences in communities between disease states.
+To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.
 
-## Diversity is Insufficient for Virome-Based Cancer Classification
-Microbiome and disease associations are often described as being of an altered diversity (i.e.
-"dysbiosis").
-We therefore initially evaluated the diversity of virome OGUs and their association with colorectal cancer.
-We utilized the Bray-Curtis dissimilarity metric to evaluate the differences in communities between disease states.
-To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced negative controls.
+We calculated the differences in virus alpha diversity associated with colorectal cancer.
+We found no significant alterations in either Shannon entropy or richness **(Figure \ref{betaogu} C-D)**.
+There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**).
+It is worth noting that there was a significant difference between the blank controls (those few that remained after sub-sampling) and the other study groups, further supporting the quality of our sample set (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**.
+Overall, standard diversity metrics were insufficient for capturing the differences in the virus communities between disease states.
 
-There was no statistically significant clustering of the disease groups, as visualized by NMDS ordination (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu} A**).
-An Anosim test with a post hoc multivariate Tukey test was used to calculate the statistical significance of the differences between the disease groups based on the variance around the cluster centroids **(Figure \ref{betaogu} B**).
-There were no significant differences between the disease groups, although there was a strongly significant difference between the blank controls (those few that remained after sub-sampling) and the rest of the study groups, further supporting the quality of our sample set (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**.
-
-In addition to beta diversity, we also calculated the differences in virus alpha diversity associated with colorectal cancer.
-Again we found no significant alterations in either Shannon entropy or richness of the virus communities **(Figure \ref{betaogu} C-D)**.
-Overall, standard diversity metrics were insufficient for capturing the differences in the virus communities between disease states.
-This suggested to us that a more sophisticated approach for understanding the microbial community may be required, such as machine learning classification.
-
-## Machine Learning Links the Virome to Colorectal Cancer
-Previous work has shown that 16S rRNA community signatures are effective for classifying stool samples as originating from healthy, precancerous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja].
-This is valuable because it presents a potential alternative screening approach to an invasive and expensive colonoscopy.
-The exceptional performance of bacteria in these classification models also supports a role for bacteria in colorectal cancer.
+## Virome Composition Changes in Colorectal Cancer
+Previous work has shown that 16S rRNA gene relative abundance profiles are effective features for classifying stool samples as originating from healthy, adenomatous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja].
+The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer.
 Here we built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.
@@
 We built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals.
-These models were based on virus metagenomic community signatures or bacterial 16S rRNA gene signatures.
-We also included a whole metagenomic sequence set to ensure viral metagenomic observations were not a trait of metagenomics in general.
-Each operational units' relative abundance was used in the feature set.
-The same model approach was used for all three datasets, and the only difference was the data used to learn the model.
-We confirmed that our model using bacterial 16S data replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba].
+These models were based on virus metagenomic community or bacterial 16S rRNA gene relative abundance profiles.
+We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba].
+We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance.
+The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.
 
-We compared the bacterial 16S rRNA gene model to a model built using the virome signatures.
-The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively, after fifteen random forest iterations **(Figure \ref{predmodel} A - B)**.
+To evaluate the synergistic capabilities of the bacteria and viruses within the model, we built a combinatory model that used both bacterial and viral community data.
+The combination model yielded modest but significantly improved performance beyond the separate virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, and yielded an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
+This suggested that features from the virus and bacterial communities may have had synergistic capabilities for classifying stool as belonging to cancerous individuals.
 
-To confirm that this observation was due to the virome community itself, and was not a trait of metagenomic datasets in general, we built a model using whole metagenomic community signatures.
-This model performed poorly with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`, lending support to our observation that classification performance is a trait unique to the virus metagenome **(Figure \ref{predmodel} A - B)**.
-Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs.
-Further filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**.
+We also compared these models to relative abundance profiles from bacterial metagenomic shotgun sequencing data.
+This model performed poorly, with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
+Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs **(Figure \ref{16scompare})**.
+Filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**.
 The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.
 
-To evaluate the synergistic capabilities of the bacterial and viral signatures within the model, we built a combinatory model using both bacteria community and virome data.
-The combination model yielded modest but significantly improved performance beyond the virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
-This suggests the virus and bacterial communities may have synergistic capabilities for classifying stool as belonging to cancerous individuals.
+The association between the two communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed.
+*Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu].
+The virome signature was also driven by a few operational genomic units, suggesting a role for the OGUs in cancer development **(Figure \ref{predmodel} D)**.
+The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified").
+Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant unknown viral population associated with the human virome.
+This is common in the virome; studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz].
+When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.
 
-The association between the two communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was iteratively removed.
-*Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with it's previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu].
-The virome signature was also driven primarily by a few operational genomic units, suggesting a role in cancer development **(Figure \ref{predmodel} D)**.
-The identified important viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified").
-Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant viral dark matter (uncharacterized virus genomic space) associated with the human virome.
-This is common in the virome, and studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz].
+## Phage Influence Shifts During CRC Progression
+Because our cohort included healthy, adenomatous, and cancerous colons, we were able to gain insight into virus community shifts during cancer progression.
+We evaluated community shifts between the two disease transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the sample classes around the transitions.
+We found that, while bacterial 16S rRNA gene models perform equally well for all disease class comparisons, the virome models performed worse when seperating healthy from adenomatous, and better when seperating adenomatous from cancerous **(Figure \ref{transitionmodels} A-B)**.
+Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members drove the transitions from healthy to adenomatous and adenomatous to cancerous, with one phage exception **(Figure \ref{transitionmodels} C-E)**.
+A *Myoviridae* OGU (Cluster 188) was the seventh most important phage driving the transition from a healthy to adenomatous state, and was the most important phage driving the progression from an adenomatous to carcingenic state.
+Therefore, like bacteria, there are distinct phages assocaited with the transitions from a healthy to an adenomatous colon, and from an adenomatous to a cancerous colon.
 
-When the bacterial and viral community signatures were combined, two bacterial and two viral organisms primarily drove the community association with cancer **(Figure \ref{predmodel} E)**.
-The most important microbes were two unidentified viruses, followed by *Bacteroides* and *Fusobacterium*.
-
-## The colon Virome Allows for Cancerous, Adenomatous, and Healthy Classification
-After evaluating our ability to classify samples as cancerous or healthy, we incorporated the precancerous adenoma samples into the model and evaluated our ability to classify all three states in our total sample set.
-We used three-class random forest models for the bacterial 16S and viral sample sets.
+## Inclusive Colonic Virome Classification
+After evaluating our ability to classify samples between two disease states, we performed a comprehensive three-class classification of all three disease states.
+We used a three-class random forest model for the bacterial 16S rRNA gene and viral sample sets.
 The bacterial signature model yielded an AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded an overall AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**).
-Both models were best able to classify cancer samples from healthy or precancerous samples, but struggled to distinguish precancerous from healthy or cancerous **(Figure \ref{threewaymodel} A-B)**.
+Both models were best able to classify cancer samples from healthy or precancerous samples, but struggled to distinguish adenomatous from healthy or cancerous **(Figure \ref{threewaymodel} A-B)**.
 The cancerous signal was the most discriminatory of the three sample types.
 
-Many of the microbes important for the two-class (cancer vs healthy) bacteria and virus models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
+The microbes important for the cancer vs healthy and healthy vs adenoma bacteria and virus models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
 The most important bacterium was the same *Fusobacterium* between the two and three class models, supporting its significance to the association between cancer and the bacterial communities **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**.
-Unlike the two-class virome model, the viruses most important to the three-class model were identified bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.
-A *Podoviridae* was most important, followed by two *Siphoviridae* phages.
+The viruses most important to the three-class model were identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.
 
@@
 
-## Temperate Bacteriophages Drive Link Between Virome and Colorectal Cancer
-The virome-based model was able to accurately classify stool samples as cancerous, precancerous, or healthy.
-Not only is this important for establishing an alternative diagnostic model, but it also suggests an underlying biological importance for viruses in colorectal cancer.
-Above we used our classification models to evaluate which virus OGUs were most highly linked to colorectal cancer.
-We were able to further characterize these important OGUs to better understand the underlying biological link between the virome and colorectal cancer.
-
-The role of the virome in colorectal cancer could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts.
-To better understand the types of viruses that are important for colorectal cancer, we utilized the longest sequences from the OGUs as the representative sequences to be used for taxonomic classification.
-These sequences were aligned to a set of all reference virus genomes, including bacteriophages and eukaryotic viruses.
-A strict e-value threshold of 1e-25 was used to improve our confidence in the matches between the genome sequences.
-The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel} E)**.
+## Dominance of Bacteriophages in Colorectal Cancer Virome
+Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts.
+To better understand the types of viruses that are important for colorectal cancer, we identified the virome OGUs as being similar to either eukaryotic viruses or bacteriophages.
+The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**.
 Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes.
 
-We evaluated whether the phages in the community were primarily lytic (lyse their hosts) or temperate (lysogenic; can integrate into their host's genome, as well as lyse the cell).
+We evaluated whether the phages in the community were primarily lytic (replicate by lysing their host) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell).
 We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes.
 This approach was done as described in previous work [@Minot:2011ez; @Hannigan:2015fz].
-We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{threewaymodel} E)**.
+We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
 Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes.
-These findings are consistent with previous reports suggesting the colon virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
+These findings are consistent with previous reports suggesting the gut virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
 
-## Influential Bacteria and Phages Associate Indirectly
-Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were predators of the influential bacteria, and thus influenced their relative abundance through predation (e.g.
-One of the important phages targets *Fusobacterium*).
-If this hypothesis was true, we would expect that the relative abundance between many important bacteria and phages to be correlated.
-We correlated the relative abundance values of every phage OGU and bacterial OTU and found strikingly low correlation between the bacterial and phage abundances **(Figure \ref{correlations} A,C)**.
+## Community Context of Influential Phages
+Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were predators of the influential bacteria, and thus influenced their relative abundance through predation.
+If this hypothesis were true, we would expect the relative abundance between important bacteria and phages to be correlated.
+Instead we observed a strikingly low correlation between the bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**.
 There was an overall absence of correlations between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**.
-This evidence supported our null hypothesis that the influential phages are not predators of the influential bacterial.
+This evidence supported our null hypothesis that the influential phages are not primarily predators of influential bacteria.
 
-We further investigated the infectious capabilities and potential host ranges of the influential phages by building a machine learning model to predict which phage OGUs infect which bacteria in the overall community.
-The predicted interactions were then used to build a network which was used for subsequent analysis.
-This approach was done as has been previously described (**Cite Network Preprint**).
+Our next hypothesis was that the influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the select influential bacteria.
+This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network.
+We investigated the infectious capabilities and potential host ranges of the influential phages by building a machine learning model to predict which phage OGUs infect which bacteria in the overall community.
+The predicted interactions were then used to build a network which was used for subsequent analysis, as has been previously described (**Cite Network Preprint**).
 This analysis revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**.
 We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model.
-We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), indicating that phages important in driving colorectal cancer are also community hubs **(Figure \ref{network} B)**.
-
-## Influential Phage Identities Shift During CRC Progression
-Because our cohort sampled healthy, adenomatous, and cancerous colons, we are able to gain initial insight into virus community shifts during cancer progression.
-We evaluated the community changes at the two disease transitions (healthy to adenomatous and adenomatous to cancerous) by building individual random forest models to compare only those sample classes.
-We found that, while bacterial 16S rRNA models perform equally well for all disease class comparisons, the virome models performed worst when seperating healthy from adenomatous, better when seperating adenomatous from cancerous, and the best when seperating healthy from cancerous **(Figure \ref{transitionmodels} A-B)**.
-Like bacterial communities, the most important phage OGUs provide insight into which organisms are driving the transitions between disease states.
-Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members drove the transitions from healthy to adenomatous and adenomatous to cancerous, with one phage exception **(Figure \ref{transitionmodels} C-E)**.
-A *Myoviridae* OGU (Cluster 188) was the seventh most important phage driving the transition from a healthy to adenomatous state, and was the most important phage driving the progression from an adenomatous to carcingenic state.
-Therefore, like bacteria, there are distinct phages assocaited with the transitions from a healthy to an adenomatous colon, and from an adenomatous to a cancerous colon.
+We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), indicating that phages important in driving colorectal cancer were also community hubs **(Figure \ref{network} B)**.
+Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.
 
@@
 Because some bacteria also have oncogenic properties, bacteriophages may play an indirect role in promoting carcinogenesis by altering bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
-Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colon communities (the human colon virome) and colorectal cancer has yet to be evaluated.
-Here we show that, like colon bacterial communities, the colon virome is associated with colorectal cancer.
+Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colorectal communities and colorectal cancer has yet to be evaluated.
+Here we show that, like colonic bacterial communities, the colon virome is associated with colorectal cancer.
 Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.
 
-We found that the basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states.
-This led us to implement a more sophisticated machine learning approach (random forest classificaiton model) that allowed us to detect strong associations between the colon virus communities and colorectal cancer.
-Unlike other human cancer types that are often assocaited with human viruses, colorectal cancer is primarily assocaited with altered bacteriophage communities.
-These phage communities were not predators of the most influential bacteria (e.g.
+Together our data allowed us to begin modeling the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**.
+We found basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states.
+By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus communities and colorectal cancer.
+Colorectal cancer was primarily assocaited with altered bacteriophage communities.
+These phage communities were not exclusive predators of the most influential bacteria (e.g.
 *Fusobacterium* linked to decreases in *Fusobacterium* phages), as demonstrated by the lack of correlation between the abundance of these entities.
-Additionally, we identified a correlation in which those phages strongly linked to colorectal cancer were also identified as influential hubs within the overall phage-bacteria network, linking the influence of a phage on the community as a whole to its importance to cancer progression.
-Our previous work has shown that modifying colon bacterial communities with antibiotics alters colorectal cancer progression and tumor burden in mice, and suggests phage-mediated community alterations may have a similar effect, especially if those phages were broadly infectious [@Zackular:2016en].
-Together this data supports a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
+Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead directly modulating the influential bacteria.
+Our previous work has shown that modifying colon bacterial communities with antibiotics alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en].
+This provides a precident for phage indirectly influencing colorectal cancer progression y altering the bacterial community composition.
+Overall our data supports a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
+Although our evidence suggests phages indirectly influence colorectal cancer developement, we are not able to rule out the role of phages directly interacting with the human host.
 
-To better understand the potential mechanisms by which phages modulate the colon bacterial community, we built a working hypothesis that incorporated our findings into the Flynn model for colorectal cancer developement **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
-We hypothesized that the broadly infectious phages in the colon began lysing, and thereby disrupting, the bacterial communities to open a niche in which opportunistic bacteria were able to colonize, such as *Fusobacterium nucleatum*.
-Once the "driver" bacteria had established themselves in the epithelium, other opporunistic "passenger" bacteria were able to adhere to the driver, colonize, and begin establishing a biofilm.
-We feel it is likely that the phages played a role in biofilm dispersal and growth, by lysing bacteria within the biofilm, a process shown to be important for effective biofilm growth.
+In addition to modeling the basic potential connections between virus communities, bacteria communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur.
+This was done by incorporating our findings into the Flynn model for colorectal cancer developement **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
+We hypothesize that the broadly infectious phages in the colon began lysing, and thereby disrupting, the bacterial communities to open a niche in which opportunistic bacteria (such as *Fusobacterium nucleatum*) were able to colonize.
+Once the influential "driver" bacteria had established themselves in the epithelium, other opporunistic "passenger" bacteria were able to adhere to the driver, colonize, and begin establishing a biofilm.
+Phages may have played a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process shown to be important for effective biofilm growth [@Rossmann:2015cj].
 The oncogenic bacteria were then able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response.
@@
 
-Although we have evidence for the interaction between the bacteriophage and bacterial community, we are not able to address whether the phages are directly interacting with the human host **(Figure \ref{modelsummary} A)**.
-Like most foreign entities, bacteriophages are known to be capable of acting as human antigens and thereby initiate or contribute to an immune response.
-In the context of colorectal cancer, some phages may have been induced out of their bacterial host (most phages were temperate) and interacted with the human host to promote the cancer-associated inflammatory response.
-Because we are not able to address this behavior with our dataset, we noted it as a known unknown, and another plausible mechanism by which some phages could be associated with colorectal cancer developement.
-
 A notable observation from our analysis was the lack of performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences.
 This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies, and the advantage of 16S rRNA over whole metagenomic shotgun sequencing.
-We found that there were six bacterial OTUs that drove the performance of the 16S rRNA classiciation model, and these OTUs were all sparsely present and lowly abundant.
+We found that there were six bacterial OTUs that drove the performance of the 16S rRNA classification model, and these OTUs were all sparsely present and lowly abundant.
 Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model.
 The bacterial metagenomic OGUs represented only the most abundant taxa, which was uninformative for this application.
-There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but these approaches did not utilize OGU clustering like we did here, and the models only performed **as well** as the 16S rRNA model.
+There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but these approaches did not utilize OGU clustering like we did here, and the models only performed **as well** as the 16S rRNA model [@Zeller:2014ix].
 Thus the targeted 16S rRNA sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples.
-Despite a loss of enthusiasm in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
+Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
 
@@
 Alpha and beta diversity were calculated using the operational genomic unit abundance profiles for each sample.
-Sequences were sub-sampled down to 100,000 sequences.
+Sequences were rarefied to 100,000 sequences.
 Samples with less than the cutoff were removed from the analysis.
@@
 
-![*Cohort and sample processing outline.
-Thirty subject stool samples were collected from healthy, adenoma (pre-cancer), and carcinoma (cancer) patients.
-Stool samples were split into two aliquots, the first of which was used for bacterial sequencing and the second which was used for virus sequencing.
-Bacterial sequencing was done using both 16S rRNA amplicon and whole metagenomic shotgun sequencing techniques.
-Virus samples were purified for viruses using filtration and a combination of chloroform (bacterial lysis) and DNase (exposed genomic DNA degradation).
-The resulting encapsulated virus DNA was sequenced using whole metagenomic shotgun sequencing.* \label{sampleproc}](../figures/sample-processing.pdf){ width=100% }
-
-\newpage
-
 ![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S signatures, whole metagenomic signatures, and a combination of virome and 16S signatures.
@@
 
-![*ROC curves from A) virome and B) bacterial 16S three-class random forest models tuned on mean AUC.
-Each curve represents the ability of the specified class to be classified against the other two classes.
-C) Quantification of the mean AUC variation for each model based on 10 model iterations.
-A pairwise Wilcoxon test with a Bonferroni multiple hypothesis correction demonstrated that the models are significantly different (alpha = 0.01).
-D) Mean decrease in accuracy when virome operational genomic units and E) bacterial 16S OTUs are removed from the respective three-class classification models.
-Results based on 25 iterations.
-F) Relative abundance of the six most important virome OGUs in the model, with the most important on the right.
-Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=100% }
+![*Relative abundance correlations between bacterial OTUs and virome OGUs.
+A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OGUs (y-axis) with blue being positively correlated and red being negatively correlated.
+Operational units are organized by importance in their colorectal cancer classification models, such that the most important units are in the top left corner.
+B) Magnification of the boxed region in pannel (A), highlighting the correlation between the most important bacterial OTUs and virome OGUs.
+The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as pannel (A).
+C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OGUs.*\label{correlations}](../figures/phage-bacteria-cor.pdf){ width=100% }
 
@@
 B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OGU.
-A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}](../figures/NetworkAndScatter.pdf){ width=80% }
+A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, r = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}](../figures/NetworkAndScatter.pdf){ width=80% }
 
@@
 \beginsupplement
+
+![*Cohort and sample processing outline.
+Thirty subject stool samples were collected from healthy, adenoma (pre-cancer), and carcinoma (cancer) patients.
+Stool samples were split into two aliquots, the first of which was used for bacterial sequencing and the second which was used for virus sequencing.
+Bacterial sequencing was done using both 16S rRNA amplicon and whole metagenomic shotgun sequencing techniques.
+Virus samples were purified for viruses using filtration and a combination of chloroform (bacterial lysis) and DNase (exposed genomic DNA degradation).
+The resulting encapsulated virus DNA was sequenced using whole metagenomic shotgun sequencing.* \label{sampleproc}](../figures/sample-processing.pdf){ width=100% }
+
+\newpage
 
@@
 
-![*Lysogenic phage relative abundance in disease states.
-Phage OGUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
-No disease groups were statistically significant.*\label{replicationstyles}](../figures/lysogenic-relabund.pdf){ width=50% }
-
-\newpage
-
-![*Relative abundance correlations between bacterial OTUs and virome OGUs.
-A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OGUs (y-axis) with blue being positively correlated and red being negatively correlated.
-Operational units are organized by importance in their colorectal cancer classification models, such that the most important units are in the top left corner.
-B) Magnification of the boxed region in pannel (A), highlighting the correlation between the most important bacterial OTUs and virome OGUs.
-The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as pannel (A).
-C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OGUs.*\label{correlations}](../figures/phage-bacteria-cor.pdf){ width=100% }
-
-\newpage
-
 ![*Comparison of bacterial 16S rRNA classification models with and without OTUs whose median relative abundance are greater than zero.
@@
 
+![*ROC curves from A) virome and B) bacterial 16S three-class random forest models tuned on mean AUC.
+Each curve represents the ability of the specified class to be classified against the other two classes.
+C) Quantification of the mean AUC variation for each model based on 10 model iterations.
+A pairwise Wilcoxon test with a Bonferroni multiple hypothesis correction demonstrated that the models are significantly different (alpha = 0.01).
+D) Mean decrease in accuracy when virome operational genomic units and E) bacterial 16S OTUs are removed from the respective three-class classification models.
+Results based on 25 iterations.
+F) Relative abundance of the six most important virome OGUs in the model, with the most important on the right.
+Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=80% }
+
+\newpage
+
+![*Lysogenic phage relative abundance in disease states.
+Phage OGUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
+No disease groups were statistically significant.*\label{replicationstyles}](../figures/lysogenic-relabund.pdf){ width=50% }
+
+\newpage
+
 # References

### 2017-03-14 doc/manuscript.Rmd
Meh, title.

@@
 ---
-title: Characterizing the Role of the Human Virome in Colorectal Cancer
+title: The Colorectal Cancer Virome
 author: 

### 2017-03-16 bin/CompareBacteriaFilters.R
Huge push.

@@ -1,0 +1,367 @@
+gcinfo(FALSE)
+
+library("optparse")
+library("ggplot2")
+library("plotROC")
+library("reshape2")
+library("wesanderson")
+library("vegan")
+library("plyr")
+library("dplyr")
+library("caret")
+library("cowplot")
+library("matrixStats")
+library("parallel")
+
+minavgrelabund <- 0
+minsamps <- 30
+iterationcount <- 15
+
+
+rarefybig <- function(x, subdepth) {
+    print("Creating empty matrix")
+
+    randnorep <- function(count, min, max) {
+        if((max - min + 1) < count) {
+            write("ERROR: Sample is out of range!", stderr())
+            break
+        }
+        uct <- 0
+        rounded <- NULL
+        while(uct < count) {
+            itrr <- round(runif((count - uct), min = min, max = max))
+            rounded <- c(rounded, itrr)
+            rounded <- unique(rounded)
+            uct <- length(rounded)
+        }
+        rounded
+    }
+
+    m <- as.matrix(x)
+    m[] <- 0
+    m <- as.data.frame(m)
+
+    df <- data.frame(value = c(0, 0))
+    xcount <- 0
+    finalcount <- 0
+    # Create reference table
+    print("Creating reference table")
+    for (i in 1:length(x)) {
+        cval <- x[,i]
+        name <- colnames(x)[i]
+        finalcount <- xcount + cval
+        df[,name] <- c(xcount, finalcount)
+        xcount <- finalcount
+    }
+    df <- as.data.frame(t(df[,-1]))
+    df$names <- rownames(df)
+
+    randdist <- randnorep(subdepth, 1, sum(x))
+
+    print("Annotating random distribution")
+    getValue <- function(x, data) {
+        tmp <- data %>%
+        filter(x <= V2, x > V1)
+        return(tmp$names)
+    }
+
+    row <- sapply(randdist, getValue, data=df)
+    row <- table(row)
+    ind <- names(row)
+    x[] <- 0
+    x[,ind] <- row
+
+    return(x)
+}
+
+caretmodel <- function(x) {
+  fitControl <- trainControl(method = "repeatedcv",
+	repeats = 5,
+	number=5,
+	classProbs = TRUE,
+	summaryFunction = twoClassSummary,
+	savePredictions = TRUE)
[…]

### 2017-03-16 bin/ConnectionImportanceCorr.R

@@ -62,6 +62,17 @@ caretmodel <- function(x) {
   return(model)
 }
 
+tcar <- function(x) {
+  fitControl <- trainControl(method = "repeatedcv",
+  repeats = 5,
+  number=5,
+  classProbs = TRUE,
+  summaryFunction = twoClassSummary,
+  savePredictions = TRUE)
+  model <- train(V30~., data=x, trControl=fitControl, method="rf", metric="ROC", tuneLength=5)
+  return(model)
+}
+
 plotnetwork <- function (nodeframe=nodeout, edgeframe=edgeout) {
   write("Preparing Data for Plotting", stderr())
   # Pull out the data for clustering
@@ -80,9 +91,15 @@ plotnetwork <- function (nodeframe=nodeout, edgeframe=edgeout) {
   return(outputgraph)
 }
 
-GetAverageImportance <- function(x, y) {
+GetAverageImportance <- function(x, y, twomodel = FALSE) {
   write(y, stderr())
-  functionmodel <- caretmodel(x)
+  if(twomodel == TRUE) {
+    print("Running two class model.")
+    functionmodel <- tcar(x)
+  } else {
+    print("Running multiclass model.")
+    functionmodel <- caretmodel(x)
+  }
   resultvardf <- data.frame(varImp(functionmodel$finalModel))
   resultvardf$categories <- rownames(resultvardf)
   resultvardf$categories <- factor(resultvardf$categories, levels = resultvardf$categories)
@@ -194,3 +211,50 @@ scorw <- data.frame("names" = c("pval", "rho"), "scores" = c(scor$p.value, as.nu
 
 write.table(scorw, file = "./rtables/cor-stats.tsv", quote = FALSE, sep = "\t", row.names = FALSE, col.names = TRUE)
 
+
+##############################################
+# Compare Importance Centrality Across Times #
+##############################################
+### Healthy Vs Adenoma
+abssubset <- castmerge[!c(castmerge$V30 %in% "Negative"),]
+abssubset <- abssubset[!c(abssubset$V30 %in% "Cancer"),]
+abssubset$V30 <- factor(abssubset$V30)
+abssubset <- abssubset[,c(colSums(abssubset != 0) > 30)]
+# Get rid of the IDs
+absmissingid <- abssubset[,-which(names(abssubset) %in% c("V22"))]
+
[…]

### 2017-03-16 bin/IdentLyticPhages.sh

@@ -15,6 +15,7 @@ export aclame="./data/metadata/aclame_proteins_prophages_0.4.fa"
 export bacteriadb="./data/metadata/BacteriaReference.fa"
 
 mkdir -p ./data/tmpidlytic
+mkdir -p ./data/repcycle
 
 RunBlastx () {
 	# 1 = Query Seqs
@@ -70,23 +71,35 @@ grep -A 1 -f ./data/tmpidlytic/contiglist.tsv ${fastafile} \
 RunBlastx \
 	./data/tmpidlytic/contigrepset.fa \
 	${integrase} \
-	./data/tmpidlytic/intblastout.tsv
+	./data/repcycle/intblastout.tsv
 
 # Align contigs to ACLAME database
 RunBlastx \
 	./data/tmpidlytic/contigrepset.fa \
 	${aclame} \
-	./data/tmpidlytic/aclameblastout.tsv
+	./data/repcycle/aclameblastout.tsv
 
 # Align contigs to bacterial database
 RunBlastn \
 	./data/tmpidlytic/contigrepset.fa \
 	${bacteriadb} \
-	./data/tmpidlytic/phagebacteriablastout.tsv
+	./data/repcycle/phagebacteriablastout.tsv
 
 # Combine the blast hits
 cat \
-	./data/tmpidlytic/intblastout.tsv \
-	./data/tmpidlytic/aclameblastout.tsv \
-	./data/tmpidlytic/phagebacteriablastout.tsv
+	./data/repcycle/intblastout.tsv \
+	./data/repcycle/aclameblastout.tsv \
+	./data/repcycle/phagebacteriablastout.tsv \
+	| cut -f 1 \
+	| sort \
+	| uniq \
+	> ./data/repcycle/phagereplicationcycle.tsv
+
+# Add the contig IDs to the table
+awk -F "\t" 'FNR==NR { a[$1] = $3; next } { for( i in a ) if($1 ~ i) {print "Cluster_"a[$1]"\tLysogenic"} next }' \
+	./data/contigclustersidentity/longestcontigsvirus.tsv \
+	./data/repcycle/phagereplicationcycle.tsv \
+	> ./data/repcycle/repclust.tsv
+
+
 

### 2017-03-16 bin/QualityControlStats.R

@@ -75,8 +75,7 @@ depthplot <- ggplot(inputmerge, aes(x = V2, y = V1, fill = V30)) +
     coord_flip() +
     scale_fill_manual(values = wes_palette("Royal1")[c(1,2,4,3)]) +
     ylab("VLP Genomic DNA Yield (ng/uL)") +
-    xlab("") +
-    geom_hline(yintercept = opt$sdepth, linetype="dashed")
+    xlab("")
 
 gridplot <- plot_grid(dnaconcplot, depthplot, labels = c("A", "B"))
 

### 2017-03-16 bin/phagebacteriacorr.R

@@ -12,6 +12,7 @@ library("caret")
 library("cowplot")
 library("matrixStats")
 library("parallel")
+library("cowplot")
 
 minavgrelabund <- 0
 minsamps <- 30
@@ -27,7 +28,7 @@ caretmodel <- function(x) {
   return(model)
 }
 
-impcalcs <- function(x, iterationcount = 5) {
+impcalcs <- function(x, iterationcount = 10) {
 	avgimportance <- lapply(c(1:iterationcount), function(i) GetAverageImportance(x, i))
 	avgimportancedf <- ldply(avgimportance, data.frame)
 	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall))
@@ -39,7 +40,7 @@ impcalcs <- function(x, iterationcount = 5) {
 	importaverage$V7 <- as.character(importaverage$V7)
 	importaverage <- importaverage[!c(importaverage$Overall %in% NA),]
 	importaverage[is.na(importaverage)] <- "Unknown"
-	importaverage <- importaverage[order(importaverage$mean, decreasing = FALSE),]
+	importaverage <- importaverage[order(importaverage$mean, decreasing = TRUE),]
 	importaverage$categories <- factor(importaverage$categories, levels = importaverage$categories)
 	importaverage$V7 <- factor(importaverage$V7, levels = importaverage$V7)
 	
@@ -167,60 +168,115 @@ virus <- virus[,-length(virus)]
 bacteria <- bacteria[,-1]
 
 cordf <- melt(cor(virus, bacteria))
-fcordf <- cordf[c(cordf$value > 0.75 | cordf$value < -0.75),]
 
-hm <- ggplot(cordf, aes(x = Var2, y = Var1, fill = value)) +
-	theme_classic() +
+# Iterate to get average importance plot
+importancevalsvirus <- impcalcs(absmissingid)
+importancevalsbacteria <- impcalcs(pasubset)
+
+virusclean <- unique(importancevalsvirus[, c(1,4)])
+colnames(virusclean) <- c("categories", "virusmean")
+bacteriaclean <- unique(importancevalsbacteria[, c(1,4)])
+colnames(bacteriaclean) <- c("categories", "bacteriamean")
+cordmerge <- merge(cordf, virusclean, by.x = "Var1", by.y = "categories")
[…]

### 2017-03-16 bin/predictProgression.R

@@ -1,0 +1,446 @@
+gcinfo(FALSE)
+
+library("optparse")
+library("ggplot2")
+library("plotROC")
+library("reshape2")
+library("wesanderson")
+library("vegan")
+library("plyr")
+library("dplyr")
+library("caret")
+library("cowplot")
+library("matrixStats")
+library("parallel")
+
+minavgrelabund <- 0
+minsamps <- 30
+iterationcount <- 10
+
+
+rarefybig <- function(x, subdepth) {
+    print("Creating empty matrix")
+
+    randnorep <- function(count, min, max) {
+        if((max - min + 1) < count) {
+            write("ERROR: Sample is out of range!", stderr())
+            break
+        }
+        uct <- 0
+        rounded <- NULL
+        while(uct < count) {
+            itrr <- round(runif((count - uct), min = min, max = max))
+            rounded <- c(rounded, itrr)
+            rounded <- unique(rounded)
+            uct <- length(rounded)
+        }
+        rounded
+    }
+
+    m <- as.matrix(x)
+    m[] <- 0
+    m <- as.data.frame(m)
+
+    df <- data.frame(value = c(0, 0))
+    xcount <- 0
+    finalcount <- 0
+    # Create reference table
+    print("Creating reference table")
+    for (i in 1:length(x)) {
+        cval <- x[,i]
+        name <- colnames(x)[i]
+        finalcount <- xcount + cval
+        df[,name] <- c(xcount, finalcount)
+        xcount <- finalcount
+    }
+    df <- as.data.frame(t(df[,-1]))
+    df$names <- rownames(df)
+
+    randdist <- randnorep(subdepth, 1, sum(x))
+
+    print("Annotating random distribution")
+    getValue <- function(x, data) {
+        tmp <- data %>%
+        filter(x <= V2, x > V1)
+        return(tmp$names)
+    }
+
+    row <- sapply(randdist, getValue, data=df)
+    row <- table(row)
+    ind <- names(row)
+    x[] <- 0
+    x[,ind] <- row
+
+    return(x)
+}
+
+caretmodel <- function(x) {
+  fitControl <- trainControl(method = "repeatedcv",
+	repeats = 5,
+	number=5,
+	classProbs = TRUE,
[…]

### 2017-03-16 bin/predictThreeClass.R

@@ -1,4 +1,3 @@
-library("optparse")
 library("ggplot2")
 library("plotROC")
 library("reshape2")
@@ -62,12 +61,12 @@ GetAverageImportance <- function(x, y) {
 plotimportance <- function(x, iterationcount = 25, topcount = 10) {
 	avgimportance <- lapply(c(1:iterationcount), function(i) GetAverageImportance(x, i))
 	avgimportancedf <- ldply(avgimportance, data.frame)
-	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall))
+	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall), stder = sd(Overall)/sqrt(length(Overall)))
 	importaverage <- merge(avgimportancedf, import, by = "categories")
 
-	virustax <- virustax[,c(1,3,7)]
+	virustaxdf <- virustax[,c(1,3,7)]
 
-	importaverage <- merge(importaverage, virustax, by.x = "categories", by.y = "V1", all = TRUE)
+	importaverage <- merge(importaverage, virustaxdf, by.x = "categories", by.y = "V1", all = TRUE)
 	importaverage$V7 <- as.character(importaverage$V7)
 	importaverage <- importaverage[!c(importaverage$Overall %in% NA),]
 	importaverage[is.na(importaverage)] <- "Unknown"
@@ -81,20 +80,52 @@ plotimportance <- function(x, iterationcount = 25, topcount = 10) {
 	dfplot <- importaverage[(length(importaverage[,1])-numberto):(length(importaverage[,1])),]
 	
 	importanceplot <- ggplot(dfplot, aes(x=categories, y=Overall)) +
-	  theme_classic() +
-	  theme(
-	    axis.line.x = element_line(colour = "black"),
-	    axis.line.y = element_line(colour = "black")
-	  ) +
-	  geom_dotplot(fill=wes_palette("Royal1")[2], binaxis = "y", stackdir = "center", dotsize = 0.25, stackratio = 0.5) +
-	  xlab("Virus Identity") +
-	  ylab("Mean Accuracy Decrease") +
-	  stat_summary(fun.y = mean, fun.ymin = mean, fun.ymax = mean, geom = "crossbar", width = 0.75) +
-	  geom_vline(xintercept=binlength,color="grey") +
[…]

### 2017-03-16 bin/predictVirome.R

@@ -17,70 +17,70 @@ minavgrelabund <- 0
 minsamps <- 30
 
 rarefybig <- function(x, subdepth) {
-    print("Creating empty matrix")
+	print("Creating empty matrix")
 
-    randnorep <- function(count, min, max) {
-        if((max - min + 1) < count) {
-            write("ERROR: Sample is out of range!", stderr())
-            break
-        }
-        uct <- 0
-        rounded <- NULL
-        while(uct < count) {
-            itrr <- round(runif((count - uct), min = min, max = max))
-            rounded <- c(rounded, itrr)
-            rounded <- unique(rounded)
-            uct <- length(rounded)
-        }
-        rounded
-    }
+	randnorep <- function(count, min, max) {
+		if((max - min + 1) < count) {
+			write("ERROR: Sample is out of range!", stderr())
+			break
+		}
+		uct <- 0
+		rounded <- NULL
+		while(uct < count) {
+			itrr <- round(runif((count - uct), min = min, max = max))
+			rounded <- c(rounded, itrr)
+			rounded <- unique(rounded)
+			uct <- length(rounded)
+		}
+		rounded
+	}
 
-    m <- as.matrix(x)
-    m[] <- 0
-    m <- as.data.frame(m)
+	m <- as.matrix(x)
+	m[] <- 0
+	m <- as.data.frame(m)
 
-    df <- data.frame(value = c(0, 0))
-    xcount <- 0
-    finalcount <- 0
-    # Create reference table
-    print("Creating reference table")
-    for (i in 1:length(x)) {
-        cval <- x[,i]
-        name <- colnames(x)[i]
-        finalcount <- xcount + cval
-        df[,name] <- c(xcount, finalcount)
-        xcount <- finalcount
-    }
-    df <- as.data.frame(t(df[,-1]))
-    df$names <- rownames(df)
+	df <- data.frame(value = c(0, 0))
+	xcount <- 0
+	finalcount <- 0
+	# Create reference table
+	print("Creating reference table")
+	for (i in 1:length(x)) {
+		cval <- x[,i]
+		name <- colnames(x)[i]
+		finalcount <- xcount + cval
+		df[,name] <- c(xcount, finalcount)
+		xcount <- finalcount
+	}
[…]

### 2017-03-16 bin/repsty.R

@@ -1,0 +1,85 @@
+gcinfo(FALSE)
+
+library("optparse")
+library("ggplot2")
+library("plotROC")
+library("reshape2")
+library("wesanderson")
+library("vegan")
+library("plyr")
+library("dplyr")
+library("caret")
+library("cowplot")
+library("matrixStats")
+library("parallel")
+
+minavgrelabund <- -1
+minsamps <- 0
+
+input <- read.delim("./data/VirusClusteredContigAbund.tsv", header=TRUE, sep="\t")
+tempteratelist <- read.delim("./data/repcycle/repclust.tsv", header = FALSE, sep = "\t")
+datadisease <- read.delim("./data/metadata/MasterMeta.tsv", header=FALSE, sep="\t")[,c(2,30,22)]
+
+# Rarefy input table
+minimumsubsample <- 1000000
+inputcast <- dcast(input, V1 ~ V2)
+inputcast[is.na(inputcast)] <- 0
+inputcast[,-1] <-round(inputcast[,-1],0)
+row.names(inputcast) <- inputcast[,1]
+inputcast <- inputcast[,-1]
+inputcast <- as.data.frame(t(inputcast))
+counter <- 1
+rarefunction <- function(y) {
+	if(sum(y) >= 1e9) {
+		return(rarefybig(y, minimumsubsample))
+	} else {
+		return(rrarefy(y, minimumsubsample))
+	}
+}
+rareoutput <- lapply(c(1:length(inputcast[,1])), function(i) {
+	write(counter, stderr())
+	counter <<- counter + 1; 
+	subsetdf <- inputcast[i,]
+	if(sum(subsetdf) >= minimumsubsample) {
+		return(rarefunction(subsetdf))
+	}
+})
+rareoutputbind <- as.data.frame(do.call(rbind, rareoutput))
+length(rareoutputbind[,1])
+
+inputmelt <- melt(as.matrix(rareoutputbind))
+colnames(inputmelt) <- c("V2", "V1", "sum")
+inputmelt$sum <- ifelse(inputmelt$sum >= (minimumsubsample)^-1, inputmelt$sum, 0)
+
+inputrelabund <- data.frame(inputmelt %>% group_by(V2) %>% mutate(RelAbund = 100 * sum / sum(sum)))
+relabundcast <- dcast(inputrelabund, V2 ~ V1, value.var = "RelAbund")
[…]

### 2017-03-16 bin/QualityControlStats.R
Fix the colors

@@ -57,7 +57,7 @@ dnaconcplot <- ggplot(metadata, aes(x = V2, y = V26, fill = V30)) +
 	) +
 	geom_bar(stat="identity") +
 	coord_flip() +
-	scale_fill_manual(values = wes_palette("Royal1")[c(1,2,4,3)]) +
+	scale_fill_manual(values = c(wes_palette("Royal1")[c(1,2,4)], wes_palette("Darjeeling")[5]), name = "Disease") +
 	ylab("VLP Genomic DNA Yield (ng/uL)") +
 	xlab("Prepared Samples")
 
@@ -73,7 +73,7 @@ depthplot <- ggplot(inputmerge, aes(x = V2, y = V1, fill = V30)) +
     ) +
     geom_bar(stat="identity") +
     coord_flip() +
-    scale_fill_manual(values = wes_palette("Royal1")[c(1,2,4,3)]) +
+    scale_fill_manual(values = c(wes_palette("Royal1")[c(1,2,4)], wes_palette("Darjeeling")[5]), name = "Disease") +
     ylab("VLP Genomic DNA Yield (ng/uL)") +
     xlab("")
 

### 2017-03-17 bin/QualityControlStats.R
Updated QC axis

@@ -74,7 +74,7 @@ depthplot <- ggplot(inputmerge, aes(x = V2, y = V1, fill = V30)) +
     geom_bar(stat="identity") +
     coord_flip() +
     scale_fill_manual(values = c(wes_palette("Royal1")[c(1,2,4)], wes_palette("Darjeeling")[5]), name = "Disease") +
-    ylab("VLP Genomic DNA Yield (ng/uL)") +
+    ylab("VLP Sequence Yield (ng/uL)") +
     xlab("")
 
 gridplot <- plot_grid(dnaconcplot, depthplot, labels = c("A", "B"))

### 2017-03-17 bin/predictVirome.R

@@ -453,7 +453,7 @@ GetAverageAUC <- function(x, y) {
 	return(resultdf)
 }
 
-iterationcount <- 15
+iterationcount <- 25
 
 viromeauc <- lapply(c(1:iterationcount), function(i) GetAverageAUC(absmissingid, i))
 viromeaucdf <- ldply(viromeauc, data.frame)
@@ -523,7 +523,7 @@ boundmodel <- rbind(subsetmodel$pred, outmodel$pred, combomodel$pred, metaoutmod
 boundplot <- ggplot(boundmodel, aes(d = obs, m = Healthy, color = class)) +
 	geom_roc(n.cuts = 0) +
 	style_roc() +
-	scale_color_manual(values = wes_palette("Royal1"), name = "Disease") +
+	scale_color_manual(values = c(wes_palette("Royal1")[c(1,2,4)], wes_palette("Darjeeling")[5]), name = "Disease") +
 	theme(
 		legend.position = c(0.75, 0.25),
 		legend.background = element_rect(colour = "black")

### 2017-03-17 doc/manuscript.Rmd
Huge update.

@@
 ---
-title: The Colorectal Cancer Virome
+title: Viral and Bacterial Communities of Colorectal Cancer
 author: 
@@
   - code: 1
-    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48108
+    address: Department of Microbiology & Immunology, University of Michigan, Ann Arbor, Michigan, 48109
   - code: 2
-    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48108
+    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
   - code: 3
@@
     variant: markdown
+fontsize: 11pt
+mainfont: 'Helvetica'
 ---
@@
 ***Keywords***: Colorectal Cancer, Virome, Machine Learning
-
-***Text Length***: 34,182 / ~39,000 Characters
 
@@
 Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world.
-Although the majority of colorectal cancer case causes remains unclear, it's progression has been linked to colonic bacterial community composition.
-Viruses are another important component of the colonic microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential.
+Although the majority of colorectal cancer case causes remain unclear, its progression has been linked to changes in colonic bacterial community composition.
+Viruses are another important component of the colonic microbial community, but have yet to be studied in colorectal cancer, despite their oncogenic potential.
 We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons.
 We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community.
-We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated virome-based random forest models identified significant changes in the virus community.
+We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated virome-based random forest models identified striking changes in the virus community.
 The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting the community was indirectly linked to colorectal cancer by modulating bacterial community structure and functionality.
 These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
-Together our findings add to the existing model for the role the microbiome plays in colorectal cancer development, thus providing us with a more complete understanding of colorectal cancer etiology.
+Together our findings add to the existing model for the microbiome's role in colorectal cancer development, thus providing a more complete understanding of colorectal cancer etiology.
 
-**Word Count**: 226
+**Word Count**: 226 / 250
 
@@
 Its progression and severity have been linked to colonic bacterial community composition.
-Little is known about cancer-associated colon virus communities and their influence on bacteria.
-We began addressing this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients, and how they compare to bacterial community changes.
-The altered colorectal cancer virome was used in virome-based machine learning models to accurately classify patient cancer status.
-The cancer-virus classifier was driven primarily by bacteriophages (bacterial viruses).
-Wider phage host ranges were more strongly associated with colorectal cancer.
-The results suggest an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.
+Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities.
+We began addressing this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients, and how they compared to bacterial community changes.
+The results suggested an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.
+These findings both support a biological role for viruses in colorectal cancer, as well as provide a new understanding of basic colorectal cancer etiology.
 
-**Word Count**: 125
+**Word Count**: 113 / 120
 
@@
 # Introduction
-Due to their mutagenic abilities and their propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
+Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
 Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer.
 The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g.
-cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
-The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggest that colorectal cancer may be linked to altered virus communities.
+cancer), and has already been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
+The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer together suggest that colorectal cancer may be associated with altered virus communities.
 
@@
 The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans will have died from the disease [@Siegel:2014jo].
-Although the majority of colorectal cancer case causes remains unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
-This work has led to the proposal of disease models in which bacteria colonize the colon, develop biofilms, promote infiltration, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
+Although the majority of colorectal cancer case causes remain unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
+This work has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote infiltration, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
 This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja].
@@
 Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.
-We evaluated colorectal cancer disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer developement: healthy, adenomatous, and cancerous.
+We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.
 
@@
 Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr], and survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz].
-Thus work such as this, which aims to power early detection and prevention of progression from early cancer stages, has a great potential to inform therapeutic development.
+Thus work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.
 
-In this study we report that colorectal cancer is associated with variation in colonic virus communities.
+Here we begin addressing the knowledge gap in whether virus communities are altered in colorectal cancer and, if they are, how those changes might impact cancer progression and severity.
+We also aim to evaluate their diagnostic potential.
+We report that colorectal cancer is associated with variation in colonic virus community composition.
 The majority of viruses identified within the virome were temperate bacteriophages.
-Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was also driven by a subset of influential phages.
-Our data suggest that the influential phages do not exclusively infect influential bacterial (e.g.
-*Fusobacterium* phages infecting the *Fusobacterium*), but rather act through the community as a whole.
+Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was driven by a subset of influential phages.
+Our data suggest that the influential phages do not exclusively infect influential bacterial, but rather act through the community as a whole.
 The implications of these findings are threefold.
@@
 
-## The Colorectal Cancer Virome Cohort
+## Cohort Design, Sample Collection, and Processing
 The study cohort consisted of 90 human subjects, 30 of which had healthy colons, 30 of which had adenomas, and 30 which had carcinomas **(Figure \ref{sampleproc})**.
 Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques.
-The other half of the stool samples were purified for virus like particles (VLPs) and consequent genomic DNA extraction, followed by shotgun metagenomic sequencing.
-The VLP purification allowed us to observe the *active virome* because we only sequenced those viruses that are encapsulated.
+The other half of the stool samples were purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing.
+The VLP purification allowed us to observe the *active virome* because we only sequenced those viruses that were encapsulated.
 
-Virus DNA was purified before sequencing, which allowed us to analyze DNA from within virus capsids **(Figure \ref{sampleproc})**.
-Each extraction protocol was performed with a blank control to detect any contaminants from reagents.
+Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**.
+Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources.
 Only one of the nine controls contained detectable DNA, which was of a minimal concentration, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
@@
 The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**.
-Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs; conceptually similar to 16S rRNA gene operational taxonomic units) with the majority containing hundreds of related contigs **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
+Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
 
-## Virus Diversity and Colorectal Cancer
+## Unaltered Virome Diversity in Colorectal Cancer
 Microbiome and disease associations are often described as being of an altered diversity (i.e.
 "dysbiotic").
-We therefore initially evaluated the diversity of virome OGUs and their association with colorectal cancer.
-We utilized the Bray-Curtis dissimilarity metric to evaluate differences in communities between disease states.
+We therefore initially evaluated the changes in virome OGU diversity in colorectal cancer.
+We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics.
 To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.
 
-We calculated the differences in virus alpha diversity associated with colorectal cancer.
-We found no significant alterations in either Shannon entropy or richness **(Figure \ref{betaogu} C-D)**.
+We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**.
 There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**).
-It is worth noting that there was a significant difference between the blank controls (those few that remained after sub-sampling) and the other study groups, further supporting the quality of our sample set (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**.
-Overall, standard diversity metrics were insufficient for capturing the differences in the virus communities between disease states.
+It is worth noting that there was a significant difference between the few blank controls that remained after subsampling, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**.
+This further supported the quality of our sample set.
+Overall, standard diversity metrics were insufficient for capturing virus community differences between disease states.
 
-## Virome Composition Changes in Colorectal Cancer
-Previous work has shown that 16S rRNA gene relative abundance profiles are effective features for classifying stool samples as originating from healthy, adenomatous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja].
+## Altered Virome Composition in Colorectal Cancer
+16S rRNA gene relative abundance profiles are effective feature sets for classifying stool samples as originating from healthy, adenomatous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja].
 The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer.
-Here we built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.
+We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.
 
-We built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals.
-These models were based on virus metagenomic community or bacterial 16S rRNA gene relative abundance profiles.
+To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals.
+These models were based on virus metagenomic community relative abundance profiles and were compared to 16S rRNA gene models.
 We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba].
 We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance.
-The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.
+Statistically, the viral model performed better than the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`), although the results were practically similar, with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.
 
-To evaluate the synergistic capabilities of the bacteria and viruses within the model, we built a combinatory model that used both bacterial and viral community data.
-The combination model yielded modest but significantly improved performance beyond the separate virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, and yielded an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
+To evaluate the synergistic capabilities of the bacteria and viruses, we built a combinatory model that used both bacterial and viral community data.
+The combination model yielded statistically significant but minor improved performance beyond the separate virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
 This suggested that features from the virus and bacterial communities may have had synergistic capabilities for classifying stool as belonging to cancerous individuals.
 
-We also compared these models to relative abundance profiles from bacterial metagenomic shotgun sequencing data.
-This model performed poorly, with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
+To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data.
+This model performed more poorly than the viral model, with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
+This performance was also much poorer than that observed in the 16S rRNA gene model.
 Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs **(Figure \ref{16scompare})**.
@@
 
-## Phage Influence Shifts During CRC Progression
-Because our cohort included healthy, adenomatous, and cancerous colons, we were able to gain insight into virus community shifts during cancer progression.
-We evaluated community shifts between the two disease transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the sample classes around the transitions.
-We found that, while bacterial 16S rRNA gene models perform equally well for all disease class comparisons, the virome models performed worse when seperating healthy from adenomatous, and better when seperating adenomatous from cancerous **(Figure \ref{transitionmodels} A-B)**.
-Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members drove the transitions from healthy to adenomatous and adenomatous to cancerous, with one phage exception **(Figure \ref{transitionmodels} C-E)**.
-A *Myoviridae* OGU (Cluster 188) was the seventh most important phage driving the transition from a healthy to adenomatous state, and was the most important phage driving the progression from an adenomatous to carcingenic state.
-Therefore, like bacteria, there are distinct phages assocaited with the transitions from a healthy to an adenomatous colon, and from an adenomatous to a cancerous colon.
+## Shifted Phage Influence Across Cancer Stages
+Because our cohort included healthy, adenomatous, and cancerous colons, we were able to evaluate shifts in influential phage identities during cancer progression.
+We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the sample classes around the transitions.
+While bacterial 16S rRNA gene models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**.
+Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important in the transitions from healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.
 
-## Inclusive Colonic Virome Classification
-After evaluating our ability to classify samples between two disease states, we performed a comprehensive three-class classification of all three disease states.
-We used a three-class random forest model for the bacterial 16S rRNA gene and viral sample sets.
-The bacterial signature model yielded an AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded an overall AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**).
-Both models were best able to classify cancer samples from healthy or precancerous samples, but struggled to distinguish adenomatous from healthy or cancerous **(Figure \ref{threewaymodel} A-B)**.
-The cancerous signal was the most discriminatory of the three sample types.
+After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states.
+The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**).
 
-The microbes important for the cancer vs healthy and healthy vs adenoma bacteria and virus models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
+The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
 The most important bacterium was the same *Fusobacterium* between the two and three class models, supporting its significance to the association between cancer and the bacterial communities **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**.
-The viruses most important to the three-class model were identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.
+The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.
 
 The classification model determined cancer state by incorporating the relative abundance profiles of the microbes within each community.
-The signatures ranged from notably high abundance of some OGUs, low abundance of some OGUs, and an absence of other OGUs **(Figure \ref{threewaymodel} F)**.
-Not all important OGUs were of increased abundance.
+Influential phage relative abundances were both higher and lower in cancer **(Figure \ref{threewaymodel} F)**.
+Not all important OGUs were of increased abundance in the diseased state.
 The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.
 
-## Dominance of Bacteriophages in Colorectal Cancer Virome
+## Bacteriophage Dominance in Colorectal Cancer Virome
 Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts.
@@
 
-We evaluated whether the phages in the community were primarily lytic (replicate by lysing their host) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell).
-We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes.
-This approach was done as described in previous work [@Minot:2011ez; @Hannigan:2015fz].
+We evaluated whether the phages in the community were primarily lytic (replicate by lysing their hosts) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell).
+We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz].
 We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
@@
 ## Community Context of Influential Phages
-Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were predators of the influential bacteria, and thus influenced their relative abundance through predation.
-If this hypothesis were true, we would expect the relative abundance between important bacteria and phages to be correlated.
+Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were primarily predators of the influential bacteria, and thus influenced their relative abundance through predation.
+If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacteria and phages.
 Instead we observed a strikingly low correlation between the bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**.
-There was an overall absence of correlations between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**.
-This evidence supported our null hypothesis that the influential phages are not primarily predators of influential bacteria.
+There was an overall absence of correlation between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**.
+This evidence supported our null hypothesis that the influential phages were not primarily predators of influential bacteria.
 
-Our next hypothesis was that the influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the select influential bacteria.
+Given these findings, we hypothesized that the influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
 This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network.
-We investigated the infectious capabilities and potential host ranges of the influential phages by building a machine learning model to predict which phage OGUs infect which bacteria in the overall community.
-The predicted interactions were then used to build a network which was used for subsequent analysis, as has been previously described (**Cite Network Preprint**).
-This analysis revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**.
+We investigated the infectious capabilities and potential host ranges of all phage OGUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**).
+The predicted interactions were then used to identify community hub phages.
+This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**.
 We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model.
-We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), indicating that phages important in driving colorectal cancer were also community hubs **(Figure \ref{network} B)**.
+The phages with high centrality values were defined as community hubs.
+We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**.
 Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.
@@
 Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Because some bacteria also have oncogenic properties, bacteriophages may play an indirect role in promoting carcinogenesis by altering bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
+Some bacteria also have oncogenic properties, meaning bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
 Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colorectal communities and colorectal cancer has yet to be evaluated.
-Here we show that, like colonic bacterial communities, the colon virome is associated with colorectal cancer.
+Here we show that, like colonic bacterial communities, the colon virome was altered in colorectal cancer.
 Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.
 
-Together our data allowed us to begin modeling the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**.
+Our data allowed us to begin delineating the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**.
 We found basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states.
-By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus communities and colorectal cancer.
-Colorectal cancer was primarily assocaited with altered bacteriophage communities.
-These phage communities were not exclusive predators of the most influential bacteria (e.g.
-*Fusobacterium* linked to decreases in *Fusobacterium* phages), as demonstrated by the lack of correlation between the abundance of these entities.
-Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead directly modulating the influential bacteria.
+By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer.
+The colorectal cancer virome was composed primarily of bacteriophages.
+These phage communities were not exclusive predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundance of these entities.
+Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria.
 Our previous work has shown that modifying colon bacterial communities with antibiotics alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en].
-This provides a precident for phage indirectly influencing colorectal cancer progression y altering the bacterial community composition.
-Overall our data supports a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
-Although our evidence suggests phages indirectly influence colorectal cancer developement, we are not able to rule out the role of phages directly interacting with the human host.
+This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition.
+Overall, our data supports a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
+Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host.
 
 In addition to modeling the basic potential connections between virus communities, bacteria communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur.
-This was done by incorporating our findings into the Flynn model for colorectal cancer developement **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
+This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
 We hypothesize that the broadly infectious phages in the colon began lysing, and thereby disrupting, the bacterial communities to open a niche in which opportunistic bacteria (such as *Fusobacterium nucleatum*) were able to colonize.
-Once the influential "driver" bacteria had established themselves in the epithelium, other opporunistic "passenger" bacteria were able to adhere to the driver, colonize, and begin establishing a biofilm.
-Phages may have played a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process shown to be important for effective biofilm growth [@Rossmann:2015cj].
+Once the initial influential bacteria had established themselves in the epithelium, other opportunistic bacteria were able to adhere to the driver, colonize, and begin establishing a biofilm.
+Phages may have played a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj].
 The oncogenic bacteria were then able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response.
-As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria important to our cancer classification model.
-As the bacteria entered their oncogenic synergy with the epithelium, we hypothesize the phages could continue mediating biofilm dispersal, as well as support the colonized oncogenic bacteria by lysing competing cells to maintain the niche and provide nutrients to the other bacteria.
-In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide the future, mostly functional studies, by our group and others, into the role the virome plays in colorectal cancer.
+As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundance was most influential.
+As the bacteria entered their oncogenic synergy with the epithelium, we hypothesize the phages could have continued mediating biofilm dispersal, as well as support the colonized oncogenic bacteria by lysing competing cells to maintain the niche and provide nutrients to the other bacteria.
+In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future into the role the virome plays in colorectal cancer.
 
 A notable observation from our analysis was the lack of performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences.
-This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies, and the advantage of 16S rRNA over whole metagenomic shotgun sequencing.
-We found that there were six bacterial OTUs that drove the performance of the 16S rRNA classification model, and these OTUs were all sparsely present and lowly abundant.
+This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies, and the advantage of 16S rRNA gene sequencing over whole metagenomic shotgun sequencing.
+We found that there were six bacterial OTUs that drove the performance of the 16S rRNA gene classification model, and these OTUs were all sparsely present and lowly abundant.
 Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model.
 The bacterial metagenomic OGUs represented only the most abundant taxa, which was uninformative for this application.
-There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but these approaches did not utilize OGU clustering like we did here, and the models only performed **as well** as the 16S rRNA model [@Zeller:2014ix].
-Thus the targeted 16S rRNA sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples.
+There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but these approaches did not utilize OGU clustering like we did here, and the models only performed **as well** as the 16S rRNA gene model [@Zeller:2014ix].
+Thus the targeted 16S rRNA gene sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples.
 Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
@@
 ## Analysis Source Code & Availability
-All associated source code and Makefile are available for review at the following GitHub repository: https://github.com/SchlossLab/Hannigan-2016-ColonCancerVirome.
+All associated source code is available at the following GitHub repository: https://github.com/SchlossLab/Hannigan-2016-ColonCancerVirome.
 
@@
 B) Quantification of the AUC variation for each model, and how it compares to each of the other models based on 15 iterations.
-A pairwise Wilcoxon test with a Bonferroni multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01).
-C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S classification model when removed from the classification model.
-Results based on 25 iterations.
-OTU features are colored by taxonomic identity.
+A pairwise Wilcoxon test with a false discovery rate multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01).
+C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S rRNA gene classification model when removed from the classification model.
+Results based on 15 iterations.
+Mean is represented by a point, and bars represent standard error.
 D) Mean decrease in accuracy of each operational genomic unit in the virome classification model.
-E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S and virome features.*\label{predmodel}](../figures/predmodel-viromebacteria.pdf){ width=100% }
+E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S rRNA gene and virome features.*\label{predmodel}](../figures/predmodel-viromebacteria.pdf){ width=100% }
 
@@
 B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OGU.
-A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, r = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}](../figures/NetworkAndScatter.pdf){ width=80% }
+A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}](../figures/NetworkAndScatter.pdf){ width=80% }
 
@@
 
-![*Working hypothesis of how the bacteriophage community is associated with colorectal cancer and the associated bacterial community.*\label{modelsummary}](../figures/WorkingHypothesis.pdf){ width=100% }
+![*Final model and working hypothesis from this study.
+A) Basic model illustrating the connections between the virome, bacterial communities, and colorectal cancer.
+B) Working hypothesis of how the bacteriophage community is associated with colorectal cancer and the associated bacterial community.*\label{modelsummary}](../figures/WorkingHypothesis.pdf){ width=100% }
 
@@
 B) Density plot (analogous to histogram) of the number of virome operational genomic units containing the specific number of contigs, as indicated by the x-axis.
-C-D) Sample plots as panels C and D, but for the whole metagenomic sample set.*\label{clustercontigqc}](../figures/contiglengthbycluster.pdf){ width=75% }
+C-D) Sample plots as panels C and D, but for the whole metagenomic sample set.*\label{clustercontigqc}](../figures/contiglengthbycluster.png){ width=75% }
 
@@
 
-\newpage
-
-![*Transition of colorectal cancer importance through disease progression.*\label{transitionmodels}](../figures/temporalmodels.pdf){ width=80% }
+![*Transition of colorectal cancer importance through disease progression.
+A) Virus and B) 16S rRNA gene model performance (AUC) when discriminating all binary combinations of disease types.
+Blue line represents mean performance from multiple random iterations.
+C-E) Top ten important phage OGUs when classifying each combination of disease state, as measured by the mean decrease in accuracy metric.
+Mean is represented by a point, and bars represent standard error.
+Disease comparison is specified in the top left corner of each panel.
+F-H) Top ten important bacterial 16S rRNA gene OTUs for classifying each disease state combination.*\label{transitionmodels}](../figures/temporalmodels.pdf){ width=75% }
 

### 2017-03-26 doc/manuscript.Rmd

@@
 Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world.
-Although the majority of colorectal cancer case causes remain unclear, its progression has been linked to changes in colonic bacterial community composition.
-Viruses are another important component of the colonic microbial community, but have yet to be studied in colorectal cancer, despite their oncogenic potential.
+Although etiology for the majority of colorectal cancer cases remain unclear, its progression has been linked to changes in colonic bacterial community composition.
+Viruses are another important component of the colonic microbial community.
+However, they have yet to be studied in colorectal cancer, despite their oncogenic potential.
 We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons.
 We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community.
-We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated virome-based random forest models identified striking changes in the virus community.
-The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting the community was indirectly linked to colorectal cancer by modulating bacterial community structure and functionality.
+We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated random forest models identified striking changes in the virus community.
+The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function.
 These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
-Together our findings add to the existing model for the microbiome's role in colorectal cancer development, thus providing a more complete understanding of colorectal cancer etiology.
 
-**Word Count**: 226 / 250
+**Word Count**: 201 / 250
 
@@
 Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities.
-We began addressing this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients, and how they compared to bacterial community changes.
+We addressed this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients and how they compared to bacterial community changes.
 The results suggested an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.
@@
 
-**Word Count**: 113 / 120
+**Word Count**: 111 / 120
 
@@
 The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g.
-cancer), and has already been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
-The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer together suggest that colorectal cancer may be associated with altered virus communities.
+cancer).[Comment: the gut virome isn't associted with the following diseases...] The human virome has already been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
+The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggests that colorectal cancer may be associated with altered virus communities.
 
 Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo].
-The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans will have died from the disease [@Siegel:2014jo].
-Although the majority of colorectal cancer case causes remain unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
-This work has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote infiltration, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
+The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans died from the disease [@Siegel:2014jo].
+Although the etiology for [cite specific number - 80%?] of colorectal cancer cases remains unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
+This work has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
 This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja].
@@
 
-Here we begin addressing the knowledge gap in whether virus communities are altered in colorectal cancer and, if they are, how those changes might impact cancer progression and severity.
-We also aim to evaluate their diagnostic potential.
+Here we address the gap in knowledge of whether virus communities are altered in colorectal cancer and, if they are, how those changes might impact cancer progression and severity.
+We also aimed to evaluate their potential as a reservoir of diagnostic biomarkers.
 We report that colorectal cancer is associated with variation in colonic virus community composition.
@@
 *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.
+[Comment: I worry that from "We report..." to the end of the paragraph is too much a rehash of the Results.
+I would suggest moving this section to the abstract since the abstract is a bit light on results.
+]
 
@@
 Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources.
-Only one of the nine controls contained detectable DNA, which was of a minimal concentration, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
-As was expected, these controls were sparsely sequenced and were mostly removed while sub-sampling to even depths **(Figure \ref{qualcontrol} B)**.
+Only one of the nine controls contained DNA at a minimal concentration (XX ng/ul), thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
+As was expected, these controls yielded few sequences and were mostly removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
 The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**.
 Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
+[Comment: What do you think about using operational viral units (OVU) rather than OGU?
+I think of OGU as representing bacterial genomes.
+Adding phage to the name would make it more clear what we are refering to]
 
@@
 We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics.
-To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.
-
-We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**.
+~~To control for uneven sequencing depths, we rarefied to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.~~ [Comment: put this in the methods section rarefied != sub-sampled, make sure you rarefied (i.e.
+subsampled 1000 times)] We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**.
 There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**).
-It is worth noting that there was a significant difference between the few blank controls that remained after subsampling, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**.
+It is worth noting that there was a significant difference between the few blank controls that remained after rarefying the data, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**.
 This further supported the quality of our sample set.
-Overall, standard diversity metrics were insufficient for capturing virus community differences between disease states.
+Similar to what was previously seen when analyzing the 16S rRNA gene sequences from the same samples [@Zackular:2014fba], standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states.
 
 ## Altered Virome Composition in Colorectal Cancer
-16S rRNA gene relative abundance profiles are effective feature sets for classifying stool samples as originating from healthy, adenomatous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja].
+16S rRNA gene relative abundance profiles are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
 The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer.
-We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.
+We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared their performance to models built using bacterial community signatures.
 
-To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals.
-These models were based on virus metagenomic community relative abundance profiles and were compared to 16S rRNA gene models.
+To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons.
 We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba].
@@
 
-To evaluate the synergistic capabilities of the bacteria and viruses, we built a combinatory model that used both bacterial and viral community data.
-The combination model yielded statistically significant but minor improved performance beyond the separate virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
-This suggested that features from the virus and bacterial communities may have had synergistic capabilities for classifying stool as belonging to cancerous individuals.
+[Comment:
+1A: Get rid of the legend title "Disease". I feel like the 0.1 and 0.9 marks on the ROC curves adds clutter without improving interpretation.
+1B: Can you order: Bacteria, Virus, Combined, Metagenome - why is the combined AUC > virus when it seems the opposite in A? Can you put the AUCs in the legend for 1A and get rid of 1B?
+1C: Format OTU names to be like "OTU 4" rather than "Otu00004"
+1D: Format cluster  names to be like "Cluster 115" rather than "Cluster_115"
+1E: See comments for 1C and 1D
+Can you justify the y-axis in C, D, and E so they are vertically aligned and the x-axes are the same lengths?
+]
+
+To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data.
+The combined model yielded statistically significant but small performance improvement beyond the viral (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
+This suggested that features from the virus and bacterial communities interact to improve the ability to classify stool as belonging to individuals with cancerous colons.
 
 To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data.
-This model performed more poorly than the viral model, with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
-This performance was also much poorer than that observed in the 16S rRNA gene model.
-Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs **(Figure \ref{16scompare})**.
-Filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**.
+This model performed more poorly than any of the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`) **(Figure \ref{predmodel} A - B)**.
+Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**.
+Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**.
 The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.
+[Comment: Can you add a comment about the nature of the features in the viral model?]
 
 The association between the two communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed.
-*Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu].
-The virome signature was also driven by a few operational genomic units, suggesting a role for the OGUs in cancer development **(Figure \ref{predmodel} D)**.
+*Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu].
+The virome signature was also driven by a few OGU, suggesting a role for the OGUs in cancer development **(Figure \ref{predmodel} D)**.
 The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified").
@@
 ## Shifted Phage Influence Across Cancer Stages
-Because our cohort included healthy, adenomatous, and cancerous colons, we were able to evaluate shifts in influential phage identities during cancer progression.
-We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the sample classes around the transitions.
+Because our cohort included people with healthy, adenomatous, and cancerous colons we were able to evaluate shifts in influential phage identities during cancer progression.
+We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions.
 While bacterial 16S rRNA gene models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**.
@@
 The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**).
+The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
+The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**.
+The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.
+{Influential phage relative abundances were both higher and lower in cancer **(Figure \ref{threewaymodel} F)**.
+Not all important OGUs were of increased abundance in the diseased state.
+The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.} [Comment: I don't feel like the text between the {}'s says anything.
+It's like word salad.]
 
-The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
-The most important bacterium was the same *Fusobacterium* between the two and three class models, supporting its significance to the association between cancer and the bacterial communities **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**.
-The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.
-
-The classification model determined cancer state by incorporating the relative abundance profiles of the microbes within each community.
-Influential phage relative abundances were both higher and lower in cancer **(Figure \ref{threewaymodel} F)**.
-Not all important OGUs were of increased abundance in the diseased state.
-The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.
+[Comment: Can't distinguish colors in Fig S9F (and axis label overlaps panel label)]
 
@@
 We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
-Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes.
-These findings are consistent with previous reports suggesting the gut virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
+Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses indicating that the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes.
+These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
 
@@
 
-Given these findings, we hypothesized that the influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
+[ Comment: how are the rows and columns sorted in Figure 2? Looks like by importance, how about sorting them by co-occurrence? ]
+
+Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
 This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network.
@@
 The phages with high centrality values were defined as community hubs.
-We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**.
+Phage OGU centrality was significantly and positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**.
 Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.
@@
 The colorectal cancer virome was composed primarily of bacteriophages.
-These phage communities were not exclusive predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundance of these entities.
+These phage communities were not exclusive predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations.
 Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria.
-Our previous work has shown that modifying colon bacterial communities with antibiotics alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en].
+Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter(the Microbiome article)].
 This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition.
@@
 As the bacteria entered their oncogenic synergy with the epithelium, we hypothesize the phages could have continued mediating biofilm dispersal, as well as support the colonized oncogenic bacteria by lysing competing cells to maintain the niche and provide nutrients to the other bacteria.
-In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future into the role the virome plays in colorectal cancer.
+In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role of the virome in colorectal cancer.
 
-A notable observation from our analysis was the lack of performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences.
+A notable observation from our analysis was the poor performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences.
 This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies, and the advantage of 16S rRNA gene sequencing over whole metagenomic shotgun sequencing.
@@
 Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model.
-The bacterial metagenomic OGUs represented only the most abundant taxa, which was uninformative for this application.
+The bacterial metagenomic OGUs represented only the most abundant taxa, which was not informative for this application.
 There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but these approaches did not utilize OGU clustering like we did here, and the models only performed **as well** as the 16S rRNA gene model [@Zeller:2014ix].
@@
 
-In addition to the therapeutic ramifications for understanding the colorectal cancer microbiome, our findings provide a proof-of-principle that viruses, while under-appreciated and understudied in the human microbiome, are an important contributer to human disease that has the potential to provide an abundance of information that supplements that of bacterial communities.
+In addition to the therapeutic ramifications for understanding the colorectal cancer microbiome, our findings provide a proof-of-principle that viruses, while under-appreciated and understudied in the human microbiome, are an important contributor to human disease that has the potential to provide an abundance of information that supplements that of bacterial communities.
 Evidence has suggested that the virome is a crucial component to the microbiome and that bacteriophages are important players.
@@
 
-## 16S Data Acquisition & Processing
+## 16S rRNA Gene Sequence Data Acquisition & Processing
 The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba].
 Sequence (fastq) and metadata files were downloaded from http://www.mothur.org/MicrobiomeBiomarkerCRC.
-The 16S rRNA gene sequences were analyzed as described previously, relying on the Mothur analytical toolkit (v1.37.0) [@Sze:2016iw; @Schloss:2009do].
-Briefly, the sequences were de-replicated, screened for chimeras using UCHIME [@Edgar:2011gy] and the SILVA database [@Pruesse:2007jc], and binned into operational taxonomic units (OTUS) using a 97% similarity threshold.
-Abundance was normalized for uneven sequencing depth by randomly sub-sampling to 10,000 sequences, as previously reported [@Baxter:2016dja].
+The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:XXXXXX; @Schloss:2009do].
+Briefly, the sequences were de-replicated, aligned to the SILVA database [@Pruesse:2007jc], screened for chimeras using UCHIME [@Edgar:2011gy], and binned into operational taxonomic units (OTUs) using a 97% similarity threshold.
+Abundances were normalized for uneven sequencing depth by randomly sub-sampling to 10,000 sequences, as previously reported [@Baxter:2016dja].
 
@@
 Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz].
-Briefly, an aliquot of stool (~0.1g) was resuspended in SM buffer and vortexed to facilitate resuspension.
-The resuspended stool was centrifuged to remove major particulate debris, followed by filtering through a 0.22µm filter to remove smaller contaminants.
+Briefly, an aliquot of stool (ca.
+0.1g) was resuspended in SM [Comment: Define SM] buffer and vortexed to facilitate resuspension.
+The resuspended stool was centrifuged to remove major particulate debris, followed by filtering through a 0.22-µm filter to remove smaller contaminants.
 The filtered supernatant was treated with chloroform to lyse contaminating cells including bacteria, human, fungi, etc.
@@
 Abundance was corrected for contig reference length and the number of contigs included in each operational genomic unit.
-Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1e6 and 5e5 reads, respectively, and removing samples with less total samples than the threshold.
+Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1,000,000 and 500,000 reads, respectively, and removing samples with less total samples than the threshold.
 Thresholds were set for maximizing sequence information while minimizing numbers of lost samples.
@@
 Alpha diversity was calculated using the Shannon Entropy and Richness metrics.
-Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (Anosim) with a post-hoc multivariate Tukey test.
+Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (ANOSIM) with a post-hoc multivariate Tukey test.
 All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE.
@@
 # Acknowledgments
-We thank the members of the Schloss lab for their underlying contributions. GD Hannigan was supported in part by the Michigan Molecular Mechanisms of Microbial Pathogenesis Fellowship.
+The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
+GD Hannigan was supported by the Michigan Molecular Mechanisms of Microbial Pathogenesis Fellowship.
+PD Schloss was supported by funding from the National Institutes of Health (P30DK034933).
+MT Ruffin was supported by funding from the National Institutes of Health (5U01CA86400).
 
@@
 
-![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S signatures, whole metagenomic signatures, and a combination of virome and 16S signatures.
-A) ROC curve for visualizing the performance of each of the models for classifying stool as coming from either a cancerous or healthy individual.
-B) Quantification of the AUC variation for each model, and how it compares to each of the other models based on 15 iterations.
+![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S rRNA gene sequence signatures, whole metagenomic signatures, and a combination of virome and 16S rRNA gene sequence signatures.
+A) ROC curve for visualizing the performance of each of the models for classifying stool as coming from either an individual with a cancerous or healthy colon.
+B) Quantification of the AUC variation for each model, and how it compared to each of the other models based on 15 iterations.
 A pairwise Wilcoxon test with a false discovery rate multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01).
@@
 A) NMDS ordination of community samples, colored for cancerous (green), pre-cancerous (red), and healthy (yellow).
-B) Differences in means between disease group centroids with 95% confidence intervals based on an Anosim test with a post hoc multivariate Tukey test.
+B) Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test.
 Comparisons (indicated on y-axis) in which the intervals cross the zero mean difference line (dashed line) were not significantly different.
@@
 A) NMDS ordination of community samples, colored by disease state.
-B) Differences in means between disease group centroids with 95% confidence intervals based on an Anosim test with a post hoc multivariate Tukey test.
+B) Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test.
 Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}](../figures/diversity-beta-ogu-negative.pdf){ width=100% }

### 2017-03-26 bin/predictVirome.R
Updated code.

@@ -453,7 +453,7 @@ GetAverageAUC <- function(x, y) {
 	return(resultdf)
 }
 
-iterationcount <- 25
+iterationcount <- 15
 
 viromeauc <- lapply(c(1:iterationcount), function(i) GetAverageAUC(absmissingid, i))
 viromeaucdf <- ldply(viromeauc, data.frame)
@@ -473,7 +473,7 @@ metagenomeaucdf$class <- "Metagenomic"
 
 megatron <- rbind(viromeaucdf, bacteriaaucdf, metagenomeaucdf, comboaucdf)
 
-aucstat <- pairwise.wilcox.test(x=megatron$highAUC, g=megatron$class, p.adjust.method="bonferroni")
+aucstat <- pairwise.wilcox.test(x=megatron$highAUC, g=megatron$class, p.adjust.method="BH")
 aucpv <- melt(aucstat$p.value)
 da <- as.data.frame(sort(unique(c(as.character(aucpv[,1]), as.character(aucpv[,2])))))
 colnames(da) <- "name"

### 2017-03-28 doc/manuscript.Rmd
Huge updates incorporating Pats comments.

@@
 The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function.
+Our data suggested that the influential phages do not exclusively infect influential bacterial, but rather act through the community as a whole.
 These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
 
-**Word Count**: 201 / 250
+**Word Count**: 224 / 250
 
@@
 The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g.
-cancer).[Comment: the gut virome isn't associted with the following diseases...] The human virome has already been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
+cancer).
+Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
 The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggests that colorectal cancer may be associated with altered virus communities.
@@
 We also aimed to evaluate their potential as a reservoir of diagnostic biomarkers.
-We report that colorectal cancer is associated with variation in colonic virus community composition.
-The majority of viruses identified within the virome were temperate bacteriophages.
-Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was driven by a subset of influential phages.
-Our data suggest that the influential phages do not exclusively infect influential bacterial, but rather act through the community as a whole.
-The implications of these findings are threefold.
+The implications of this study are threefold.
 *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process.
@@
 *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.
-[Comment: I worry that from "We report..." to the end of the paragraph is too much a rehash of the Results.
-I would suggest moving this section to the abstract since the abstract is a bit light on results.
-]
 
@@
 Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources.
-Only one of the nine controls contained DNA at a minimal concentration (XX ng/ul), thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
+Only one of the nine controls contained DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
 As was expected, these controls yielded few sequences and were mostly removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
@@
 We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics.
-~~To control for uneven sequencing depths, we rarefied to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.~~ [Comment: put this in the methods section rarefied != sub-sampled, make sure you rarefied (i.e.
-subsampled 1000 times)] We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**.
+We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**.
 There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**).
@@
 
+<!--
 [Comment:
@@
 ]
+-->
 
@@
 The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.
-[Comment: Can you add a comment about the nature of the features in the viral model?]
+Although the features in the viral model were also of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.
 
@@
 The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.
-{Influential phage relative abundances were both higher and lower in cancer **(Figure \ref{threewaymodel} F)**.
-Not all important OGUs were of increased abundance in the diseased state.
-The viral classification model depended on the unique signatures of these different abundance profiles to accurately classify each sample.} [Comment: I don't feel like the text between the {}'s says anything.
-It's like word salad.]
+Not all important OGUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.
 
@@
 Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (ANOSIM) with a post-hoc multivariate Tukey test.
-All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE.
+All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE].
 
@@
 ![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S rRNA gene sequence signatures, whole metagenomic signatures, and a combination of virome and 16S rRNA gene sequence signatures.
-A) ROC curve for visualizing the performance of each of the models for classifying stool as coming from either an individual with a cancerous or healthy colon.
+A) An example ROC curve for visualizing the performance of each of the models for classifying stool as coming from either an individual with a cancerous or healthy colon.
 B) Quantification of the AUC variation for each model, and how it compared to each of the other models based on 15 iterations.

### 2017-03-28 doc/manuscript.Rmd
Made updates.

@@
 Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world.
-Although etiology for the majority of colorectal cancer cases remain unclear, its progression has been linked to changes in colonic bacterial community composition.
-Viruses are another important component of the colonic microbial community.
-However, they have yet to be studied in colorectal cancer, despite their oncogenic potential.
+Its progression has been linked to changes in colonic bacterial community composition.
+Viruses are another important component of the colonic microbial community, however they have yet to be studied in colorectal cancer despite their oncogenic potential.
 We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons.
@@
 The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function.
-Our data suggested that the influential phages do not exclusively infect influential bacterial, but rather act through the community as a whole.
+Our data suggested that the influential phages did not exclusively infect influential bacterial, but rather acted through the community as a whole.
 These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
 
-**Word Count**: 224 / 250
+**Word Count**: 213 / 250
 
@@
 
-**Word Count**: 111 / 120
+**Word Count**: 112 / 120
 
@@
 Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
-The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggests that colorectal cancer may be associated with altered virus communities.
+The strong association of bacterial communities with colorectal cancer and the precedence for the virome in impacting other human diseases together suggest that colorectal cancer may be associated with altered virus communities.
 
 Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo].
-The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans died from the disease [@Siegel:2014jo].
-Although the etiology for [cite specific number - 80%?] of colorectal cancer cases remains unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
+The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans have died from the disease [@Siegel:2014jo].
+One component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
 This work has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
@@
 
-Here we address the gap in knowledge of whether virus communities are altered in colorectal cancer and, if they are, how those changes might impact cancer progression and severity.
+Here we address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those changes might impact cancer progression and severity.
 We also aimed to evaluate their potential as a reservoir of diagnostic biomarkers.
@@
 Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources.
-Only one of the nine controls contained DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
+Only one of the nine controls contained detectable DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
 As was expected, these controls yielded few sequences and were mostly removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
 The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**.
-Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
-[Comment: What do you think about using operational viral units (OVU) rather than OGU?
-I think of OGU as representing bacterial genomes.
-Adding phage to the name would make it more clear what we are refering to]
+Because contigs only represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
 
@@
 "dysbiotic").
-We therefore initially evaluated the changes in virome OGU diversity in colorectal cancer.
+We therefore initially evaluated the changes in virome OVU diversity in colorectal cancer.
 We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics.
@@
 
-<!--
-[Comment:
-1A: Get rid of the legend title "Disease". I feel like the 0.1 and 0.9 marks on the ROC curves adds clutter without improving interpretation.
-1B: Can you order: Bacteria, Virus, Combined, Metagenome - why is the combined AUC > virus when it seems the opposite in A? Can you put the AUCs in the legend for 1A and get rid of 1B?
-1C: Format OTU names to be like "OTU 4" rather than "Otu00004"
-1D: Format cluster  names to be like "Cluster 115" rather than "Cluster_115"
-1E: See comments for 1C and 1D
-Can you justify the y-axis in C, D, and E so they are vertically aligned and the x-axes are the same lengths?
-]
--->
-
 To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data.
@@
 *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu].
-The virome signature was also driven by a few OGU, suggesting a role for the OGUs in cancer development **(Figure \ref{predmodel} D)**.
+The virome signature was also driven by a few OVUs, suggesting a role for the viruses in cancer development **(Figure \ref{predmodel} D)**.
 The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified").
@@
 The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.
-Not all important OGUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.
-
-[Comment: Can't distinguish colors in Fig S9F (and axis label overlaps panel label)]
+Not all important OVUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.
 
@@
 Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts.
-To better understand the types of viruses that are important for colorectal cancer, we identified the virome OGUs as being similar to either eukaryotic viruses or bacteriophages.
+To better understand the types of viruses that are important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages.
 The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**.
-Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes.
+Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes.
 
 We evaluated whether the phages in the community were primarily lytic (replicate by lysing their hosts) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell).
-We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz].
+We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz].
 We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
-Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses indicating that the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes.
+Thus the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that can lie dormant in bacterial genomes.
 These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
@@
 Instead we observed a strikingly low correlation between the bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**.
-There was an overall absence of correlation between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**.
+There was an overall absence of correlation between the most influential phage OVUs and bacterial OTUs **(Figure \ref{correlations} B)**.
 This evidence supported our null hypothesis that the influential phages were not primarily predators of influential bacteria.
-
-[ Comment: how are the rows and columns sorted in Figure 2? Looks like by importance, how about sorting them by co-occurrence? ]
 
@@
 This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network.
-We investigated the infectious capabilities and potential host ranges of all phage OGUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**).
+We investigated the infectious capabilities and potential host ranges of all phage OVUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**).
 The predicted interactions were then used to identify community hub phages.
 This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**.
-We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model.
+We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OVU's connection to the rest of the network, and compared the centrality to the importance of each OVU in the colorectal cancer classification model.
 The phages with high centrality values were defined as community hubs.
-Phage OGU centrality was significantly and positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**.
+Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**.
 Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.
@@
 Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria.
-Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter(the Microbiome article)].
+Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb].
 This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition.
@@
 ## Analysis Source Code & Availability
-All associated source code is available at the following GitHub repository: https://github.com/SchlossLab/Hannigan-2016-ColonCancerVirome.
+All associated source code is available at the following GitHub repository:
+
+https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017
 
@@
 Sequence (fastq) and metadata files were downloaded from http://www.mothur.org/MicrobiomeBiomarkerCRC.
-The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:XXXXXX; @Schloss:2009do].
+The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:2013db; @Schloss:2009do].
 Briefly, the sequences were de-replicated, aligned to the SILVA database [@Pruesse:2007jc], screened for chimeras using UCHIME [@Edgar:2011gy], and binned into operational taxonomic units (OTUs) using a 97% similarity threshold.
@@
 Briefly, an aliquot of stool (ca.
-0.1g) was resuspended in SM [Comment: Define SM] buffer and vortexed to facilitate resuspension.
+0.1g) was resuspended in standard SM buffer (Crystalgen) and vortexed to facilitate resuspension.
 The resuspended stool was centrifuged to remove major particulate debris, followed by filtering through a 0.22-µm filter to remove smaller contaminants.
@@
 ## Operational Genomic Unit Classification
-Much like operational taxonomic units (OGUs) are used as an operational definition of similar 16S rRNA gene sequences in absence of taxonomic identification, we operationally defined closely related contig sequences as operational genomic units (OGUs) in the absence of taxonomic identity.
-OGUs were defined with the CONCOCT algorithm (v0.4.0) which bins related contigs by similar tetra-mer and co-abundance profiles within samples using a variational Bayesian approach [@Alneberg:2014fc].
+Much like operational taxonomic units (OTUs) are used as an operational definition of similar 16S rRNA gene sequences in absence of taxonomic identification, we operationally defined closely related bacterial contig sequences as operational genomic units (OGUs) and virus contigs as operational viral units (OVUs) in the absence of taxonomic identity.
+OGUs and OVUs were defined with the CONCOCT algorithm (v0.4.0) which bins related contigs by similar tetra-mer and co-abundance profiles within samples using a variational Bayesian approach [@Alneberg:2014fc].
 CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria due to computational limitations.
@@
 ## Taxonomic Identification of Operational Genomic Units
-Viral operational genomic units (OGUs) were identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
+Operational viral units (OVUs) were identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
 The longest contiguous sequence in each operational genomic unit was used as a representative sequence for classification.
@@
 ## Phage Replication Style Identification
-Phage OGU replication sytle was identified using methods described previously [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
-Briefly, we identified lysogenic phage OGUs as representative contigs containing at least one of three genomic markers: 1) phage integrase genes, 2) prophage genes from the ACLAME database, 3) genomic similarity to bacterial reference genomes.
-Integrase genes were identified in phage OGU representative contigs by aligning the contigs to a reference database of all known phage integrase genes from the Uniprot database (Uniprot search term: "organism:phage gene:int NOT putative").
+OVU replication sytle was identified using methods described previously [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
+Briefly, we identified lysogenic OVUs as representative contigs containing at least one of three genomic markers: 1) phage integrase genes, 2) prophage genes from the ACLAME database, 3) genomic similarity to bacterial reference genomes.
+Integrase genes were identified in phage OVU representative contigs by aligning the contigs to a reference database of all known phage integrase genes from the Uniprot database (Uniprot search term: "organism:phage gene:int NOT putative").
 Prophage genes were identified in the same way, using the ACLAME set of reference prophage genes.
@@
 
-![*Relative abundance correlations between bacterial OTUs and virome OGUs.
-A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OGUs (y-axis) with blue being positively correlated and red being negatively correlated.
-Operational units are organized by importance in their colorectal cancer classification models, such that the most important units are in the top left corner.
-B) Magnification of the boxed region in pannel (A), highlighting the correlation between the most important bacterial OTUs and virome OGUs.
+![*Relative abundance correlations between bacterial OTUs and virome OVUs.
+A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OVUs (y-axis) with blue being positively correlated and red being negatively correlated.
+Bar plots indicate the viral (left) and bacterial (bottom) operational unit importance in their colorectal cancer classification models, such that the most important units are in the top left corner.
+B) Magnification of the boxed region in pannel (A), highlighting the correlation between the most important bacterial OTUs and virome OVUs.
 The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as pannel (A).
-C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OGUs.*\label{correlations}](../figures/phage-bacteria-cor.pdf){ width=100% }
+C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OVUs.*\label{correlations}](../figures/phage-bacteria-cor.pdf){ width=100% }
 
@@
 A) Visualization of the community network for our colorectal cancer cohort.
-B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OGU.
+B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OVU.
 A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}](../figures/NetworkAndScatter.pdf){ width=80% }
@@
 A) Heated scatter plot demonstrating the distribution of contig coverage (number of sequences mapping to each contig) and contig length for the virus metagenomic sample set.
-B) Scatter plot illustrating the distribution of operational genomic unit (OGU) length and sequence coverage for the virus metagenomic sample set.
+B) Scatter plot illustrating the distribution of operational viral unit (OVU) length and sequence coverage for the virus metagenomic sample set.
 C) Heated scatter plot demonstrating the distribution of contig coverage and length for the whole metagenomic sample set.
@@
 Blue line represents mean performance from multiple random iterations.
-C-E) Top ten important phage OGUs when classifying each combination of disease state, as measured by the mean decrease in accuracy metric.
+C-E) Top ten important phage OVUs when classifying each combination of disease state, as measured by the mean decrease in accuracy metric.
 Mean is represented by a point, and bars represent standard error.
@@
 Results based on 25 iterations.
-F) Relative abundance of the six most important virome OGUs in the model, with the most important on the right.
+F) Relative abundance of the six most important virome OVUs in the model, with the most important on the right.
 Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=80% }
@@
 ![*Lysogenic phage relative abundance in disease states.
-Phage OGUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
+Phage OVUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
 No disease groups were statistically significant.*\label{replicationstyles}](../figures/lysogenic-relabund.pdf){ width=50% }

### 2017-03-29 doc/manuscript.Rmd
More manuscript updates.

@@
 The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function.
-Our data suggested that the influential phages did not exclusively infect influential bacterial, but rather acted through the community as a whole.
+Our data suggested that the influential phages did not exclusively infect influential bacteria, but rather acted through the community as a whole.
 These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
@@
 # Significance Statement
+
 Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide.
@@
 We addressed this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients and how they compared to bacterial community changes.
-The results suggested an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.
-These findings both support a biological role for viruses in colorectal cancer, as well as provide a new understanding of basic colorectal cancer etiology.
+The results suggested an indirect role for the virome in impacting colorectal cancer by modulating their associated bacterial community.
+These findings both support a biological role for viruses in colorectal cancer and provide a new understanding of basic colorectal cancer etiology.
 
-**Word Count**: 112 / 120
+**Word Count**: 111 / 120
 
@@
 Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
-The strong association of bacterial communities with colorectal cancer and the precedence for the virome in impacting other human diseases together suggest that colorectal cancer may be associated with altered virus communities.
+The strong association of bacterial communities with colorectal cancer and the precedence for the virome in impacting other human diseases suggest that colorectal cancer may be associated with altered virus communities.
 
@@
 The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans have died from the disease [@Siegel:2014jo].
-One component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
-This work has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
+An important component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
+Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
 This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja].
-While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome.
+While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome.
 Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.
@@
 Here we address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those changes might impact cancer progression and severity.
-We also aimed to evaluate their potential as a reservoir of diagnostic biomarkers.
+We also aimed to evaluate their potential for use as diagnostic biomarkers.
 The implications of this study are threefold.
-*First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process.
+*First*, this work will support a biological role for the virome in colorectal cancer development and suggest that more than bacteria are involved in the process.
 *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples.
@@
 ## Cohort Design, Sample Collection, and Processing
-The study cohort consisted of 90 human subjects, 30 of which had healthy colons, 30 of which had adenomas, and 30 which had carcinomas **(Figure \ref{sampleproc})**.
+Our study cohort consisted of 90 human subjects, 30 of which had healthy colons, 30 of which had adenomas, and 30 which had carcinomas **(Figure \ref{sampleproc})**.
 Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques.
@@
 Only one of the nine controls contained detectable DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
-As was expected, these controls yielded few sequences and were mostly removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
+As was expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
 The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**.
@@
 We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance.
-Statistically, the viral model performed better than the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`), although the results were practically similar, with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.
+The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`), with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.
 
 To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data.
-The combined model yielded statistically significant but small performance improvement beyond the viral (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
-This suggested that features from the virus and bacterial communities interact to improve the ability to classify stool as belonging to individuals with cancerous colons.
+The combined model yielded a statistically significant but minor performance improvement beyond the viral (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
+This suggested that the combined features from the virus and bacterial communities improve our ability to classify stool as belonging to individuals with cancerous colons.
 
 To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data.
-This model performed more poorly than any of the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`) **(Figure \ref{predmodel} A - B)**.
+This model performed more poorly than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`) **(Figure \ref{predmodel} A - B)**.
 Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**.
@@
 
-The association between the two communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed.
+The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed.
 *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu].
@@
 If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacteria and phages.
-Instead we observed a strikingly low correlation between the bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**.
-There was an overall absence of correlation between the most influential phage OVUs and bacterial OTUs **(Figure \ref{correlations} B)**.
+Instead we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**.
+There was an overall absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**.
 This evidence supported our null hypothesis that the influential phages were not primarily predators of influential bacteria.
@@
 This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition.
-Overall, our data supports a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
+Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
 Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host.
@@
 As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundance was most influential.
-As the bacteria entered their oncogenic synergy with the epithelium, we hypothesize the phages could have continued mediating biofilm dispersal, as well as support the colonized oncogenic bacteria by lysing competing cells to maintain the niche and provide nutrients to the other bacteria.
-In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role of the virome in colorectal cancer.
+As the bacteria entered their oncogenic synergy with the epithelium, we hypothesize the phages could have continued mediating biofilm dispersal, as well as support the colonized oncogenic bacteria by lysing competing cells to maintain the niche and provide nutrients to other bacteria.
+In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role the virome plays colorectal cancer.
 
@@
 Briefly, whole evacuated stool was collected from patients who were 18 years of age or older, able to provide informed consent, have had colonoscopy and histologically confirmed colonic disease status, had not had surgery, had not had chemotherapy or radiation, and were free of known co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and inflammatory bowel disease.
-Samples were collected from four locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor (Michigan, USA).
+Samples were collected from four geographic locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor (Michigan, USA).
 Ninety patients were recruited to the study, thirty of which were designated healthy, thirty with detected adenomas, and thirty with detected carcinomas.
@@
 Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz].
-Briefly, an aliquot of stool (ca.
-0.1g) was resuspended in standard SM buffer (Crystalgen) and vortexed to facilitate resuspension.
+Briefly, an aliquot of stool (~0.1g) was resuspended in standard SM buffer (Crystalgen) and vortexed to facilitate resuspension.
 The resuspended stool was centrifuged to remove major particulate debris, followed by filtering through a 0.22-µm filter to remove smaller contaminants.
@@
 Disease classes were staggered across purification runs to prevent run variation as a confounding factor.
-Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT library preparation kit according to the standard kit protocol.
+Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT preparation kit according to the standard kit protocol.
 The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
@@
 Samples with less than the cutoff were removed from the analysis.
-Alpha diversity was calculated using the Shannon Entropy and Richness metrics.
+Alpha diversity was calculated using the Shannon entropy and richness metrics.
 Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (ANOSIM) with a post-hoc multivariate Tukey test.
@@
 Classification modeling was performed in R using the Caret package [@caretClassificatio:_U2Lit_1].
-OTU and OGU abundance data was preprocessed by removing features (OTUs and OGUs) that were present in less than half of the samples.
+OTU, OVU, and OGU abundance data was preprocessed by removing features (OTUs, OVUs, and OGUs) that were present in less than thirty of the samples.
 This served both as an effective feature reduction technique and made the calculations computationally feasible.
@@
 bacteria, virus), the sample model parameters were used for each group.
-The maximum AUC during training was recorded across 10 iterations of each group model creation to test the significance of the differences between feature set performance.
+The maximum AUC during training was recorded across twenty iterations of each group model creation to test the significance of the differences between feature set performance.
 Statistical significance was evaluated using a Wilcoxon test between two categories, or a pairwise Wilcoxon test with Bonferroni corrected p-values when comparing more than two categories.
@@
 # Acknowledgments
-The authors thank the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
+The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
 GD Hannigan was supported by the Michigan Molecular Mechanisms of Microbial Pathogenesis Fellowship.

### 2017-04-11 doc/manuscript.Rmd
Updates and edits.

@@
 ```{r loadfiles, echo=FALSE}
+psigfig = 1
+asigfig = 3
 sigfig = 3
+
 alphadf <- read.delim(file = "../rtables/diversity.tsv", header = TRUE, sep = "\t")
@@
 corstats <- read.delim(file = "../rtables/cor-stats.tsv", header = TRUE, sep = "\t")
+
+pvalcalc <- function(x) {
+  return(ifelse(x < 0.001, "< 0.001", paste("=", signif(x, digits = psigfig), sep = " ")))
+}
 ```
@@
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer.
+Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer.
 The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g.
 cancer).
-Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
+Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
 The strong association of bacterial communities with colorectal cancer and the precedence for the virome in impacting other human diseases suggest that colorectal cancer may be associated with altered virus communities.
@@
 Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
-This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja].
+This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix].
 While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome.
@@
 adenoma) in the large intestine [@Fearon:2011go].
-If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death.
-Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr], and survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz].
+If left untreated, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death.
+Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr].
+Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz].
 Thus work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.
@@
 Here we address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those changes might impact cancer progression and severity.
-We also aimed to evaluate their potential for use as diagnostic biomarkers.
+We also aimed to evaluate the virome's potential for use as a diagnostic biomarker.
 The implications of this study are threefold.
-*First*, this work will support a biological role for the virome in colorectal cancer development and suggest that more than bacteria are involved in the process.
+*First*, this work supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process.
 *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples.
@@
 ## Cohort Design, Sample Collection, and Processing
-Our study cohort consisted of 90 human subjects, 30 of which had healthy colons, 30 of which had adenomas, and 30 which had carcinomas **(Figure \ref{sampleproc})**.
-Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques.
-The other half of the stool samples were purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing.
-The VLP purification allowed us to observe the *active virome* because we only sequenced those viruses that were encapsulated.
+Our study cohort consisted of 90 human subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30 who had carcinomas **(Figure \ref{sampleproc})**.
+Half of each stool sample was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques.
+The other half of each stool sample was purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing.
+The VLP purification allowed us to observe the *active and extracellular virome* because we only sequenced those viruses that were encapsulated.
 
 Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**.
-Each extraction protocol was performed with a blank control to detect contaminants from reagents or other unintentional sources.
-Only one of the nine controls contained detectable DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
+Each extraction was performed with a blank buffer control to detect contaminants from reagents or other unintentional sources.
+Only one of the nine controls contained detectable DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful purification of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
 As was expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
@@
 We therefore initially evaluated the changes in virome OVU diversity in colorectal cancer.
-We evaluated differences in communities between disease states using the Bray-Curtis, Shannon entropy, and richness diversity metrics.
+We evaluated differences in communities between disease states using the Shannon entropy, richness, and Bray-Curtis.
 We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**.
-There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**).
-It is worth noting that there was a significant difference between the few blank controls that remained after rarefying the data, and the other study groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**.
+There was also no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**).
+It is worth noting that there was a significant difference between the few blank controls that remained after rarefying the data, and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**.
 This further supported the quality of our sample set.
-Similar to what was previously seen when analyzing the 16S rRNA gene sequences from the same samples [@Zackular:2014fba], standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states.
+Similar to what was previously seen when analyzing the 16S rRNA gene sequences (from the same samples) [@Zackular:2014fba] and metagenomic samples [@Zeller:2014ix], standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states.
 
@@
 We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance.
-The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`), with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.
+The viral model performed as well as the bacterial model (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"])`), with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = asigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = asigfig)`, respectively **(Figure \ref{predmodel} A - B)**.
 
 To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data.
-The combined model yielded a statistically significant but minor performance improvement beyond the viral (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**.
+The combined model yielded a statistically significant but minor performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**.
 This suggested that the combined features from the virus and bacterial communities improve our ability to classify stool as belonging to individuals with cancerous colons.
@@
 To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data.
-This model performed more poorly than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`) **(Figure \ref{predmodel} A - B)**.
+This model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**.
 Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**.
@@
 Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant unknown viral population associated with the human virome.
-This is common in the virome; studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz].
+This is common in the virome; studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa].
 When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.
@@
 ## Shifted Phage Influence Across Cancer Stages
-Because our cohort included people with healthy, adenomatous, and cancerous colons we were able to evaluate shifts in influential phage identities during cancer progression.
+We were able to evaluate shifts in influential phage identities during cancer progression because our cohort included people with healthy, adenomatous, and cancerous colons.
 We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions.
@@
 After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states.
-The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**).
+The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = asigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = asigfig)` (p-value `r pvalcalc(threesig)`, **Figure \ref{threewaymodel} A-C**).
 The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
@@
 Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts.
-To better understand the types of viruses that are important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages.
+To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages.
 The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**.
@@
 We evaluated whether the phages in the community were primarily lytic (replicate by lysing their hosts) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell).
-We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz].
+We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
 We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
@@
 The phages with high centrality values were defined as community hubs.
-Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**.
+Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**.
 Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.
@@
 The bacterial metagenomic OGUs represented only the most abundant taxa, which was not informative for this application.
-There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but these approaches did not utilize OGU clustering like we did here, and the models only performed **as well** as the 16S rRNA gene model [@Zeller:2014ix].
+There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but the previous approach has not utilize OGU clustering like we did here, it relied on lowly abundant signatures, and the models only performed as well as the 16S rRNA gene model [@Zeller:2014ix].
 Thus the targeted 16S rRNA gene sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples.
@@
 
-## Analysis Source Code & Availability
+## Analysis Source Code & Data Availability
+All study sequences are available on the NCBI Sequence Read Archive under the following accession IDs:
+
 All associated source code is available at the following GitHub repository:
@@
 ## Virus Metagenomic Library Preparation & Sequencing
-Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz].
+Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz; @Thurber:2009dn; @Kleiner:2015kd; @Minot:2011ez].
 Briefly, an aliquot of stool (~0.1g) was resuspended in standard SM buffer (Crystalgen) and vortexed to facilitate resuspension.
@@
 The exposed genomic DNA from the lysed cells was degraded by treating the samples with DNase.
-The DNA was extracted from the purified VLPs using the Wizard PCR Purification Preparation Kit (Promega).
+The DNA was extracted from the purified VLPs using the Wizard PCR Purification Preparation Kit (Promega) [@Brum:2016jl].
 Disease classes were staggered across purification runs to prevent run variation as a confounding factor.
@@
 Both the viral and whole metagenomic sample sets were subjected to the same quality control procedures.
-The sequences were obtained as de-multiplexed fastq files from the HiSeq platform and subjected to 5' and 3' adapter trimming using the CutAdapt program (v1.9.1) with an error rate of 0.1 and an overlap of 10 [@Martin:2011eu].
+The sequences were obtained as de-multiplexed fastq files and subjected to 5' and 3' adapter trimming using the CutAdapt program (v1.9.1) with an error rate of 0.1 and an overlap of 10 [@Martin:2011eu].
 The FastX toolkit (v0.0.14) was used to quality trim the reads to a minimum length of 75bp and a minimum quality score of 30 [@FASTXToolkit:wr].
@@
 Contigs were assembled using paired end read files that were purged of sequences without a corresponding pair (e.g.
-One read removed due to low quality).
+one read removed due to low quality).
 The Megahit program (v1.0.6) was used to assemble contigs for each sample using a minimum contig length of 1000 bp and iterating assemblies from 21-mers to 101-mers by 20 [@Li:2016kd].
@@
 ## Diversity
-Alpha and beta diversity were calculated using the operational genomic unit abundance profiles for each sample.
+Alpha and beta diversity were calculated using the operational viral unit abundance profiles for each sample.
 Sequences were rarefied to 100,000 sequences.
@@
 Both were validated using five-fold cross validation.
-Each training set was repeated five times, and the model was tuned across five iterations of mtry values.
+Each training set was repeated five times, and the model was tuned for mtry values.
 For consistency and accurate comparison between feature groups (e.g.
-bacteria, virus), the sample model parameters were used for each group.
-The maximum AUC during training was recorded across twenty iterations of each group model creation to test the significance of the differences between feature set performance.
+bacteria, viruses), the sample model parameters were used for each group.
+The maximum AUC during training was recorded across twenty iterations of each group model to test the significance of the differences between feature set performance.
 Statistical significance was evaluated using a Wilcoxon test between two categories, or a pairwise Wilcoxon test with Bonferroni corrected p-values when comparing more than two categories.
@@
 ## Taxonomic Identification of Operational Genomic Units
-Operational viral units (OVUs) were identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
-The longest contiguous sequence in each operational genomic unit was used as a representative sequence for classification.
+Operational viral units (OVUs) weretaxonomically identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
+The longest contiguous sequence in each operational genomic unit was used as a representative sequence for classification, as described previously [@Guidi:2016kf].
 Each representative sequence was aligned to the reference genome database using the tblastx alignment algorithm (v2.2.27) and a strict similarity threshold (e-value < 1e-25) [@Camacho:2009fc].
@@
 Prophage genes were identified in the same way, using the ACLAME set of reference prophage genes.
-In both cases, the blastx algorithm was used with an e-value of 10e-5.
+In both cases, the blastx algorithm was used with an e-value threshold of 10e-5.
 Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes.

### 2017-04-11 doc/manuscript.Rmd
I wanted to add a couple more citations.

@@
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer.
+Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer.
 The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g.
@@
 The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans have died from the disease [@Siegel:2014jo].
-An important component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb].
+An important component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb; @Dejea:2014fz; @Arthur:2012kl].
 Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].

### 2017-04-25 doc/manuscript.Rmd
Added Macks comments.

@@
 Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world.
-Its progression has been linked to changes in colonic bacterial community composition.
+Its risk has been linked to changes in colonic bacterial community composition.
 Viruses are another important component of the colonic microbial community, however they have yet to be studied in colorectal cancer despite their oncogenic potential.
-We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons.
+We evaluated the colorectal cancer virome (virus community) in stool using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons.
 We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community.
@@
 Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide.
-Its progression and severity have been linked to colonic bacterial community composition.
+Its risk and severity have been linked to colonic bacterial community composition.
 Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities.
-We addressed this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients and how they compared to bacterial community changes.
+We addressed this knowledge gap by identifying changes in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community changes.
 The results suggested an indirect role for the virome in impacting colorectal cancer by modulating their associated bacterial community.
@@
 
-## Shifted Phage Influence Across Cancer Stages
-We were able to evaluate shifts in influential phage identities during cancer progression because our cohort included people with healthy, adenomatous, and cancerous colons.
+## Shifted Phage Influence Between Cancer Progression Stages
+Because previous work has identified shifts in bacterial importance in colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we evaluated shifts in influential phage identities between healthy, adenomatous, and cancerous colons.
 We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions.
 While bacterial 16S rRNA gene models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**.
-Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important in the transitions from healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.
+Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important between the healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.
 
@@
 Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
-This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network.
+This hypothesis would be supported by showing that the influential bacteriophages were community hubs (central members) within the bacteria and phage interactive network.
 We investigated the infectious capabilities and potential host ranges of all phage OVUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**).

### 2017-06-09 doc/manuscript.Rmd
A couple more changes.

@@
 We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community.
-We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated random forest models identified striking changes in the virus community.
+We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer <!--not sure what this is trying to say...guessing no detectable difference in virome a and b diversity between cancer and healthy samples?-->, but more sophisticated random forest models identified striking changes in the virus community.
 The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function.
@@
 Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer.
-The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g.
-cancer).
+The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g., cancer).
 Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
-The strong association of bacterial communities with colorectal cancer and the precedence for the virome in impacting other human diseases suggest that colorectal cancer may be associated with altered virus communities.
+The strong association of bacterial communities with colorectal cancer and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.
 
 Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo].
-The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans have died from the disease [@Siegel:2014jo].
+The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016 and over 500,000 Americans have died from the disease [@Siegel:2014jo].
 An important component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb; @Dejea:2014fz; @Arthur:2012kl].
 Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
-This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix].
+This association also has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix].
 While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome.
@@
 ## Cohort Design, Sample Collection, and Processing
-Our study cohort consisted of 90 human subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30 who had carcinomas **(Figure \ref{sampleproc})**.
+Our study cohort consisted of 90 human subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30 of whom had carcinomas **(Figure \ref{sampleproc})**.
 Half of each stool sample was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques.
 The other half of each stool sample was purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing.
-The VLP purification allowed us to observe the *active and extracellular virome* because we only sequenced those viruses that were encapsulated.
+In the VLP purification, cells were disrupted and extracellular DNA degraded **(Figure \ref{sampleproc})** to allow the exclusive analysis of viral DNA within virus capsids.
+In this manner the *extracellular virome* of encapsulated viruses was targeted.
 
-Virus DNA was purified prior to sequencing, allowing us to analyze DNA exclusively from within virus capsids **(Figure \ref{sampleproc})**.
 Each extraction was performed with a blank buffer control to detect contaminants from reagents or other unintentional sources.
-Only one of the nine controls contained detectable DNA, at a minimal concentration of 0.011 ng/µl, thus providing initial evidence of successful purification of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
+Only one of the nine controls contained detectable DNA at a minimal concentration of 0.011 ng/µl, thus providing evidence of the enrichment and purification of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
 As was expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
-The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**.
-Because contigs only represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})**.
+The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1 kb **(Figure \ref{contigqc})**<!-- scale/notation units in A and C axes don't match; same for B and D x-axes; it would also be awesome if you could label directly on the plot which is viral and which is whole-->.
+Because contigs only represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})** to approximate organismal units.
 
 ## Unaltered Virome Diversity in Colorectal Cancer
-Microbiome and disease associations are often described as being of an altered diversity (i.e.
+Microbiome and disease associations often are described as being of an altered diversity (i.e.
 "dysbiotic").
-We therefore initially evaluated the changes in virome OVU diversity in colorectal cancer.
-We evaluated differences in communities between disease states using the Shannon entropy, richness, and Bray-Curtis.
-We observed no significant alterations in either Shannon entropy or richness in the diseased state **(Figure \ref{betaogu} C-D)**.
-There was also no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**).
-It is worth noting that there was a significant difference between the few blank controls that remained after rarefying the data, and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**.
-This further supported the quality of our sample set.
-Similar to what was previously seen when analyzing the 16S rRNA gene sequences (from the same samples) [@Zackular:2014fba] and metagenomic samples [@Zeller:2014ix], standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states.
+Therefore, first we evaluated the influence of colorectal cancer on virome OVU diversity.
+We evaluated differences in communities between disease states using the Shannon entropy, richness, and Bray-Curtis metrics.
+We observed no significant alterations in either Shannon entropy or richness in the diseased state as compared to the healthy state **(Figure \ref{betaogu} C-D)**.
+There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**<!--would be pretty nice/logical to have the treatment colors consistent across panels-->).
+Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})** <!--I'd leave out panel A here, it doesn't add anything/is distracting-->, further supporting the quality of the sample set.
+In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**.
+This is consistent with what has been observed when the same metrics are applied to 16S rRNA gene sequences (from the same samples) [@Zackular:2014fba] and metagenomic samples [@Zeller:2014ix] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.<!--am I interpretting/re-interpretting this correctly?-->
 
 ## Altered Virome Composition in Colorectal Cancer
-16S rRNA gene relative abundance profiles are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
+As opposed to the diversity metrics discussed above, 16S rRNA gene relative abundance profiles are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
 The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer.
@@
 To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons.
-We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba].
+We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]<!--need to more specifically label Fig 1 components in the legend.
+Maybe as Bacterial 16S, Virome, combined 16S+virome, whole metagenome; the inconsistency is that the "Virus" is metagenome too; maybe label Panel E as "Combined Model IDs"...microbe is a bit ambiguous, not intuitively interpretted as bacteria+virus; could make C and D consistent as "Bacterial Model ID", "Viral Model ID"-->.
 We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance.
@@
 To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data.
-The combined model yielded a statistically significant but minor performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**.
-This suggested that the combined features from the virus and bacterial communities improve our ability to classify stool as belonging to individuals with cancerous colons.
+The combined model yielded a statistically significant, but minor, performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**.
+The combined features from the virus and bacterial communities improved our ability to classify stool as belonging to individuals with cancerous colons.
 
@@
 The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.
-Although the features in the viral model were also of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.
+Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.
 
@@
 *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu].
-The virome signature was also driven by a few OVUs, suggesting a role for the viruses in cancer development **(Figure \ref{predmodel} D)**.
-The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified").
-Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant unknown viral population associated with the human virome.
-This is common in the virome; studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa].
+The virome signature also was driven by a few OVUs, suggesting a role for these viruses in cancer development **(Figure \ref{predmodel} D)**.
+The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and an "unclassified" phage taxon<!--"orphan" is a co-opted term, better to just leave is as unclassified-->.
+Many of the important viruses were unidentifiable (denoted "unknown").
+This is common in viromes across habitats; studies can have as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa].
 When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.
@@
 ## Shifted Phage Influence Between Cancer Progression Stages
-Because previous work has identified shifts in bacterial importance in colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we evaluated shifts in influential phage identities between healthy, adenomatous, and cancerous colons.
+Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of specific phages could be detected between healthy, adenomatous, and cancerous colons.
 We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions.
@@
 After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states.
-The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = asigfig)` and outperformed the viral community model which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = asigfig)` (p-value `r pvalcalc(threesig)`, **Figure \ref{threewaymodel} A-C**).
-The microbes important for the cancer vs healthy and healthy vs adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
+The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = asigfig)` and outperformed the viral community model, which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = asigfig)` (p-value `r pvalcalc(threesig)`, **Figure \ref{threewaymodel} A-C**).
+The microbes important for the cancer versus healthy and healthy versus adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
 The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**.
-The viruses most important to the three-class model were also identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.
-Not all important OVUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.
+The viruses most important to the three-class model were identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**, but not all important OVUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.
 
@@
 To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages.
-The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**.
-Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes.
+The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**<!--however, remember the role that the chloroform treatment had on altering the E:B virus ratio...as there are more enveloped eukaryotic viruses than bacterial viruses, these will have lipid bilayers in their capsid and will have been disrupted and lost during the purification step-->.
+Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes.
 
-We evaluated whether the phages in the community were primarily lytic (replicate by lysing their hosts) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell).
+We evaluated whether the phages in the community were primarily lytic (obligately lyse their hosts after replication) or temperate (able to integrate into their host's genome to form a lysogen, as well as to transition to a lytic mode).
 We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
-We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
-Thus the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that can lie dormant in bacterial genomes.
+We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
+Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that <!--we should be careful here not to propogate the false notion that intergrated phages are dormant...quite the contrary, they may be playing a huge role in altering the expression/function of their hosts, potentially more impactful ont he community that lytic phage-->have the ability to integrate into host bacterial genomes.
 These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
@@
 ## Community Context of Influential Phages
-Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were primarily predators of the influential bacteria, and thus influenced their relative abundance through predation.
+Because the link between colorectal cancer and the virome was driven by bacteriophages <!--just noting again the potnetial bias that chloroform treatment could play, could simply play down that driven by bacteriopahge versus other viruses was a primariy finding.
+Could just take this out and start with "We hypotehsized..."-->, we hypothesized that the influential phages were primarily predators of the influential bacteria, and thus influenced their relative abundance through predation.
 If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacteria and phages.
-Instead we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**.
-There was an overall absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**.
+Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**.
+Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**.
 This evidence supported our null hypothesis that the influential phages were not primarily predators of influential bacteria.
@@
 Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
-This hypothesis would be supported by showing that the influential bacteriophages were community hubs (central members) within the bacteria and phage interactive network.
-We investigated the infectious capabilities and potential host ranges of all phage OVUs using a random forest model to predict which phages infected which bacteria in the overall community (**Cite Network Preprint**).
-The predicted interactions were then used to identify community hub phages.
-This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**.
-We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OVU's connection to the rest of the network, and compared the centrality to the importance of each OVU in the colorectal cancer classification model.
+In other words, we hypothesized that the influential bacteriophages were community hubs (central members) within the bacteria and phage interactive network.
+We investigated the <!--not sure what "infectious capabilities" means here or how it is different from host range/who infects whom--> potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community (**Cite Network Preprint**).
+The predicted interactions were then used to identify phage community hubs.
+This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)** <!-- do you mean high prevalence of phages wide host ranges?
+or range of possible tropisms (as it technically reads now, I think)...which would mean some pahge infect one host, some infect a hgue number of hosts...if so, what is the max number of hosts predicted, or the distribution of host ranges?
+This is an interesting feature to report.
+I'd also rephrase or state expicitly what wide tropism range means with repsect to the network representation.--> .
+We calculated the alpha centrality (measure of importance in the ecological network) of each phage OVU's connection to the rest of the network.
 The phages with high centrality values were defined as community hubs.
+Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model.
 Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**.
@@
 Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Some bacteria also have oncogenic properties, meaning bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
+Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
 Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colorectal communities and colorectal cancer has yet to be evaluated.
 Here we show that, like colonic bacterial communities, the colon virome was altered in colorectal cancer.
-Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.
+Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition <!--can you isolate this scenario from the oncogenesis by bacteria-modulated phage commumity composition?-->.
 
-Our data allowed us to begin delineating the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**.
+Here, we have begun to delineate the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**.
 We found basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states.
 By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer.
-The colorectal cancer virome was composed primarily of bacteriophages.
-These phage communities were not exclusive predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations.
+The colorectal cancer virome was composed primarily of bacteriophages <!--again, have to nuance this, biases by chloroform treatment...lipid layer-containing euk viruses are disproportionately more influenced by chlor than phages-->.
+These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations.
 Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria.
@@
 This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
-We hypothesize that the broadly infectious phages in the colon began lysing, and thereby disrupting, the bacterial communities to open a niche in which opportunistic bacteria (such as *Fusobacterium nucleatum*) were able to colonize.
-Once the initial influential bacteria had established themselves in the epithelium, other opportunistic bacteria were able to adhere to the driver, colonize, and begin establishing a biofilm.
+We hypothesize that the broadly infectious phages in the colon began lysing <!--began?
+relative to what/when?
+instigated by what?
+oncogenesis?
+if so, this means something else, unknown(?) initiates the shift?--> and thereby disrupted the bacterial communities.
+This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize.
+Once the initial influential founder bacteria established themselves in the epithelium, secondary opportunistic bacteria were able to adhere to the founders, colonize, and begin establishing a biofilm.
 Phages may have played a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj].
-The oncogenic bacteria were then able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response.
-As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundance was most influential.
-As the bacteria entered their oncogenic synergy with the epithelium, we hypothesize the phages could have continued mediating biofilm dispersal, as well as support the colonized oncogenic bacteria by lysing competing cells to maintain the niche and provide nutrients to other bacteria.
-In addition to highlighting the most likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role the virome plays colorectal cancer.
+The oncogenic bacteria may then have been able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response.
+As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundances were most influential.
+As the bacteria entered their oncogenic synergy with the epithelium, we conjecture that the phages continued mediating biofilm dispersal.
+This process would thereby support the colonized oncogenic bacteria by lysing competing cells and releasing nutrients to other bacteria in the form of cellular lysates.
+In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role the virome plays colorectal cancer.
 
-A notable observation from our analysis was the poor performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences.
-This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies, and the advantage of 16S rRNA gene sequencing over whole metagenomic shotgun sequencing.
-We found that there were six bacterial OTUs that drove the performance of the 16S rRNA gene classification model, and these OTUs were all sparsely present and lowly abundant.
+A notable finding was the poor performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences.
+This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies and the advantage of 16S rRNA gene sequencing over whole metagenomic shotgun sequencing <!--do you ever lay out the logic that this is what you think explains this (I don't remember that discussion yet)?
+if not, should do that first here, briefly-->.
+Six bacterial OTUs  drove the performance of the 16S rRNA gene classification model and these OTUs were all sparsely present and lowly abundant.
 Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model.
 The bacterial metagenomic OGUs represented only the most abundant taxa, which was not informative for this application.
-There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification, but the previous approach has not utilize OGU clustering like we did here, it relied on lowly abundant signatures, and the models only performed as well as the 16S rRNA gene model [@Zeller:2014ix].
-Thus the targeted 16S rRNA gene sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples.
+There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification [@Zeller:2014ix], but this previous approach relied on lowly abundant signatures and did not utilize OGU clustering, as done here.
+In that former case, the models only performed as well as the 16S rRNA gene model [@Zeller:2014ix].
+Thus, the targeted 16S rRNA gene sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples.
 Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
+<!--alternatively, if making the case that rare community members are important (and detectable by 16S and not metaG), maybe they are disproportionately active, so perhaps metatrascriptomics could be informative as an alternate data type?-->
 
-In addition to the therapeutic ramifications for understanding the colorectal cancer microbiome, our findings provide a proof-of-principle that viruses, while under-appreciated and understudied in the human microbiome, are an important contributor to human disease that has the potential to provide an abundance of information that supplements that of bacterial communities.
+In addition to the therapeutic ramifications for understanding the colorectal cancer microbiome, our findings provide a proof-of-principle that viruses, while understudied and currently under-appreciated in the human microbiome, are an important contributor to human disease.
+Viral communities have the potential to provide an abundance of information to supplement that of bacterial communities.
 Evidence has suggested that the virome is a crucial component to the microbiome and that bacteriophages are important players.
-Bacteriophage and bacterial communities cannot thrive without each other [@RodriguezValera:2009cr].
-Not only is the human virome an important part of human health and disease, but it appears to have a particular significance in cancer research.
+Bacteriophage and bacterial communities cannot thrive <!--qualify "thrive"...maybe bring in the concepts from the mSystems paper about community stability, e.g., the Koskella review...the Rodriguez Valera one is a kind fo weird one for communities "thriving"--> without one another [@RodriguezValera:2009cr].
+Not only is the human virome an important element to consider in human health and disease <!--insert refs, again, a handful of those from this intro and mSystems paper...Hannigan, Modi, etc.-->, but it is likely to have a significant impact on cancer etiology and progression.
 
@@
 The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
+<!--one lane for 90 samples?-->
 
@@
 Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz; @Thurber:2009dn; @Kleiner:2015kd; @Minot:2011ez].
-Briefly, an aliquot of stool (~0.1g) was resuspended in standard SM buffer (Crystalgen) and vortexed to facilitate resuspension.
-The resuspended stool was centrifuged to remove major particulate debris, followed by filtering through a 0.22-µm filter to remove smaller contaminants.
+Briefly, an aliquot of stool (~0.1 g) was resuspended in SM buffer (<!--give molarity of constituents-->; Crystalgen) and vortexed to facilitate resuspension.
+The resuspended stool was centrifuged to remove major particulate debris then filtered through a 0.22-µm filter to remove smaller contaminants.
 The filtered supernatant was treated with chloroform to lyse contaminating cells including bacteria, human, fungi, etc.
-The exposed genomic DNA from the lysed cells was degraded by treating the samples with DNase.
-The DNA was extracted from the purified VLPs using the Wizard PCR Purification Preparation Kit (Promega) [@Brum:2016jl].
+The exposed genomic DNA from the lysed cells was degraded by treating the samples with DNase <!--concentration and time of incubation; how was the inactivation of the DNase performed?-->.
+The DNA was extracted from the purified virus-like particles (VLPs) using the Wizard PCR Purification Preparation Kit (Promega) <!--I don't think the ref is needed there for the kit use, that is pretty standard-->.
 Disease classes were staggered across purification runs to prevent run variation as a confounding factor.
-Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT preparation kit according to the standard kit protocol.
+As for whole community metagenomes, purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT preparation kit according to the standard kit protocol.
 The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
-The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz].
+The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz] <!--if cycle number was 12 for whole community, should state that in previous paragraph to make it explicit-->.
 The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
@@
 ## Metagenome Quality Control
-Both the viral and whole metagenomic sample sets were subjected to the same quality control procedures.
+Both the viral and whole community metagenomic sample sets were subjected to the same quality control procedures.
 The sequences were obtained as de-multiplexed fastq files and subjected to 5' and 3' adapter trimming using the CutAdapt program (v1.9.1) with an error rate of 0.1 and an overlap of 10 [@Martin:2011eu].
-The FastX toolkit (v0.0.14) was used to quality trim the reads to a minimum length of 75bp and a minimum quality score of 30 [@FASTXToolkit:wr].
+The FastX toolkit (v0.0.14) was used to quality trim the reads to a minimum length of 75 bp and a minimum quality score of 30 [@FASTXToolkit:wr].
 Reads mapping to the human genome were removed using the DeconSeq algorithm (v0.4.3) and default parameters [@Schmieder:2011fo].
@@
 Abundance was corrected for contig reference length and the number of contigs included in each operational genomic unit.
-Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1,000,000 and 500,000 reads, respectively, and removing samples with less total samples than the threshold.
+Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1,000,000 and 500,000 reads <!--what was the distribution of reads per smaple?
+or at least max/min/median?
+not sure how to think about these thresholds without some grounding, quick supp fig or table addition?-->, respectively, and by removing samples with fewer total samples than the threshold <!--I don't understand this clause "by removing samples with fewer total samples than the threshold", meant as fewer total reads than the threshold?-->.
 Thresholds were set for maximizing sequence information while minimizing numbers of lost samples.
@@
 ## Operational Genomic Unit Classification
-Much like operational taxonomic units (OTUs) are used as an operational definition of similar 16S rRNA gene sequences in absence of taxonomic identification, we operationally defined closely related bacterial contig sequences as operational genomic units (OGUs) and virus contigs as operational viral units (OVUs) in the absence of taxonomic identity.
+Much like operational taxonomic units (OTUs) are used as an operational definition of similar 16S rRNA gene sequences, we defined closely related bacterial contig sequences as operational genomic units (OGUs) and virus contigs as operational viral units (OVUs) in the absence of taxonomic identity.
 OGUs and OVUs were defined with the CONCOCT algorithm (v0.4.0) which bins related contigs by similar tetra-mer and co-abundance profiles within samples using a variational Bayesian approach [@Alneberg:2014fc].
-CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria due to computational limitations.
+CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria.<!--I took out "due to computational limitations", as it implies that is a tradeoff with possible consequences.
+Maybe better not to open that can of worms and jsut state unapologetically that those were the thresholds used.-->
 
@@
 This served both as an effective feature reduction technique and made the calculations computationally feasible.
-The binary random forest model was trained using the Area Under the ROC Curve (AUC) and the three-class random forest model was trained using the mean AUC.
+The binary random forest model was trained using the Area Under the receiver operating characteristic Curve (AUC) and the three-class random forest model was trained using the mean AUC.
 Both were validated using five-fold cross validation.
-Each training set was repeated five times, and the model was tuned for mtry values.
-For consistency and accurate comparison between feature groups (e.g.
-bacteria, viruses), the sample model parameters were used for each group.
+Each training set was repeated five times, and the model was tuned for mtry <!--what is this?--> values.
+For consistency and accurate comparison between feature groups (e.g., bacteria, viruses), the sample model parameters were used for each group.
 The maximum AUC during training was recorded across twenty iterations of each group model to test the significance of the differences between feature set performance.
@@
 ## Taxonomic Identification of Operational Genomic Units
-Operational viral units (OVUs) weretaxonomically identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
+Operational viral units (OVUs) were taxonomically identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
 The longest contiguous sequence in each operational genomic unit was used as a representative sequence for classification, as described previously [@Guidi:2016kf].
@@
 ## Ecological Network Analysis & Correlations
-The ecological network of the bacterial and phage operational genomic units were constructed and analyzed as previously described (cite network preprint here).
+The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described (**Cite Network Preprint**).
 Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1).
@@
 ## Phage Replication Style Identification
-OVU replication sytle was identified using methods described previously [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
-Briefly, we identified lysogenic OVUs as representative contigs containing at least one of three genomic markers: 1) phage integrase genes, 2) prophage genes from the ACLAME database, 3) genomic similarity to bacterial reference genomes.
+Phage OVU replication mode was predicted using methods described previously [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
+Briefly, we identified temperate OVUs as representative contigs containing at least one of three genomic markers: 1) phage integrase genes, 2) prophage genes from the ACLAME database, or 3) genomic similarity to bacterial reference genomes.
 Integrase genes were identified in phage OVU representative contigs by aligning the contigs to a reference database of all known phage integrase genes from the Uniprot database (Uniprot search term: "organism:phage gene:int NOT putative").
 Prophage genes were identified in the same way, using the ACLAME set of reference prophage genes.
-In both cases, the blastx algorithm was used with an e-value threshold of 10e-5.
-Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes.
+In both cases, the blastx algorithm was used with an e-value threshold of 10e-5 <!--I know there is not much to do aboutt his now, but this is pretty lenient-->.
+Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes <!--and anything to ensure that they weren't actually bacteria?
+steps to identify rRNA genes or other "diagnostic" bacterial genes, e.g., single-copy or housekeeping genes that haven't really been reported in phage, etc.
+-->.
 To accomplish this, representative phage contigs were aligned to the European Nucleotide Archive bacterial genome reference set using the blastn algorithm (e-value < 10e-25).
@@
 The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
-GD Hannigan was supported by the Michigan Molecular Mechanisms of Microbial Pathogenesis Fellowship.
+GD Hannigan was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528).
 PD Schloss was supported by funding from the National Institutes of Health (P30DK034933).

### 2017-06-10 doc/manuscript.Rmd
Updated with edits and SRA

@@
 We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community.
-We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer <!--not sure what this is trying to say...guessing no detectable difference in virome a and b diversity between cancer and healthy samples?-->, but more sophisticated random forest models identified striking changes in the virus community.
+We identified no detectable difference in diversity (alpha or beta) between healthy, adenomatous, or cancerous colonic samples, but more sophisticated random forest models identified striking changes in the virus community.
 The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function.
@@
 Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer.
-The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g., cancer).
+The gut virome (the virus community of the gut, including eukaryotic and prokaryotic viruses) therefore has the potential to impact health and disease (e.g.
+cancer).
 Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
@@
 
-Colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e.
+Colorectal cancer progresses in a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e.
 adenoma) in the large intestine [@Fearon:2011go].
@@
 As was expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
-The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1 kb **(Figure \ref{contigqc})**<!-- scale/notation units in A and C axes don't match; same for B and D x-axes; it would also be awesome if you could label directly on the plot which is viral and which is whole-->.
+The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1 kb **(Figure \ref{contigqc})**.
 Because contigs only represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})** to approximate organismal units.
@@
 ## Unaltered Virome Diversity in Colorectal Cancer
-Microbiome and disease associations often are described as being of an altered diversity (i.e.
+Microbiome and disease associations are often described as being of an altered diversity (i.e.
 "dysbiotic").
-Therefore, first we evaluated the influence of colorectal cancer on virome OVU diversity.
+We therefore first evaluated the influence of colorectal cancer on virome OVU diversity.
 We evaluated differences in communities between disease states using the Shannon entropy, richness, and Bray-Curtis metrics.
-We observed no significant alterations in either Shannon entropy or richness in the diseased state as compared to the healthy state **(Figure \ref{betaogu} C-D)**.
-There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**<!--would be pretty nice/logical to have the treatment colors consistent across panels-->).
-Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})** <!--I'd leave out panel A here, it doesn't add anything/is distracting-->, further supporting the quality of the sample set.
+We observed no significant alterations in either Shannon entropy or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**.
+There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**).
+Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**, further supporting the quality of the sample set.
 In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**.
-This is consistent with what has been observed when the same metrics are applied to 16S rRNA gene sequences (from the same samples) [@Zackular:2014fba] and metagenomic samples [@Zeller:2014ix] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.<!--am I interpretting/re-interpretting this correctly?-->
+This is consistent with what has been observed when the same metrics were applied to metagenomic samples [@Zeller:2014ix] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.
 
@@
 To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons.
-We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]<!--need to more specifically label Fig 1 components in the legend.
-Maybe as Bacterial 16S, Virome, combined 16S+virome, whole metagenome; the inconsistency is that the "Virus" is metagenome too; maybe label Panel E as "Combined Model IDs"...microbe is a bit ambiguous, not intuitively interpretted as bacteria+virus; could make C and D consistent as "Bacterial Model ID", "Viral Model ID"-->.
+We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba].
 We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance.
@@
 To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data.
-The combined model yielded a statistically significant, but minor, performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**.
+The combined model yielded a modest but statistically significant performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**.
 The combined features from the virus and bacterial communities improved our ability to classify stool as belonging to individuals with cancerous colons.
@@
 Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**.
-The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.
+The majority of these OTUs had a relative abundance lower than 1% across the samples **(Figure \ref{16scompare} B)**.
 Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.
@@
 The virome signature also was driven by a few OVUs, suggesting a role for these viruses in cancer development **(Figure \ref{predmodel} D)**.
-The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and an "unclassified" phage taxon<!--"orphan" is a co-opted term, better to just leave is as unclassified-->.
+The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and "unclassified" phage taxa.
 Many of the important viruses were unidentifiable (denoted "unknown").
-This is common in viromes across habitats; studies can have as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa].
+This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa].
 When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.
@@
 After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states.
-The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = asigfig)` and outperformed the viral community model, which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = asigfig)` (p-value `r pvalcalc(threesig)`, **Figure \ref{threewaymodel} A-C**).
+The 16S rRNA gene model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = asigfig)` and outperformed the viral community model, which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = asigfig)` (p-value `r pvalcalc(threesig)`, **Figure \ref{threewaymodel} A-C**).
 The microbes important for the cancer versus healthy and healthy versus adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
@@
 To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages.
-The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**<!--however, remember the role that the chloroform treatment had on altering the E:B virus ratio...as there are more enveloped eukaryotic viruses than bacterial viruses, these will have lipid bilayers in their capsid and will have been disrupted and lost during the purification step-->.
+The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**.
 Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes.
+It is important to note that this could have been influenced by our methodological biases against enveloped viruses, due to chloroform and DNase treatment for purification.
 
@@
 We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
-Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that <!--we should be careful here not to propogate the false notion that intergrated phages are dormant...quite the contrary, they may be playing a huge role in altering the expression/function of their hosts, potentially more impactful ont he community that lytic phage-->have the ability to integrate into host bacterial genomes.
+Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that have the ability to integrate into host bacterial genomes.
 These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
@@
 ## Community Context of Influential Phages
-Because the link between colorectal cancer and the virome was driven by bacteriophages <!--just noting again the potnetial bias that chloroform treatment could play, could simply play down that driven by bacteriopahge versus other viruses was a primariy finding.
-Could just take this out and start with "We hypotehsized..."-->, we hypothesized that the influential phages were primarily predators of the influential bacteria, and thus influenced their relative abundance through predation.
+Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were primarily predators of the influential bacteria, and thus influenced their relative abundance through predation.
 If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacteria and phages.
@@
 In other words, we hypothesized that the influential bacteriophages were community hubs (central members) within the bacteria and phage interactive network.
-We investigated the <!--not sure what "infectious capabilities" means here or how it is different from host range/who infects whom--> potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community (**Cite Network Preprint**).
+We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2017cj].
 The predicted interactions were then used to identify phage community hubs.
-This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)** <!-- do you mean high prevalence of phages wide host ranges?
-or range of possible tropisms (as it technically reads now, I think)...which would mean some pahge infect one host, some infect a hgue number of hosts...if so, what is the max number of hosts predicted, or the distribution of host ranges?
-This is an interesting feature to report.
-I'd also rephrase or state expicitly what wide tropism range means with repsect to the network representation.--> .
 We calculated the alpha centrality (measure of importance in the ecological network) of each phage OVU's connection to the rest of the network.
@@
 Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
+Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
 Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colorectal communities and colorectal cancer has yet to be evaluated.
 Here we show that, like colonic bacterial communities, the colon virome was altered in colorectal cancer.
-Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition <!--can you isolate this scenario from the oncogenesis by bacteria-modulated phage commumity composition?-->.
+Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.
 
@@
 By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer.
-The colorectal cancer virome was composed primarily of bacteriophages <!--again, have to nuance this, biases by chloroform treatment...lipid layer-containing euk viruses are disproportionately more influenced by chlor than phages-->.
+The colorectal cancer virome was composed primarily of bacteriophages.
 These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations.
@@
 This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
-We hypothesize that the broadly infectious phages in the colon began lysing <!--began?
-relative to what/when?
-instigated by what?
-oncogenesis?
-if so, this means something else, unknown(?) initiates the shift?--> and thereby disrupted the bacterial communities.
+We hypothesize that the process began with broadly infectious phages in the colon lysing and thereby disrupted the existing bacterial communities.
 This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize.
@@
 Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
-<!--alternatively, if making the case that rare community members are important (and detectable by 16S and not metaG), maybe they are disproportionately active, so perhaps metatrascriptomics could be informative as an alternate data type?-->
 
@@
 Evidence has suggested that the virome is a crucial component to the microbiome and that bacteriophages are important players.
-Bacteriophage and bacterial communities cannot thrive <!--qualify "thrive"...maybe bring in the concepts from the mSystems paper about community stability, e.g., the Koskella review...the Rodriguez Valera one is a kind fo weird one for communities "thriving"--> without one another [@RodriguezValera:2009cr].
-Not only is the human virome an important element to consider in human health and disease <!--insert refs, again, a handful of those from this intro and mSystems paper...Hannigan, Modi, etc.-->, but it is likely to have a significant impact on cancer etiology and progression.
+Bacteriophage and bacterial communities cannot maintain stability and co-evolution without one another [@RodriguezValera:2009cr; @Brockhurst:2013iq].
+Not only is the human virome an important element to consider in human health and disease [@Ly:2014ew; @Monaco:2016ita; @Willner:2009dq; @Abeles:2015dy; @Modi:2013fi; @SantiagoRodriguez:2015gd; @Norman:2015kb], but it is likely to have a significant impact on cancer etiology and progression as well.
 
@@
 ## Analysis Source Code & Data Availability
-All study sequences are available on the NCBI Sequence Read Archive under the following accession IDs:
+All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`.
 
@@
 DNA was extracted from stool samples using the PowerSoil-htp 96 Well Soil DNA Isolation Kit (Mo Bio Laboratories) using an EPMotion 5075 pipetting system.
-Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT library preparation kit according to the standard kit protocol.
+Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT library preparation kit according to the standard kit protocol, including 12 cycles of limited cycle PCR.
 The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
 The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
-<!--one lane for 90 samples?-->
 
@@
 Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz; @Thurber:2009dn; @Kleiner:2015kd; @Minot:2011ez].
-Briefly, an aliquot of stool (~0.1 g) was resuspended in SM buffer (<!--give molarity of constituents-->; Crystalgen) and vortexed to facilitate resuspension.
+Briefly, an aliquot of stool (~0.1 g) was resuspended in SM buffer (Crystalgen; Catalog #: 221-179) and vortexed to facilitate resuspension.
 The resuspended stool was centrifuged to remove major particulate debris then filtered through a 0.22-µm filter to remove smaller contaminants.
-The filtered supernatant was treated with chloroform to lyse contaminating cells including bacteria, human, fungi, etc.
-The exposed genomic DNA from the lysed cells was degraded by treating the samples with DNase <!--concentration and time of incubation; how was the inactivation of the DNase performed?-->.
-The DNA was extracted from the purified virus-like particles (VLPs) using the Wizard PCR Purification Preparation Kit (Promega) <!--I don't think the ref is needed there for the kit use, that is pretty standard-->.
+The filtered supernatant was treated with chloroform for ten minutes with gentle shaking, so as to lyse contaminating cells including bacteria, human, fungi, etc.
+The exposed genomic DNA from the lysed cells was degraded by treating the samples with 5U of DNase for one hour at 37C.
+DNase was deactivated by incubating the sample at 75C for ten minutes.
+The DNA was extracted from the purified virus-like particles (VLPs) using the Wizard PCR Purification Preparation Kit (Promega).
 Disease classes were staggered across purification runs to prevent run variation as a confounding factor.
@@
 The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
-The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz] <!--if cycle number was 12 for whole community, should state that in previous paragraph to make it explicit-->.
+The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz].
 The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
@@
 or at least max/min/median?
-not sure how to think about these thresholds without some grounding, quick supp fig or table addition?-->, respectively, and by removing samples with fewer total samples than the threshold <!--I don't understand this clause "by removing samples with fewer total samples than the threshold", meant as fewer total reads than the threshold?-->.
+not sure how to think about these thresholds without some grounding, quick supp fig or table addition?-->, respectively, and by removing samples with fewer total reads than the threshold.
 Thresholds were set for maximizing sequence information while minimizing numbers of lost samples.
@@
 OGUs and OVUs were defined with the CONCOCT algorithm (v0.4.0) which bins related contigs by similar tetra-mer and co-abundance profiles within samples using a variational Bayesian approach [@Alneberg:2014fc].
-CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria.<!--I took out "due to computational limitations", as it implies that is a tradeoff with possible consequences.
-Maybe better not to open that can of worms and jsut state unapologetically that those were the thresholds used.-->
+CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria.
 
@@
 Both were validated using five-fold cross validation.
-Each training set was repeated five times, and the model was tuned for mtry <!--what is this?--> values.
+Each training set was repeated five times, and the model was tuned for mtry values.
 For consistency and accurate comparison between feature groups (e.g., bacteria, viruses), the sample model parameters were used for each group.
@@
 Prophage genes were identified in the same way, using the ACLAME set of reference prophage genes.
-In both cases, the blastx algorithm was used with an e-value threshold of 10e-5 <!--I know there is not much to do aboutt his now, but this is pretty lenient-->.
+In both cases, the blastx algorithm was used with an e-value threshold of 10e-5.
 Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes <!--and anything to ensure that they weren't actually bacteria?
@@
 ![*Beta-diversity comparing disease states and the study negative controls.
-A) NMDS ordination of community samples, colored by disease state.
-B) Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test.
+Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test.
 Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}](../figures/diversity-beta-ogu-negative.pdf){ width=100% }

### 2017-06-17 doc/manuscript.Rmd

@@
 This association also has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix].
-While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome.
+While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and for identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome.
 Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.
 We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.
+<!-- I would change the structure of the first two paragraphs to be more linear, but I understand if you want to put it out there as is for time, and because you are probably getting tired of my comments.
+On high level it reads as "1.
+Viruses play a role in cancer and evidence points to a role in colorectal cancer.
+2.
+Colorectal cancer is a big deal.
+We know about the relationship with microbes.
+3.
+We should study viruses of colorectal cancer." To be less circular, I'd line it up as "1.
+Colorectal cancer is a big deal.
+We know about the relationship with microbes.
+2.
+Viruses play a role in cancer and evidence points to a role in colorectal cancer too.
+3.
+We should study viruses of colorectal cancer." -->
 
-Colorectal cancer progresses in a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e.
-adenoma) in the large intestine [@Fearon:2011go].
+Colorectal cancer progresses in a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e., adenoma) in the large intestine [@Fearon:2011go].
 If left untreated, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death.
@@
 Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz].
-Thus work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.
+Thus, work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.
+<!--I feel this fits oddly here, one solution could be to put these stages of progression/need for early detection above when you introduce why colorectal cancer is a big deal...and how this argument of early detection motivates microbiome research...and now we have the logic to say that virome research could further improve upon this, as we know nothing of the role of viruses...that is the knowledge gap, then leads in well to this last paragraph.-->
 
@@
 The implications of this study are threefold.
-*First*, this work supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process.
-*Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples.
+*First*, this work supports a biological role for the virome in colorectal cancer development and suggests that more than the bacterial members of the associated microbial communities are involved in the process.
+*Second*, we present a supplementary, or even alternative, virus-based approach for classification modeling of colorectal cancer using stool samples.
 *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.
@@
 In the VLP purification, cells were disrupted and extracellular DNA degraded **(Figure \ref{sampleproc})** to allow the exclusive analysis of viral DNA within virus capsids.
-In this manner the *extracellular virome* of encapsulated viruses was targeted.
+In this manner, the *extracellular virome* of encapsulated viruses was targeted.
 
@@
 The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1 kb **(Figure \ref{contigqc})**.
-Because contigs only represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})** to approximate organismal units.
+Because contigs represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})** to approximate organismal units.
 
 ## Unaltered Virome Diversity in Colorectal Cancer
-Microbiome and disease associations are often described as being of an altered diversity (i.e.
-"dysbiotic").
+Microbiome and disease associations are often described as being of an altered diversity (i.e., "dysbiotic").
 We therefore first evaluated the influence of colorectal cancer on virome OVU diversity.
@@
 Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes.
-It is important to note that this could have been influenced by our methodological biases against enveloped viruses, due to chloroform and DNase treatment for purification.
+It is important to note that this could have been influenced by our methodological biases against enveloped viruses (more common of eukaryotic viruses than bacteriophage), due to chloroform and DNase treatment for purification.
 
-We evaluated whether the phages in the community were primarily lytic (obligately lyse their hosts after replication) or temperate (able to integrate into their host's genome to form a lysogen, as well as to transition to a lytic mode).
+We evaluated whether the phages in the community were primarily lytic (obligately lyse their hosts after replication) or temperate (able to integrate into their host's genome to form a lysogen, and subsequently transition to a lytic mode).
 We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
-We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
+<!-- which of these features was most responsible for identifying the temperate phages here?
+Guessing #3, but hoping for #1 or #2--> We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
 Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that have the ability to integrate into host bacterial genomes.
@@
 Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model.
-Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**.
+Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**.
 Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.
@@
 
-Here, we have begun to delineate the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**.
-We found basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states.
+Here, we have begun to delineate the role the colonic virome plays <!--I think ok to leave this as present tense, as then it reads more generally...we have begun to delineate something general and bigger than this individual study-->in colorectal cancer **(Figure \ref{modelsummary} A)**.
+We found that basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states.
 By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer.
@@
 Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
-Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host.
+Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host <!--ref for this?
+I know i've read it someone, if not shown explicitely, then in a review, like https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3760236/ and https://www.ncbi.nlm.nih.gov/pubmed/22748808 and https://www.nature.com/articles/srep14802-->.
 
-In addition to modeling the basic potential connections between virus communities, bacteria communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur.
+In addition to modeling the potential connections between virus communities, bacteria communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur.
 This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
-We hypothesize that the process began with broadly infectious phages in the colon lysing and thereby disrupted the existing bacterial communities.
-This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize.
+We hypothesize that the process began with broadly infectious phages in the colon lysing and thereby disrupting the existing bacterial communities.
+<!--do we have the time resolution to suggest that the phages started this?
+and what could trigger that?
+What about the alternative hypothesis that the microbes began the shift and their phages responded, or (a more specific instance of your proposed case) that integrated prophges responded to cues from an altered environment, e.g., mediated byv a shift in transcriptional regulation of their hosts, that caused them to switch to lysis and drive a community shift.
+--> This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize.
 Once the initial influential founder bacteria established themselves in the epithelium, secondary opportunistic bacteria were able to adhere to the founders, colonize, and begin establishing a biofilm.
@@
 A notable finding was the poor performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences.
-This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies and the advantage of 16S rRNA gene sequencing over whole metagenomic shotgun sequencing <!--do you ever lay out the logic that this is what you think explains this (I don't remember that discussion yet)?
-if not, should do that first here, briefly-->.
-Six bacterial OTUs  drove the performance of the 16S rRNA gene classification model and these OTUs were all sparsely present and lowly abundant.
+We believe this observation speaks to the importance of sequencing coverage in microbial community studies and the advantage of the high coverage of 16S rRNA gene sequencing relative to the lower per OTU coverage possible using whole metagenomic shotgun sequencing.
+To demonstrate this concept, consider that six bacterial OTUs drove the performance of the 16S rRNA gene classification model and these OTUs were all sparsely present and lowly abundant.
 Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model.
@@
 In addition to the therapeutic ramifications for understanding the colorectal cancer microbiome, our findings provide a proof-of-principle that viruses, while understudied and currently under-appreciated in the human microbiome, are an important contributor to human disease.
-Viral communities have the potential to provide an abundance of information to supplement that of bacterial communities.
+Viral community dynamices have the potential to provide an abundance of information to supplement those of bacterial communities.
 Evidence has suggested that the virome is a crucial component to the microbiome and that bacteriophages are important players.
 Bacteriophage and bacterial communities cannot maintain stability and co-evolution without one another [@RodriguezValera:2009cr; @Brockhurst:2013iq].
-Not only is the human virome an important element to consider in human health and disease [@Ly:2014ew; @Monaco:2016ita; @Willner:2009dq; @Abeles:2015dy; @Modi:2013fi; @SantiagoRodriguez:2015gd; @Norman:2015kb], but it is likely to have a significant impact on cancer etiology and progression as well.
+Not only is the human virome an important element to consider in human health and disease [@Ly:2014ew; @Monaco:2016ita; @Willner:2009dq; @Abeles:2015dy; @Modi:2013fi; @SantiagoRodriguez:2015gd; @Norman:2015kb], but our findings support that it is likely to have a significant impact on cancer etiology and progression.
 
@@
 Abundance was corrected for contig reference length and the number of contigs included in each operational genomic unit.
-Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1,000,000 and 500,000 reads <!--what was the distribution of reads per smaple?
-or at least max/min/median?
-not sure how to think about these thresholds without some grounding, quick supp fig or table addition?-->, respectively, and by removing samples with fewer total reads than the threshold.
+Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1,000,000 and 500,000 reads, respectively, and by removing samples with fewer total reads than the threshold.
 Thresholds were set for maximizing sequence information while minimizing numbers of lost samples.
@@
 In both cases, the blastx algorithm was used with an e-value threshold of 10e-5.
-Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes <!--and anything to ensure that they weren't actually bacteria?
-steps to identify rRNA genes or other "diagnostic" bacterial genes, e.g., single-copy or housekeeping genes that haven't really been reported in phage, etc.
--->.
+Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes.
 To accomplish this, representative phage contigs were aligned to the European Nucleotide Archive bacterial genome reference set using the blastn algorithm (e-value < 10e-25).

### 2017-06-18 doc/manuscript.Rmd
Added edits from MD.

@@
     variant: markdown
-fontsize: 11pt
+fontsize: 12pt
 mainfont: 'Helvetica'
@@
 We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.
-<!-- I would change the structure of the first two paragraphs to be more linear, but I understand if you want to put it out there as is for time, and because you are probably getting tired of my comments.
-On high level it reads as "1.
-Viruses play a role in cancer and evidence points to a role in colorectal cancer.
-2.
-Colorectal cancer is a big deal.
-We know about the relationship with microbes.
-3.
-We should study viruses of colorectal cancer." To be less circular, I'd line it up as "1.
-Colorectal cancer is a big deal.
-We know about the relationship with microbes.
-2.
-Viruses play a role in cancer and evidence points to a role in colorectal cancer too.
-3.
-We should study viruses of colorectal cancer." -->
 
@@
 Thus, work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.
-<!--I feel this fits oddly here, one solution could be to put these stages of progression/need for early detection above when you introduce why colorectal cancer is a big deal...and how this argument of early detection motivates microbiome research...and now we have the logic to say that virome research could further improve upon this, as we know nothing of the role of viruses...that is the knowledge gap, then leads in well to this last paragraph.-->
 
@@
 We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
-<!-- which of these features was most responsible for identifying the temperate phages here?
-Guessing #3, but hoping for #1 or #2--> We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
+We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
 Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that have the ability to integrate into host bacterial genomes.
@@
 
-Here, we have begun to delineate the role the colonic virome plays <!--I think ok to leave this as present tense, as then it reads more generally...we have begun to delineate something general and bigger than this individual study-->in colorectal cancer **(Figure \ref{modelsummary} A)**.
+Here, we have begun to delineate the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**.
 We found that basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states.
@@
 Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
-Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host <!--ref for this?
-I know i've read it someone, if not shown explicitely, then in a review, like https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3760236/ and https://www.ncbi.nlm.nih.gov/pubmed/22748808 and https://www.nature.com/articles/srep14802-->.
+Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host [@Lengeling:2013ia; @Grski:2012fa].
 
@@
 We hypothesize that the process began with broadly infectious phages in the colon lysing and thereby disrupting the existing bacterial communities.
-<!--do we have the time resolution to suggest that the phages started this?
-and what could trigger that?
-What about the alternative hypothesis that the microbes began the shift and their phages responded, or (a more specific instance of your proposed case) that integrated prophges responded to cues from an altered environment, e.g., mediated byv a shift in transcriptional regulation of their hosts, that caused them to switch to lysis and drive a community shift.
---> This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize.
+This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize.
 Once the initial influential founder bacteria established themselves in the epithelium, secondary opportunistic bacteria were able to adhere to the founders, colonize, and begin establishing a biofilm.

### 2017-06-18 doc/manuscript.Rmd
A few more minor edits.

@@
 We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
-Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that have the ability to integrate into host bacterial genomes.
 These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
@@
 ## Ecological Network Analysis & Correlations
-The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described (**Cite Network Preprint**).
+The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described [@Hannigan:2017cj].
 Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1).
@@
 C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S rRNA gene classification model when removed from the classification model.
-Results based on 15 iterations.
 Mean is represented by a point, and bars represent standard error.
-D) Mean decrease in accuracy of each operational genomic unit in the virome classification model.
+D) Mean decrease in accuracy of each operational virus unit in the virome classification model.
 E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S rRNA gene and virome features.*\label{predmodel}](../figures/predmodel-viromebacteria.pdf){ width=100% }

### 2017-06-19 doc/manuscript.Rmd
Forgot to push these too.

@@
 
-| ***Journal***: PNAS ([*Preparation Details*](http://www.pnas.org/site/authors/procedures.xhtml))
 | ***Major Classification***: Biological Sciences
@@
 ***Keywords***: Colorectal Cancer, Virome, Machine Learning
-
-\* *Figures included for internal editing purposes*
 
@@
 
-**Word Count**: 213 / 250
+\newpage
 
@@
 These findings both support a biological role for viruses in colorectal cancer and provide a new understanding of basic colorectal cancer etiology.
-
-**Word Count**: 111 / 120
 
@@
 
-# Methods
-
-## Analysis Source Code & Data Availability
+# Materials & Methods
+This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent.
+A detailed description of protocols used for sample collection, processing, and analysis is provided in SI Materials and Methods.
 All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`.
-
 All associated source code is available at the following GitHub repository:
 
-https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017
-
-## Study Design and Patient Sampling
-This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent.
-Design and sampling of this sample set have been reported previously [@Zackular:2014fba].
-Briefly, whole evacuated stool was collected from patients who were 18 years of age or older, able to provide informed consent, have had colonoscopy and histologically confirmed colonic disease status, had not had surgery, had not had chemotherapy or radiation, and were free of known co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and inflammatory bowel disease.
-Samples were collected from four geographic locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor (Michigan, USA).
-Ninety patients were recruited to the study, thirty of which were designated healthy, thirty with detected adenomas, and thirty with detected carcinomas.
-
-## 16S rRNA Gene Sequence Data Acquisition & Processing
-The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba].
-Sequence (fastq) and metadata files were downloaded from http://www.mothur.org/MicrobiomeBiomarkerCRC.
-The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:2013db; @Schloss:2009do].
-Briefly, the sequences were de-replicated, aligned to the SILVA database [@Pruesse:2007jc], screened for chimeras using UCHIME [@Edgar:2011gy], and binned into operational taxonomic units (OTUs) using a 97% similarity threshold.
-Abundances were normalized for uneven sequencing depth by randomly sub-sampling to 10,000 sequences, as previously reported [@Baxter:2016dja].
-
-## Whole Metagenomic Library Preparation & Sequencing
-DNA was extracted from stool samples using the PowerSoil-htp 96 Well Soil DNA Isolation Kit (Mo Bio Laboratories) using an EPMotion 5075 pipetting system.
-Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT library preparation kit according to the standard kit protocol, including 12 cycles of limited cycle PCR.
-The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
-The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
-
-## Virus Metagenomic Library Preparation & Sequencing
-Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz; @Thurber:2009dn; @Kleiner:2015kd; @Minot:2011ez].
-Briefly, an aliquot of stool (~0.1 g) was resuspended in SM buffer (Crystalgen; Catalog #: 221-179) and vortexed to facilitate resuspension.
-The resuspended stool was centrifuged to remove major particulate debris then filtered through a 0.22-µm filter to remove smaller contaminants.
-The filtered supernatant was treated with chloroform for ten minutes with gentle shaking, so as to lyse contaminating cells including bacteria, human, fungi, etc.
-The exposed genomic DNA from the lysed cells was degraded by treating the samples with 5U of DNase for one hour at 37C.
-DNase was deactivated by incubating the sample at 75C for ten minutes.
-The DNA was extracted from the purified virus-like particles (VLPs) using the Wizard PCR Purification Preparation Kit (Promega).
-Disease classes were staggered across purification runs to prevent run variation as a confounding factor.
-As for whole community metagenomes, purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT preparation kit according to the standard kit protocol.
-The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
-The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz].
-The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
-
-## Metagenome Quality Control
-Both the viral and whole community metagenomic sample sets were subjected to the same quality control procedures.
-The sequences were obtained as de-multiplexed fastq files and subjected to 5' and 3' adapter trimming using the CutAdapt program (v1.9.1) with an error rate of 0.1 and an overlap of 10 [@Martin:2011eu].
-The FastX toolkit (v0.0.14) was used to quality trim the reads to a minimum length of 75 bp and a minimum quality score of 30 [@FASTXToolkit:wr].
-Reads mapping to the human genome were removed using the DeconSeq algorithm (v0.4.3) and default parameters [@Schmieder:2011fo].
-
-## Contig Assembly & Abundance
-Contigs were assembled using paired end read files that were purged of sequences without a corresponding pair (e.g.
-one read removed due to low quality).
-The Megahit program (v1.0.6) was used to assemble contigs for each sample using a minimum contig length of 1000 bp and iterating assemblies from 21-mers to 101-mers by 20 [@Li:2016kd].
-Contigs from the virus and whole metagenomic sample sets were concatenated within their respective groups.
-Abundance of the contigs within each sample was calculated by aligning sequences back to the concatenated contig files using the bowtie2 global aligner (v2.2.1), with a 25 bp seed length and an allowance of one mismatch [@Langmead:2012jh].
-Abundance was corrected for contig reference length and the number of contigs included in each operational genomic unit.
-Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1,000,000 and 500,000 reads, respectively, and by removing samples with fewer total reads than the threshold.
-Thresholds were set for maximizing sequence information while minimizing numbers of lost samples.
-
-## Operational Genomic Unit Classification
-Much like operational taxonomic units (OTUs) are used as an operational definition of similar 16S rRNA gene sequences, we defined closely related bacterial contig sequences as operational genomic units (OGUs) and virus contigs as operational viral units (OVUs) in the absence of taxonomic identity.
-OGUs and OVUs were defined with the CONCOCT algorithm (v0.4.0) which bins related contigs by similar tetra-mer and co-abundance profiles within samples using a variational Bayesian approach [@Alneberg:2014fc].
-CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria.
-
-## Diversity
-Alpha and beta diversity were calculated using the operational viral unit abundance profiles for each sample.
-Sequences were rarefied to 100,000 sequences.
-Samples with less than the cutoff were removed from the analysis.
-Alpha diversity was calculated using the Shannon entropy and richness metrics.
-Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (ANOSIM) with a post-hoc multivariate Tukey test.
-All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE].
-
-## Classification Modeling
-Classification modeling was performed in R using the Caret package [@caretClassificatio:_U2Lit_1].
-OTU, OVU, and OGU abundance data was preprocessed by removing features (OTUs, OVUs, and OGUs) that were present in less than thirty of the samples.
-This served both as an effective feature reduction technique and made the calculations computationally feasible.
-The binary random forest model was trained using the Area Under the receiver operating characteristic Curve (AUC) and the three-class random forest model was trained using the mean AUC.
-Both were validated using five-fold cross validation.
-Each training set was repeated five times, and the model was tuned for mtry values.
-For consistency and accurate comparison between feature groups (e.g., bacteria, viruses), the sample model parameters were used for each group.
-The maximum AUC during training was recorded across twenty iterations of each group model to test the significance of the differences between feature set performance.
-Statistical significance was evaluated using a Wilcoxon test between two categories, or a pairwise Wilcoxon test with Bonferroni corrected p-values when comparing more than two categories.
-
-## Taxonomic Identification of Operational Genomic Units
-Operational viral units (OVUs) were taxonomically identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
-The longest contiguous sequence in each operational genomic unit was used as a representative sequence for classification, as described previously [@Guidi:2016kf].
-Each representative sequence was aligned to the reference genome database using the tblastx alignment algorithm (v2.2.27) and a strict similarity threshold (e-value < 1e-25) [@Camacho:2009fc].
-Annotation was interpreted as phage, eukaryotic virus, or unknown.
-
-## Ecological Network Analysis & Correlations
-The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described [@Hannigan:2017cj].
-Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1).
-The degree of phage centrality was quantified using the alpha centrality metric in the igraph CRAN package.
-A Spearman correlation was performed between model importance and phage centrality scores.
-
-## Phage Replication Style Identification
-Phage OVU replication mode was predicted using methods described previously [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
-Briefly, we identified temperate OVUs as representative contigs containing at least one of three genomic markers: 1) phage integrase genes, 2) prophage genes from the ACLAME database, or 3) genomic similarity to bacterial reference genomes.
-Integrase genes were identified in phage OVU representative contigs by aligning the contigs to a reference database of all known phage integrase genes from the Uniprot database (Uniprot search term: "organism:phage gene:int NOT putative").
-Prophage genes were identified in the same way, using the ACLAME set of reference prophage genes.
-In both cases, the blastx algorithm was used with an e-value threshold of 10e-5.
-Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes.
-To accomplish this, representative phage contigs were aligned to the European Nucleotide Archive bacterial genome reference set using the blastn algorithm (e-value < 10e-25).
+https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017.
 
@@
 
+# Supplemental Materials & Methods
+
+## Analysis Source Code & Data Availability
+All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`.
+
+All associated source code is available at the following GitHub repository:
+
+https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017
+
+## Study Design and Patient Sampling
+This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent.
+Design and sampling of this sample set have been reported previously [@Zackular:2014fba].
+Briefly, whole evacuated stool was collected from patients who were 18 years of age or older, able to provide informed consent, have had colonoscopy and histologically confirmed colonic disease status, had not had surgery, had not had chemotherapy or radiation, and were free of known co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and inflammatory bowel disease.
+Samples were collected from four geographic locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor (Michigan, USA).
+Ninety patients were recruited to the study, thirty of which were designated healthy, thirty with detected adenomas, and thirty with detected carcinomas.
+
+## 16S rRNA Gene Sequence Data Acquisition & Processing
+The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba].
+Sequence (fastq) and metadata files were downloaded from http://www.mothur.org/MicrobiomeBiomarkerCRC.
+The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:2013db; @Schloss:2009do].
+Briefly, the sequences were de-replicated, aligned to the SILVA database [@Pruesse:2007jc], screened for chimeras using UCHIME [@Edgar:2011gy], and binned into operational taxonomic units (OTUs) using a 97% similarity threshold.
+Abundances were normalized for uneven sequencing depth by randomly sub-sampling to 10,000 sequences, as previously reported [@Baxter:2016dja].
+
+## Whole Metagenomic Library Preparation & Sequencing
+DNA was extracted from stool samples using the PowerSoil-htp 96 Well Soil DNA Isolation Kit (Mo Bio Laboratories) using an EPMotion 5075 pipetting system.
+Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT library preparation kit according to the standard kit protocol, including 12 cycles of limited cycle PCR.
+The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
+The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
+
+## Virus Metagenomic Library Preparation & Sequencing
+Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz; @Thurber:2009dn; @Kleiner:2015kd; @Minot:2011ez].
+Briefly, an aliquot of stool (~0.1 g) was resuspended in SM buffer (Crystalgen; Catalog #: 221-179) and vortexed to facilitate resuspension.
+The resuspended stool was centrifuged to remove major particulate debris then filtered through a 0.22-µm filter to remove smaller contaminants.
+The filtered supernatant was treated with chloroform for ten minutes with gentle shaking, so as to lyse contaminating cells including bacteria, human, fungi, etc.
+The exposed genomic DNA from the lysed cells was degraded by treating the samples with 5U of DNase for one hour at 37C.
+DNase was deactivated by incubating the sample at 75C for ten minutes.
+The DNA was extracted from the purified virus-like particles (VLPs) using the Wizard PCR Purification Preparation Kit (Promega).
+Disease classes were staggered across purification runs to prevent run variation as a confounding factor.
+As for whole community metagenomes, purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT preparation kit according to the standard kit protocol.
+The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
+The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz].
+The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
+
+## Metagenome Quality Control
+Both the viral and whole community metagenomic sample sets were subjected to the same quality control procedures.
+The sequences were obtained as de-multiplexed fastq files and subjected to 5' and 3' adapter trimming using the CutAdapt program (v1.9.1) with an error rate of 0.1 and an overlap of 10 [@Martin:2011eu].
+The FastX toolkit (v0.0.14) was used to quality trim the reads to a minimum length of 75 bp and a minimum quality score of 30 [@FASTXToolkit:wr].
+Reads mapping to the human genome were removed using the DeconSeq algorithm (v0.4.3) and default parameters [@Schmieder:2011fo].
+
+## Contig Assembly & Abundance
+Contigs were assembled using paired end read files that were purged of sequences without a corresponding pair (e.g.
+one read removed due to low quality).
+The Megahit program (v1.0.6) was used to assemble contigs for each sample using a minimum contig length of 1000 bp and iterating assemblies from 21-mers to 101-mers by 20 [@Li:2016kd].
+Contigs from the virus and whole metagenomic sample sets were concatenated within their respective groups.
+Abundance of the contigs within each sample was calculated by aligning sequences back to the concatenated contig files using the bowtie2 global aligner (v2.2.1), with a 25 bp seed length and an allowance of one mismatch [@Langmead:2012jh].
+Abundance was corrected for contig reference length and the number of contigs included in each operational genomic unit.
+Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1,000,000 and 500,000 reads, respectively, and by removing samples with fewer total reads than the threshold.
+Thresholds were set for maximizing sequence information while minimizing numbers of lost samples.
+
+## Operational Genomic Unit Classification
+Much like operational taxonomic units (OTUs) are used as an operational definition of similar 16S rRNA gene sequences, we defined closely related bacterial contig sequences as operational genomic units (OGUs) and virus contigs as operational viral units (OVUs) in the absence of taxonomic identity.
+OGUs and OVUs were defined with the CONCOCT algorithm (v0.4.0) which bins related contigs by similar tetra-mer and co-abundance profiles within samples using a variational Bayesian approach [@Alneberg:2014fc].
+CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria.
+
+## Diversity
+Alpha and beta diversity were calculated using the operational viral unit abundance profiles for each sample.
+Sequences were rarefied to 100,000 sequences.
+Samples with less than the cutoff were removed from the analysis.
+Alpha diversity was calculated using the Shannon entropy and richness metrics.
+Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (ANOSIM) with a post-hoc multivariate Tukey test.
+All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE].
+
+## Classification Modeling
+Classification modeling was performed in R using the Caret package [@caretClassificatio:_U2Lit_1].
+OTU, OVU, and OGU abundance data was preprocessed by removing features (OTUs, OVUs, and OGUs) that were present in less than thirty of the samples.
+This served both as an effective feature reduction technique and made the calculations computationally feasible.
+The binary random forest model was trained using the Area Under the receiver operating characteristic Curve (AUC) and the three-class random forest model was trained using the mean AUC.
+Both were validated using five-fold cross validation.
+Each training set was repeated five times, and the model was tuned for mtry values.
+For consistency and accurate comparison between feature groups (e.g., bacteria, viruses), the sample model parameters were used for each group.
+The maximum AUC during training was recorded across twenty iterations of each group model to test the significance of the differences between feature set performance.
+Statistical significance was evaluated using a Wilcoxon test between two categories, or a pairwise Wilcoxon test with Bonferroni corrected p-values when comparing more than two categories.
+
+## Taxonomic Identification of Operational Genomic Units
+Operational viral units (OVUs) were taxonomically identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
+The longest contiguous sequence in each operational genomic unit was used as a representative sequence for classification, as described previously [@Guidi:2016kf].
+Each representative sequence was aligned to the reference genome database using the tblastx alignment algorithm (v2.2.27) and a strict similarity threshold (e-value < 1e-25) [@Camacho:2009fc].
+Annotation was interpreted as phage, eukaryotic virus, or unknown.
+
+## Ecological Network Analysis & Correlations
+The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described [@Hannigan:2017cj].
+Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1).
+The degree of phage centrality was quantified using the alpha centrality metric in the igraph CRAN package.
+A Spearman correlation was performed between model importance and phage centrality scores.
+
+## Phage Replication Style Identification
+Phage OVU replication mode was predicted using methods described previously [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
+Briefly, we identified temperate OVUs as representative contigs containing at least one of three genomic markers: 1) phage integrase genes, 2) prophage genes from the ACLAME database, or 3) genomic similarity to bacterial reference genomes.
+Integrase genes were identified in phage OVU representative contigs by aligning the contigs to a reference database of all known phage integrase genes from the Uniprot database (Uniprot search term: "organism:phage gene:int NOT putative").
+Prophage genes were identified in the same way, using the ACLAME set of reference prophage genes.
+In both cases, the blastx algorithm was used with an e-value threshold of 10e-5.
+Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes.
+To accomplish this, representative phage contigs were aligned to the European Nucleotide Archive bacterial genome reference set using the blastn algorithm (e-value < 10e-25).
+
+\newpage
+
 # References

### 2017-06-20 doc/manuscript.Rmd
Updated with pat edits and format.

@@
 
-***Keywords***: Colorectal Cancer, Virome, Machine Learning
+***Keywords***: Colorectal Cancer, Virome, Machine Learning, Bacteriophage, Phage, Microbiome, Microbiota
 
@@
 Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world.
-Its risk has been linked to changes in colonic bacterial community composition.
+Colorectal cancer development has been linked to differences in colonic bacterial community composition.
 Viruses are another important component of the colonic microbial community, however they have yet to be studied in colorectal cancer despite their oncogenic potential.
@@
 We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community.
-We identified no detectable difference in diversity (alpha or beta) between healthy, adenomatous, or cancerous colonic samples, but more sophisticated random forest models identified striking changes in the virus community.
+We identified no detectable difference in viral diversity (alpha or beta) between healthy, adenomatous, or cancerous colonic samples, but more sophisticated random forest models identified striking differences in the virome.
 The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function.
-Our data suggested that the influential phages did not exclusively infect influential bacteria, but rather acted through the community as a whole.
+Our data suggest that the influential phages do not exclusively infect influential bacteria, but rather act through the community as a whole.
 These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
@@
 Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities.
-We addressed this knowledge gap by identifying changes in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community changes.
+We addressed this knowledge gap by identifying differences in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community differences.
 The results suggested an indirect role for the virome in impacting colorectal cancer by modulating their associated bacterial community.
@@
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer.
-The gut virome (the virus community of the gut, including eukaryotic and prokaryotic viruses) therefore has the potential to impact health and disease (e.g.
-cancer).
+Because bacteriophages (viruses that specifically infect bacteria) are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer.
+The gut virome (the virus community of the gut) therefore has the potential to impact health and disease.
 Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
@@
 Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo].
-The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016 and over 500,000 Americans have died from the disease [@Siegel:2014jo].
-An important component of colorectal cancer etiology is variation in colorectal bacterial community composition [@Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb; @Dejea:2014fz; @Arthur:2012kl].
+The US National Cancer Institute estimates over 1.5 million Americans were diagnosed with colorectal cancer in 2016 and over 500,000 Americans died from the disease [@Siegel:2014jo].
+Growing evidence suggests that an important component of colorectal cancer etiology may be perturbations in the colonic bacterial community [@Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb; @Dejea:2014fz; @Arthur:2012kl].
 Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
@@
 Colorectal cancer progresses in a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e., adenoma) in the large intestine [@Fearon:2011go].
-If left untreated, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death.
+If not removed, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death.
 Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr].
@@
 
-Here we address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those changes might impact cancer progression and severity.
+Here we address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those differences might impact cancer progression and severity.
 We also aimed to evaluate the virome's potential for use as a diagnostic biomarker.
@@
 Half of each stool sample was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques.
+The 16S rRNA gene sequencing was performed for a previous study, and the sequences were re-analyzed using contemporary methods [@Zackular:2014fba].
 The other half of each stool sample was purified for virus like particles (VLPs) before genomic DNA extraction and shotgun metagenomic sequencing.
@@
 We therefore first evaluated the influence of colorectal cancer on virome OVU diversity.
-We evaluated differences in communities between disease states using the Shannon entropy, richness, and Bray-Curtis metrics.
-We observed no significant alterations in either Shannon entropy or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**.
+We evaluated differences in communities between disease states using the Shannon diversity, richness, and Bray-Curtis metrics.
+We observed no significant alterations in either Shannon diversity or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**.
 There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**).
@@
 In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**.
-This is consistent with what has been observed when the same metrics were applied to metagenomic samples [@Zeller:2014ix] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.
+This is consistent with what has been observed when the same metrics were applied to 16S rRNA sequenced and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.
 
 ## Altered Virome Composition in Colorectal Cancer
-As opposed to the diversity metrics discussed above, 16S rRNA gene relative abundance profiles are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
+As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
 The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer.
@@
 We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba].
-We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance.
-The viral model performed as well as the bacterial model (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"])`), with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = asigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = asigfig)`, respectively **(Figure \ref{predmodel} A - B)**.
-
+We then compared the bacterial OTU model to a model built using OVU relative abundances.
+The viral model performed as well as the bacterial model (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"])`), with the viral and bacterial models achieving mean area under the curve (AUC) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = asigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = asigfig)`, respectively **(Figure \ref{predmodel} A - B)**.
 To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data.
@@
 
-To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data.
+To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using OGU relative abundance profiles from bacterial metagenomic shotgun sequencing data.
 This model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**.
-Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**.
+Because the coverage provided by the metagenomic sequencing was not as deep as the equivalent 16S rRNA gene sequencing, we attempted to compare the approaches at a common sequencing depth.
+This investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**.
 Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**.
@@
 
-The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed.
-*Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu].
-The virome signature also was driven by a few OVUs, suggesting a role for these viruses in cancer development **(Figure \ref{predmodel} D)**.
+The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes.
+*Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu].
+The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**.
 The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and "unclassified" phage taxa.
@@
 We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions.
-While bacterial 16S rRNA gene models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**.
+While bacterial OTU models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**.
 Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important between the healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.
@@
 The 16S rRNA gene model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = asigfig)` and outperformed the viral community model, which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = asigfig)` (p-value `r pvalcalc(threesig)`, **Figure \ref{threewaymodel} A-C**).
-The microbes important for the cancer versus healthy and healthy versus adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
+The microbes important for the healthy versus cancer and healthy versus adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**.
 The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**.
@@
 ## Bacteriophage Dominance in Colorectal Cancer Virome
-Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts.
+Differences in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts.
 To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages.
@@
 
-We evaluated whether the phages in the community were primarily lytic (obligately lyse their hosts after replication) or temperate (able to integrate into their host's genome to form a lysogen, and subsequently transition to a lytic mode).
-We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
-We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
+We evaluated whether the phages in the community were primarily lytic (i.e.
+obligately lyse their hosts after replication) or temperate (i.e.
+able to integrate into their host's genome to form a lysogen, and subsequently transition to a lytic mode).
+We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
+We found that the majority of the phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
 These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
@@
 Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**.
-Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.
+Together these findings supported our hypothesis that influential phages were hubs within their microbial communities and had broad host ranges.
 
@@
 Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
-Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colorectal communities and colorectal cancer has yet to be evaluated.
-Here we show that, like colonic bacterial communities, the colon virome was altered in colorectal cancer.
+Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, a mechanistic link between virus colorectal communities and colorectal cancer has yet to be evaluated.
+Here we show that, like colonic bacterial communities, the colon virome was altered in patients with colorectal cancer relative to those with healthy colons.
 Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.
@@
 Here, we have begun to delineate the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**.
-We found that basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states.
+We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states.
 By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer.
@@
 In that former case, the models only performed as well as the 16S rRNA gene model [@Zeller:2014ix].
-Thus, the targeted 16S rRNA gene sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples.
+Thus, the targeted 16S rRNA gene sequencing approach, which represented only a fraction of the bacterial metagenomic sequencing depth, was more effective for detecting colorectal cancer in stool samples.
 Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
 
-In addition to the therapeutic ramifications for understanding the colorectal cancer microbiome, our findings provide a proof-of-principle that viruses, while understudied and currently under-appreciated in the human microbiome, are an important contributor to human disease.
-Viral community dynamices have the potential to provide an abundance of information to supplement those of bacterial communities.
+In addition to the diagnostic ramifications for understanding the colorectal cancer microbiome, our findings suggest that viruses, while understudied and currently under-appreciated in the human microbiome, are likely to be an important contributor to human disease.
+Viral community dynamics have the potential to provide an abundance of information to supplement those of bacterial communities.
 Evidence has suggested that the virome is a crucial component to the microbiome and that bacteriophages are important players.
@@
 https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017.
-
-# Conflicts of Interest
-The authors declare no conflicts of interest.
 
@@
 B) Working hypothesis of how the bacteriophage community is associated with colorectal cancer and the associated bacterial community.*\label{modelsummary}](../figures/WorkingHypothesis.pdf){ width=100% }
+
+\newpage
+
+# Supplemental Information
+
+# Supplemental Materials & Methods
+
+## Analysis Source Code & Data Availability
+All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`.
+
+All associated source code is available at the following GitHub repository:
+
+https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017
+
+## Study Design and Patient Sampling
+This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent.
+Design and sampling of this sample set have been reported previously [@Zackular:2014fba].
+Briefly, whole evacuated stool was collected from patients who were 18 years of age or older, able to provide informed consent, have had colonoscopy and histologically confirmed colonic disease status, had not had surgery, had not had chemotherapy or radiation, and were free of known co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and inflammatory bowel disease.
+Samples were collected from four geographic locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor (Michigan, USA).
+Ninety patients were recruited to the study, thirty of which were designated healthy, thirty with detected adenomas, and thirty with detected carcinomas.
+
+## 16S rRNA Gene Sequence Data Acquisition & Processing
+The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba].
+Sequence (fastq) and metadata files were downloaded from http://www.mothur.org/MicrobiomeBiomarkerCRC.
+The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:2013db; @Schloss:2009do].
+Briefly, the sequences were de-replicated, aligned to the SILVA database [@Pruesse:2007jc], screened for chimeras using UCHIME [@Edgar:2011gy], and binned into operational taxonomic units (OTUs) using a 97% similarity threshold.
+Abundances were normalized for uneven sequencing depth by randomly sub-sampling to 10,000 sequences, as previously reported [@Baxter:2016dja].
+
+## Whole Metagenomic Library Preparation & Sequencing
+DNA was extracted from stool samples using the PowerSoil-htp 96 Well Soil DNA Isolation Kit (Mo Bio Laboratories) using an EPMotion 5075 pipetting system.
+Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT library preparation kit according to the standard kit protocol, including 12 cycles of limited cycle PCR.
+The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
+The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
+
+## Virus Metagenomic Library Preparation & Sequencing
+Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz; @Thurber:2009dn; @Kleiner:2015kd; @Minot:2011ez].
+Briefly, an aliquot of stool (~0.1 g) was resuspended in SM buffer (Crystalgen; Catalog #: 221-179) and vortexed to facilitate resuspension.
+The resuspended stool was centrifuged to remove major particulate debris then filtered through a 0.22-µm filter to remove smaller contaminants.
+The filtered supernatant was treated with chloroform for ten minutes with gentle shaking, so as to lyse contaminating cells including bacteria, human, fungi, etc.
+The exposed genomic DNA from the lysed cells was degraded by treating the samples with 5U of DNase for one hour at 37C.
+DNase was deactivated by incubating the sample at 75C for ten minutes.
+The DNA was extracted from the purified virus-like particles (VLPs) using the Wizard PCR Purification Preparation Kit (Promega).
+Disease classes were staggered across purification runs to prevent run variation as a confounding factor.
+As for whole community metagenomes, purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT preparation kit according to the standard kit protocol.
+The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
+The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz].
+The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
+
+## Metagenome Quality Control
+Both the viral and whole community metagenomic sample sets were subjected to the same quality control procedures.
+The sequences were obtained as de-multiplexed fastq files and subjected to 5' and 3' adapter trimming using the CutAdapt program (v1.9.1) with an error rate of 0.1 and an overlap of 10 [@Martin:2011eu].
+The FastX toolkit (v0.0.14) was used to quality trim the reads to a minimum length of 75 bp and a minimum quality score of 30 [@FASTXToolkit:wr].
+Reads mapping to the human genome were removed using the DeconSeq algorithm (v0.4.3) and default parameters [@Schmieder:2011fo].
+
+## Contig Assembly & Abundance
+Contigs were assembled using paired end read files that were purged of sequences without a corresponding pair (e.g.
+one read removed due to low quality).
+The Megahit program (v1.0.6) was used to assemble contigs for each sample using a minimum contig length of 1000 bp and iterating assemblies from 21-mers to 101-mers by 20 [@Li:2016kd].
+Contigs from the virus and whole metagenomic sample sets were concatenated within their respective groups.
+Abundance of the contigs within each sample was calculated by aligning sequences back to the concatenated contig files using the bowtie2 global aligner (v2.2.1), with a 25 bp seed length and an allowance of one mismatch [@Langmead:2012jh].
+Abundance was corrected for contig reference length and the number of contigs included in each operational genomic unit.
+Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1,000,000 and 500,000 reads, respectively, and by removing samples with fewer total reads than the threshold.
+Thresholds were set for maximizing sequence information while minimizing numbers of lost samples.
+
+## Operational Genomic Unit Classification
+Much like operational taxonomic units (OTUs) are used as an operational definition of similar 16S rRNA gene sequences, we defined closely related bacterial contig sequences as operational genomic units (OGUs) and virus contigs as operational viral units (OVUs) in the absence of taxonomic identity.
+OGUs and OVUs were defined with the CONCOCT algorithm (v0.4.0) which bins related contigs by similar tetra-mer and co-abundance profiles within samples using a variational Bayesian approach [@Alneberg:2014fc].
+CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria.
+
+## Diversity
+Alpha and beta diversity were calculated using the operational viral unit abundance profiles for each sample.
+Sequences were rarefied to 100,000 sequences.
+Samples with less than the cutoff were removed from the analysis.
+Alpha diversity was calculated using the Shannon diversity and richness metrics.
+Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (ANOSIM) with a post-hoc multivariate Tukey test.
+All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE].
+
+## Classification Modeling
+Classification modeling was performed in R using the Caret package [@caretClassificatio:_U2Lit_1].
+OTU, OVU, and OGU abundance data was preprocessed by removing features (OTUs, OVUs, and OGUs) that were present in less than thirty of the samples.
+This served both as an effective feature reduction technique and made the calculations computationally feasible.
+The binary random forest model was trained using the Area Under the receiver operating characteristic Curve (AUC) and the three-class random forest model was trained using the mean AUC.
+Both were validated using five-fold cross validation.
+Each training set was repeated five times, and the model was tuned for mtry values.
+For consistency and accurate comparison between feature groups (e.g., bacteria, viruses), the sample model parameters were used for each group.
+The maximum AUC during training was recorded across twenty iterations of each group model to test the significance of the differences between feature set performance.
+Statistical significance was evaluated using a Wilcoxon test between two categories, or a pairwise Wilcoxon test with Bonferroni corrected p-values when comparing more than two categories.
+
+## Taxonomic Identification of Operational Genomic Units
+Operational viral units (OVUs) were taxonomically identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
+The longest contiguous sequence in each operational genomic unit was used as a representative sequence for classification, as described previously [@Guidi:2016kf].
+Each representative sequence was aligned to the reference genome database using the tblastx alignment algorithm (v2.2.27) and a strict similarity threshold (e-value < 1e-25) [@Camacho:2009fc].
+Annotation was interpreted as phage, eukaryotic virus, or unknown.
+
+## Ecological Network Analysis & Correlations
+The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described [@Hannigan:2017cj].
+Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1).
+The degree of phage centrality was quantified using the alpha centrality metric in the igraph CRAN package.
+A Spearman correlation was performed between model importance and phage centrality scores.
+
+## Phage Replication Style Identification
+Phage OVU replication mode was predicted using methods described previously [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
+Briefly, we identified temperate OVUs as representative contigs containing at least one of three genomic markers: 1) phage integrase genes, 2) prophage genes from the ACLAME database, or 3) genomic similarity to bacterial reference genomes.
+Integrase genes were identified in phage OVU representative contigs by aligning the contigs to a reference database of all known phage integrase genes from the Uniprot database (Uniprot search term: "organism:phage gene:int NOT putative").
+Prophage genes were identified in the same way, using the ACLAME set of reference prophage genes.
+In both cases, the blastx algorithm was used with an e-value threshold of 10e-5.
+Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes.
+To accomplish this, representative phage contigs were aligned to the European Nucleotide Archive bacterial genome reference set using the blastn algorithm (e-value < 10e-25).
 
@@
 
-# Supplemental Materials & Methods
-
-## Analysis Source Code & Data Availability
-All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`.
-
-All associated source code is available at the following GitHub repository:
-
-https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017
-
-## Study Design and Patient Sampling
-This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent.
-Design and sampling of this sample set have been reported previously [@Zackular:2014fba].
-Briefly, whole evacuated stool was collected from patients who were 18 years of age or older, able to provide informed consent, have had colonoscopy and histologically confirmed colonic disease status, had not had surgery, had not had chemotherapy or radiation, and were free of known co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and inflammatory bowel disease.
-Samples were collected from four geographic locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor (Michigan, USA).
-Ninety patients were recruited to the study, thirty of which were designated healthy, thirty with detected adenomas, and thirty with detected carcinomas.
-
-## 16S rRNA Gene Sequence Data Acquisition & Processing
-The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba].
-Sequence (fastq) and metadata files were downloaded from http://www.mothur.org/MicrobiomeBiomarkerCRC.
-The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:2013db; @Schloss:2009do].
-Briefly, the sequences were de-replicated, aligned to the SILVA database [@Pruesse:2007jc], screened for chimeras using UCHIME [@Edgar:2011gy], and binned into operational taxonomic units (OTUs) using a 97% similarity threshold.
-Abundances were normalized for uneven sequencing depth by randomly sub-sampling to 10,000 sequences, as previously reported [@Baxter:2016dja].
-
-## Whole Metagenomic Library Preparation & Sequencing
-DNA was extracted from stool samples using the PowerSoil-htp 96 Well Soil DNA Isolation Kit (Mo Bio Laboratories) using an EPMotion 5075 pipetting system.
-Purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT library preparation kit according to the standard kit protocol, including 12 cycles of limited cycle PCR.
-The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
-The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
-
-## Virus Metagenomic Library Preparation & Sequencing
-Genomic DNA was extracted from purified virus-like particles (VLPs) from stool samples, using a modified version of a previously published protocol [@Hannigan:2015fz; @Thurber:2009dn; @Kleiner:2015kd; @Minot:2011ez].
-Briefly, an aliquot of stool (~0.1 g) was resuspended in SM buffer (Crystalgen; Catalog #: 221-179) and vortexed to facilitate resuspension.
-The resuspended stool was centrifuged to remove major particulate debris then filtered through a 0.22-µm filter to remove smaller contaminants.
-The filtered supernatant was treated with chloroform for ten minutes with gentle shaking, so as to lyse contaminating cells including bacteria, human, fungi, etc.
-The exposed genomic DNA from the lysed cells was degraded by treating the samples with 5U of DNase for one hour at 37C.
-DNase was deactivated by incubating the sample at 75C for ten minutes.
-The DNA was extracted from the purified virus-like particles (VLPs) using the Wizard PCR Purification Preparation Kit (Promega).
-Disease classes were staggered across purification runs to prevent run variation as a confounding factor.
-As for whole community metagenomes, purified DNA was used to prepare a shotgun sequencing library using the Illumina Nextera XT preparation kit according to the standard kit protocol.
-The tagmentation time was increased from five minutes to ten minutes to improve DNA fragment length distribution.
-The PCR cycle number was increased from twelve to eighteen cycles to address the low biomass of the samples, as has been described previously [@Hannigan:2015fz].
-The library was sequenced using one lane of the Illumina HiSeq4000 platform and yielded 125 bp paired end reads.
-
-## Metagenome Quality Control
-Both the viral and whole community metagenomic sample sets were subjected to the same quality control procedures.
-The sequences were obtained as de-multiplexed fastq files and subjected to 5' and 3' adapter trimming using the CutAdapt program (v1.9.1) with an error rate of 0.1 and an overlap of 10 [@Martin:2011eu].
-The FastX toolkit (v0.0.14) was used to quality trim the reads to a minimum length of 75 bp and a minimum quality score of 30 [@FASTXToolkit:wr].
-Reads mapping to the human genome were removed using the DeconSeq algorithm (v0.4.3) and default parameters [@Schmieder:2011fo].
-
-## Contig Assembly & Abundance
-Contigs were assembled using paired end read files that were purged of sequences without a corresponding pair (e.g.
-one read removed due to low quality).
-The Megahit program (v1.0.6) was used to assemble contigs for each sample using a minimum contig length of 1000 bp and iterating assemblies from 21-mers to 101-mers by 20 [@Li:2016kd].
-Contigs from the virus and whole metagenomic sample sets were concatenated within their respective groups.
-Abundance of the contigs within each sample was calculated by aligning sequences back to the concatenated contig files using the bowtie2 global aligner (v2.2.1), with a 25 bp seed length and an allowance of one mismatch [@Langmead:2012jh].
-Abundance was corrected for contig reference length and the number of contigs included in each operational genomic unit.
-Abundance was also corrected for uneven sampling depth by randomly sub-sampling virome and whole metagenomes to 1,000,000 and 500,000 reads, respectively, and by removing samples with fewer total reads than the threshold.
-Thresholds were set for maximizing sequence information while minimizing numbers of lost samples.
-
-## Operational Genomic Unit Classification
-Much like operational taxonomic units (OTUs) are used as an operational definition of similar 16S rRNA gene sequences, we defined closely related bacterial contig sequences as operational genomic units (OGUs) and virus contigs as operational viral units (OVUs) in the absence of taxonomic identity.
-OGUs and OVUs were defined with the CONCOCT algorithm (v0.4.0) which bins related contigs by similar tetra-mer and co-abundance profiles within samples using a variational Bayesian approach [@Alneberg:2014fc].
-CONCOCT was used with a length threshold of 1000 bp for virus contigs and 2000 bp for bacteria.
-
-## Diversity
-Alpha and beta diversity were calculated using the operational viral unit abundance profiles for each sample.
-Sequences were rarefied to 100,000 sequences.
-Samples with less than the cutoff were removed from the analysis.
-Alpha diversity was calculated using the Shannon entropy and richness metrics.
-Beta diversity was calculated using the Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the statistical significance between the disease state clusters was assessed using an analysis of similarity (ANOSIM) with a post-hoc multivariate Tukey test.
-All diversity calculations were performed in R using the Vegan package [@veganCommunityEco:xdLliqSE].
-
-## Classification Modeling
-Classification modeling was performed in R using the Caret package [@caretClassificatio:_U2Lit_1].
-OTU, OVU, and OGU abundance data was preprocessed by removing features (OTUs, OVUs, and OGUs) that were present in less than thirty of the samples.
-This served both as an effective feature reduction technique and made the calculations computationally feasible.
-The binary random forest model was trained using the Area Under the receiver operating characteristic Curve (AUC) and the three-class random forest model was trained using the mean AUC.
-Both were validated using five-fold cross validation.
-Each training set was repeated five times, and the model was tuned for mtry values.
-For consistency and accurate comparison between feature groups (e.g., bacteria, viruses), the sample model parameters were used for each group.
-The maximum AUC during training was recorded across twenty iterations of each group model to test the significance of the differences between feature set performance.
-Statistical significance was evaluated using a Wilcoxon test between two categories, or a pairwise Wilcoxon test with Bonferroni corrected p-values when comparing more than two categories.
-
-## Taxonomic Identification of Operational Genomic Units
-Operational viral units (OVUs) were taxonomically identified using a reference database consisting of all bacteriophage and eukaryotic virus genomes present in the European Nucleotide Archives.
-The longest contiguous sequence in each operational genomic unit was used as a representative sequence for classification, as described previously [@Guidi:2016kf].
-Each representative sequence was aligned to the reference genome database using the tblastx alignment algorithm (v2.2.27) and a strict similarity threshold (e-value < 1e-25) [@Camacho:2009fc].
-Annotation was interpreted as phage, eukaryotic virus, or unknown.
-
-## Ecological Network Analysis & Correlations
-The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described [@Hannigan:2017cj].
-Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1).
-The degree of phage centrality was quantified using the alpha centrality metric in the igraph CRAN package.
-A Spearman correlation was performed between model importance and phage centrality scores.
-
-## Phage Replication Style Identification
-Phage OVU replication mode was predicted using methods described previously [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
-Briefly, we identified temperate OVUs as representative contigs containing at least one of three genomic markers: 1) phage integrase genes, 2) prophage genes from the ACLAME database, or 3) genomic similarity to bacterial reference genomes.
-Integrase genes were identified in phage OVU representative contigs by aligning the contigs to a reference database of all known phage integrase genes from the Uniprot database (Uniprot search term: "organism:phage gene:int NOT putative").
-Prophage genes were identified in the same way, using the ACLAME set of reference prophage genes.
-In both cases, the blastx algorithm was used with an e-value threshold of 10e-5.
-Representative contigs were also identified as potential lysogenic phages by having a high genomic similarity to bacterial genomes.
-To accomplish this, representative phage contigs were aligned to the European Nucleotide Archive bacterial genome reference set using the blastn algorithm (e-value < 10e-25).
-
-\newpage
-
 # References

### 2017-06-27 doc/manuscript.Rmd
Updated for new format.

@@
 ---
-title: Viral and Bacterial Communities of Colorectal Cancer
+title: Diagnostic Potential & Interactive Dynamics of Viral & Bacterial Communities in Colorectal Cancer
 author:
@@
   - code: 2
-    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
+    address: Department of Ecology & Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
   - code: 3
-    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
+    address: Department of Family & Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
   - code: \*
@@
     variant: markdown
+abstract: Colorectal cancer is a primary cause of morbidity and mortality throughout the world.
+Colorectal cancer development has been linked to differences in colonic bacterial community composition, however viruses have yet to be studied in colorectal cancer despite their oncogenic potential.
+We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition.
+Through random forest modeling we identified differences in the healthy and colorectal cancer virome.
+The cancer-associated virome consisted primarily of temperate bacteriophages that were also bacteria-virus community network hubs.
+These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
+font: Times
 fontsize: 12pt
-mainfont: 'Helvetica'
 ---
-
-&nbsp;
-
-| ***Corresponding Author Information***
-| Patrick D Schloss, PhD
-| 1150 W Medical Center Dr. 1526 MSRB I
-| Ann Arbor, Michigan 48109
-| Phone: (734) 647-5801
-| Email: pschloss@umich.edu
-
-| ***Major Classification***: Biological Sciences
-| ***Minor Classification***: Microbiology
-
-***Keywords***: Colorectal Cancer, Virome, Machine Learning, Bacteriophage, Phage, Microbiome, Microbiota
 
@@
 
-# Abstract
-
-Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world.
-Colorectal cancer development has been linked to differences in colonic bacterial community composition.
-Viruses are another important component of the colonic microbial community, however they have yet to be studied in colorectal cancer despite their oncogenic potential.
-We evaluated the colorectal cancer virome (virus community) in stool using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons.
-We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community.
-We identified no detectable difference in viral diversity (alpha or beta) between healthy, adenomatous, or cancerous colonic samples, but more sophisticated random forest models identified striking differences in the virome.
-The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function.
-Our data suggest that the influential phages do not exclusively infect influential bacteria, but rather act through the community as a whole.
-These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
-
-\newpage
-
-# Significance Statement
-
-Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide.
-Its risk and severity have been linked to colonic bacterial community composition.
-Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities.
-We addressed this knowledge gap by identifying differences in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community differences.
-The results suggested an indirect role for the virome in impacting colorectal cancer by modulating their associated bacterial community.
-These findings both support a biological role for viruses in colorectal cancer and provide a new understanding of basic colorectal cancer etiology.
-
-\newpage
-
-# Introduction
+## Introduction
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
@@
 *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.
-
-# Results
 
@@
 Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.
+Thus, the targeted 16S rRNA gene sequencing approach, which represented only a fraction of the bacterial metagenomic sequencing depth, was more effective for detecting colorectal cancer in stool samples.
+Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
 
@@
 
-# Discussion
+## Working Model for Virome & Cancer Progression
 Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
@@
 
-Here, we have begun to delineate the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**.
+We have begun to delineate the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**.
 We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states.
@@
 
-A notable finding was the poor performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences.
-We believe this observation speaks to the importance of sequencing coverage in microbial community studies and the advantage of the high coverage of 16S rRNA gene sequencing relative to the lower per OTU coverage possible using whole metagenomic shotgun sequencing.
-To demonstrate this concept, consider that six bacterial OTUs drove the performance of the 16S rRNA gene classification model and these OTUs were all sparsely present and lowly abundant.
-Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model.
-The bacterial metagenomic OGUs represented only the most abundant taxa, which was not informative for this application.
-There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification [@Zeller:2014ix], but this previous approach relied on lowly abundant signatures and did not utilize OGU clustering, as done here.
-In that former case, the models only performed as well as the 16S rRNA gene model [@Zeller:2014ix].
-Thus, the targeted 16S rRNA gene sequencing approach, which represented only a fraction of the bacterial metagenomic sequencing depth, was more effective for detecting colorectal cancer in stool samples.
-Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
+## Conclusions
 
@@
 
-# Materials & Methods
-This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent.
-A detailed description of protocols used for sample collection, processing, and analysis is provided in SI Materials and Methods.
-All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`.
-All associated source code is available at the following GitHub repository:
-
-https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017.
-
-# Acknowledgments
+## Acknowledgments
 The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
@@
 MT Ruffin was supported by funding from the National Institutes of Health (5U01CA86400).
+The authors declare no competing interests.
 
@@
 
-# Figures
+## Figures
 
@@
 
-# Supplemental Information
+# Supplemental Materials
 
-# Supplemental Materials & Methods
+# Materials & Methods
 
@@
 
-https://github.com/SchlossLab/Hannigan_CRCVirome_PNAS_2017
+https://github.com/SchlossLab/Hannigan_CRCVirome_Science_2017
 
@@
 ## 16S rRNA Gene Sequence Data Acquisition & Processing
-The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba].
-Sequence (fastq) and metadata files were downloaded from http://www.mothur.org/MicrobiomeBiomarkerCRC.
+The 16S rRNA gene sequences associated with this study were previously reported [@Zackular:2014fba]. Sequence (fastq) and metadata files were downloaded from:
+
+http://www.mothur.org/MicrobiomeBiomarkerCRC
+
 The 16S rRNA gene sequences were analyzed as described previously, relying on the mothur software package (v1.37.0) [@Kozich:2013db; @Schloss:2009do].
@@
 F) Relative abundance of the six most important virome OVUs in the model, with the most important on the right.
-Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=80% }
+Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=75% }
 

### 2017-06-27 doc/manuscript.Rmd
Updated manuscript with pat edits.

@@
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Because bacteriophages (viruses that specifically infect bacteria) are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer.
-The gut virome (the virus community of the gut) therefore has the potential to impact health and disease.
+Because bacteriophages (i.e.
+viruses that specifically infect bacteria) are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer.
+The gut virome (i.e.
+the virus community of the gut) therefore has the potential to impact health and disease.
 Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
@@
 Only one of the nine controls contained detectable DNA at a minimal concentration of 0.011 ng/µl, thus providing evidence of the enrichment and purification of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**.
-As was expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
+As expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**.
 The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1 kb **(Figure \ref{contigqc})**.
@@
 Microbiome and disease associations are often described as being of an altered diversity (i.e., "dysbiotic").
-We therefore first evaluated the influence of colorectal cancer on virome OVU diversity.
+Therefore, we first evaluated the influence of colorectal cancer on virome OVU diversity.
 We evaluated differences in communities between disease states using the Shannon diversity, richness, and Bray-Curtis metrics.
@@
 Thus, the targeted 16S rRNA gene sequencing approach, which represented only a fraction of the bacterial metagenomic sequencing depth, was more effective for detecting colorectal cancer in stool samples.
-Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
+Despite the recent loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
 
@@
 Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
-In other words, we hypothesized that the influential bacteriophages were community hubs (central members) within the bacteria and phage interactive network.
+In other words, we hypothesized that the influential bacteriophages were community hubs (i.e.
+central members) within the bacteria and phage interactive network.
 We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2017cj].
 The predicted interactions were then used to identify phage community hubs.
-We calculated the alpha centrality (measure of importance in the ecological network) of each phage OVU's connection to the rest of the network.
+We calculated the alpha centrality (i.e.
+measure of importance in the ecological network) of each phage OVU's connection to the rest of the network.
 The phages with high centrality values were defined as community hubs.
@@
 
-In addition to modeling the potential connections between virus communities, bacteria communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur.
+In addition to modeling the potential connections between virus communities, bacterial communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur.
 This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
-We hypothesize that the process began with broadly infectious phages in the colon lysing and thereby disrupting the existing bacterial communities.
-This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize.
-Once the initial influential founder bacteria established themselves in the epithelium, secondary opportunistic bacteria were able to adhere to the founders, colonize, and begin establishing a biofilm.
-Phages may have played a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj].
-The oncogenic bacteria may then have been able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response.
+We hypothesize that the process begins with broadly infectious phages in the colon lysing and thereby disrupting the existing bacterial communities.
+This shift opens novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize.
+Once the initial influential founder bacteria establish themselves in the epithelium, secondary opportunistic bacteria are able to adhere to the founders, colonize, and establish a biofilm.
+Phages may play a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj].
+The oncogenic bacteria may then be able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response.
 As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundances were most influential.
-As the bacteria entered their oncogenic synergy with the epithelium, we conjecture that the phages continued mediating biofilm dispersal.
+As the bacteria enter their oncogenic synergy with the epithelium, we conjecture that the phages continue mediating biofilm dispersal.
 This process would thereby support the colonized oncogenic bacteria by lysing competing cells and releasing nutrients to other bacteria in the form of cellular lysates.
-In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role the virome plays colorectal cancer.
+In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities this model will guide future research investigations of the role the virome plays colorectal cancer.
 

### 2017-07-11 doc/manuscript.Rmd
Updated man and cl.

@@
 ---
-title: Diagnostic Potential & Interactive Dynamics of Viral & Bacterial Communities in Colorectal Cancer
+title: Diagnostics & Interactive Dynamics of the Colorectal Cancer Virome
 author:
@@
     variant: markdown
-abstract: Colorectal cancer is a primary cause of morbidity and mortality throughout the world.
-Colorectal cancer development has been linked to differences in colonic bacterial community composition, however viruses have yet to be studied in colorectal cancer despite their oncogenic potential.
+abstract: Viruses are assocaited with many human cancers, largely due to their mutagenic and functionally manipulative abilities.
+Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses.
+We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions while the virome role remains unknown.
 We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition.
@@
 
-## Cohort Design, Sample Collection, and Processing
+## Sample Collection and Processing
 Our study cohort consisted of 90 human subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30 of whom had carcinomas **(Figure \ref{sampleproc})**.
@@
 
-## Unaltered Virome Diversity in Colorectal Cancer
+## Unaltered Diversity in Colorectal Cancer
 Microbiome and disease associations are often described as being of an altered diversity (i.e., "dysbiotic").
@@
 
-## Altered Virome Composition in Colorectal Cancer
+## Virome Composition in Colorectal Cancer
 As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
@@
 
-## Shifted Phage Influence Between Cancer Progression Stages
+## Phage Influence Between CRC Stages
 Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of specific phages could be detected between healthy, adenomatous, and cancerous colons.
@@
 
-## Bacteriophage Dominance in Colorectal Cancer Virome
+## Phage Dominance in CRC Virome
 Differences in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts.
@@
 
-## Working Model for Virome & Cancer Progression
+## Model for Virome & Cancer Progression
 Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
@@
 Not only is the human virome an important element to consider in human health and disease [@Ly:2014ew; @Monaco:2016ita; @Willner:2009dq; @Abeles:2015dy; @Modi:2013fi; @SantiagoRodriguez:2015gd; @Norman:2015kb], but our findings support that it is likely to have a significant impact on cancer etiology and progression.
+
+## Materials and Methods
+This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent.
+All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`.
+All associated source code is available at the following GitHub repository.
+Further detailed methods can be found in the supplemental methods.
+
+https://github.com/SchlossLab/Hannigan_CRCVirome_Nature_2017
 
@@
 
-https://github.com/SchlossLab/Hannigan_CRCVirome_Science_2017
+https://github.com/SchlossLab/Hannigan_CRCVirome_Nature_2017
 

### 2017-10-21 figures/pdf2eps.sh
Some files are redone and converted to appropriate format.

@@ -1,0 +1,14 @@
+pdftops -eps sample-processing.pdf
+pdftops -eps predmodel-viromebacteria.pdf
+pdftops -eps phage-bacteria-cor.pdf
+pdftops -eps lysogenic-relabund.pdf
+pdftops -eps WorkingHypothesis.pdf
+pdftops -eps qualitycontrol.pdf
+pdftops -eps qc-contiglengthcoverage.pdf
+pdftops -eps diversity-beta-ogu.pdf
+pdftops -eps diversity-beta-ogu-negative.pdf
+pdftops -eps filtered16S.pdf
+pdftops -eps temporalmodels.pdf
+pdftops -eps predmodel-threewayclassification.pdf
+pdftops -eps NetworkAndScatter.pdf
+convert ./contiglengthbycluster.png contiglengthbycluster.tiff

### 2017-10-21 doc/manuscript.Rmd
Updated document for submission.

@@
 ---
-title: Diagnostics & Interactive Dynamics of the Colorectal Cancer Virome
+title: Diagnostic Potential & the Interactive Dynamics of the Colorectal Cancer Virome
 author:
@@
   - code: 2
-    address: Department of Ecology & Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
+    address: Department of Ecology and Evolutionary Biology, University of Michigan, Ann Arbor, Michigan, 48109
   - code: 3
-    address: Department of Family & Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
+    address: Department of Family and Community Medicine, Pennsylvania State University Hershey Medical Center, Hershey, Pennsylvania, 17033
   - code: \*
@@
     variant: markdown
-abstract: Viruses are assocaited with many human cancers, largely due to their mutagenic and functionally manipulative abilities.
-Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses.
-We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions while the virome role remains unknown.
-We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition.
-Through random forest modeling we identified differences in the healthy and colorectal cancer virome.
-The cancer-associated virome consisted primarily of temperate bacteriophages that were also bacteria-virus community network hubs.
-These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
-font: Times
 fontsize: 12pt
+mainfont: 'Helvetica'
 ---
+
+&nbsp;
+
+| ***Corresponding Author Information***
+| Patrick D Schloss, PhD
+| 1150 W Medical Center Dr. 1526 MSRB I
+| Ann Arbor, Michigan 48109
+| Phone: (734) 647-5801
+| Email: pschloss@umich.edu
+
+| **Article Class**:  Research Article
 
@@
 
+# Abstract
+
+Viruses are associated with many human cancers, largely due to their mutagenic and functionally manipulative abilities.
+Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses.
+We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions while the virome role remains unknown.
+We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition.
+Through random forest modeling we identified differences in the healthy and colorectal cancer virome.
+The cancer-associated virome consisted primarily of temperate bacteriophages that were also predicted to be bacteria-virus community network hubs.
+These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
+
+# Importance
+
+Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide.
+Its risk and severity have been linked to colonic bacterial community composition.
+Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities.
+We addressed this knowledge gap by identifying differences in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community differences.
+The results suggested an indirect role for the virome in impacting colorectal cancer by modulating their associated bacterial community.
+These findings both support a biological role for viruses in colorectal cancer and provide a new understanding of basic colorectal cancer etiology.
+
 \newpage
 
-## Introduction
+# Introduction
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
@@
 *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.
+
+# Results
 
@@
 To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data.
-The combined model yielded a modest but statistically significant performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**.
-The combined features from the virus and bacterial communities improved our ability to classify stool as belonging to individuals with cancerous colons.
+The combined model did not yield a statistically significant performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**.
 
@@
 
-## Model for Virome & Cancer Progression
+# Discussion
 Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
@@
 
-We have begun to delineate the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**.
+Using our findings here, we have begun to delineate the role the colonic virome plays in colorectal cancer in the form of a working hypothesis that will guide our future studies **(Figure \ref{modelsummary} A)**.
 We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states.
@@
 
-## Conclusions
-
 In addition to the diagnostic ramifications for understanding the colorectal cancer microbiome, our findings suggest that viruses, while understudied and currently under-appreciated in the human microbiome, are likely to be an important contributor to human disease.
@@
 
-## Materials and Methods
-This study was approved by the University of Michigan Institutional Review Board and all subjects provided informed consent.
-All study sequences are available on the NCBI Sequence Read Archive under the BioProject ID `PRJNA389927`.
-All associated source code is available at the following GitHub repository.
-Further detailed methods can be found in the supplemental methods.
-
-https://github.com/SchlossLab/Hannigan_CRCVirome_Nature_2017
-
-## Acknowledgments
-The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
-GD Hannigan was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528).
-PD Schloss was supported by funding from the National Institutes of Health (P30DK034933).
-MT Ruffin was supported by funding from the National Institutes of Health (5U01CA86400).
-The authors declare no competing interests.
-
-\newpage
-
-## Figures
-
-![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S rRNA gene sequence signatures, whole metagenomic signatures, and a combination of virome and 16S rRNA gene sequence signatures.
-A) An example ROC curve for visualizing the performance of each of the models for classifying stool as coming from either an individual with a cancerous or healthy colon.
-B) Quantification of the AUC variation for each model, and how it compared to each of the other models based on 15 iterations.
-A pairwise Wilcoxon test with a false discovery rate multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01).
-C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S rRNA gene classification model when removed from the classification model.
-Mean is represented by a point, and bars represent standard error.
-D) Mean decrease in accuracy of each operational virus unit in the virome classification model.
-E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S rRNA gene and virome features.*\label{predmodel}](../figures/predmodel-viromebacteria.pdf){ width=100% }
-
-\newpage
-
-![*Relative abundance correlations between bacterial OTUs and virome OVUs.
-A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OVUs (y-axis) with blue being positively correlated and red being negatively correlated.
-Bar plots indicate the viral (left) and bacterial (bottom) operational unit importance in their colorectal cancer classification models, such that the most important units are in the top left corner.
-B) Magnification of the boxed region in pannel (A), highlighting the correlation between the most important bacterial OTUs and virome OVUs.
-The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as pannel (A).
-C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OVUs.*\label{correlations}](../figures/phage-bacteria-cor.pdf){ width=100% }
-
-\newpage
-
-![*Community network analysis utilizing predicted interactions between bacteria and phage operational genomic units.
-A) Visualization of the community network for our colorectal cancer cohort.
-B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OVU.
-A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}](../figures/NetworkAndScatter.pdf){ width=80% }
-
-\newpage
-
-![*Final model and working hypothesis from this study.
-A) Basic model illustrating the connections between the virome, bacterial communities, and colorectal cancer.
-B) Working hypothesis of how the bacteriophage community is associated with colorectal cancer and the associated bacterial community.*\label{modelsummary}](../figures/WorkingHypothesis.pdf){ width=100% }
-
-\newpage
-
-# Supplemental Materials
-
-# Materials & Methods
+# Materials and Methods
 
@@
 
-https://github.com/SchlossLab/Hannigan_CRCVirome_Nature_2017
+https://github.com/SchlossLab/Hannigan_CRCVirome_mBio_2017
 
@@
 The binary random forest model was trained using the Area Under the receiver operating characteristic Curve (AUC) and the three-class random forest model was trained using the mean AUC.
-Both were validated using five-fold cross validation.
+Both were validated using five-fold nested cross validation to prevent over-fitting on the tuning paramters.
 Each training set was repeated five times, and the model was tuned for mtry values.
@@
 
+# Funding Information
+GD Hannigan was supported in part by the Molecular Mechanisms in Microbial Pathogenesis Training Program (T32 AI007528).
+PD Schloss was supported by funding from the National Institutes of Health (P30DK034933).
+MT Ruffin was supported by funding from the National Institutes of Health (5U01CA86400).
+The authors declare no competing interests.
+
+# Acknowledgments
+The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
+
 \newpage
 
-# Supplemental Figures
-
-\beginsupplement
+# Figure Legends
 
@@
 Virus samples were purified for viruses using filtration and a combination of chloroform (bacterial lysis) and DNase (exposed genomic DNA degradation).
-The resulting encapsulated virus DNA was sequenced using whole metagenomic shotgun sequencing.* \label{sampleproc}](../figures/sample-processing.pdf){ width=100% }
+The resulting encapsulated virus DNA was sequenced using whole metagenomic shotgun sequencing.* \label{sampleproc}]()
 
 \newpage
+
+![*Results from healthy vs cancer classification models built using virome signatures, bacterial 16S rRNA gene sequence signatures, whole metagenomic signatures, and a combination of virome and 16S rRNA gene sequence signatures.
+A) An example ROC curve for visualizing the performance of each of the models for classifying stool as coming from either an individual with a cancerous or healthy colon.
+B) Quantification of the AUC variation for each model, and how it compared to each of the other models based on 15 iterations.
+A pairwise Wilcoxon test with a false discovery rate multiple hypothesis correction demonstrated that all models are significantly different from each other (p-value < 0.01).
+C) Mean decrease in accuracy (measurement of importance) of each operational taxonomic unit within the 16S rRNA gene classification model when removed from the classification model.
+Mean is represented by a point, and bars represent standard error.
+D) Mean decrease in accuracy of each operational virus unit in the virome classification model.
+E) Mean decrease in accuracy of each operational genomic unit and operational taxonomic unit in the model using both 16S rRNA gene and virome features.*\label{predmodel}]()
+
+\newpage
+
+![*Relative abundance correlations between bacterial OTUs and virome OVUs.
+A) Pearson correlation coefficient values between all bacterial OTUs (x-axis) and viral OVUs (y-axis) with blue being positively correlated and red being negatively correlated.
+Bar plots indicate the viral (left) and bacterial (bottom) operational unit importance in their colorectal cancer classification models, such that the most important units are in the top left corner.
+B) Magnification of the boxed region in panel (A), highlighting the correlation between the most important bacterial OTUs and virome OVUs.
+The most important operational units are in the top left corner of the heatmap, and the correlation scale is the same as panel (A).
+C) Histogram quantifying the frequencies of Pearson correlation coefficients between all bacterial OTUs and virome OVUs.*\label{correlations}]()
+
+\newpage
+
+![*Lysogenic phage relative abundance in disease states.
+Phage OVUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
+No disease groups were statistically significant.*\label{replicationstyles}](){ width=50% }
+
+\newpage
+
+![*Final model and working hypothesis from this study.
+A) Basic model illustrating the connections between the virome, bacterial communities, and colorectal cancer.
+B) Working hypothesis of how the bacteriophage community is associated with colorectal cancer and the associated bacterial community.*\label{modelsummary}]()
+
+\newpage
+
+# Supplemental Figure Legends
+
+\beginsupplement
 
@@
 Each bar represents a sample which is grouped and colored by its associated disease group.
-B) Sequence yield following quality control including quality score filtering and human decontamination.*\label{qualcontrol}](../figures/qualitycontrol.pdf){ width=100% }
+B) Sequence yield following quality control including quality score filtering and human decontamination.*\label{qualcontrol}]()
 
@@
 C) Heated scatter plot demonstrating the distribution of contig coverage and length for the whole metagenomic sample set.
-D) Scatter plot illustrating the distribution of operational genomic unit (OGU) length and sequence coverage for the whole metagenomic sample set.*\label{contigqc}](../figures/qc-contiglengthcoverage.pdf){ width=75% }
+D) Scatter plot illustrating the distribution of operational genomic unit (OGU) length and sequence coverage for the whole metagenomic sample set.*\label{contigqc}]()
 
@@
 B) Density plot (analogous to histogram) of the number of virome operational genomic units containing the specific number of contigs, as indicated by the x-axis.
-C-D) Sample plots as panels C and D, but for the whole metagenomic sample set.*\label{clustercontigqc}](../figures/contiglengthbycluster.png){ width=75% }
+C-D) Sample plots as panels C and D, but for the whole metagenomic sample set.*\label{clustercontigqc}]()
 
@@
 Comparisons (indicated on y-axis) in which the intervals cross the zero mean difference line (dashed line) were not significantly different.
-C) Shannon diversity and D) richness alpha diversity quantification comparing pre-cancerous (grey), cancerous (red), and healthy (tan) states.*\label{betaogu}](../figures/diversity-beta-ogu.pdf){ width=100% }
+C) Shannon diversity and D) richness alpha diversity quantification comparing pre-cancerous (grey), cancerous (red), and healthy (tan) states.*\label{betaogu}]()
 
@@
 Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test.
-Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}](../figures/diversity-beta-ogu-negative.pdf){ width=100% }
+Comparisons in which the intervals cross the zero mean difference line (dashed line) were not significantly different.*\label{betaogunegative}]()
 
@@
 B) Relative abundance of the six bacterial OTUs removed when filtered for OTUs with median relative abundance of zero.
-OTU relative abundance is seperated by healthy (red) and cancerous (grey) samples.
-Relative abundance of 1% is marked by the dashed line.*\label{16scompare}](../figures/filtered16S.pdf){ width=100% }
+OTU relative abundance is separated by healthy (red) and cancerous (grey) samples.
+Relative abundance of 1% is marked by the dashed line.*\label{16scompare}]()
+
+\newpage
 
@@
 Disease comparison is specified in the top left corner of each panel.
-F-H) Top ten important bacterial 16S rRNA gene OTUs for classifying each disease state combination.*\label{transitionmodels}](../figures/temporalmodels.pdf){ width=75% }
+F-H) Top ten important bacterial 16S rRNA gene OTUs for classifying each disease state combination.*\label{transitionmodels}]()
 
@@
 F) Relative abundance of the six most important virome OVUs in the model, with the most important on the right.
-Line indicates abundance mean.*\label{threewaymodel}](../figures/predmodel-threewayclassification.pdf){ width=75% }
+Line indicates abundance mean.*\label{threewaymodel}](){}
 
@@
 
-![*Lysogenic phage relative abundance in disease states.
-Phage OVUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
-No disease groups were statistically significant.*\label{replicationstyles}](../figures/lysogenic-relabund.pdf){ width=50% }
+![*Community network analysis utilizing predicted interactions between bacteria and phage operational genomic units.
+A) Visualization of the community network for our colorectal cancer cohort.
+B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OVU.
+A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}]()
 

### 2017-10-21 bin/CompareBacteriaFilters.R
Code on code on code.

@@ -75,14 +75,14 @@ rarefybig <- function(x, subdepth) {
 }
 
 caretmodel <- function(x) {
-  fitControl <- trainControl(method = "repeatedcv",
+fitControl <- trainControl(method = "repeatedcv",
 	repeats = 5,
 	number=5,
 	classProbs = TRUE,
 	summaryFunction = twoClassSummary,
 	savePredictions = TRUE)
-  model <- train(V30~., data=x, trControl=fitControl, method="C5.0", metric="ROC", tuneLength=5)
-  return(model)
+model <- train(V30~., data=x, trControl=fitControl, method="rf", metric="ROC", tuneLength=5)
+return(model)
 }
 
 GetAverageImportance <- function(x, y) {
@@ -203,16 +203,54 @@ pcomb <- function(x, iterationcount = 25, topcount = 10) {
 	  return(importanceplot)
 }
 
+nestedcv <- function(x, iterations = 5, split = 0.75) {
+  outlist <- lapply(1:iterations, function(i) {
+    write(i, stdout())
+
+    trainIndex <- createDataPartition(
+      x$V30,
+      p = split,
+      list = FALSE, 
+      times = 1
+    )
+
+    dftrain <- x[trainIndex,]
+    dftest <- x[-trainIndex,]
+    
+    outmodel <- caretmodel(dftrain)
+    
+    x_test <- dftest[,-length(dftest)]
+    y_test <- dftest[,length(dftest)]
+    
+    outpred <- predict(outmodel, x_test, type="prob")
+    outpred$pred <- predict(outmodel, x_test)
+    outpred$obs <- y_test
+
+    # confusionMatrix(outpred, y_test)
+    # postResample(pred = outpred$pred, obs = outpred$obs)
+    sumout <- twoClassSummary(outpred, lev = levels(outpred$obs))
+    sumroc <- sumout[[1]]
+    sumsens <- sumout[[2]]
+    sumspec <- sumout[[3]]
+    return(list(sumroc, outpred, sumsens, sumspec))
+  })
+  
+  # Get the max and min values
+  rocpositions <- sapply(outlist,`[`,1)
+  maxl <- outlist[[match(max(unlist(rocpositions)), rocpositions)]]
+  medl <- outlist[[match(median(unlist(rocpositions)), rocpositions)]]
[…]

### 2017-10-21 bin/ContigLengthHits.R

@@ -50,7 +50,8 @@ viruscontigplot <- ggplot(mergedabund, aes(x=Length, y=sum)) +
     scale_y_log10() +
     scale_x_log10() +
     xlab("Contig Length (log bp)") +
-    ylab("Sequences Mapping to Contigs (log)")
+    ylab("Sequences Mapping to Contigs (log)") +
+    ggtitle("Virus")
 
 mergedclusters <- merge(mergedabund, clusters, by = "ContigID")
 
@@ -61,9 +62,10 @@ virusclusterplot <- ggplot(clusterstats, aes(x=avglength, y=sumcount)) +
     geom_point() +
     scale_fill_gradientn(colours=c("steelblue4", "steelblue3", "steelblue2","steelblue1","slategray1")) +
     scale_y_log10() +
-    scale_x_log10() +
+    scale_x_log10(limits = 10^c(3,5.5)) +
     xlab("Average Cluster Length (log bp)") +
-    ylab("Sequences Mapping to Clusters (log)")
+    ylab("Sequences Mapping to Clusters (log)") +
+    ggtitle("Virus")
 
 # Bacteria
 
@@ -85,7 +87,8 @@ bacteriacontigplot <- ggplot(mergedabundbac, aes(x=Length, y=sum)) +
     scale_y_log10() +
     scale_x_log10() +
     xlab("Contig Length (log bp)") +
-    ylab("Sequences Mapping to Contigs (log)")
+    ylab("Sequences Mapping to Contigs (log)") +
+    ggtitle("Bacteria")
 
 mergedclustersbac <- merge(mergedabundbac, clustersbac, by = "ContigID")
 
@@ -94,11 +97,11 @@ clustersbactats <- ddply(mergedclustersbac, "Cluster", summarize, sumcount = sum
 bacteriaclusterplot <- ggplot(clustersbactats, aes(x=avglength, y=sumcount)) +
     theme_classic() +
     geom_point() +
-    scale_fill_gradientn(colours=c("steelblue4", "steelblue3", "steelblue2","steelblue1","slategray1")) +
     scale_y_log10() +
-    scale_x_log10() +
+    scale_x_log10(limits = 10^c(3,5.5)) +
     xlab("Average Cluster Length (log bp)") +
-    ylab("Sequences Mapping to Clusters (log)")
+    ylab("Sequences Mapping to Clusters (log)") +
+    ggtitle("Bacteria")
 
[…]

### 2017-10-21 bin/diversity-beta.R

@@ -183,7 +183,7 @@ shannonplot <- ggplot(mergedshannon, aes(x=V30, y=ShannonDiv, fill=V30)) +
     geom_boxplot(notch=FALSE) +
     xlab("Health Status") +
     ylab("Shannon Diversity") +
-    scale_fill_manual(values = alpha(wes_palette("Royal1"), 0.75))
+    scale_fill_manual(values = alpha(wes_palette("Darjeeling"), 0.75))
 pairwise.wilcox.test(x=mergedshannon$ShannonDiv, g=mergedshannon$V30, p.adjust.method="bonferroni")
 
 inputrich <- data.frame(rownames(negcast), rarefy(negcast, sample=100000))
@@ -196,7 +196,7 @@ richplot <- ggplot(mergedrich, aes(x=V30, y=Rich, fill=V30)) +
     geom_boxplot(notch=FALSE) +
     xlab("Health Status") +
     ylab("Richness") +
-    scale_fill_manual(values = alpha(wes_palette("Royal1"), 0.75))
+    scale_fill_manual(values = alpha(wes_palette("Darjeeling"), 0.75))
 pairwise.wilcox.test(x=mergedrich$Rich, g=mergedrich$V30, p.adjust.method="bonferroni")
 
 
@@ -208,8 +208,8 @@ pdf(file="./figures/diversity-beta-ogu.pdf", width=14, height=4)
     gridplotnoneg
 dev.off()
 
-pdf(file="./figures/diversity-beta-ogu-negative.pdf", width=10, height=4)
-    gridplot
+pdf(file="./figures/diversity-beta-ogu-negative.pdf", width=6, height=4)
+    plotdiffs
 dev.off()
 
 fortable <- data.frame(rbind(anovastatwithneg, anovastatwithoutneg))

### 2017-10-21 bin/predictProgression.R

@@ -95,8 +95,8 @@ GetAverageImportance <- function(x, y) {
 	return(resultvardf)
 }
 
-plotimportance <- function(x, iterationcount = 10, topcount = 10) {
-	avgimportance <- lapply(c(1:iterationcount), function(i) GetAverageImportance(x, i))
+plotimportance <- function(x, iterationcount = 10, topcount = 10, corecount = 5) {
+	avgimportance <- mclapply(c(1:iterationcount), mc.cores = corecount, function(i) GetAverageImportance(x, i))
 	avgimportancedf <- ldply(avgimportance, data.frame)
 	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall))
 	importaverage <- merge(avgimportancedf, import, by = "categories")
@@ -199,13 +199,55 @@ pbi <- function(x, iterationcount = 10, topcount = 10) {
 
 GetAverageAUC <- function(x, y) {
 	write(y, stderr())
-	functionmodel <- caretmodel(x)
-	highAUC <- functionmodel$results$ROC[order(functionmodel$results$ROC, decreasing = TRUE)[1]]
-	highSpec <- functionmodel$results$Spec[order(functionmodel$results$Spec, decreasing = TRUE)[1]]
-	highSens <- functionmodel$results$Sens[order(functionmodel$results$Sens, decreasing = TRUE)[1]]
-	resultdf <- data.frame(y,highAUC, highSpec, highSens)
+	functionmodel <- nestedcv(x)
+	avgAUC <- functionmodel[[5]]
+	avgSpec <- functionmodel[[8]]
+	avgSens <- functionmodel[[7]]
+	resultdf <- data.frame(y, avgAUC, avgSpec, avgSens)
 	return(resultdf)
 }
+
+nestedcv <- function(x, iterations = 5, split = 0.75) {
+  outlist <- lapply(1:iterations, function(i) {
+    write(i, stdout())
+
+    trainIndex <- createDataPartition(
+      x$V30,
+      p = split,
+      list = FALSE, 
+      times = 1
+    )
+
+    dftrain <- x[trainIndex,]
+    dftest <- x[-trainIndex,]
+    
+    outmodel <- caretmodel(dftrain)
+    
+    x_test <- dftest[,-length(dftest)]
[…]

### 2017-10-21 bin/predictThreeClass.R

@@ -8,6 +8,7 @@ library("dplyr")
 library("caret")
 library("cowplot")
 library("matrixStats")
+library("parallel")
 
 minavgrelabund <- 0
 minsamps <- 30
@@ -27,7 +28,7 @@ caretmodel <- function(x) {
 }
 
 makemeanroc <- function(x, keep = "Cancer", remove = "Adenoma", one = "Healthy") {
-	plotx <- x$pred
+	plotx <- x
 	plotx[plotx==remove]<-keep
 	plotx$all <- plotx[, keep] + plotx[, remove]
 	plotx$one <- plotx[, one]
@@ -58,8 +59,8 @@ GetAverageImportance <- function(x, y) {
 	return(resultvardf)
 }
 
-plotimportance <- function(x, iterationcount = 25, topcount = 10) {
-	avgimportance <- lapply(c(1:iterationcount), function(i) GetAverageImportance(x, i))
+plotimportance <- function(x, iterationcount = 25, topcount = 10, corecount = 5) {
+	avgimportance <- mclapply(c(1:iterationcount), mc.cores = corecount, function(i) GetAverageImportance(x, i))
 	avgimportancedf <- ldply(avgimportance, data.frame)
 	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall), stder = sd(Overall)/sqrt(length(Overall)))
 	importaverage <- merge(avgimportancedf, import, by = "categories")
@@ -178,10 +179,51 @@ pbi <- function(x, iterationcount = 25, topcount = 10) {
 
 GetAverageAUC <- function(x, y) {
 	write(y, stderr())
-	functionmodel <- caretmodel(x)
-	highAUC <- functionmodel$results$Mean_AUC[order(functionmodel$results$Mean_AUC, decreasing = TRUE)[1]]
-	resultdf <- data.frame(y,highAUC)
+	functionmodel <- nestedcv(x)
+	avgAUC <- functionmodel[[5]]
+	avgSpec <- functionmodel[[8]]
+	avgSens <- functionmodel[[7]]
+	resultdf <- data.frame(y, avgAUC, avgSpec, avgSens)
 	return(resultdf)
+}
+
+nestedcv <- function(x, iterations = 5, split = 0.75) {
+  outlist <- lapply(1:iterations, function(i) {
+    write(i, stdout())
+
+    trainIndex <- createDataPartition(
+      x$V30,
[…]

### 2017-10-21 bin/predictVirome.R

@@ -93,8 +93,8 @@ GetAverageImportance <- function(x, y) {
 	return(resultvardf)
 }
 
-plotimportance <- function(x, iterationcount = 25, topcount = 10) {
-	avgimportance <- lapply(c(1:iterationcount), function(i) GetAverageImportance(x, i))
+plotimportance <- function(x, iterationcount = 25, topcount = 10, corecount = 5) {
+	avgimportance <- mclapply(c(1:iterationcount), mc.cores = corecount, function(i) GetAverageImportance(x, i))
 	avgimportancedf <- ldply(avgimportance, data.frame)
 	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall))
 	importaverage <- merge(avgimportancedf, import, by = "categories")
@@ -161,8 +161,8 @@ plotimportance <- function(x, iterationcount = 25, topcount = 10) {
 	return(importanceplot)
 }
 
-pbi <- function(x, iterationcount = 25, topcount = 10) {
-	avgimportance <- lapply(c(1:iterationcount), function(i) GetAverageImportance(x, i))
+pbi <- function(x, iterationcount = 25, topcount = 10, corecount = 5) {
+	avgimportance <- mclapply(c(1:iterationcount), mc.cores = corecount, function(i) GetAverageImportance(x, i))
 	avgimportancedf <- ldply(avgimportance, data.frame)
 	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall))
 	importaverage <- merge(avgimportancedf, import, by = "categories")
@@ -208,8 +208,8 @@ pbi <- function(x, iterationcount = 25, topcount = 10) {
 	return(importanceplot)
 }
 
-pcomb <- function(x, iterationcount = 25, topcount = 10) {
-	avgimportance <- lapply(c(1:iterationcount), function(i) GetAverageImportance(x, i))
+pcomb <- function(x, iterationcount = 25, topcount = 10, corecount = 5) {
+	avgimportance <- mclapply(c(1:iterationcount), mc.cores = corecount, function(i) GetAverageImportance(x, i))
 	avgimportancedf <- ldply(avgimportance, data.frame)
[…]

### 2017-10-24 doc/manuscript.Rmd

@@
 In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**.
-This is consistent with what has been observed when the same metrics were applied to 16S rRNA sequenced and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.
+This is consistent with what has been observed when the same metrics were applied to 16S rRNA gene sequences and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.
 
 ## Virome Composition in Colorectal Cancer
-As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective feature sets for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
+As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
 The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer.

### 2018-05-14 doc/manuscript.Rmd
Rendered edits and moved the references.
Some doc updates.

@@
 
-Viruses are associated with many human cancers, largely due to their mutagenic and functionally manipulative abilities.
+Human viruses (excluding bacteriophages) have been associated with many human cancers, largely due to their mutagenic and functionally manipulative abilities.
 Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses.
@@
 Its risk and severity have been linked to colonic bacterial community composition.
-Although viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities.
+Although human-specific viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities.
 We addressed this knowledge gap by identifying differences in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community differences.
@@
 # Introduction
+The human gut virome is the community of all viruses found in the gut, including bacteriophages (viruses that only infect bacteria) and eukaryotic viruses (viruses that only infect human and other eukaryotic cells).
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Because bacteriophages (i.e.
-viruses that specifically infect bacteria) are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer.
-The gut virome (i.e.
-the virus community of the gut) therefore has the potential to impact health and disease.
-Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
-The strong association of bacterial communities with colorectal cancer and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.
+Because bacteriophages, another major component of the virome, are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and members of those bacterial communities have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer as well.
+The gut virome therefore has a potential to be associated with, and potentially impact, human cancer.
+Altered human virome composition and diversity have already been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb].
+The strong association of bacterial communities with colorectal cancer, the previous identification of human-specific viruses that cause cancer, and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.
 
@@
 ## Sample Collection and Processing
-Our study cohort consisted of 90 human subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30 of whom had carcinomas **(Figure \ref{sampleproc})**.
+Our study cohort consisted of stool samples collected from 90 human subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30 of whom had carcinomas **(Figure \ref{sampleproc})**.
 Half of each stool sample was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques.
@@
 
-To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using OGU relative abundance profiles from bacterial metagenomic shotgun sequencing data.
-This model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**.
-Because the coverage provided by the metagenomic sequencing was not as deep as the equivalent 16S rRNA gene sequencing, we attempted to compare the approaches at a common sequencing depth.
-This investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**.
+We compared viral metagenomic methods to bacterial metagenomic methods by building a viral model and a model built using OGU relative abundance profiles from bacterial metagenomic shotgun sequencing data.
+This bacterial model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**.
+To determine the cause of the discrepency between the two bacterial sequencing methods, we attempted to compare the approaches at a common sequencing depth.
+This revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**.
 Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**.
@@
 Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.
-Thus, the targeted 16S rRNA gene sequencing approach, which represented only a fraction of the bacterial metagenomic sequencing depth, was more effective for detecting colorectal cancer in stool samples.
-Despite the recent loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
 
@@
 # Discussion
-Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
-Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, a mechanistic link between virus colorectal communities and colorectal cancer has yet to be evaluated.
+Because of their propensity for mutagenesis and capacity for modulating their host functionality, many human viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
+Some bacteria also have oncogenic properties, suggesting that bacteriophages, a component of the human virome in addition to human-specific viruses, may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb].
+Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, a link between virus colorectal communities and colorectal cancer has yet to be evaluated.
 Here we show that, like colonic bacterial communities, the colon virome was altered in patients with colorectal cancer relative to those with healthy colons.
@@
 Briefly, whole evacuated stool was collected from patients who were 18 years of age or older, able to provide informed consent, have had colonoscopy and histologically confirmed colonic disease status, had not had surgery, had not had chemotherapy or radiation, and were free of known co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and inflammatory bowel disease.
+Healthy subjects entered the clinical for the study and did not present as a result of co-morbities.
 Samples were collected from four geographic locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor (Michigan, USA).
@@
 The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
+
+\newpage
+
+# References
+
+<div id = 'refs'></div>
 
@@
 A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}]()
-
-\newpage
-
-# References

### 2018-05-20 bin/predictProgression.R
A bunch of additions for revisions.

@@ -1,8 +1,6 @@
 gcinfo(FALSE)
 
-library("optparse")
 library("ggplot2")
-library("plotROC")
 library("reshape2")
 library("wesanderson")
 library("vegan")
@@ -96,11 +94,14 @@ GetAverageImportance <- function(x, y) {
 }
 
 plotimportance <- function(x, iterationcount = 10, topcount = 10, corecount = 5) {
+	write("Getting Avg Importance", stderr())
 	avgimportance <- mclapply(c(1:iterationcount), mc.cores = corecount, function(i) GetAverageImportance(x, i))
+	write("Formatting Importance", stderr())
 	avgimportancedf <- ldply(avgimportance, data.frame)
 	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall))
 	importaverage <- merge(avgimportancedf, import, by = "categories")
 
+	write("Getting Virus Tax", stderr())
 	virustax <- virustax[,c(1,3,7)]
 
 	importaverage <- merge(importaverage, virustax, by.x = "categories", by.y = "V1", all = TRUE)
@@ -108,8 +109,8 @@ plotimportance <- function(x, iterationcount = 10, topcount = 10, corecount = 5)
 	importaverage <- importaverage[!c(importaverage$Overall %in% NA),]
 	importaverage[is.na(importaverage)] <- "Unknown"
 	importaverage <- importaverage[order(importaverage$mean, decreasing = FALSE),]
-	importaverage$categories <- factor(importaverage$categories, levels = importaverage$categories)
-	importaverage$V7 <- factor(importaverage$V7, levels = importaverage$V7)
+	importaverage$categories <- factor(importaverage$categories, levels = unique(importaverage$categories))
+	importaverage$V7 <- factor(importaverage$V7)
 	
 	binlength <- c(1:topcount) + 0.5
 
@@ -149,6 +150,8 @@ plotimportance <- function(x, iterationcount = 10, topcount = 10, corecount = 5)
 	  return(importanceplot)
 }
 
+hvai <- plotimportance(absmissingid, iterationcount = 3)
+
 pbi <- function(x, iterationcount = 10, topcount = 10) {
[…]

### 2018-05-20 bin/QualityControlStats.R
Some code updates but going to be overwritten.

@@ -18,16 +18,30 @@ option_list <- list(
   make_option(c("-i", "--input"),
     type = "character",
     default = NULL,
-    help = "Input count file.",
-  metavar = "character"),  make_option(c("-m", "--metadata"),
+    help = "Input count file."),
+  make_option(c("-c", "--contamination"),
+    type = "character",
+    default = NULL,
+    help = "Input count file for contaminants."),
+  make_option(c("-m", "--metadata"),
     type = "character",
     default = NULL,
     help = "Metadata table with disease states.",
+    metavar = "character"),
+  make_option(c("-t", "--metadatat"),
+    type = "character",
+    default = NULL,
+    help = "Metadata table with disease states for bacteria.",
     metavar = "character"),
   make_option(c("-o", "--out"),
     type = "character",
     default = NULL,
     help = "Output pdf to draw resulting plot.",
+    metavar = "character"),
+  make_option(c("-a", "--contout"),
+    type = "character",
+    default = NULL,
+    help = "Output pdf to draw resulting contamination plot.",
     metavar = "character"),
   make_option(c("-s", "--sdepth"),
     type = "integer",
@@ -42,44 +56,76 @@ opt <- parse_args(opt_parser);
 ################
 # Run Analysis #
 ################
-input <- read.delim(opt$input, header=FALSE, sep="\t")
+input <- read.delim(opt$input, header=TRUE, sep="\t")
+hc <- read.delim(opt$contamination, header=TRUE, sep="\t")
 metadata <- read.delim(opt$metadata, header=FALSE, sep="\t")[,c(2,26,27,30)]
+metabac <- read.delim(opt$metadatat, header=FALSE, sep="\t")[,c(2,26,27,30)]
+
+metadata$class <- "virus"
+metabac$class <- "bacteria"
+metabac$V2 <- gsub(".cont", "", metabac$V2)
 head(metadata)
+head(metabac)
+metatot <- rbind(metadata, metabac)
 
-# DNA concentration
-dnaconcplot <- ggplot(metadata, aes(x = V2, y = V26, fill = V30)) +
-	theme_classic() +
-	theme(
-		axis.text.y=element_blank(),
-		axis.ticks.y=element_blank(),
[…]

### 2018-05-20 bin/predictProgression.R
Some other stuff that is going to get overwirtten.

@@ -261,6 +261,13 @@ taxonomy$Taxonomy <- sub(".+\\;(.+)\\(\\d+\\)\\;$", "\\1", taxonomy$Taxonomy, pe
 virustax$V1 <- sub("^", "Cluster_", virustax$V1, perl = TRUE)
 virustax <- unique(virustax)
 
+# Remove the bacteria-not-virus clusters
+removaltable <- read.delim("./data/contigclustersidentity/BacteriaNotVirus.tsv", header = FALSE, sep = "\t")
+removaltable$V1 <- gsub("^", "Cluster_", removaltable$V1, perl = TRUE)
+# Clean input
+input <- input[!c(input$V1 %in% removaltable$V1),]
+virustax <- virustax[!c(virustax$V1 %in% removaltable$V1),]
+
 # Rarefy input table
 minimumsubsample <- 1000000
 inputcast <- dcast(input, V1 ~ V2)

### 2018-05-20 doc/manuscript.Rmd

@@
 
-Human viruses (excluding bacteriophages) have been associated with many human cancers, largely due to their mutagenic and functionally manipulative abilities.
+Human viruses (those that infect human cells) have been associated with many cancers, largely due to their mutagenic and functionally manipulative abilities.
 Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses.
-We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions while the virome role remains unknown.
+We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions but an unknown association with the virome.
 We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition.
@@
 # Introduction
-The human gut virome is the community of all viruses found in the gut, including bacteriophages (viruses that only infect bacteria) and eukaryotic viruses (viruses that only infect human and other eukaryotic cells).
+The human gut virome is the community of all viruses found in the gut, including bacteriophages (viruses that only infect bacteria) and eukaryotic viruses (viruses that only infect eukaryotic cells), and human-specific viruses (viruses that only infect human cells).
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
@@
 
-Here we address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those differences might impact cancer progression and severity.
+Here we begin to address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those differences might impact cancer progression and severity.
 We also aimed to evaluate the virome's potential for use as a diagnostic biomarker.
@@
 ## Phage Influence Between CRC Stages
-Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of specific phages could be detected between healthy, adenomatous, and cancerous colons.
+Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of phages could be detected between healthy, adenomatous, and cancerous colons.
 We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions.
@@
 We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky].
-We found that the majority of the phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**.
+We found that the majority of the phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles})**.
 These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].
@@
 ## Community Context of Influential Phages
-Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were primarily predators of the influential bacteria, and thus influenced their relative abundance through predation.
-If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacteria and phages.
+Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the bacteriophage signal may be a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal.
+If this hypothesis was true, we would expect a correlation between the relative abundances of influential bacteria and phages.
 Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**.
 Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**.
-This evidence supported our null hypothesis that the influential phages were not primarily predators of influential bacteria.
+This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria, and suggest a more complex relationship between the bacterial and phage communities.
 
@@
 We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states.
-By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer.
+By implementing a machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer.
 The colorectal cancer virome was composed primarily of bacteriophages.
@@
 
+At this time, our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level.
+There are many different ways in which this system can be operating, and our working hypothesis is one.
+For example, it is possible that the bacterial communities cause a change in the virome instead of the virome altering the bacterial communities.
+Reviewing every possible mechanism and characteristic of the system is beyond the scope of this paper, but the results we present provide a step toward the answer.
+Important future directions will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and even attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.
+
+As with most microbiome studies it is important to note that, despite including data for both bacteria and viruses, we are still only studying a subset of the microbiome.
+Here we are building on the bacterial-centric literature around colorectal cancer by beginning to characterize the virome.
+The caveat here is that we are technically limited to studying only the DNA virome, and therefore are not measuring RNA viruses.
+This will be an important direction for future studies, as well as beginning to characterize other additional microbial communities such as fungi, archaea, etc.
+
 In addition to the diagnostic ramifications for understanding the colorectal cancer microbiome, our findings suggest that viruses, while understudied and currently under-appreciated in the human microbiome, are likely to be an important contributor to human disease.
@@
 Briefly, whole evacuated stool was collected from patients who were 18 years of age or older, able to provide informed consent, have had colonoscopy and histologically confirmed colonic disease status, had not had surgery, had not had chemotherapy or radiation, and were free of known co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and inflammatory bowel disease.
-Healthy subjects entered the clinical for the study and did not present as a result of co-morbities.
+Healthy subjects entered the clinic for the study and did not present as a result of co-morbities.
 Samples were collected from four geographic locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor (Michigan, USA).
@@
 Annotation was interpreted as phage, eukaryotic virus, or unknown.
+As an additional quality control step, these OVUs were also aligned to the bacterial reference genome set from the European Nucleotide Archives using the blastn algorithm (e-value < 1e-25) and OVUs with similarity to bacterial genomes and not viral genomes were removed from analysis.
 
@@
 
+![*Lysogenic phage relative abundance in disease states.
+Phage OVUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
+No disease groups were statistically significant.*\label{replicationstyles}](){ width=50% }
+
+\newpage
+
 ![*Relative abundance correlations between bacterial OTUs and virome OVUs.
@@
 
-![*Lysogenic phage relative abundance in disease states.
-Phage OVUs were predicted to be either lytic or lysogenic, and the relative abundance of lysogenic phages was quantified and represented as a boxplot.
-No disease groups were statistically significant.*\label{replicationstyles}](){ width=50% }
-
-\newpage
-
-![*Final model and working hypothesis from this study.
+![*Final working hypothesis from this study.
+These panels summarize our thoughts on our results and represent interesting future directions that we predict will build on the presented work.
 A) Basic model illustrating the connections between the virome, bacterial communities, and colorectal cancer.

### 2018-05-27 doc/manuscript.Rmd
Final doc edits

@@
 Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses.
-We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions but an unknown association with the virome.
+We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions but with an unknown association with the gut virome.
 We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition.
@@
 The cancer-associated virome consisted primarily of temperate bacteriophages that were also predicted to be bacteria-virus community network hubs.
-These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.
+These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and potentially impact cancer progression by altering the bacterial host communities.
 
@@
 # Introduction
-The human gut virome is the community of all viruses found in the gut, including bacteriophages (viruses that only infect bacteria) and eukaryotic viruses (viruses that only infect eukaryotic cells), and human-specific viruses (viruses that only infect human cells).
+The human gut virome is the community of all viruses found in the gut, including bacteriophages (viruses that only infect bacteria), eukaryotic viruses (viruses that only infect eukaryotic cells), and human-specific viruses (viruses that only infect human cells).
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
-Because bacteriophages, another major component of the virome, are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and members of those bacterial communities have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer as well.
+Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and members of those bacterial communities have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer as well.
 The gut virome therefore has a potential to be associated with, and potentially impact, human cancer.
@@
 *First*, this work supports a biological role for the virome in colorectal cancer development and suggests that more than the bacterial members of the associated microbial communities are involved in the process.
-*Second*, we present a supplementary, or even alternative, virus-based approach for classification modeling of colorectal cancer using stool samples.
+*Second*, we present a supplementary virus-based approach for classification modeling of colorectal cancer using stool samples.
 *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.
@@
 ## Unaltered Diversity in Colorectal Cancer
-Microbiome and disease associations are often described as being of an altered diversity (i.e., "dysbiotic").
+Microbiome and disease associations are often described as being of an altered diversity (i.e.
+"dysbiotic").
 Therefore, we first evaluated the influence of colorectal cancer on virome OVU diversity.
@@
 In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**.
-This is consistent with what has been observed when the same metrics were applied to 16S rRNA gene sequences and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.
+This is consistent with what has been observed when the same metrics were applied to 16S rRNA gene sequences and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these community structures.
 
@@
 As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
-The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer.
+By using classification models instead of attempting to identify single differentially abundant OTUs, these and other studies have been successful in capturing complex community relationships in which differences in taxonomic relative abundance are considered in the context of other taxa.
+The exceptional performance of bacteria in these classification models supports a role for bacterial functionality in colorectal cancer.
 We built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared their performance to models built using bacterial community signatures.
@@
 This bacterial model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**.
-To determine the cause of the discrepency between the two bacterial sequencing methods, we attempted to compare the approaches at a common sequencing depth.
+To determine the cause of the discrepancy between the two bacterial sequencing methods, we attempted to compare the approaches at a common sequencing depth.
 This revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**.
@@
 The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**.
-The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and "unclassified" phage taxa.
-Many of the important viruses were unidentifiable (denoted "unknown").
+The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa who could not be confidently idetnfied beyond their broad phage identification (i.e.
+"unclassified").
+Many of the important viruses, who were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown").
 This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa].
@@
 Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of phages could be detected between healthy, adenomatous, and cancerous colons.
-We evaluated community shifts between the two disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions.
+We evaluated community shifts between the disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions.
 While bacterial OTU models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**.
@@
 ## Phage Dominance in CRC Virome
-Differences in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts.
+Differences in the colorectal cancer virome could have been driven by eukaryotic viruses or by bacteriophages.
 To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages.
@@
 central members) within the bacteria and phage interactive network.
-We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2017cj].
+We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2018gv].
 The predicted interactions were then used to identify phage community hubs.
@@
 
-Using our findings here, we have begun to delineate the role the colonic virome plays in colorectal cancer in the form of a working hypothesis that will guide our future studies **(Figure \ref{modelsummary} A)**.
+Using our findings here, we have begun a path toward the delineation of the role the colonic virome plays in colorectal cancer in the form of a working hypothesis that will guide our future studies **(Figure \ref{modelsummary} A)**.
 We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states.
@@
 Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb].
-This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition.
+This provides a precedent for phages indirectly influencing colorectal cancer progression by altering the bacterial community composition.
 Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**.
@@
 In addition to modeling the potential connections between virus communities, bacterial communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur.
-This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu].
-We hypothesize that the process begins with broadly infectious phages in the colon lysing and thereby disrupting the existing bacterial communities.
+This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu], although it is important to note that there are also many other alternative hypotheses by which the system could be operating.
+We hypothesize that the process begins with broadly infectious phages in the colon lysing, and thereby disrupting, the existing bacterial communities.
 This shift opens novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize.
@@
 For example, it is possible that the bacterial communities cause a change in the virome instead of the virome altering the bacterial communities.
-Reviewing every possible mechanism and characteristic of the system is beyond the scope of this paper, but the results we present provide a step toward the answer.
+Reviewing every possible mechanism and characteristic of the system is beyond the scope of this manuscript, but the results we present provide a step toward the answer.
 Important future directions will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and even attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.
@@
 
-https://github.com/SchlossLab/Hannigan_CRCVirome_mBio_2017
+https://github.com/SchlossLab/Hannigan_CRCVirome_mBio_2018
 
@@
 ## Ecological Network Analysis & Correlations
-The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described [@Hannigan:2017cj].
+The ecological network of the bacterial and phage operational genomic units was constructed and analyzed as previously described [@Hannigan:2018gv].
 Briefly, a random forest model was used to predict interactions between bacterial and phage genomic units, and those interactions were recorded in a graph database using *neo4j* graph databasing software (v2.3.1).
@@
 The authors thank the Schloss lab members for their underlying contributions, and the Great Lakes-New England Early Detection Research Network for providing the fecal samples that were used in this study.
+The authors also thank the participants in the study cohort.
 
@@
 
+![*Community network analysis utilizing predicted interactions between bacteria and phage operational genomic units.
+A) Visualization of the community network for our colorectal cancer cohort.
+B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OVU.
+A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}]()
+
+\newpage
+
 ![*Final working hypothesis from this study.
@@
 Each bar represents a sample which is grouped and colored by its associated disease group.
-B) Sequence yield following quality control including quality score filtering and human decontamination.*\label{qualcontrol}]()
+B) Sequence yield following quality control including quality score filtering and human decontamination.
+Dashed line shows rarefaction line in which all samples with lower sequence yield were excluded from downstream analysis.*\label{qualcontrol}]()
 
@@
 
-![*Beta-diversity comparing disease states and the study negative controls.
+![*Beta-diversity comparing disease states and the study negative controls which remained following rarefaction.
 Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test.
@@
 Line indicates abundance mean.*\label{threewaymodel}](){}
-
-\newpage
-
-![*Community network analysis utilizing predicted interactions between bacteria and phage operational genomic units.
-A) Visualization of the community network for our colorectal cancer cohort.
-B) Scatter plot illustrating the correlation between importance (mean decrease in accuracy) and the degree of centrality for each OVU.
-A linear regression line was fit to illustrate the correlation (blue) which was found to be statistically significantly and weakly correlated (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`).*\label{network}]()

### 2018-05-27 doc/manuscript.tex
Final doc edits

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
    \setmainfont[]{Helvetica}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Diagnostic Potential \& the Interactive Dynamics of the Colorectal Cancer Virome},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\usepackage{setspace}
\doublespacing
\usepackage[vmargin=1in,hmargin=1in]{geometry}
\usepackage{lineno}
\linenumbers
\usepackage{parskip}
\setlength{\parskip}{7.5pt}
\exhyphenpenalty=10000
\hyphenpenalty=10000

\newcommand{\beginsupplement}{%
        \setcounter{table}{0}
        \renewcommand{\thetable}{S\arabic{table}}%
        \setcounter{figure}{0}
        \renewcommand{\thefigure}{S\arabic{figure}}%
     }

  \title{Diagnostic Potential \& the Interactive Dynamics of the Colorectal
Cancer Virome}
    \usepackage{authblk}
    \font\myfont=Helvetica at 12pt
    \font\nextfont=Helvetica at 10pt
                        \author[1]{\myfont Geoffrey D Hannigan}
                    \author[2]{\myfont Melissa B Duhaime}
                    \author[3]{\myfont Mack T Ruffin IV}
                    \author[1]{\myfont Charlie C Koumpouras}
                    \author[1,*]{\myfont Patrick D Schloss}
                            \affil[1]{\nextfont Department of Microbiology \& Immunology, University of Michigan, Ann
Arbor, Michigan, 48109}
                    \affil[2]{\nextfont Department of Ecology and Evolutionary Biology, University of Michigan,
Ann Arbor, Michigan, 48109}
                    \affil[3]{\nextfont Department of Family and Community Medicine, Pennsylvania State
University Hershey Medical Center, Hershey, Pennsylvania, 17033}
                    \affil[*]{\nextfont To whom correspondence may be addressed.}
            \date{}

\begin{document}
\maketitle

~

\textbf{\emph{Corresponding Author Information}}\\
Patrick D Schloss, PhD\\
1150 W Medical Center Dr.~1526 MSRB I\\
Ann Arbor, Michigan 48109\\
Phone: (734) 647-5801\\
Email: pschloss@umich.edu

\textbf{Article Class}: Research Article

\newpage

\section{Abstract}\label{abstract}

Human viruses (those that infect human cells) have been associated with
many cancers, largely due to their mutagenic and functionally
manipulative abilities. Despite this, cancer microbiome studies have
almost exclusively focused on bacteria instead of viruses. We began
evaluating the cancer virome by focusing on colorectal cancer, a primary
cause of morbidity and mortality throughout the world, and a cancer
linked to altered colonic bacterial community compositions but with an
unknown association with the gut virome. We used 16S rRNA gene, whole
shotgun metagenomic, and purified virus metagenomic sequencing of stool
to evaluate the differences in human colorectal cancer virus and
bacterial community composition. Through random forest modeling we
identified differences in the healthy and colorectal cancer virome. The
cancer-associated virome consisted primarily of temperate bacteriophages
that were also predicted to be bacteria-virus community network hubs.
These results provide foundational evidence that bacteriophage
communities are associated with colorectal cancer and potentially impact
cancer progression by altering the bacterial host communities.

\section{Importance}\label{importance}

Colorectal cancer is a leading cause of cancer-related death in the
United States and worldwide. Its risk and severity have been linked to
colonic bacterial community composition. Although human-specific viruses
have been linked to other cancers and diseases, little is known about
colorectal cancer virus communities. We addressed this knowledge gap by
identifying differences in colonic virus communities in the stool of
colorectal cancer patients and how they compared to bacterial community
differences. The results suggested an indirect role for the virome in
impacting colorectal cancer by modulating their associated bacterial
community. These findings both support a biological role for viruses in
colorectal cancer and provide a new understanding of basic colorectal
cancer etiology.

\newpage

\section{Introduction}\label{introduction}

The human gut virome is the community of all viruses found in the gut,
including bacteriophages (viruses that only infect bacteria), eukaryotic
viruses (viruses that only infect eukaryotic cells), and human-specific
viruses (viruses that only infect human cells). Due to their mutagenic
abilities and propensity for functional manipulation, human viruses are
strongly associated with, and in many cases cause, cancer (1--4).
Because bacteriophages are crucial for bacterial community stability and
composition (5--7), and members of those bacterial communities have been
implicated as oncogenic agents (8--11), bacteriophages have the
potential to indirectly impact cancer as well. The gut virome therefore
has a potential to be associated with, and potentially impact, human
cancer. Altered human virome composition and diversity have already been
identified in diseases including periodontal disease (12), HIV (13),
cystic fibrosis (14), antibiotic exposure (15, 16), urinary tract
infections (17), and inflammatory bowel disease (18). The strong
association of bacterial communities with colorectal cancer, the
previous identification of human-specific viruses that cause cancer, and
the precedent for the virome to impact other human diseases suggest that
colorectal cancer may be associated with altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths
in the United States (19). The US National Cancer Institute estimates
over 1.5 million Americans were diagnosed with colorectal cancer in 2016
and over 500,000 Americans died from the disease (19). Growing evidence
suggests that an important component of colorectal cancer etiology may
be perturbations in the colonic bacterial community (8, 10, 11, 20, 21).
Work in this area has led to a proposed disease model in which bacteria
colonize the colon, develop biofilms, promote inflammation, and enter an
oncogenic synergy with the cancerous human cells (22). This association
also has allowed researchers to leverage bacterial community signatures
as biomarkers to provide accurate, noninvasive colorectal cancer
detection from stool (8, 23, 24). While an understanding of colorectal
cancer bacterial communities has proven fruitful both for disease
classification and for identifying the underlying disease etiology,
bacteria are only a subset of the colon microbiome. Viruses are another
important component of the colon microbial community that have yet to be
studied in the context of colorectal cancer. We evaluated disruptions in
virus and bacterial community composition in a human cohort whose stool
was sampled at the three relevant stages of cancer development: healthy,
adenomatous, and cancerous.

Colorectal cancer progresses in a stepwise process that begins when
healthy tissue develops into a precancerous polyp (i.e., adenoma) in the
large intestine (25). If not removed, the adenoma may develop into a
cancerous lesion that can invade and metastasize, leading to severe
illness and death. Progression to cancer can be prevented when
precancerous adenomas are detected and removed during routine screening
(26, 27). Survival for colorectal cancer patients may exceed 90\% when
the lesions are detected early and removed (26). Thus, work that aims to
facilitate early detection and prevention of progression beyond early
cancer stages has great potential to inform therapeutic development.

Here we begin to address the knowledge gap of whether virus community
composition is altered in colorectal cancer and, if it is, how those
differences might impact cancer progression and severity. We also aimed
to evaluate the virome's potential for use as a diagnostic biomarker.
The implications of this study are threefold. \emph{First}, this work
supports a biological role for the virome in colorectal cancer
development and suggests that more than the bacterial members of the
associated microbial communities are involved in the process.
\emph{Second}, we present a supplementary virus-based approach for
classification modeling of colorectal cancer using stool samples.
\emph{Third}, we provide initial support for the importance of studying
the virome as a component of the microbiome ecological network,
especially in cancer.

\section{Results}\label{results}

\subsection{Sample Collection and
Processing}\label{sample-collection-and-processing}

Our study cohort consisted of stool samples collected from 90 human
subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30
of whom had carcinomas \textbf{(Figure \ref{sampleproc})}. Half of each
stool sample was used to sequence the bacterial communities using both
16S rRNA gene and shotgun sequencing techniques. The 16S rRNA gene
sequencing was performed for a previous study, and the sequences were
re-analyzed using contemporary methods (8). The other half of each stool
sample was purified for virus like particles (VLPs) before genomic DNA
extraction and shotgun metagenomic sequencing. In the VLP purification,
cells were disrupted and extracellular DNA degraded \textbf{(Figure
\ref{sampleproc})} to allow the exclusive analysis of viral DNA within
virus capsids. In this manner, the \emph{extracellular virome} of
encapsulated viruses was targeted.

Each extraction was performed with a blank buffer control to detect
contaminants from reagents or other unintentional sources. Only one of
the nine controls contained detectable DNA at a minimal concentration of
0.011 ng/µl, thus providing evidence of the enrichment and purification
of VLP genomic DNA over potential contaminants \textbf{(Figure
\ref{qualcontrol} A)}. As expected, these controls yielded few sequences
and were almost entirely removed while rarefying the datasets to a
common number of sequences \textbf{(Figure \ref{qualcontrol} B)}. The
high quality phage and bacterial sequences were assembled into highly
covered contigs longer than 1 kb \textbf{(Figure \ref{contigqc})}.
Because contigs represent genome fragments, we further clustered related
bacterial contigs into operational genomic units (OGUs) and viral
contigs into operational viral units (OVUs) \textbf{(Figure
\ref{contigqc} - \ref{clustercontigqc})} to approximate organismal
units.

\subsection{Unaltered Diversity in Colorectal
Cancer}\label{unaltered-diversity-in-colorectal-cancer}

Microbiome and disease associations are often described as being of an
altered diversity (i.e. ``dysbiotic''). Therefore, we first evaluated
the influence of colorectal cancer on virome OVU diversity. We evaluated
differences in communities between disease states using the Shannon
diversity, richness, and Bray-Curtis metrics. We observed no significant
alterations in either Shannon diversity or richness in the diseased
states as compared to the healthy state \textbf{(Figure \ref{betaogu}
C-D)}. There was no statistically significant clustering of the disease
groups (ANOSIM p-value = 0.6, \textbf{Figure \ref{betaogu}}). Notably,
there was a significant difference between the few blank controls that
remained after rarefying the data and the other study groups (ANOSIM
p-value \textless{} 0.001, \textbf{Figure \ref{betaogunegative})},
further supporting the quality of the sample set. In summary, standard
alpha and beta diversity metrics were insufficient for capturing virus
community differences between disease states \textbf{(Figure
\ref{betaogu})}. This is consistent with what has been observed when the
same metrics were applied to 16S rRNA gene sequences and metagenomic
samples (8, 23, 24) and points to the need for alternate approaches to
detect the impact of colorectal cancer disease state on these community
structures.

\subsection{Virome Composition in Colorectal
Cancer}\label{virome-composition-in-colorectal-cancer}

As opposed to the diversity metrics discussed above, OTU-based relative
abundance profiles generated from 16S rRNA gene sequences are effective
for classifying stool samples as originating from individuals with
healthy, adenomatous, or cancerous colons (8, 23). By using
classification models instead of attempting to identify single
differentially abundant OTUs, these and other studies have been
successful in capturing complex community relationships in which
differences in taxonomic relative abundance are considered in the
context of other taxa. The exceptional performance of bacteria in these
classification models supports a role for bacterial functionality in
colorectal cancer. We built off of these findings by evaluating the
ability of virus community signatures to classify stool samples and
compared their performance to models built using bacterial community
signatures.

To identify the altered virus communities associated with colorectal
cancer, we built and tested random forest models for classifying stool
samples as belonging to individuals with either cancerous or healthy
colons. We confirmed that our bacterial 16S rRNA gene model replicated
the performance of the original report which used logit models instead
of random forest models \textbf{(Figure \ref{predmodel} A)} (8). We then
compared the bacterial OTU model to a model built using OVU relative
abundances. The viral model performed as well as the bacterial model
(corrected p-value = 0.4), with the viral and bacterial models achieving
mean area under the curve (AUC) values of 0.792 and 0.809, respectively
\textbf{(Figure \ref{predmodel} A - B)}. To evaluate the ability of both
bacterial and viral biomarkers to classify samples, we built a combined
model that used both bacterial and viral community data. The combined
model did not yield a statistically significant performance improvement
beyond the viral (corrected p-value = 0.4) and bacterial (corrected
p-value = 0.08) models, yielding an AUC of 0.768 \textbf{(Figure
\ref{predmodel} A - B)}.

We compared viral metagenomic methods to bacterial metagenomic methods
by building a viral model and a model built using OGU relative abundance
profiles from bacterial metagenomic shotgun sequencing data. This
bacterial model performed worse than the other models (mean AUC = 0.474)
\textbf{(Figure \ref{predmodel} A - B)}. To determine the cause of the
discrepancy between the two bacterial sequencing methods, we attempted
to compare the approaches at a common sequencing depth. This revealed
that the bacterial 16S rRNA gene model was strongly driven by sparse and
low abundance OTUs \textbf{(Figure \ref{16scompare})}. Removal of OTUs
with a median abundance of zero resulted in the removal of six OTUs, and
a loss of model performance down to what was observed in the
metagenome-based model \textbf{(Figure \ref{16scompare} A)}. The
majority of these OTUs had a relative abundance lower than 1\% across
the samples \textbf{(Figure \ref{16scompare} B)}. Although the features
in the viral model also were of low abundance \textbf{(Figure
\ref{threewaymodel} F)}, the coverage was sufficient for high model
performance, likely because viral genomes are orders of magnitude
smaller than bacterial genomes.

The association between the bacterial and viral communities and
colorectal cancer was driven by a few important microbes.
\emph{Fusobacterium} was the primary driver of the bacterial association
with colorectal cancer, which is consistent with its previously
described oncogenic potential \textbf{(Figure \ref{predmodel} C)}(22).
The virome signature also was driven by a few OVUs, suggesting a role
for these viruses in tumorigenesis \textbf{(Figure \ref{predmodel} D)}.
The identified viruses were bacteriophages, belonging to
\emph{Siphoviridae}, \emph{Myoviridae}, and phage taxa who could not be
confidently idetnfied beyond their broad phage identification (i.e.
``unclassified''). Many of the important viruses, who were confirmed to
not have genomic similarity to known bacterial genomes, were
unidentifiable (denoted ``unknown''). This is common in viromes across
habitats; studies have reported as much as 95\% of virus sequences
belonging to unknown genomic units (14, 28--30). When the bacterial and
viral community signatures were combined, both bacterial and viral
organisms drove the community association with cancer \textbf{(Figure
\ref{predmodel} E)}.

\subsection{Phage Influence Between CRC
Stages}\label{phage-influence-between-crc-stages}

Because previous work has identified shifts in which bacteria were most
important at different stages of colorectal cancer (8, 20, 22), we
explored whether shifts in the relative influence of phages could be
detected between healthy, adenomatous, and cancerous colons. We
evaluated community shifts between the disease stage transitions
(healthy to adenomatous and adenomatous to cancerous) by building random
forest models to compare only the diagnosis groups around the
transitions. While bacterial OTU models performed equally well for all
disease class comparisons, the virome model performances differed
\textbf{(Figure \ref{transitionmodels} A-B)}. Like bacteria
\textbf{(Figure \ref{transitionmodels} F-H)}, different virome members
were important between the healthy to adenomatous and adenomatous to
cancerous stages \textbf{(Figure \ref{transitionmodels} C-E)}.

After evaluating our ability to classify samples between two disease
states, we performed a three-class random forest model including all
disease states. The 16S rRNA gene model yielded a mean AUC of 0.765 and
outperformed the viral community model, which yielded a mean AUC of
0.658 (p-value \textless{} 0.001, \textbf{Figure \ref{threewaymodel}
A-C}). The microbes important for the healthy versus cancer and healthy
versus adenoma models were also important for the three-class model
\textbf{(Figure \ref{threewaymodel} D-E)}. The most important bacterium
in the two and three class models was the same \emph{Fusobacterium} (OTU
4) \textbf{(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)}.
The viruses most important to the three-class model were identified as
bacteriophages \textbf{(Figure \ref{predmodel} D, Figure
\ref{threewaymodel} E)}, but not all important OVUs were of increased
abundance in the diseased state \textbf{(Figure \ref{threewaymodel} F)}.

\subsection{Phage Dominance in CRC
Virome}\label{phage-dominance-in-crc-virome}

Differences in the colorectal cancer virome could have been driven by
eukaryotic viruses or by bacteriophages. To better understand the types
of viruses that were important for colorectal cancer, we identified the
virome OVUs as being similar to either eukaryotic viruses or
bacteriophages. The most important viruses to the classification model
were identified as bacteriophages (\textbf{Figure \ref{threewaymodel})}.
Overall, we were able to identify 78.8\% of the OVUs as known viruses,
and 93.8\% of those viral OVUs aligned to bacteriophage reference
genomes. It is important to note that this could have been influenced by
our methodological biases against enveloped viruses (more common of
eukaryotic viruses than bacteriophage), due to chloroform and DNase
treatment for purification.

We evaluated whether the phages in the community were primarily lytic
(i.e.~obligately lyse their hosts after replication) or temperate
(i.e.~able to integrate into their host's genome to form a lysogen, and
subsequently transition to a lytic mode). We accomplished this by
identifying three markers for temperate phages in the OVU representative
sequences: 1) presence of phage integrase genes, 2) presence of known
prophage genes, according the the ACLAME (A CLAssification of Mobile
genetic Elements) database, and 3) nucleotide similarity to regions of
bacterial genomes (29, 31, 32). We found that the majority of the phages
were temperate and that the overall fraction of temperate phages
remained consistent throughout the healthy, adenomatous, and cancerous
stages \textbf{(Figure \ref{replicationstyles})}. These findings were
consistent with previous reports suggesting the gut virome is primarily
composed of temperate phages (13, 18, 31, 33).

\subsection{Community Context of Influential
Phages}\label{community-context-of-influential-phages}

Because the link between colorectal cancer and the virome was driven by
bacteriophages, we hypothesized that the bacteriophage signal may be a
mere reflection of the bacterial signal, and thus highly correlated with
the bacterial signal. If this hypothesis was true, we would expect a
correlation between the relative abundances of influential bacteria and
phages. Instead, we observed a strikingly low correlation between
bacterial and phage relative abundances \textbf{(Figure
\ref{correlations} A,C)}. Overall, there was an absence of correlation
between the most influential OVUs and bacterial OTUs \textbf{(Figure
\ref{correlations} B)}. This evidence supported our null hypothesis that
the influential phages were not primarily reflections of influential
bacteria, and suggest a more complex relationship between the bacterial
and phage communities.

Given these findings, we hypothesized that the most influential phages
were acting by infecting a wide range of bacteria in the overall
community, instead of just the influential bacteria. In other words, we
hypothesized that the influential bacteriophages were community hubs
(i.e.~central members) within the bacteria and phage interactive
network. We investigated the potential host ranges of all phage OVUs
using a previously developed random forest model that relies on sequence
features to predict which phages infected which bacteria in the
community \textbf{(Figure \ref{network} A)} (34). The predicted
interactions were then used to identify phage community hubs. We
calculated the alpha centrality (i.e.~measure of importance in the
ecological network) of each phage OVU's connection to the rest of the
network. The phages with high centrality values were defined as
community hubs. Next, the centrality of each OVU was compared to its
importance in the colorectal cancer classification model. Phage OVU
centrality was significantly and positively correlated with importance
to the disease model (p-value = 0.004, R = 0.176), suggesting that
phages important in driving colorectal cancer also were more likely to
be community hubs \textbf{(Figure \ref{network} B)}. Together these
findings supported our hypothesis that influential phages were hubs
within their microbial communities and had broad host ranges.

\section{Discussion}\label{discussion}

Because of their propensity for mutagenesis and capacity for modulating
their host functionality, many human viruses are oncogenic (1--4). Some
bacteria also have oncogenic properties, suggesting that bacteriophages,
a component of the human virome in addition to human-specific viruses,
may play an indirect role in promoting carcinogenesis by influencing
bacterial community composition and dynamics (8--10). Despite their
carcinogenic potential and the strong association between bacteria and
colorectal cancer, a link between virus colorectal communities and
colorectal cancer has yet to be evaluated. Here we show that, like
colonic bacterial communities, the colon virome was altered in patients
with colorectal cancer relative to those with healthy colons. Our
findings support a working hypothesis for oncogenesis by phage-modulated
bacterial community composition.

Using our findings here, we have begun a path toward the delineation of
the role the colonic virome plays in colorectal cancer in the form of a
working hypothesis that will guide our future studies \textbf{(Figure
\ref{modelsummary} A)}. We found that basic diversity metrics of alpha
diversity (richness and Shannon diversity) and beta diversity
(Bray-Curtis dissimilarity) were insufficient for identifying virome
community differences between healthy and cancerous states. By
implementing a machine learning approach (random forest classification),
we detected strong associations between the colon virus community
composition and colorectal cancer. The colorectal cancer virome was
composed primarily of bacteriophages. These phage communities were not
exclusively predators of the most influential bacteria, as demonstrated
by the lack of correlation between the abundances of the bacterial and
phage populations. Instead, we identified influential phages as being
community hubs, suggesting phages influence cancer by altering the
greater bacterial community instead of directly modulating the
influential bacteria. Our previous work has shown that modifying colon
bacterial communities alters colorectal cancer progression and tumor
burden in mice (10, 20). This provides a precedent for phages indirectly
influencing colorectal cancer progression by altering the bacterial
community composition. Overall, our data support a model in which the
bacteriophage community modulates the bacterial community, and through
those interactions indirectly influences the bacteria driving colorectal
cancer progression \textbf{(Figure \ref{modelsummary} A)}. Although our
evidence suggested phages indirectly influenced colorectal cancer
development, we were not able to rule out the role of phages directly
interacting with the human host (35, 36).

In addition to modeling the potential connections between virus
communities, bacterial communities, and colorectal cancer, we also used
our data and existing knowledge of phage biology to develop a working
hypothesis for the mechanisms by which this may occur. This was done by
incorporating our findings into the current model for colorectal cancer
development \textbf{(Figure \ref{modelsummary} B)} (22), although it is
important to note that there are also many other alternative hypotheses
by which the system could be operating. We hypothesize that the process
begins with broadly infectious phages in the colon lysing, and thereby
disrupting, the existing bacterial communities. This shift opens novel
niche space that enabled opportunistic bacteria (such as
\emph{Fusobacterium nucleatum}) to colonize. Once the initial
influential founder bacteria establish themselves in the epithelium,
secondary opportunistic bacteria are able to adhere to the founders,
colonize, and establish a biofilm. Phages may play a role in biofilm
dispersal and growth by lysing bacteria within the biofilm, a process
important for effective biofilm growth (37). The oncogenic bacteria may
then be able to transform the epithelial cells and disrupt tight
junctions to infiltrate the epithelium, thereby initiating an
inflammatory immune response. As the adenomatous polyps developed and
progressed towards carcinogenesis, we observed a shift in the phages and
bacteria whose relative abundances were most influential. As the
bacteria enter their oncogenic synergy with the epithelium, we
conjecture that the phages continue mediating biofilm dispersal. This
process would thereby support the colonized oncogenic bacteria by lysing
competing cells and releasing nutrients to other bacteria in the form of
cellular lysates. In addition to highlighting the likely mechanisms by
which the colorectal cancer virome is interacting with the bacterial
communities this model will guide future research investigations of the
role the virome plays colorectal cancer.

At this time, our working hypothesis represents a conceptualization of
areas for future work, which will be required to characterize the
colorectal cancer microbiome at the functional, mechanistic level. There
are many different ways in which this system can be operating, and our
working hypothesis is one. For example, it is possible that the
bacterial communities cause a change in the virome instead of the virome
altering the bacterial communities. Reviewing every possible mechanism
and characteristic of the system is beyond the scope of this manuscript,
but the results we present provide a step toward the answer. Important
future directions will include larger cohort human studies, further
\emph{in vitro} and \emph{in vivo} mechanistic experimentation, and even
attempts at community studies using absolute abundance values instead of
relative abundance, which would allow for more accurate community
dynamic modeling.

As with most microbiome studies it is important to note that, despite
including data for both bacteria and viruses, we are still only studying
a subset of the microbiome. Here we are building on the
bacterial-centric literature around colorectal cancer by beginning to
characterize the virome. The caveat here is that we are technically
limited to studying only the DNA virome, and therefore are not measuring
RNA viruses. This will be an important direction for future studies, as
well as beginning to characterize other additional microbial communities
such as fungi, archaea, etc.

In addition to the diagnostic ramifications for understanding the
colorectal cancer microbiome, our findings suggest that viruses, while
understudied and currently under-appreciated in the human microbiome,
are likely to be an important contributor to human disease. Viral
community dynamics have the potential to provide an abundance of
information to supplement those of bacterial communities. Evidence has
suggested that the virome is a crucial component to the microbiome and
that bacteriophages are important players. Bacteriophage and bacterial
communities cannot maintain stability and co-evolution without one
another (6, 38). Not only is the human virome an important element to
consider in human health and disease (12--18), but our findings support
that it is likely to have a significant impact on cancer etiology and
progression.

\section{Materials and Methods}\label{materials-and-methods}

\subsection{Analysis Source Code \& Data
Availability}\label{analysis-source-code-data-availability}

All study sequences are available on the NCBI Sequence Read Archive
under the BioProject ID \texttt{PRJNA389927}.

All associated source code is available at the following GitHub
repository:

https://github.com/SchlossLab/Hannigan\_CRCVirome\_mBio\_2018

\subsection{Study Design and Patient
Sampling}\label{study-design-and-patient-sampling}

This study was approved by the University of Michigan Institutional
Review Board and all subjects provided informed consent. Design and
sampling of this sample set have been reported previously (8). Briefly,
whole evacuated stool was collected from patients who were 18 years of
age or older, able to provide informed consent, have had colonoscopy and
histologically confirmed colonic disease status, had not had surgery,
had not had chemotherapy or radiation, and were free of known
co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and
inflammatory bowel disease. Healthy subjects entered the clinic for the
study and did not present as a result of co-morbities. Samples were
collected from four geographic locations: Toronto (Ontario, Canada),
Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor
(Michigan, USA). Ninety patients were recruited to the study, thirty of
which were designated healthy, thirty with detected adenomas, and thirty
with detected carcinomas.

\subsection{16S rRNA Gene Sequence Data Acquisition \&
Processing}\label{s-rrna-gene-sequence-data-acquisition-processing}

The 16S rRNA gene sequences associated with this study were previously
reported (8). Sequence (fastq) and metadata files were downloaded from:

http://www.mothur.org/MicrobiomeBiomarkerCRC

The 16S rRNA gene sequences were analyzed as described previously,
relying on the mothur software package (v1.37.0) (39, 40). Briefly, the
sequences were de-replicated, aligned to the SILVA database (41),
screened for chimeras using UCHIME (42), and binned into operational
taxonomic units (OTUs) using a 97\% similarity threshold. Abundances
were normalized for uneven sequencing depth by randomly sub-sampling to
10,000 sequences, as previously reported (23).

\subsection{Whole Metagenomic Library Preparation \&
Sequencing}\label{whole-metagenomic-library-preparation-sequencing}

DNA was extracted from stool samples using the PowerSoil-htp 96 Well
Soil DNA Isolation Kit (Mo Bio Laboratories) using an EPMotion 5075
pipetting system. Purified DNA was used to prepare a shotgun sequencing
library using the Illumina Nextera XT library preparation kit according
to the standard kit protocol, including 12 cycles of limited cycle PCR.
The tagmentation time was increased from five minutes to ten minutes to
improve DNA fragment length distribution. The library was sequenced
using one lane of the Illumina HiSeq4000 platform and yielded 125 bp
paired end reads.

\subsection{Virus Metagenomic Library Preparation \&
Sequencing}\label{virus-metagenomic-library-preparation-sequencing}

Genomic DNA was extracted from purified virus-like particles (VLPs) from
stool samples, using a modified version of a previously published
protocol (29, 31, 43, 44). Briefly, an aliquot of stool
(\textasciitilde{}0.1 g) was resuspended in SM buffer (Crystalgen;
Catalog \#: 221-179) and vortexed to facilitate resuspension. The
resuspended stool was centrifuged to remove major particulate debris
then filtered through a 0.22-µm filter to remove smaller contaminants.
The filtered supernatant was treated with chloroform for ten minutes
with gentle shaking, so as to lyse contaminating cells including
bacteria, human, fungi, etc. The exposed genomic DNA from the lysed
cells was degraded by treating the samples with 5U of DNase for one hour
at 37C. DNase was deactivated by incubating the sample at 75C for ten
minutes. The DNA was extracted from the purified virus-like particles
(VLPs) using the Wizard PCR Purification Preparation Kit (Promega).
Disease classes were staggered across purification runs to prevent run
variation as a confounding factor. As for whole community metagenomes,
purified DNA was used to prepare a shotgun sequencing library using the
Illumina Nextera XT preparation kit according to the standard kit
protocol. The tagmentation time was increased from five minutes to ten
minutes to improve DNA fragment length distribution. The PCR cycle
number was increased from twelve to eighteen cycles to address the low
biomass of the samples, as has been described previously (29). The
library was sequenced using one lane of the Illumina HiSeq4000 platform
and yielded 125 bp paired end reads.

\subsection{Metagenome Quality
Control}\label{metagenome-quality-control}

Both the viral and whole community metagenomic sample sets were
subjected to the same quality control procedures. The sequences were
obtained as de-multiplexed fastq files and subjected to 5' and 3'
adapter trimming using the CutAdapt program (v1.9.1) with an error rate
of 0.1 and an overlap of 10 (45). The FastX toolkit (v0.0.14) was used
to quality trim the reads to a minimum length of 75 bp and a minimum
quality score of 30 (46). Reads mapping to the human genome were removed
using the DeconSeq algorithm (v0.4.3) and default parameters (47).

\subsection{Contig Assembly \&
Abundance}\label{contig-assembly-abundance}

Contigs were assembled using paired end read files that were purged of
sequences without a corresponding pair (e.g.~one read removed due to low
quality). The Megahit program (v1.0.6) was used to assemble contigs for
each sample using a minimum contig length of 1000 bp and iterating
assemblies from 21-mers to 101-mers by 20 (48). Contigs from the virus
and whole metagenomic sample sets were concatenated within their
respective groups. Abundance of the contigs within each sample was
calculated by aligning sequences back to the concatenated contig files
using the bowtie2 global aligner (v2.2.1), with a 25 bp seed length and
an allowance of one mismatch (49). Abundance was corrected for contig
reference length and the number of contigs included in each operational
genomic unit. Abundance was also corrected for uneven sampling depth by
randomly sub-sampling virome and whole metagenomes to 1,000,000 and
500,000 reads, respectively, and by removing samples with fewer total
reads than the threshold. Thresholds were set for maximizing sequence
information while minimizing numbers of lost samples.

\subsection{Operational Genomic Unit
Classification}\label{operational-genomic-unit-classification}

Much like operational taxonomic units (OTUs) are used as an operational
definition of similar 16S rRNA gene sequences, we defined closely
related bacterial contig sequences as operational genomic units (OGUs)
and virus contigs as operational viral units (OVUs) in the absence of
taxonomic identity. OGUs and OVUs were defined with the CONCOCT
algorithm (v0.4.0) which bins related contigs by similar tetra-mer and
co-abundance profiles within samples using a variational Bayesian
approach (50). CONCOCT was used with a length threshold of 1000 bp for
virus contigs and 2000 bp for bacteria.

\subsection{Diversity}\label{diversity}

Alpha and beta diversity were calculated using the operational viral
unit abundance profiles for each sample. Sequences were rarefied to
100,000 sequences. Samples with less than the cutoff were removed from
the analysis. Alpha diversity was calculated using the Shannon diversity
and richness metrics. Beta diversity was calculated using the
Bray-Curtis metric (mean of 25 random sub-sampling iterations), and the
statistical significance between the disease state clusters was assessed
using an analysis of similarity (ANOSIM) with a post-hoc multivariate
Tukey test. All diversity calculations were performed in R using the
Vegan package (51).

\subsection{Classification Modeling}\label{classification-modeling}

Classification modeling was performed in R using the Caret package (52).
OTU, OVU, and OGU abundance data was preprocessed by removing features
(OTUs, OVUs, and OGUs) that were present in less than thirty of the
samples. This served both as an effective feature reduction technique
and made the calculations computationally feasible. The binary random
forest model was trained using the Area Under the receiver operating
characteristic Curve (AUC) and the three-class random forest model was
trained using the mean AUC. Both were validated using five-fold nested
cross validation to prevent over-fitting on the tuning paramters. Each
training set was repeated five times, and the model was tuned for mtry
values. For consistency and accurate comparison between feature groups
(e.g., bacteria, viruses), the sample model parameters were used for
each group. The maximum AUC during training was recorded across twenty
iterations of each group model to test the significance of the
differences between feature set performance. Statistical significance
was evaluated using a Wilcoxon test between two categories, or a
pairwise Wilcoxon test with Bonferroni corrected p-values when comparing
more than two categories.

\subsection{Taxonomic Identification of Operational Genomic
Units}\label{taxonomic-identification-of-operational-genomic-units}

Operational viral units (OVUs) were taxonomically identified using a
reference database consisting of all bacteriophage and eukaryotic virus
genomes present in the European Nucleotide Archives. The longest
contiguous sequence in each operational genomic unit was used as a
representative sequence for classification, as described previously
(53). Each representative sequence was aligned to the reference genome
database using the tblastx alignment algorithm (v2.2.27) and a strict
similarity threshold (e-value \textless{} 1e-25) (54). Annotation was
interpreted as phage, eukaryotic virus, or unknown. As an additional
quality control step, these OVUs were also aligned to the bacterial
reference genome set from the European Nucleotide Archives using the
blastn algorithm (e-value \textless{} 1e-25) and OVUs with similarity to
bacterial genomes and not viral genomes were removed from analysis.

\subsection{Ecological Network Analysis \&
Correlations}\label{ecological-network-analysis-correlations}

The ecological network of the bacterial and phage operational genomic
units was constructed and analyzed as previously described (34).
Briefly, a random forest model was used to predict interactions between
bacterial and phage genomic units, and those interactions were recorded
in a graph database using \emph{neo4j} graph databasing software
(v2.3.1). The degree of phage centrality was quantified using the alpha
centrality metric in the igraph CRAN package. A Spearman correlation was
performed between model importance and phage centrality scores.

\subsection{Phage Replication Style
Identification}\label{phage-replication-style-identification}

Phage OVU replication mode was predicted using methods described
previously (29, 31, 32). Briefly, we identified temperate OVUs as
representative contigs containing at least one of three genomic markers:
1) phage integrase genes, 2) prophage genes from the ACLAME database, or
3) genomic similarity to bacterial reference genomes. Integrase genes
were identified in phage OVU representative contigs by aligning the
contigs to a reference database of all known phage integrase genes from
the Uniprot database (Uniprot search term: ``organism:phage gene:int NOT
putative''). Prophage genes were identified in the same way, using the
ACLAME set of reference prophage genes. In both cases, the blastx
algorithm was used with an e-value threshold of 10e-5. Representative
contigs were also identified as potential lysogenic phages by having a
high genomic similarity to bacterial genomes. To accomplish this,
representative phage contigs were aligned to the European Nucleotide
Archive bacterial genome reference set using the blastn algorithm
(e-value \textless{} 10e-25).

\section{Funding Information}\label{funding-information}

GD Hannigan was supported in part by the Molecular Mechanisms in
Microbial Pathogenesis Training Program (T32 AI007528). PD Schloss was
supported by funding from the National Institutes of Health
(P30DK034933). MT Ruffin was supported by funding from the National
Institutes of Health (5U01CA86400). The authors declare no competing
interests.

\section{Acknowledgments}\label{acknowledgments}

The authors thank the Schloss lab members for their underlying
contributions, and the Great Lakes-New England Early Detection Research
Network for providing the fecal samples that were used in this study.
The authors also thank the participants in the study cohort.

\newpage

\section{References}\label{references}

\hypertarget{refs}{}
\hypertarget{ref-Feng:2008kr}{}
1. Feng H, Shuda M, Chang Y, Moore PS. 2008. Clonal integration of a
polyomavirus in human Merkel cell carcinoma. Science 319:1096--1100.

\hypertarget{ref-Shuda:2011gf}{}
2. Shuda M, Kwun HJ, Feng H, Chang Y, Moore PS. 2011. Human Merkel cell
polyomavirus small T antigen is an oncoprotein targeting the 4E-BP1
translation regulator. Journal of Clinical Investigation 121:3623--3634.

\hypertarget{ref-Schiller:2012ba}{}
3. Schiller JT, Castellsagué X, Garland SM. 2012. A review of clinical
trials of human papillomavirus prophylactic vaccines. Vaccine 30 Suppl
5:F123--38.

\hypertarget{ref-Chang:1994up}{}
4. Chang Y, Cesarman E, Pessin MS, Lee F, Culpepper J, Knowles DM, Moore
PS. 1994. Identification of herpesvirus-like DNA sequences in
AIDS-associated Kaposi's sarcoma. Science 266:1865--1869.

\hypertarget{ref-Harcombe:2005fd}{}
5. Harcombe WR, Bull JJ. 2005. Impact of phages on two-species bacterial
communities. Applied and Environmental Microbiology 71:5254--5259.

\hypertarget{ref-RodriguezValera:2009cr}{}
6. Rodriguez-Valera F, Martin-Cuadrado A-B, Rodriguez-Brito B, Pašić L,
Thingstad TF, Rohwer F, Mira A. 2009. Explaining microbial population
genomics through phage predation. Nature Reviews Microbiology
7:828--836.

\hypertarget{ref-Cortez:2014bk}{}
7. Cortez MH, Weitz JS. 2014. Coevolution can reverse predator-prey
cycles. Proceedings of the National Academy of Sciences of the United
States of America 111:7486--7491.

\hypertarget{ref-Zackular:2014fba}{}
8. Zackular JP, Rogers MAM, Ruffin MT, Schloss PD. 2014. The human gut
microbiome as a screening tool for colorectal cancer. Cancer prevention
research (Philadelphia, Pa) 7:1112--1121.

\hypertarget{ref-Garrett:2015fg}{}
9. Garrett WS. 2015. Cancer and the microbiota. Science 348:80--86.

\hypertarget{ref-Baxter:2014hb}{}
10. Baxter NT, Zackular JP, Chen GY, Schloss PD. 2014. Structure of the
gut microbiome following colonization with human feces determines
colonic tumor burden. Microbiome 2:20.

\hypertarget{ref-Arthur:2012kl}{}
11. Arthur JC, Perez-Chanona E, Mühlbauer M, Tomkovich S, Uronis JM, Fan
T-J, Campbell BJ, Abujamel T, Dogan B, Rogers AB, Rhodes JM, Stintzi A,
Simpson KW, Hansen JJ, Keku TO, Fodor AA, Jobin C. 2012. Intestinal
inflammation targets cancer-inducing activity of the microbiota. Science
338:120--123.

\hypertarget{ref-Ly:2014ew}{}
12. Ly M, Abeles SR, Boehm TK, Robles-Sikisaka R, Naidu M,
Santiago-Rodriguez T, Pride DT. 2014. Altered Oral Viral Ecology in
Association with Periodontal Disease. mBio 5:e01133--14--e01133--14.

\hypertarget{ref-Monaco:2016ita}{}
13. Monaco CL, Gootenberg DB, Zhao G, Handley SA, Ghebremichael MS, Lim
ES, Lankowski A, Baldridge MT, Wilen CB, Flagg M, Norman JM, Keller BC,
Luévano JM, Wang D, Boum Y, Martin JN, Hunt PW, Bangsberg DR, Siedner
MJ, Kwon DS, Virgin HW. 2016. Altered Virome and Bacterial Microbiome in
Human Immunodeficiency Virus-Associated Acquired Immunodeficiency
Syndrome. Cell Host and Microbe 19:311--322.

\hypertarget{ref-Willner:2009dq}{}
14. Willner D, Furlan M, Haynes M, Schmieder R, Angly FE, Silva J,
Tammadoni S, Nosrat B, Conrad D, Rohwer F. 2009. Metagenomic analysis of
respiratory tract DNA viral communities in cystic fibrosis and
non-cystic fibrosis individuals. PLOS ONE 4:e7370.

\hypertarget{ref-Abeles:2015dy}{}
15. Abeles SR, Ly M, Santiago-Rodriguez TM, Pride DT. 2015. Effects of
Long Term Antibiotic Therapy on Human Oral and Fecal Viromes. PLOS ONE
10:e0134941.

\hypertarget{ref-Modi:2013fi}{}
16. Modi SR, Lee HH, Spina CS, Collins JJ. 2013. Antibiotic treatment
expands the resistance reservoir and ecological network of the phage
metagenome. Nature 499:219--222.

\hypertarget{ref-SantiagoRodriguez:2015gd}{}
17. Santiago-Rodriguez TM, Ly M, Bonilla N, Pride DT. 2015. The human
urine virome in association with urinary tract infections. Frontiers in
Microbiology 6:14.

\hypertarget{ref-Norman:2015kb}{}
18. Norman JM, Handley SA, Baldridge MT, Droit L, Liu CY, Keller BC,
Kambal A, Monaco CL, Zhao G, Fleshner P, Stappenbeck TS, McGovern DPB,
Keshavarzian A, Mutlu EA, Sauk J, Gevers D, Xavier RJ, Wang D, Parkes M,
Virgin HW. 2015. Disease-specific alterations in the enteric virome in
inflammatory bowel disease. Cell 160:447--460.

\hypertarget{ref-Siegel:2014jo}{}
19. Siegel R, Desantis C, Jemal A. 2014. Colorectal cancer statistics,
2014. CA: a cancer journal for clinicians 64:104--117.

\hypertarget{ref-Zackular:2016en}{}
20. Zackular JP, Baxter NT, Chen GY, Schloss PD. 2016. Manipulation of
the Gut Microbiota Reveals Role in Colon Tumorigenesis. mSphere
1:e00001--15.

\hypertarget{ref-Dejea:2014fz}{}
21. Dejea CM, Wick EC, Hechenbleikner EM, White JR, Mark Welch JL,
Rossetti BJ, Peterson SN, Snesrud EC, Borisy GG, Lazarev M, Stein E,
Vadivelu J, Roslani AC, Malik AA, Wanyiri JW, Goh KL, Thevambiga I, Fu
K, Wan F, Llosa N, Housseau F, Romans K, Wu X, McAllister FM, Wu S,
Vogelstein B, Kinzler KW, Pardoll DM, Sears CL. 2014. Microbiota
organization is a distinct feature of proximal colorectal cancers.
Proceedings of the National Academy of Sciences of the United States of
America 111:18321--18326.

\hypertarget{ref-Flynn:2016iu}{}
22. Flynn KJ, Baxter NT, Schloss PD. 2016. Metabolic and Community
Synergy of Oral Bacteria in Colorectal Cancer. mSphere 1:e00102--16.

\hypertarget{ref-Baxter:2016dja}{}
23. Baxter NT, Ruffin MT, Rogers MAM, Schloss PD. 2016. Microbiota-based
model improves the sensitivity of fecal immunochemical test for
detecting colonic lesions. Genome medicine 8:37.

\hypertarget{ref-Zeller:2014ix}{}
24. Zeller G, Tap J, Voigt AY, Sunagawa S, Kultima JR, Costea PI, Amiot
A, Böhm J, Brunetti F, Habermann N, Hercog R, Koch M, Luciani A, Mende
DR, Schneider MA, Schrotz-King P, Tournigand C, Tran Van Nhieu J, Yamada
T, Zimmermann J, Benes V, Kloor M, Ulrich CM, Knebel Doeberitz M von,
Sobhani I, Bork P. 2014. Potential of fecal microbiota for early-stage
detection of colorectal cancer. Molecular systems biology 10:766--766.

\hypertarget{ref-Fearon:2011go}{}
25. Fearon ER. 2011. Molecular genetics of colorectal cancer. Annual
review of pathology 6:479--507.

\hypertarget{ref-Levin:2008kz}{}
26. Levin B, Lieberman DA, McFarland B, Smith RA, Brooks D, Andrews KS,
Dash C, Giardiello FM, Glick S, Levin TR, Pickhardt P, Rex DK, Thorson
A, Winawer SJ, American Cancer Society Colorectal Cancer Advisory Group,
US Multi-Society Task Force, American College of Radiology Colon Cancer
Committee. 2008. Screening and surveillance for the early detection of
colorectal cancer and adenomatous polyps, 2008: a joint guideline from
the American Cancer Society, the US Multi-Society Task Force on
Colorectal Cancer, and the American College of Radiology., pp. 130--160.
\emph{In} CA: A cancer journal for clinicians. The University of Texas
MD Anderson Cancer Center, Houston, TX, USA. John Wiley \& Sons, Ltd.

\hypertarget{ref-Zauber:2015cr}{}
27. Zauber AG. 2015. The impact of screening on colorectal cancer
mortality and incidence: has it really made a difference? Digestive
diseases and sciences 60:681--691.

\hypertarget{ref-Pedulla:2003tu}{}
28. Pedulla ML, Ford ME, Houtz JM, Karthikeyan T, Wadsworth C, Lewis JA,
Jacobs-Sera D, Falbo J, Gross J, Pannunzio NR, Brucker W, Kumar V,
Kandasamy J, Keenan L, Bardarov S, Kriakov J, Lawrence JG, Jacobs WR,
Hendrix RW, Hatfull GF. 2003. Origins of highly mosaic mycobacteriophage
genomes. Cell 113:171--182.

\hypertarget{ref-Hannigan:2015fz}{}
29. Hannigan GD, Meisel JS, Tyldsley AS, Zheng Q, Hodkinson BP,
SanMiguel AJ, Minot S, Bushman FD, Grice EA. 2015. The Human Skin
Double-Stranded DNA Virome: Topographical and Temporal Diversity,
Genetic Enrichment, and Dynamic Associations with the Host Microbiome.
mBio 6:e01578--15.

\hypertarget{ref-Brum:2015iaa}{}
30. Brum JR, Ignacio-Espinoza JC, Roux S, Doulcier G, Acinas SG, Alberti
A, Chaffron S, Cruaud C, Vargas C de, Gasol JM, Gorsky G, Gregory AC,
Guidi L, Hingamp P, Iudicone D, Not F, Ogata H, Pesant S, Poulos BT,
Schwenck SM, Speich S, Dimier C, Kandels-Lewis S, Picheral M, Searson S,
Tara Oceans Coordinators, Bork P, Bowler C, Sunagawa S, Wincker P,
Karsenti E, Sullivan MB. 2015. Ocean plankton. Patterns and ecological
drivers of ocean viral communities. Science 348:1261498--1261498.

\hypertarget{ref-Minot:2011ez}{}
31. Minot S, Sinha R, Chen J, Li H, Keilbaugh SA, Wu GD, Lewis JD,
Bushman FD. 2011. The human gut virome: Inter-individual variation and
dynamic response to diet. Genome Research 21:1616--1625.

\hypertarget{ref-Hannigan:2017ky}{}
32. Hannigan GD, Zheng Q, Meisel JS, Minot SS, Bushman FD, Grice EA.
2017. Evolutionary and functional implications of hypervariable loci
within the skin virome. PeerJ 5:e2959.

\hypertarget{ref-Reyes:2010cwa}{}
33. Reyes A, Haynes M, Hanson N, Angly FE, Heath AC, Rohwer F, Gordon
JI. 2010. Viruses in the faecal microbiota of monozygotic twins and
their mothers. Nature 466:334--338.

\hypertarget{ref-Hannigan:2018gv}{}
34. Hannigan GD, Duhaime MB, Koutra D, Schloss PD. 2018. Biogeography
and environmental conditions shape bacteriophage-bacteria networks
across the human microbiome. PLoS Computational Biology 14:e1006099.

\hypertarget{ref-Lengeling:2013ia}{}
35. Lengeling A, Mahajan A, Gally DL. 2013. Bacteriophages as Pathogens
and Immune Modulators? mBio 4:e00868--13--e00868--13.

\hypertarget{ref-Grski:2012fa}{}
36. G rski A, Mi dzybrodzki R, Borysowski J, D browska K, Wierzbicki P,
Ohams M, Korczak-Kowalska G yna, Olszowska-Zaremba N, usiak-Szelachowska
M, K ak M, Jo czyk E, Kaniuga E, Go a A, Purchla S, Weber-D browska B,
Letkiewicz S awomir, Fortuna W, Szufnarowski K, Pawe czyk Z aw, Rog P, K
osowska D. 2012. Phage as a Modulator of Immune Responses, pp. 41--71.
\emph{In} Bacteriophages, part b. Bacteriophage Laboratory, Ludwik
Hirszfeld Institute of Immunology; Experimental Therapy, Polish Academy
of Sciences, Wrocław, Poland. agorski@ikp.pl; Elsevier.

\hypertarget{ref-Rossmann:2015cj}{}
37. Rossmann FS, Racek T, Wobser D, Puchalka J, Rabener EM, Reiger M,
Hendrickx APA, Diederich A-K, Jung K, Klein C, Huebner J. 2015.
Phage-mediated Dispersal of Biofilm and Distribution of Bacterial
Virulence Genes Is Induced by Quorum Sensing. PLoS Pathogens
11:e1004653--17.

\hypertarget{ref-Brockhurst:2013iq}{}
38. Brockhurst MA, Koskella B. 2013. Experimental coevolution of species
interactions. Trends in ecology \& evolution 28:367--375.

\hypertarget{ref-Kozich:2013db}{}
39. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq Illumina sequencing
platform. Applied and Environmental Microbiology 79:5112--5120.

\hypertarget{ref-Schloss:2009do}{}
40. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
Thallinger GG, Van Horn DJ, Weber CF. 2009. Introducing mothur:
open-source, platform-independent, community-supported software for
describing and comparing microbial communities. Applied and
Environmental Microbiology 75:7537--7541.

\hypertarget{ref-Pruesse:2007jc}{}
41. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J,
Glöckner FO. 2007. SILVA: a comprehensive online resource for quality
checked and aligned ribosomal RNA sequence data compatible with ARB.
Nucleic Acids Research 35:7188--7196.

\hypertarget{ref-Edgar:2011gy}{}
42. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. 2011. UCHIME
improves sensitivity and speed of chimera detection. Bioinformatics
27:2194--2200.

\hypertarget{ref-Thurber:2009dn}{}
43. Thurber RV, Haynes M, Breitbart M, Wegley L, Rohwer F. 2009.
Laboratory procedures to generate viral metagenomes. Nature protocols
4:470--483.

\hypertarget{ref-Kleiner:2015kd}{}
44. Kleiner M, Hooper LV, Duerkop BA. 2015. Evaluation of methods to
purify virus-like particles for metagenomic sequencing of intestinal
viromes. BMC Genomics 16:7.

\hypertarget{ref-Martin:2011eu}{}
45. Martin M. 2011. Cutadapt removes adapter sequences from
high-throughput sequencing reads. EMBnetjournal 17:10.

\hypertarget{ref-FASTXToolkit:wr}{}
46. Hannon GJ. 2010. FASTX-Toolkit GNU Affero General Public License.

\hypertarget{ref-Schmieder:2011fo}{}
47. Schmieder R, Edwards R. 2011. Fast identification and removal of
sequence contamination from genomic and metagenomic datasets. PLOS ONE
6:e17288.

\hypertarget{ref-Li:2016kd}{}
48. Li D, Luo R, Liu C-M, Leung C-M, Ting H-F, Sadakane K, Yamashita H,
Lam T-W. 2016. MEGAHIT v1.0: A fast and scalable metagenome assembler
driven by advanced methodologies and community practices. METHODS
102:3--11.

\hypertarget{ref-Langmead:2012jh}{}
49. Langmead B, Salzberg SL. 2012. Fast gapped-read alignment with
Bowtie 2. Nature Methods 9:357--359.

\hypertarget{ref-Alneberg:2014fc}{}
50. Alneberg J, Bjarnason BS aacute ri, Bruijn I de, Schirmer M, Quick
J, Ijaz UZ, Lahti L, Loman NJ, Andersson AF, Quince C. 2014. Binning
metagenomic contigs by coverage and composition. Nature Methods 1--7.

\hypertarget{ref-veganCommunityEco:xdLliqSE}{}
51. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
Minchin PR, OHara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
Wagner H. vegan: Community Ecology Package.

\hypertarget{ref-caretClassificatio:ux5fU2Litux5f1}{}
52. Kuhn M. 2016. caret: Classification and Regression Training. CRAN.

\hypertarget{ref-Guidi:2016kf}{}
53. Guidi L, Chaffron S, Bittner L, Eveillard D, Larhlimi A, Roux S,
Darzi Y, Audic S, Berline L, Brum JR, Coelho LP, Espinoza JCI, Malviya
S, Sunagawa S, Dimier C, Kandels-Lewis S, Picheral M, Poulain J, Searson
S, Tara Oceans Consortium Coordinators, Stemmann L, Not F, Hingamp P,
Speich S, Follows M, Karp-Boss L, Boss E, Ogata H, Pesant S, Weissenbach
J, Wincker P, Acinas SG, Bork P, Vargas C de, Iudicone D, Sullivan MB,
Raes J, Karsenti E, Bowler C, Gorsky G. 2016. Plankton networks driving
carbon export in the oligotrophic ocean. Nature 532:465--470.

\hypertarget{ref-Camacho:2009fc}{}
54. Camacho C, Coulouris G, Avagyan V, Ma N, Papadopoulos J, Bealer K,
Madden TL. 2009. BLAST+: architecture and applications. BMC
Bioinformatics 10:1.

\newpage

\section{Figure Legends}\label{figure-legends}

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Cohort and sample processing outline. Thirty subject
stool samples were collected from healthy, adenoma (pre-cancer), and
carcinoma (cancer) patients. Stool samples were split into two aliquots,
the first of which was used for bacterial sequencing and the second
which was used for virus sequencing. Bacterial sequencing was done using
both 16S rRNA amplicon and whole metagenomic shotgun sequencing
techniques. Virus samples were purified for viruses using filtration and
a combination of chloroform (bacterial lysis) and DNase (exposed genomic
DNA degradation). The resulting encapsulated virus DNA was sequenced
using whole metagenomic shotgun sequencing.} \label{sampleproc}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Results from healthy vs cancer classification models
built using virome signatures, bacterial 16S rRNA gene sequence
signatures, whole metagenomic signatures, and a combination of virome
and 16S rRNA gene sequence signatures. A) An example ROC curve for
visualizing the performance of each of the models for classifying stool
as coming from either an individual with a cancerous or healthy colon.
B) Quantification of the AUC variation for each model, and how it
compared to each of the other models based on 15 iterations. A pairwise
Wilcoxon test with a false discovery rate multiple hypothesis correction
demonstrated that all models are significantly different from each other
(p-value \textless{} 0.01). C) Mean decrease in accuracy (measurement of
importance) of each operational taxonomic unit within the 16S rRNA gene
classification model when removed from the classification model. Mean is
represented by a point, and bars represent standard error. D) Mean
decrease in accuracy of each operational virus unit in the virome
classification model. E) Mean decrease in accuracy of each operational
genomic unit and operational taxonomic unit in the model using both 16S
rRNA gene and virome features.}\label{predmodel}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics[width=0.50000\textwidth]{}
\caption{\emph{Lysogenic phage relative abundance in disease states.
Phage OVUs were predicted to be either lytic or lysogenic, and the
relative abundance of lysogenic phages was quantified and represented as
a boxplot. No disease groups were statistically
significant.}\label{replicationstyles}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Relative abundance correlations between bacterial OTUs
and virome OVUs. A) Pearson correlation coefficient values between all
bacterial OTUs (x-axis) and viral OVUs (y-axis) with blue being
positively correlated and red being negatively correlated. Bar plots
indicate the viral (left) and bacterial (bottom) operational unit
importance in their colorectal cancer classification models, such that
the most important units are in the top left corner. B) Magnification of
the boxed region in panel (A), highlighting the correlation between the
most important bacterial OTUs and virome OVUs. The most important
operational units are in the top left corner of the heatmap, and the
correlation scale is the same as panel (A). C) Histogram quantifying the
frequencies of Pearson correlation coefficients between all bacterial
OTUs and virome OVUs.}\label{correlations}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Community network analysis utilizing predicted
interactions between bacteria and phage operational genomic units. A)
Visualization of the community network for our colorectal cancer cohort.
B) Scatter plot illustrating the correlation between importance (mean
decrease in accuracy) and the degree of centrality for each OVU. A
linear regression line was fit to illustrate the correlation (blue)
which was found to be statistically significantly and weakly correlated
(p-value = 0.00409, R = 0.176).}\label{network}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Final working hypothesis from this study. These panels
summarize our thoughts on our results and represent interesting future
directions that we predict will build on the presented work. A) Basic
model illustrating the connections between the virome, bacterial
communities, and colorectal cancer. B) Working hypothesis of how the
bacteriophage community is associated with colorectal cancer and the
associated bacterial community.}\label{modelsummary}}
\end{figure}

\newpage

\section{Supplemental Figure Legends}\label{supplemental-figure-legends}

\beginsupplement

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Basic Quality Control Metrics. A) VLP genomic DNA yield
from all sequenced samples. Each bar represents a sample which is
grouped and colored by its associated disease group. B) Sequence yield
following quality control including quality score filtering and human
decontamination. Dashed line shows rarefaction line in which all samples
with lower sequence yield were excluded from downstream
analysis.}\label{qualcontrol}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Length and coverage statistics. A) Heated scatter plot
demonstrating the distribution of contig coverage (number of sequences
mapping to each contig) and contig length for the virus metagenomic
sample set. B) Scatter plot illustrating the distribution of operational
viral unit (OVU) length and sequence coverage for the virus metagenomic
sample set. C) Heated scatter plot demonstrating the distribution of
contig coverage and length for the whole metagenomic sample set. D)
Scatter plot illustrating the distribution of operational genomic unit
(OGU) length and sequence coverage for the whole metagenomic sample
set.}\label{contigqc}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Operational genomic unit composition stats. A) Strip
chart demonstrating the length and frequency of contigs within each
operational genomic unit of the virome sample set. The y-axis is the
operational genomic unit identifier, and x-axis is the length of each
contig, and each dot represents a contig found within the specified
operational genomic unit. B) Density plot (analogous to histogram) of
the number of virome operational genomic units containing the specific
number of contigs, as indicated by the x-axis. C-D) Sample plots as
panels C and D, but for the whole metagenomic sample
set.}\label{clustercontigqc}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Diversity calculations comparing cancer states of the
colorectal virome, based on relative abundance of operational genomic
units in each sample. A) NMDS ordination of community samples, colored
for cancerous (green), pre-cancerous (red), and healthy (yellow). B)
Differences in means between disease group centroids with 95\%
confidence intervals based on an ANOSIM test with a post hoc
multivariate Tukey test. Comparisons (indicated on y-axis) in which the
intervals cross the zero mean difference line (dashed line) were not
significantly different. C) Shannon diversity and D) richness alpha
diversity quantification comparing pre-cancerous (grey), cancerous
(red), and healthy (tan) states.}\label{betaogu}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Beta-diversity comparing disease states and the study
negative controls which remained following rarefaction. Differences in
means between disease group centroids with 95\% confidence intervals
based on an ANOSIM test with a post hoc multivariate Tukey test.
Comparisons in which the intervals cross the zero mean difference line
(dashed line) were not significantly different.}\label{betaogunegative}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Comparison of bacterial 16S rRNA classification models
with and without OTUs whose median relative abundance are greater than
zero. A) Classification model performance (measured as area under the
curve) for bacteria models using 16S rRNA data both with and without
filtering of samples whose median was zero. Significance was calculated
using a Wilcoxon rank sum test, and the resulting p-value is shown. The
random area under the curve (0.5) is marked with a dashed line. B)
Relative abundance of the six bacterial OTUs removed when filtered for
OTUs with median relative abundance of zero. OTU relative abundance is
separated by healthy (red) and cancerous (grey) samples. Relative
abundance of 1\% is marked by the dashed line.}\label{16scompare}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{Transition of colorectal cancer importance through
disease progression. A) Virus and B) 16S rRNA gene model performance
(AUC) when discriminating all binary combinations of disease types. Blue
line represents mean performance from multiple random iterations. C-E)
Top ten important phage OVUs when classifying each combination of
disease state, as measured by the mean decrease in accuracy metric. Mean
is represented by a point, and bars represent standard error. Disease
comparison is specified in the top left corner of each panel. F-H) Top
ten important bacterial 16S rRNA gene OTUs for classifying each disease
state combination.}\label{transitionmodels}}
\end{figure}

\newpage

\begin{figure}[htbp]
\centering
\includegraphics{}
\caption{\emph{ROC curves from A) virome and B) bacterial 16S
three-class random forest models tuned on mean AUC. Each curve
represents the ability of the specified class to be classified against
the other two classes. C) Quantification of the mean AUC variation for
each model based on 10 model iterations. A pairwise Wilcoxon test with a
Bonferroni multiple hypothesis correction demonstrated that the models
are significantly different (alpha = 0.01). D) Mean decrease in accuracy
when virome operational genomic units and E) bacterial 16S OTUs are
removed from the respective three-class classification models. Results
based on 25 iterations. F) Relative abundance of the six most important
virome OVUs in the model, with the most important on the right. Line
indicates abundance mean.}\label{threewaymodel}}
\end{figure}

\end{document}

### 2018-07-08 doc/manuscript.Rmd
Pushing a couple of the edits.

@@
 ## Community Context of Influential Phages
-Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the bacteriophage signal may be a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal.
+Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we hypothesized that the bacteriophage signal may be a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal.
 If this hypothesis was true, we would expect a correlation between the relative abundances of influential bacteria and phages.
@@
 Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**.
-This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria, and suggest a more complex relationship between the bacterial and phage communities.
+This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria, and suggests a more complex relationship between the bacterial and phage communities.
 
@@
 
-Using our findings here, we have begun a path toward the delineation of the role the colonic virome plays in colorectal cancer in the form of a working hypothesis that will guide our future studies **(Figure \ref{modelsummary} A)**.
+Based on our findings, we have developed a conceptual model to be tested in future studies aimed at elucidating the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**.
 We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states.
-By implementing a machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer.
-The colorectal cancer virome was composed primarily of bacteriophages.
+By implementing a machine learning approach (random forest classification) to leverage inherent, complex patterns not detected by diversity measures, we were able to detect strong associations between the colon virus community composition and colorectal cancer.
+The dsDNA virome of colorectal cancer was composed primarily of bacteriophages.
 These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations.
@@
 
-At this time, our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level.
+Our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level.
 There are many different ways in which this system can be operating, and our working hypothesis is one.
@@
 Important future directions will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and even attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.
-
-As with most microbiome studies it is important to note that, despite including data for both bacteria and viruses, we are still only studying a subset of the microbiome.
-Here we are building on the bacterial-centric literature around colorectal cancer by beginning to characterize the virome.
-The caveat here is that we are technically limited to studying only the DNA virome, and therefore are not measuring RNA viruses.
-This will be an important direction for future studies, as well as beginning to characterize other additional microbial communities such as fungi, archaea, etc.
 
@@
 A) VLP genomic DNA yield from all sequenced samples.
-Each bar represents a sample which is grouped and colored by its associated disease group.
+Each bar represents a sample which is grouped and colored by its associated disease group or no-DNA negative control.
 B) Sequence yield following quality control including quality score filtering and human decontamination.
-Dashed line shows rarefaction line in which all samples with lower sequence yield were excluded from downstream analysis.*\label{qualcontrol}]()
+Dashed line indicates rarefaction level ($10^6$ reads) in which all samples with lower sequence yields less than this level were excluded from downstream analysis.*\label{qualcontrol}]()
 
@@
 
-![*Beta-diversity comparing disease states and the study negative controls which remained following rarefaction.
+![*Beta-diversity analysis comparing Bray-Curtis dissimilarity between disease state and negative control community structures that were captured following sequence rarefaction.
 Differences in means between disease group centroids with 95% confidence intervals based on an ANOSIM test with a post hoc multivariate Tukey test.

### 2018-07-08 revfiles/internalComments.rmd
A couple others real quick.

Original comments at:

https://docs.google.com/document/d/1hW_4z9Uj_mOIxonJWkYWxQ7T9lf4Boj8IXG2JVI--wU/edit#

## COMMENT 1

<CHANGE LEGEND AND RERENDER>

Sorry I don't see the discordance between the figures. Maybe we are looking at different versions?

Love the rewording, that is much better!

## COMMENT 2

Incorporated the recommended legend.

<ADD DASHED LINE TO LEGEND>

## COMMENT 3

## COMMENT 4

We need to be careful about getting too caught up in the specific "important" phages because none of them are really that important. The importance levels are small, and multiple times less than what was observed by bacteria. So it's probably not a single phage involved. Also these are representative sequences from OVUs, and are not really meant to provide specific resolution, as described for the PLOS comp bio paper.

## COMMENT 5

Changed.

## COMMENT 6

Thanks, the intent was not to have it read as the only alternative hypothesis. We can add more, and point out that those are also not the only alternatives, because I think we could continue hypothesizing about this and coming up with more options.

<FIX THE TEXT>

## COMMENT 7

Changed.

## COMMENT 8

I like this, although I would be hesitant to say non-linear patterns because, while it excells at complex non-linear relationships, it should be able to detect them as well.

## COMMENT 9

This is great. Changed it.

## COMMENT 10

Thanks for the feedback. I'll also try to make it sound less dodgy. The intent was to do the re-analyses and textual revisions that were warranted.

## COMMENT 11

Changed it.

## COMMENT 12

### 2018-08-13 doc/manuscript.Rmd
Updates

@@
 ---
-title: Diagnostic Potential & the Interactive Dynamics of the Colorectal Cancer Virome
+title: The Diagnostic Potential & Interactive Dynamics of the Colorectal Cancer Virome
 author:
@@
 The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**.
+It is also important to note that while these viruses were driving the signature, the magnitude of their importance and the significance of those values was noticibly less than the bacterial 16S signature, suggesting that unlike what is observed in the bacteria, there are many viruses that are associated with the cancerous state.
 The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa who could not be confidently idetnfied beyond their broad phage identification (i.e.
@@
 ## Phage Dominance in CRC Virome
-Differences in the colorectal cancer virome could have been driven by eukaryotic viruses or by bacteriophages.
+Differences in the colorectal cancer virome could have been driven by eukaryotic (human) viruses or by bacteriophages.
 To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages.
@@
 ## Community Context of Influential Phages
-Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we hypothesized that the bacteriophage signal may be a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal.
+Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested a potential hypothesis that the bacteriophage signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal.
 If this hypothesis was true, we would expect a correlation between the relative abundances of influential bacteria and phages.
@@
 Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**.
-This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria, and suggests a more complex relationship between the bacterial and phage communities.
+This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria.
 
-Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
+Given these findings, we posited that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
 In other words, we hypothesized that the influential bacteriophages were community hubs (i.e.
@@
 Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model.
-Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**.
+Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages that were important in driving colorectal cancer also were more likely to be community hubs **(Figure \ref{network} B)**.
 Together these findings supported our hypothesis that influential phages were hubs within their microbial communities and had broad host ranges.
@@
 
-Based on our findings, we have developed a conceptual model to be tested in future studies aimed at elucidating the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**.
+Based on our findings, we have developed a conceptual model to be tested in our future studies aimed at elucidating the role the colonic virome plays in colorectal cancer **(Figure \ref{modelsummary} A)**.
 We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states.
@@
 In addition to modeling the potential connections between virus communities, bacterial communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur.
-This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu], although it is important to note that there are also many other alternative hypotheses by which the system could be operating.
+This was done by incorporating our findings into a current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu], although it is important to note that there are also many other alternative hypotheses by which the system could be operating.
 We hypothesize that the process begins with broadly infectious phages in the colon lysing, and thereby disrupting, the existing bacterial communities.
@@
 Our working hypothesis represents a conceptualization of areas for future work, which will be required to characterize the colorectal cancer microbiome at the functional, mechanistic level.
-There are many different ways in which this system can be operating, and our working hypothesis is one.
+There are many different ways in which this system may operating, and our working hypothesis is one.
 For example, it is possible that the bacterial communities cause a change in the virome instead of the virome altering the bacterial communities.
-Reviewing every possible mechanism and characteristic of the system is beyond the scope of this manuscript, but the results we present provide a step toward the answer.
-Important future directions will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and even attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.
+To better understand this system, future studies will include larger cohort human studies, further *in vitro* and *in vivo* mechanistic experimentation, and attempts at community studies using absolute abundance values instead of relative abundance, which would allow for more accurate community dynamic modeling.
+Overall, this study provides a conceptual foundation to direct future characterization of the colorectal cancer microbiome at the functional, mechanistic level.
 
@@
 B) Sequence yield following quality control including quality score filtering and human decontamination.
-Dashed line indicates rarefaction level ($10^6$ reads) in which all samples with lower sequence yields less than this level were excluded from downstream analysis.*\label{qualcontrol}]()
+Dashed line indicates rarefaction level ($10^6$ reads) in which all samples with lower sequence yields less than this level were excluded from downstream analysis.
+After rarefaction and removal of samples with less than $10^6$ reads, 27 healthy, 28 cancerous, 27 adenomatous, and 3 negative control samples remained.*\label{qualcontrol}]()
 

### 2018-08-13 revfiles/internalComments.rmd
Updates

@@
 
-<CHANGE LEGEND AND RERENDER>
+Sorry I don't see the discordance between the figures. Maybe we are looking at different versions? Regardless it should be fixed now.
 
-Sorry I don't see the discordance between the figures. Maybe we are looking at different versions?
-
-Love the rewording, that is much better!
+Rewording was also great.
 
@@
 
-<ADD DASHED LINE TO LEGEND>
-
 ## COMMENT 3
 
-## COMMENT 4
+Done.
 
-We need to be careful about getting too caught up in the specific "important" phages because none of them are really that important.
-The importance levels are small, and multiple times less than what was observed by bacteria.
-So it's probably not a single phage involved.
-Also these are representative sequences from OVUs, and are not really meant to provide specific resolution, as described for the PLOS comp bio paper.
+## COMMENT 4-A
+
+This is great feedback, thanks. I went through and made some revisions to hopefully make this clearer.
+
+## COMMENT 4-B
+
+I think we need to be careful about getting too caught up in the specific "important" phages because none of them are really all that important conpmared to the others.
+The importance levels are close, and multiple times less than what was observed by bacteria.
+So it's probably not a single phage involved and I don't think it really makes sense to get that caught up in a single phage (a point I probably didn't hit hard enough in the text).
+Also it is still important to note that these are representative sequences from OVUs, and are not really meant to provide specific resolution, as described for the PLOS comp bio paper.
+It could be a good idea to really dig into this if we had more compelling data (i.e.
+similar importance signal to the fusobacterium case on the CRC bacterial side of things) but here we are really only looking at community wide signatures without diving into any given specific elements.
 
@@
 We can add more, and point out that those are also not the only alternatives, because I think we could continue hypothesizing about this and coming up with more options.
-
-<FIX THE TEXT>
+I'll add in these suggestions and make it clearer that there are many options to explore.
 
@@
 
-I like this, although I would be hesitant to say non-linear patterns because, while it excells at complex non-linear relationships, it should be able to detect them as well.
+I like this, although I am hesitant to focus on non-linear patterns in the language because, while it excells at complex non-linear relationships, it should be able to a good job detecting linear relationships as well.
 
@@
 
-Thanks for the feedback. I'll also try to make it sound less dodgy. The intent was to do the re-analyses and textual revisions that were warranted.
+Thanks for the feedback. I'll also try to make it sound less dodgy.
 
@@
 ## COMMENT 12
+
+This is great and incorporated the changes.
+
+I agree that it should be self evident but this is something that some of my colleagues, who are not familiar with phages, appreciated hearing.
+
+These are also great points about how phages will interact with their bacterial hosts, and their lysogenic replication cycles. These will be important as we work to understand the system.

### 2018-09-17 doc/manuscript.Rmd
A couple minor updates

@@
 It is also important to note that while these viruses were driving the signature, the magnitude of their importance and the significance of those values was noticibly less than the bacterial 16S signature, suggesting that unlike what is observed in the bacteria, there are many viruses that are associated with the cancerous state.
-The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa who could not be confidently idetnfied beyond their broad phage identification (i.e.
+The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa that could not be confidently identified beyond their broad phage identification (i.e.
 "unclassified").
-Many of the important viruses, who were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown").
+Many of the important viruses, that were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown").
 This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa].
@@
 ## Community Context of Influential Phages
-Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested a potential hypothesis that the bacteriophage signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal.
-If this hypothesis was true, we would expect a correlation between the relative abundances of influential bacteria and phages.
-Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**.
+Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested a potential hypothesis that the virome signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal.
+If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacterial OTUs and virome OVUs.
+Instead, we observed a strikingly low correlation between bacterial and viral relative abundances **(Figure \ref{correlations} A,C)**.
 Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**.
-This evidence supported our null hypothesis that the influential phages were not primarily reflections of influential bacteria.
+This evidence supported our null hypothesis that the influential viral OVUs were not primarily reflections of influential bacteria.
 

### 2018-09-17 doc/manuscript.tex
A couple minor updates

@@
 \hypersetup{unicode=true,
-            pdftitle={Diagnostic Potential \& the Interactive Dynamics of the Colorectal Cancer Virome},
+            pdftitle={The Diagnostic Potential \& Interactive Dynamics of the Colorectal Cancer Virome},
             pdfborder={0 0 0},
@@
 
-  \title{Diagnostic Potential \& the Interactive Dynamics of the Colorectal
+  \title{The Diagnostic Potential \& Interactive Dynamics of the Colorectal
 Cancer Virome}
@@
 for these viruses in tumorigenesis \textbf{(Figure \ref{predmodel} D)}.
-The identified viruses were bacteriophages, belonging to
-\emph{Siphoviridae}, \emph{Myoviridae}, and phage taxa who could not be
-confidently idetnfied beyond their broad phage identification (i.e.
-``unclassified''). Many of the important viruses, who were confirmed to
+It is also important to note that while these viruses were driving the
+signature, the magnitude of their importance and the significance of
+those values was noticibly less than the bacterial 16S signature,
+suggesting that unlike what is observed in the bacteria, there are many
+viruses that are associated with the cancerous state. The identified
+viruses were bacteriophages, belonging to \emph{Siphoviridae},
+\emph{Myoviridae}, and phage taxa that could not be confidently
+identified beyond their broad phage identification (i.e.
+``unclassified''). Many of the important viruses, that were confirmed to
 not have genomic similarity to known bacterial genomes, were
@@
 Differences in the colorectal cancer virome could have been driven by
-eukaryotic viruses or by bacteriophages. To better understand the types
-of viruses that were important for colorectal cancer, we identified the
-virome OVUs as being similar to either eukaryotic viruses or
-bacteriophages. The most important viruses to the classification model
-were identified as bacteriophages (\textbf{Figure \ref{threewaymodel})}.
-Overall, we were able to identify 78.8\% of the OVUs as known viruses,
-and 93.8\% of those viral OVUs aligned to bacteriophage reference
-genomes. It is important to note that this could have been influenced by
-our methodological biases against enveloped viruses (more common of
-eukaryotic viruses than bacteriophage), due to chloroform and DNase
-treatment for purification.
+eukaryotic (human) viruses or by bacteriophages. To better understand
+the types of viruses that were important for colorectal cancer, we
+identified the virome OVUs as being similar to either eukaryotic viruses
+or bacteriophages. The most important viruses to the classification
+model were identified as bacteriophages (\textbf{Figure
+\ref{threewaymodel})}. Overall, we were able to identify 78.8\% of the
+OVUs as known viruses, and 93.8\% of those viral OVUs aligned to
+bacteriophage reference genomes. It is important to note that this could
+have been influenced by our methodological biases against enveloped
+viruses (more common of eukaryotic viruses than bacteriophage), due to
+chloroform and DNase treatment for purification.
 
@@
 Because the link between colorectal cancer and the virome was driven by
-bacteriophages, we hypothesized that the bacteriophage signal may be a
-mere reflection of the bacterial signal, and thus highly correlated with
-the bacterial signal. If this hypothesis was true, we would expect a
-correlation between the relative abundances of influential bacteria and
-phages. Instead, we observed a strikingly low correlation between
-bacterial and phage relative abundances \textbf{(Figure
-\ref{correlations} A,C)}. Overall, there was an absence of correlation
-between the most influential OVUs and bacterial OTUs \textbf{(Figure
-\ref{correlations} B)}. This evidence supported our null hypothesis that
-the influential phages were not primarily reflections of influential
-bacteria, and suggest a more complex relationship between the bacterial
-and phage communities.
-
-Given these findings, we hypothesized that the most influential phages
-were acting by infecting a wide range of bacteria in the overall
-community, instead of just the influential bacteria. In other words, we
+bacteriophages (as opposed to non-bacterial viruses), we tested a
+potential hypothesis that the virome signal was a mere reflection of the
+bacterial signal, and thus highly correlated with the bacterial signal.
+If this hypothesis were true, we would expect a correlation between the
+relative abundances of influential bacterial OTUs and virome OVUs.
+Instead, we observed a strikingly low correlation between bacterial and
+viral relative abundances \textbf{(Figure \ref{correlations} A,C)}.
+Overall, there was an absence of correlation between the most
+influential OVUs and bacterial OTUs \textbf{(Figure \ref{correlations}
+B)}. This evidence supported our null hypothesis that the influential
+viral OVUs were not primarily reflections of influential bacteria.
+
+Given these findings, we posited that the most influential phages were
+acting by infecting a wide range of bacteria in the overall community,
+instead of just the influential bacteria. In other words, we
 hypothesized that the influential bacteriophages were community hubs
@@
 to the disease model (p-value = 0.004, R = 0.176), suggesting that
-phages important in driving colorectal cancer also were more likely to
-be community hubs \textbf{(Figure \ref{network} B)}. Together these
-findings supported our hypothesis that influential phages were hubs
-within their microbial communities and had broad host ranges.
+phages that were important in driving colorectal cancer also were more
+likely to be community hubs \textbf{(Figure \ref{network} B)}. Together
+these findings supported our hypothesis that influential phages were
+hubs within their microbial communities and had broad host ranges.
 
@@
 
-Using our findings here, we have begun a path toward the delineation of
-the role the colonic virome plays in colorectal cancer in the form of a
-working hypothesis that will guide our future studies \textbf{(Figure
-\ref{modelsummary} A)}. We found that basic diversity metrics of alpha
-diversity (richness and Shannon diversity) and beta diversity
-(Bray-Curtis dissimilarity) were insufficient for identifying virome
-community differences between healthy and cancerous states. By
-implementing a machine learning approach (random forest classification),
-we detected strong associations between the colon virus community
-composition and colorectal cancer. The colorectal cancer virome was
-composed primarily of bacteriophages. These phage communities were not
+Based on our findings, we have developed a conceptual model to be tested
+in our future studies aimed at elucidating the role the colonic virome
+plays in colorectal cancer \textbf{(Figure \ref{modelsummary} A)}. We
+found that basic diversity metrics of alpha diversity (richness and
+Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were
+insufficient for identifying virome community differences between
+healthy and cancerous states. By implementing a machine learning
+approach (random forest classification) to leverage inherent, complex
+patterns not detected by diversity measures, we were able to detect
+strong associations between the colon virus community composition and
+colorectal cancer. The dsDNA virome of colorectal cancer was composed
+primarily of bacteriophages. These phage communities were not
 exclusively predators of the most influential bacteria, as demonstrated
@@
 hypothesis for the mechanisms by which this may occur. This was done by
-incorporating our findings into the current model for colorectal cancer
+incorporating our findings into a current model for colorectal cancer
 development \textbf{(Figure \ref{modelsummary} B)} (22), although it is
@@
 
-At this time, our working hypothesis represents a conceptualization of
-areas for future work, which will be required to characterize the
-colorectal cancer microbiome at the functional, mechanistic level. There
-are many different ways in which this system can be operating, and our
-working hypothesis is one. For example, it is possible that the
-bacterial communities cause a change in the virome instead of the virome
-altering the bacterial communities. Reviewing every possible mechanism
-and characteristic of the system is beyond the scope of this manuscript,
-but the results we present provide a step toward the answer. Important
-future directions will include larger cohort human studies, further
-\emph{in vitro} and \emph{in vivo} mechanistic experimentation, and even
-attempts at community studies using absolute abundance values instead of
-relative abundance, which would allow for more accurate community
-dynamic modeling.
-
-As with most microbiome studies it is important to note that, despite
-including data for both bacteria and viruses, we are still only studying
-a subset of the microbiome. Here we are building on the
-bacterial-centric literature around colorectal cancer by beginning to
-characterize the virome. The caveat here is that we are technically
-limited to studying only the DNA virome, and therefore are not measuring
-RNA viruses. This will be an important direction for future studies, as
-well as beginning to characterize other additional microbial communities
-such as fungi, archaea, etc.
+Our working hypothesis represents a conceptualization of areas for
+future work, which will be required to characterize the colorectal
+cancer microbiome at the functional, mechanistic level. There are many
+different ways in which this system may operating, and our working
+hypothesis is one. For example, it is possible that the bacterial
+communities cause a change in the virome instead of the virome altering
+the bacterial communities. To better understand this system, future
+studies will include larger cohort human studies, further \emph{in
+vitro} and \emph{in vivo} mechanistic experimentation, and attempts at
+community studies using absolute abundance values instead of relative
+abundance, which would allow for more accurate community dynamic
+modeling. Overall, this study provides a conceptual foundation to direct
+future characterization of the colorectal cancer microbiome at the
+functional, mechanistic level.
 
@@
 from all sequenced samples. Each bar represents a sample which is
-grouped and colored by its associated disease group. B) Sequence yield
-following quality control including quality score filtering and human
-decontamination. Dashed line shows rarefaction line in which all samples
-with lower sequence yield were excluded from downstream
-analysis.}\label{qualcontrol}}
+grouped and colored by its associated disease group or no-DNA negative
+control. B) Sequence yield following quality control including quality
+score filtering and human decontamination. Dashed line indicates
+rarefaction level (\(10^6\) reads) in which all samples with lower
+sequence yields less than this level were excluded from downstream
+analysis. After rarefaction and removal of samples with less than
+\(10^6\) reads, 27 healthy, 28 cancerous, 27 adenomatous, and 3 negative
+control samples remained.}\label{qualcontrol}}
 \end{figure}
@@
 \includegraphics{}
-\caption{\emph{Beta-diversity comparing disease states and the study
-negative controls which remained following rarefaction. Differences in
-means between disease group centroids with 95\% confidence intervals
-based on an ANOSIM test with a post hoc multivariate Tukey test.
-Comparisons in which the intervals cross the zero mean difference line
-(dashed line) were not significantly different.}\label{betaogunegative}}
+\caption{\emph{Beta-diversity analysis comparing Bray-Curtis
+dissimilarity between disease state and negative control community
+structures that were captured following sequence rarefaction.
+Differences in means between disease group centroids with 95\%
+confidence intervals based on an ANOSIM test with a post hoc
+multivariate Tukey test. Comparisons in which the intervals cross the
+zero mean difference line (dashed line) were not significantly
+different.}\label{betaogunegative}}
 \end{figure}

### 2018-09-30 doc/manuscript.Rmd

@@
 Despite this, cancer microbiome studies have almost exclusively focused on bacteria instead of viruses.
-We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world, and a cancer linked to altered colonic bacterial community compositions but with an unknown association with the gut virome.
+We began evaluating the cancer virome by focusing on colorectal cancer, a primary cause of morbidity and mortality throughout the world and a cancer linked to altered colonic bacterial community compositions but with an unknown association with the gut virome.
 We used 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing of stool to evaluate the differences in human colorectal cancer virus and bacterial community composition.
@@
 Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide.
-Its risk and severity have been linked to colonic bacterial community composition.
+Its risk and severity have been associated with colonic bacterial community composition.
 Although human-specific viruses have been linked to other cancers and diseases, little is known about colorectal cancer virus communities.
 We addressed this knowledge gap by identifying differences in colonic virus communities in the stool of colorectal cancer patients and how they compared to bacterial community differences.
-The results suggested an indirect role for the virome in impacting colorectal cancer by modulating their associated bacterial community.
+The results suggested an indirect role for the virome in colorectal cancer by modulating their associated bacterial community.
 These findings both support a biological role for viruses in colorectal cancer and provide a new understanding of basic colorectal cancer etiology.
@@
 # Introduction
-The human gut virome is the community of all viruses found in the gut, including bacteriophages (viruses that only infect bacteria), eukaryotic viruses (viruses that only infect eukaryotic cells), and human-specific viruses (viruses that only infect human cells).
+The human gut virome is the community of all viruses found in the gut, including bacteriophages (i.e.
+viruses that only infect bacteria), eukaryotic viruses (i.e.
+viruses that only infect eukaryotic cells), and human-specific viruses (i.e.
+viruses that only infect human cells).
 Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up].
@@
 Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo].
-The US National Cancer Institute estimates over 1.5 million Americans were diagnosed with colorectal cancer in 2016 and over 500,000 Americans died from the disease [@Siegel:2014jo].
+The US National Cancer Institute estimates that over 1.5 million Americans were diagnosed with colorectal cancer in 2016 and over 500,000 Americans died from the disease [@Siegel:2014jo].
 Growing evidence suggests that an important component of colorectal cancer etiology may be perturbations in the colonic bacterial community [@Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb; @Dejea:2014fz; @Arthur:2012kl].
 Work in this area has led to a proposed disease model in which bacteria colonize the colon, develop biofilms, promote inflammation, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu].
-This association also has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix].
+This association also has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja; @Zeller:2014ix; @Sze2018; @Shah2017].
 While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and for identifying the underlying disease etiology, bacteria are only a subset of the colon microbiome.
 Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.
-We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.
 
@@
 Thus, work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.
+We evaluated disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer development: healthy, adenomatous, and cancerous.
 
 Here we begin to address the knowledge gap of whether virus community composition is altered in colorectal cancer and, if it is, how those differences might impact cancer progression and severity.
-We also aimed to evaluate the virome's potential for use as a diagnostic biomarker.
+We also aimed to evaluate the virome's potential as a reservoir of diagnostic biomarkers.
 The implications of this study are threefold.
@@
 In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**.
-This is consistent with what has been observed when the same metrics were applied to 16S rRNA gene sequences and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these community structures.
+This is consistent with what has been observed when the same metrics were applied to 16S rRNA gene sequences and metagenomic samples [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja; @Sze2018; @Shah2017] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these community structures.
 
 ## Virome Composition in Colorectal Cancer
-As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zackular:2014fba; @Baxter:2016dja].
+As opposed to the diversity metrics discussed above, OTU-based relative abundance profiles generated from 16S rRNA gene sequences are effective for classifying stool samples as originating from individuals with healthy, adenomatous, or cancerous colons [@Zeller:2014ix; @Zackular:2014fba; @Baxter:2016dja; @Sze2018; @Shah2017].
 By using classification models instead of attempting to identify single differentially abundant OTUs, these and other studies have been successful in capturing complex community relationships in which differences in taxonomic relative abundance are considered in the context of other taxa.
@@
 The virome signature also was driven by a few OVUs, suggesting a role for these viruses in tumorigenesis **(Figure \ref{predmodel} D)**.
-It is also important to note that while these viruses were driving the signature, the magnitude of their importance and the significance of those values was noticibly less than the bacterial 16S signature, suggesting that unlike what is observed in the bacteria, there are many viruses that are associated with the cancerous state.
-The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa that could not be confidently identified beyond their broad phage identification (i.e.
+It is also important to note that while these viruses were driving the signature, the magnitude of their importance and the significance of those values was noticeably less than the bacterial 16S rRNA gene signature, suggesting that unlike what is observed among the bacteria, there are many viruses that are associated with the cancerous state.
+These viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and phage taxa that could not be confidently identified beyond their broad phage identification (i.e.
 "unclassified").
-Many of the important viruses, that were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown").
+Many of the important viruses, which were confirmed to not have genomic similarity to known bacterial genomes, were unidentifiable (denoted "unknown").
 This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa].
@@
 ## Phage Influence Between CRC Stages
-Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer [@Zackular:2014fba; @Zackular:2016en; @Flynn:2016iu], we explored whether shifts in the relative influence of phages could be detected between healthy, adenomatous, and cancerous colons.
+Because previous work has identified shifts in which bacteria were most important at different stages of colorectal cancer, we explored whether shifts in the relative influence of phages could be detected between healthy, adenomatous, and cancerous colons.
 We evaluated community shifts between the disease stage transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the diagnosis groups around the transitions.
-While bacterial OTU models performed equally well for all disease class comparisons, the virome model performances differed **(Figure \ref{transitionmodels} A-B)**.
+Although bacterial OTU models performed equally well for all disease class comparisons, the virome model performances varied across comparisons **(Figure \ref{transitionmodels} A-B)**.
 Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members were important between the healthy to adenomatous and adenomatous to cancerous stages **(Figure \ref{transitionmodels} C-E)**.
@@
 To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages.
-The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**.
+The viruses that were most important to the performance of the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**.
 Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes.
@@
 ## Community Context of Influential Phages
-Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested a potential hypothesis that the virome signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal.
+Because the link between colorectal cancer and the virome was driven by bacteriophages (as opposed to non-bacterial viruses), we tested the hypothesis that the virome signal was a mere reflection of the bacterial signal, and thus highly correlated with the bacterial signal.
 If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacterial OTUs and virome OVUs.
@@
 
-Given these findings, we posited that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
+Given these findings, we hypothesize that the most influential phages act by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria.
 In other words, we hypothesized that the influential bacteriophages were community hubs (i.e.
@@
 We found that basic diversity metrics of alpha diversity (richness and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community differences between healthy and cancerous states.
-By implementing a machine learning approach (random forest classification) to leverage inherent, complex patterns not detected by diversity measures, we were able to detect strong associations between the colon virus community composition and colorectal cancer.
+By implementing a machine learning approach to leverage inherent, complex patterns not detected by diversity measures, we were able to detect strong associations between the colon virus community composition and colorectal cancer.
 The dsDNA virome of colorectal cancer was composed primarily of bacteriophages.
@@
 Viral community dynamics have the potential to provide an abundance of information to supplement those of bacterial communities.
-Evidence has suggested that the virome is a crucial component to the microbiome and that bacteriophages are important players.
+The growing body of evidence suggests that the virome is a crucial component to the microbiome and that bacteriophages are important players.
 Bacteriophage and bacterial communities cannot maintain stability and co-evolution without one another [@RodriguezValera:2009cr; @Brockhurst:2013iq].

### 2018-09-30 doc/oman.tex

--- 2018-05-27 doc/manuscript.tex
@@
 
-Human viruses (those that infect human cells) have been associated with
-many cancers, largely due to their mutagenic and functionally
-manipulative abilities. Despite this, cancer microbiome studies have
-almost exclusively focused on bacteria instead of viruses. We began
-evaluating the cancer virome by focusing on colorectal cancer, a primary
-cause of morbidity and mortality throughout the world, and a cancer
-linked to altered colonic bacterial community compositions but with an
-unknown association with the gut virome. We used 16S rRNA gene, whole
-shotgun metagenomic, and purified virus metagenomic sequencing of stool
-to evaluate the differences in human colorectal cancer virus and
-bacterial community composition. Through random forest modeling we
-identified differences in the healthy and colorectal cancer virome. The
-cancer-associated virome consisted primarily of temperate bacteriophages
-that were also predicted to be bacteria-virus community network hubs.
-These results provide foundational evidence that bacteriophage
-communities are associated with colorectal cancer and potentially impact
-cancer progression by altering the bacterial host communities.
+Viruses are associated with many human cancers, largely due to their
+mutagenic and functionally manipulative abilities. Despite this, cancer
+microbiome studies have almost exclusively focused on bacteria instead
+of viruses. We began evaluating the cancer virome by focusing on
+colorectal cancer, a primary cause of morbidity and mortality throughout
+the world, and a cancer linked to altered colonic bacterial community
+compositions while the virome role remains unknown. We used 16S rRNA
+gene, whole shotgun metagenomic, and purified virus metagenomic
+sequencing of stool to evaluate the differences in human colorectal
+cancer virus and bacterial community composition. Through random forest
+modeling we identified differences in the healthy and colorectal cancer
+virome. The cancer-associated virome consisted primarily of temperate
+bacteriophages that were also predicted to be bacteria-virus community
+network hubs. These results provide foundational evidence that
+bacteriophage communities are associated with colorectal cancer and
+likely impact cancer progression by altering the bacterial host
+communities.
 
@@
 United States and worldwide. Its risk and severity have been linked to
-colonic bacterial community composition. Although human-specific viruses
-have been linked to other cancers and diseases, little is known about
-colorectal cancer virus communities. We addressed this knowledge gap by
-identifying differences in colonic virus communities in the stool of
-colorectal cancer patients and how they compared to bacterial community
+colonic bacterial community composition. Although viruses have been
+linked to other cancers and diseases, little is known about colorectal
+cancer virus communities. We addressed this knowledge gap by identifying
+differences in colonic virus communities in the stool of colorectal
+cancer patients and how they compared to bacterial community
 differences. The results suggested an indirect role for the virome in
@@
 
-The human gut virome is the community of all viruses found in the gut,
-including bacteriophages (viruses that only infect bacteria), eukaryotic
-viruses (viruses that only infect eukaryotic cells), and human-specific
-viruses (viruses that only infect human cells). Due to their mutagenic
-abilities and propensity for functional manipulation, human viruses are
-strongly associated with, and in many cases cause, cancer (1--4).
-Because bacteriophages are crucial for bacterial community stability and
-composition (5--7), and members of those bacterial communities have been
-implicated as oncogenic agents (8--11), bacteriophages have the
-potential to indirectly impact cancer as well. The gut virome therefore
-has a potential to be associated with, and potentially impact, human
-cancer. Altered human virome composition and diversity have already been
-identified in diseases including periodontal disease (12), HIV (13),
-cystic fibrosis (14), antibiotic exposure (15, 16), urinary tract
-infections (17), and inflammatory bowel disease (18). The strong
-association of bacterial communities with colorectal cancer, the
-previous identification of human-specific viruses that cause cancer, and
-the precedent for the virome to impact other human diseases suggest that
-colorectal cancer may be associated with altered virus communities.
+Due to their mutagenic abilities and propensity for functional
+manipulation, human viruses are strongly associated with, and in many
+cases cause, cancer (1--4). Because bacteriophages (i.e.~viruses that
+specifically infect bacteria) are crucial for bacterial community
+stability and composition (5--7) and have been implicated as oncogenic
+agents (8--11), bacteriophages have the potential to indirectly impact
+cancer. The gut virome (i.e.~the virus community of the gut) therefore
+has the potential to impact health and disease. Altered human virome
+composition and diversity have been identified in diseases including
+periodontal disease (12), HIV (13), cystic fibrosis (14), antibiotic
+exposure (15, 16), urinary tract infections (17), and inflammatory bowel
+disease (18). The strong association of bacterial communities with
+colorectal cancer and the precedent for the virome to impact other human
+diseases suggest that colorectal cancer may be associated with altered
+virus communities.
 
@@
 
-Here we begin to address the knowledge gap of whether virus community
-composition is altered in colorectal cancer and, if it is, how those
-differences might impact cancer progression and severity. We also aimed
-to evaluate the virome's potential for use as a diagnostic biomarker.
-The implications of this study are threefold. \emph{First}, this work
+Here we address the knowledge gap of whether virus community composition
+is altered in colorectal cancer and, if it is, how those differences
+might impact cancer progression and severity. We also aimed to evaluate
+the virome's potential for use as a diagnostic biomarker. The
+implications of this study are threefold. \emph{First}, this work
 supports a biological role for the virome in colorectal cancer
@@
 associated microbial communities are involved in the process.
-\emph{Second}, we present a supplementary virus-based approach for
-classification modeling of colorectal cancer using stool samples.
-\emph{Third}, we provide initial support for the importance of studying
-the virome as a component of the microbiome ecological network,
-especially in cancer.
+\emph{Second}, we present a supplementary, or even alternative,
+virus-based approach for classification modeling of colorectal cancer
+using stool samples. \emph{Third}, we provide initial support for the
+importance of studying the virome as a component of the microbiome
+ecological network, especially in cancer.
 
@@
 
-Our study cohort consisted of stool samples collected from 90 human
-subjects, 30 of whom had healthy colons, 30 of whom had adenomas, and 30
-of whom had carcinomas \textbf{(Figure \ref{sampleproc})}. Half of each
-stool sample was used to sequence the bacterial communities using both
-16S rRNA gene and shotgun sequencing techniques. The 16S rRNA gene
-sequencing was performed for a previous study, and the sequences were
-re-analyzed using contemporary methods (8). The other half of each stool
-sample was purified for virus like particles (VLPs) before genomic DNA
-extraction and shotgun metagenomic sequencing. In the VLP purification,
-cells were disrupted and extracellular DNA degraded \textbf{(Figure
+Our study cohort consisted of 90 human subjects, 30 of whom had healthy
+colons, 30 of whom had adenomas, and 30 of whom had carcinomas
+\textbf{(Figure \ref{sampleproc})}. Half of each stool sample was used
+to sequence the bacterial communities using both 16S rRNA gene and
+shotgun sequencing techniques. The 16S rRNA gene sequencing was
+performed for a previous study, and the sequences were re-analyzed using
+contemporary methods (8). The other half of each stool sample was
+purified for virus like particles (VLPs) before genomic DNA extraction
+and shotgun metagenomic sequencing. In the VLP purification, cells were
+disrupted and extracellular DNA degraded \textbf{(Figure
 \ref{sampleproc})} to allow the exclusive analysis of viral DNA within
@@
 Microbiome and disease associations are often described as being of an
-altered diversity (i.e. ``dysbiotic''). Therefore, we first evaluated
+altered diversity (i.e., ``dysbiotic''). Therefore, we first evaluated
 the influence of colorectal cancer on virome OVU diversity. We evaluated
@@
 C-D)}. There was no statistically significant clustering of the disease
-groups (ANOSIM p-value = 0.6, \textbf{Figure \ref{betaogu}}). Notably,
+groups (ANOSIM p-value = 0.4, \textbf{Figure \ref{betaogu}}). Notably,
 there was a significant difference between the few blank controls that
@@
 samples (8, 23, 24) and points to the need for alternate approaches to
-detect the impact of colorectal cancer disease state on these community
-structures.
+detect the impact of colorectal cancer disease state on these
+communities.
 
@@
 for classifying stool samples as originating from individuals with
-healthy, adenomatous, or cancerous colons (8, 23). By using
-classification models instead of attempting to identify single
-differentially abundant OTUs, these and other studies have been
-successful in capturing complex community relationships in which
-differences in taxonomic relative abundance are considered in the
-context of other taxa. The exceptional performance of bacteria in these
-classification models supports a role for bacterial functionality in
-colorectal cancer. We built off of these findings by evaluating the
-ability of virus community signatures to classify stool samples and
-compared their performance to models built using bacterial community
-signatures.
+healthy, adenomatous, or cancerous colons (8, 23). The exceptional
+performance of bacteria in these classification models supports a role
+for bacteria in colorectal cancer. We built off of these findings by
+evaluating the ability of virus community signatures to classify stool
+samples and compared their performance to models built using bacterial
+community signatures.
 
@@
 abundances. The viral model performed as well as the bacterial model
-(corrected p-value = 0.4), with the viral and bacterial models achieving
-mean area under the curve (AUC) values of 0.792 and 0.809, respectively
+(corrected p-value = 0.7), with the viral and bacterial models achieving
+mean area under the curve (AUC) values of 0.767 and 0.772, respectively
 \textbf{(Figure \ref{predmodel} A - B)}. To evaluate the ability of both
@@
 model did not yield a statistically significant performance improvement
-beyond the viral (corrected p-value = 0.4) and bacterial (corrected
-p-value = 0.08) models, yielding an AUC of 0.768 \textbf{(Figure
+beyond the viral (corrected p-value = 0.1) and bacterial (corrected
+p-value = 0.2) models, yielding an AUC of 0.804 \textbf{(Figure
 \ref{predmodel} A - B)}.
 
-We compared viral metagenomic methods to bacterial metagenomic methods
-by building a viral model and a model built using OGU relative abundance
-profiles from bacterial metagenomic shotgun sequencing data. This
-bacterial model performed worse than the other models (mean AUC = 0.474)
-\textbf{(Figure \ref{predmodel} A - B)}. To determine the cause of the
-discrepancy between the two bacterial sequencing methods, we attempted
-to compare the approaches at a common sequencing depth. This revealed
+To determine the advantage of viral metagenomic methods over bacterial
+metagenomic methods, we compared the viral model to a model built using
+OGU relative abundance profiles from bacterial metagenomic shotgun
+sequencing data. This model performed worse than the other models (mean
+AUC = 0.467) \textbf{(Figure \ref{predmodel} A - B)}. Because the
+coverage provided by the metagenomic sequencing was not as deep as the
+equivalent 16S rRNA gene sequencing, we attempted to compare the
+approaches at a common sequencing depth. This investigation revealed
 that the bacterial 16S rRNA gene model was strongly driven by sparse and
@@
 performance, likely because viral genomes are orders of magnitude
-smaller than bacterial genomes.
+smaller than bacterial genomes. Thus, the targeted 16S rRNA gene
+sequencing approach, which represented only a fraction of the bacterial
+metagenomic sequencing depth, was more effective for detecting
+colorectal cancer in stool samples. Despite the recent loss of
+enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic
+techniques, 16S rRNA gene sequencing is still a superior methodological
+approach for some important applications.
 
@@
 The identified viruses were bacteriophages, belonging to
-\emph{Siphoviridae}, \emph{Myoviridae}, and phage taxa who could not be
-confidently idetnfied beyond their broad phage identification (i.e.
-``unclassified''). Many of the important viruses, who were confirmed to
-not have genomic similarity to known bacterial genomes, were
-unidentifiable (denoted ``unknown''). This is common in viromes across
-habitats; studies have reported as much as 95\% of virus sequences
-belonging to unknown genomic units (14, 28--30). When the bacterial and
-viral community signatures were combined, both bacterial and viral
-organisms drove the community association with cancer \textbf{(Figure
-\ref{predmodel} E)}.
+\emph{Siphoviridae}, \emph{Myoviridae}, and ``unclassified'' phage taxa.
+Many of the important viruses were unidentifiable (denoted ``unknown'').
+This is common in viromes across habitats; studies have reported as much
+as 95\% of virus sequences belonging to unknown genomic units (14,
+28--30). When the bacterial and viral community signatures were
+combined, both bacterial and viral organisms drove the community
+association with cancer \textbf{(Figure \ref{predmodel} E)}.
 
@@
 important at different stages of colorectal cancer (8, 20, 22), we
-explored whether shifts in the relative influence of phages could be
-detected between healthy, adenomatous, and cancerous colons. We
-evaluated community shifts between the disease stage transitions
+explored whether shifts in the relative influence of specific phages
+could be detected between healthy, adenomatous, and cancerous colons. We
+evaluated community shifts between the two disease stage transitions
 (healthy to adenomatous and adenomatous to cancerous) by building random
@@
 states, we performed a three-class random forest model including all
-disease states. The 16S rRNA gene model yielded a mean AUC of 0.765 and
+disease states. The 16S rRNA gene model yielded a mean AUC of 0.774 and
 outperformed the viral community model, which yielded a mean AUC of
-0.658 (p-value \textless{} 0.001, \textbf{Figure \ref{threewaymodel}
+0.672 (p-value \textless{} 0.001, \textbf{Figure \ref{threewaymodel}
 A-C}). The microbes important for the healthy versus cancer and healthy
@@
 
-Differences in the colorectal cancer virome could have been driven by
-eukaryotic viruses or by bacteriophages. To better understand the types
-of viruses that were important for colorectal cancer, we identified the
-virome OVUs as being similar to either eukaryotic viruses or
-bacteriophages. The most important viruses to the classification model
-were identified as bacteriophages (\textbf{Figure \ref{threewaymodel})}.
-Overall, we were able to identify 78.8\% of the OVUs as known viruses,
-and 93.8\% of those viral OVUs aligned to bacteriophage reference
-genomes. It is important to note that this could have been influenced by
-our methodological biases against enveloped viruses (more common of
+Differences in the colorectal cancer virome could have been driven
+directly by eukaryotic viruses or indirectly by bacteriophages acting
+through their bacterial hosts. To better understand the types of viruses
+that were important for colorectal cancer, we identified the virome OVUs
+as being similar to either eukaryotic viruses or bacteriophages. The
+most important viruses to the classification model were identified as
+bacteriophages (\textbf{Figure \ref{threewaymodel})}. Overall, we were
+able to identify 78.8\% of the OVUs as known viruses, and 93.8\% of
+those viral OVUs aligned to bacteriophage reference genomes. It is
+important to note that this could have been influenced by our
+methodological biases against enveloped viruses (more common of
 eukaryotic viruses than bacteriophage), due to chloroform and DNase
@@
 remained consistent throughout the healthy, adenomatous, and cancerous
-stages \textbf{(Figure \ref{replicationstyles})}. These findings were
+stages \textbf{(Figure \ref{replicationstyles} E)}. These findings were
 consistent with previous reports suggesting the gut virome is primarily
@@
 Because the link between colorectal cancer and the virome was driven by
-bacteriophages, we hypothesized that the bacteriophage signal may be a
-mere reflection of the bacterial signal, and thus highly correlated with
-the bacterial signal. If this hypothesis was true, we would expect a
-correlation between the relative abundances of influential bacteria and
-phages. Instead, we observed a strikingly low correlation between
-bacterial and phage relative abundances \textbf{(Figure
-\ref{correlations} A,C)}. Overall, there was an absence of correlation
-between the most influential OVUs and bacterial OTUs \textbf{(Figure
-\ref{correlations} B)}. This evidence supported our null hypothesis that
-the influential phages were not primarily reflections of influential
-bacteria, and suggest a more complex relationship between the bacterial
-and phage communities.
+bacteriophages, we hypothesized that the influential phages were
+primarily predators of the influential bacteria, and thus influenced
+their relative abundance through predation. If this hypothesis were
+true, we would expect a correlation between the relative abundances of
+influential bacteria and phages. Instead, we observed a strikingly low
+correlation between bacterial and phage relative abundances
+\textbf{(Figure \ref{correlations} A,C)}. Overall, there was an absence
+of correlation between the most influential OVUs and bacterial OTUs
+\textbf{(Figure \ref{correlations} B)}. This evidence supported our null
+hypothesis that the influential phages were not primarily predators of
+influential bacteria.
 
@@
 features to predict which phages infected which bacteria in the
-community \textbf{(Figure \ref{network} A)} (34). The predicted
-interactions were then used to identify phage community hubs. We
-calculated the alpha centrality (i.e.~measure of importance in the
+community \textbf{(Figure \ref{network} A)} ({\textbf{???}}). The
+predicted interactions were then used to identify phage community hubs.
+We calculated the alpha centrality (i.e.~measure of importance in the
 ecological network) of each phage OVU's connection to the rest of the
@@
 centrality was significantly and positively correlated with importance
-to the disease model (p-value = 0.004, R = 0.176), suggesting that
-phages important in driving colorectal cancer also were more likely to
-be community hubs \textbf{(Figure \ref{network} B)}. Together these
+to the disease model (p-value = 0.02, R = 0.14), suggesting that phages
+important in driving colorectal cancer also were more likely to be
+community hubs \textbf{(Figure \ref{network} B)}. Together these
 findings supported our hypothesis that influential phages were hubs
@@
 Because of their propensity for mutagenesis and capacity for modulating
-their host functionality, many human viruses are oncogenic (1--4). Some
-bacteria also have oncogenic properties, suggesting that bacteriophages,
-a component of the human virome in addition to human-specific viruses,
+their host functionality, many viruses are oncogenic (1--4). Some
+bacteria also have oncogenic properties, suggesting that bacteriophages
 may play an indirect role in promoting carcinogenesis by influencing
@@
 carcinogenic potential and the strong association between bacteria and
-colorectal cancer, a link between virus colorectal communities and
-colorectal cancer has yet to be evaluated. Here we show that, like
-colonic bacterial communities, the colon virome was altered in patients
-with colorectal cancer relative to those with healthy colons. Our
-findings support a working hypothesis for oncogenesis by phage-modulated
-bacterial community composition.
-
-Using our findings here, we have begun a path toward the delineation of
-the role the colonic virome plays in colorectal cancer in the form of a
-working hypothesis that will guide our future studies \textbf{(Figure
-\ref{modelsummary} A)}. We found that basic diversity metrics of alpha
-diversity (richness and Shannon diversity) and beta diversity
-(Bray-Curtis dissimilarity) were insufficient for identifying virome
-community differences between healthy and cancerous states. By
-implementing a machine learning approach (random forest classification),
-we detected strong associations between the colon virus community
-composition and colorectal cancer. The colorectal cancer virome was
-composed primarily of bacteriophages. These phage communities were not
-exclusively predators of the most influential bacteria, as demonstrated
-by the lack of correlation between the abundances of the bacterial and
-phage populations. Instead, we identified influential phages as being
+colorectal cancer, a mechanistic link between virus colorectal
+communities and colorectal cancer has yet to be evaluated. Here we show
+that, like colonic bacterial communities, the colon virome was altered
+in patients with colorectal cancer relative to those with healthy
+colons. Our findings support a working hypothesis for oncogenesis by
+phage-modulated bacterial community composition.
+
+Using our findings here, we have begun to delineate the role the colonic
+virome plays in colorectal cancer in the form of a working hypothesis
+that will guide our future studies \textbf{(Figure \ref{modelsummary}
+A)}. We found that basic diversity metrics of alpha diversity (richness
+and Shannon diversity) and beta diversity (Bray-Curtis dissimilarity)
+were insufficient for identifying virome community differences between
+healthy and cancerous states. By implementing a more sophisticated
+machine learning approach (random forest classification), we detected
+strong associations between the colon virus community composition and
+colorectal cancer. The colorectal cancer virome was composed primarily
+of bacteriophages. These phage communities were not exclusively
+predators of the most influential bacteria, as demonstrated by the lack
+of correlation between the abundances of the bacterial and phage
+populations. Instead, we identified influential phages as being
 community hubs, suggesting phages influence cancer by altering the
@@
 bacterial communities alters colorectal cancer progression and tumor
-burden in mice (10, 20). This provides a precedent for phages indirectly
+burden in mice (10, 20). This provides a precedent for phage indirectly
 influencing colorectal cancer progression by altering the bacterial
@@
 development, we were not able to rule out the role of phages directly
-interacting with the human host (35, 36).
+interacting with the human host (34, 35).
 
@@
 incorporating our findings into the current model for colorectal cancer
-development \textbf{(Figure \ref{modelsummary} B)} (22), although it is
-important to note that there are also many other alternative hypotheses
-by which the system could be operating. We hypothesize that the process
-begins with broadly infectious phages in the colon lysing, and thereby
-disrupting, the existing bacterial communities. This shift opens novel
-niche space that enabled opportunistic bacteria (such as
-\emph{Fusobacterium nucleatum}) to colonize. Once the initial
+development \textbf{(Figure \ref{modelsummary} B)} (22). We hypothesize
+that the process begins with broadly infectious phages in the colon
+lysing and thereby disrupting the existing bacterial communities. This
+shift opens novel niche space that enabled opportunistic bacteria (such
+as \emph{Fusobacterium nucleatum}) to colonize. Once the initial
 influential founder bacteria establish themselves in the epithelium,
@@
 dispersal and growth by lysing bacteria within the biofilm, a process
-important for effective biofilm growth (37). The oncogenic bacteria may
+important for effective biofilm growth (36). The oncogenic bacteria may
 then be able to transform the epithelial cells and disrupt tight
@@
 
-At this time, our working hypothesis represents a conceptualization of
-areas for future work, which will be required to characterize the
-colorectal cancer microbiome at the functional, mechanistic level. There
-are many different ways in which this system can be operating, and our
-working hypothesis is one. For example, it is possible that the
-bacterial communities cause a change in the virome instead of the virome
-altering the bacterial communities. Reviewing every possible mechanism
-and characteristic of the system is beyond the scope of this manuscript,
-but the results we present provide a step toward the answer. Important
-future directions will include larger cohort human studies, further
-\emph{in vitro} and \emph{in vivo} mechanistic experimentation, and even
-attempts at community studies using absolute abundance values instead of
-relative abundance, which would allow for more accurate community
-dynamic modeling.
-
-As with most microbiome studies it is important to note that, despite
-including data for both bacteria and viruses, we are still only studying
-a subset of the microbiome. Here we are building on the
-bacterial-centric literature around colorectal cancer by beginning to
-characterize the virome. The caveat here is that we are technically
-limited to studying only the DNA virome, and therefore are not measuring
-RNA viruses. This will be an important direction for future studies, as
-well as beginning to characterize other additional microbial communities
-such as fungi, archaea, etc.
-
 In addition to the diagnostic ramifications for understanding the
@@
 communities cannot maintain stability and co-evolution without one
-another (6, 38). Not only is the human virome an important element to
+another (6, 37). Not only is the human virome an important element to
 consider in human health and disease (12--18), but our findings support
@@
 
-https://github.com/SchlossLab/Hannigan\_CRCVirome\_mBio\_2018
+https://github.com/SchlossLab/Hannigan\_CRCVirome\_mBio\_2017
 
@@
 co-morbidities including HIV, chronic viral hepatitis, HNPCC, FAP, and
-inflammatory bowel disease. Healthy subjects entered the clinic for the
-study and did not present as a result of co-morbities. Samples were
-collected from four geographic locations: Toronto (Ontario, Canada),
-Boston (Massachusetts, USA), Houston (Texas, USA), and Ann Arbor
-(Michigan, USA). Ninety patients were recruited to the study, thirty of
-which were designated healthy, thirty with detected adenomas, and thirty
-with detected carcinomas.
+inflammatory bowel disease. Samples were collected from four geographic
+locations: Toronto (Ontario, Canada), Boston (Massachusetts, USA),
+Houston (Texas, USA), and Ann Arbor (Michigan, USA). Ninety patients
+were recruited to the study, thirty of which were designated healthy,
+thirty with detected adenomas, and thirty with detected carcinomas.
 
@@
 The 16S rRNA gene sequences were analyzed as described previously,
-relying on the mothur software package (v1.37.0) (39, 40). Briefly, the
-sequences were de-replicated, aligned to the SILVA database (41),
-screened for chimeras using UCHIME (42), and binned into operational
+relying on the mothur software package (v1.37.0) (38, 39). Briefly, the
+sequences were de-replicated, aligned to the SILVA database (40),
+screened for chimeras using UCHIME (41), and binned into operational
 taxonomic units (OTUs) using a 97\% similarity threshold. Abundances
@@
 stool samples, using a modified version of a previously published
-protocol (29, 31, 43, 44). Briefly, an aliquot of stool
+protocol (29, 31, 42, 43). Briefly, an aliquot of stool
 (\textasciitilde{}0.1 g) was resuspended in SM buffer (Crystalgen;
@@
 adapter trimming using the CutAdapt program (v1.9.1) with an error rate
-of 0.1 and an overlap of 10 (45). The FastX toolkit (v0.0.14) was used
+of 0.1 and an overlap of 10 (44). The FastX toolkit (v0.0.14) was used
 to quality trim the reads to a minimum length of 75 bp and a minimum
-quality score of 30 (46). Reads mapping to the human genome were removed
-using the DeconSeq algorithm (v0.4.3) and default parameters (47).
+quality score of 30 (45). Reads mapping to the human genome were removed
+using the DeconSeq algorithm (v0.4.3) and default parameters (46).
 
@@
 each sample using a minimum contig length of 1000 bp and iterating
-assemblies from 21-mers to 101-mers by 20 (48). Contigs from the virus
+assemblies from 21-mers to 101-mers by 20 (47). Contigs from the virus
 and whole metagenomic sample sets were concatenated within their
@@
 using the bowtie2 global aligner (v2.2.1), with a 25 bp seed length and
-an allowance of one mismatch (49). Abundance was corrected for contig
+an allowance of one mismatch (48). Abundance was corrected for contig
 reference length and the number of contigs included in each operational
@@
 co-abundance profiles within samples using a variational Bayesian
-approach (50). CONCOCT was used with a length threshold of 1000 bp for
+approach (49). CONCOCT was used with a length threshold of 1000 bp for
 virus contigs and 2000 bp for bacteria.
@@
 Tukey test. All diversity calculations were performed in R using the
-Vegan package (51).
+Vegan package (50).
 
@@
 
-Classification modeling was performed in R using the Caret package (52).
+Classification modeling was performed in R using the Caret package (51).
 OTU, OVU, and OGU abundance data was preprocessed by removing features
@@
 representative sequence for classification, as described previously
-(53). Each representative sequence was aligned to the reference genome
+(52). Each representative sequence was aligned to the reference genome
 database using the tblastx alignment algorithm (v2.2.27) and a strict
-similarity threshold (e-value \textless{} 1e-25) (54). Annotation was
-interpreted as phage, eukaryotic virus, or unknown. As an additional
-quality control step, these OVUs were also aligned to the bacterial
-reference genome set from the European Nucleotide Archives using the
-blastn algorithm (e-value \textless{} 1e-25) and OVUs with similarity to
-bacterial genomes and not viral genomes were removed from analysis.
+similarity threshold (e-value \textless{} 1e-25) (53). Annotation was
+interpreted as phage, eukaryotic virus, or unknown.
 
@@
 The ecological network of the bacterial and phage operational genomic
-units was constructed and analyzed as previously described (34).
-Briefly, a random forest model was used to predict interactions between
-bacterial and phage genomic units, and those interactions were recorded
-in a graph database using \emph{neo4j} graph databasing software
-(v2.3.1). The degree of phage centrality was quantified using the alpha
-centrality metric in the igraph CRAN package. A Spearman correlation was
-performed between model importance and phage centrality scores.
+units was constructed and analyzed as previously described
+({\textbf{???}}). Briefly, a random forest model was used to predict
+interactions between bacterial and phage genomic units, and those
+interactions were recorded in a graph database using \emph{neo4j} graph
+databasing software (v2.3.1). The degree of phage centrality was
+quantified using the alpha centrality metric in the igraph CRAN package.
+A Spearman correlation was performed between model importance and phage
+centrality scores.
 
@@
 Network for providing the fecal samples that were used in this study.
-The authors also thank the participants in the study cohort.
 
@@
 
-\hypertarget{ref-Hannigan:2018gv}{}
-34. Hannigan GD, Duhaime MB, Koutra D, Schloss PD. 2018. Biogeography
-and environmental conditions shape bacteriophage-bacteria networks
-across the human microbiome. PLoS Computational Biology 14:e1006099.
-
 \hypertarget{ref-Lengeling:2013ia}{}
-35. Lengeling A, Mahajan A, Gally DL. 2013. Bacteriophages as Pathogens
+34. Lengeling A, Mahajan A, Gally DL. 2013. Bacteriophages as Pathogens
 and Immune Modulators? mBio 4:e00868--13--e00868--13.
@@
 \hypertarget{ref-Grski:2012fa}{}
-36. G rski A, Mi dzybrodzki R, Borysowski J, D browska K, Wierzbicki P,
+35. G rski A, Mi dzybrodzki R, Borysowski J, D browska K, Wierzbicki P,
 Ohams M, Korczak-Kowalska G yna, Olszowska-Zaremba N, usiak-Szelachowska
@@
 \hypertarget{ref-Rossmann:2015cj}{}
-37. Rossmann FS, Racek T, Wobser D, Puchalka J, Rabener EM, Reiger M,
+36. Rossmann FS, Racek T, Wobser D, Puchalka J, Rabener EM, Reiger M,
 Hendrickx APA, Diederich A-K, Jung K, Klein C, Huebner J. 2015.
@@
 \hypertarget{ref-Brockhurst:2013iq}{}
-38. Brockhurst MA, Koskella B. 2013. Experimental coevolution of species
+37. Brockhurst MA, Koskella B. 2013. Experimental coevolution of species
 interactions. Trends in ecology \& evolution 28:367--375.
@@
 \hypertarget{ref-Kozich:2013db}{}
-39. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
+38. Kozich JJ, Westcott SL, Baxter NT, Highlander SK, Schloss PD. 2013.
 Development of a dual-index sequencing strategy and curation pipeline
@@
 \hypertarget{ref-Schloss:2009do}{}
-40. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
+39. Schloss PD, Westcott SL, Ryabin T, Hall JR, Hartmann M, Hollister
 EB, Lesniewski RA, Oakley BB, Parks DH, Robinson CJ, Sahl JW, Stres B,
@@
 \hypertarget{ref-Pruesse:2007jc}{}
-41. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J,
+40. Pruesse E, Quast C, Knittel K, Fuchs BM, Ludwig W, Peplies J,
 Glöckner FO. 2007. SILVA: a comprehensive online resource for quality
@@
 \hypertarget{ref-Edgar:2011gy}{}
-42. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. 2011. UCHIME
+41. Edgar RC, Haas BJ, Clemente JC, Quince C, Knight R. 2011. UCHIME
 improves sensitivity and speed of chimera detection. Bioinformatics
@@
 \hypertarget{ref-Thurber:2009dn}{}
-43. Thurber RV, Haynes M, Breitbart M, Wegley L, Rohwer F. 2009.
+42. Thurber RV, Haynes M, Breitbart M, Wegley L, Rohwer F. 2009.
 Laboratory procedures to generate viral metagenomes. Nature protocols
@@
 \hypertarget{ref-Kleiner:2015kd}{}
-44. Kleiner M, Hooper LV, Duerkop BA. 2015. Evaluation of methods to
+43. Kleiner M, Hooper LV, Duerkop BA. 2015. Evaluation of methods to
 purify virus-like particles for metagenomic sequencing of intestinal
@@
 \hypertarget{ref-Martin:2011eu}{}
-45. Martin M. 2011. Cutadapt removes adapter sequences from
+44. Martin M. 2011. Cutadapt removes adapter sequences from
 high-throughput sequencing reads. EMBnetjournal 17:10.
@@
 \hypertarget{ref-FASTXToolkit:wr}{}
-46. Hannon GJ. 2010. FASTX-Toolkit GNU Affero General Public License.
+45. Hannon GJ. 2010. FASTX-Toolkit GNU Affero General Public License.
 
 \hypertarget{ref-Schmieder:2011fo}{}
-47. Schmieder R, Edwards R. 2011. Fast identification and removal of
+46. Schmieder R, Edwards R. 2011. Fast identification and removal of
 sequence contamination from genomic and metagenomic datasets. PLOS ONE
@@
 \hypertarget{ref-Li:2016kd}{}
-48. Li D, Luo R, Liu C-M, Leung C-M, Ting H-F, Sadakane K, Yamashita H,
+47. Li D, Luo R, Liu C-M, Leung C-M, Ting H-F, Sadakane K, Yamashita H,
 Lam T-W. 2016. MEGAHIT v1.0: A fast and scalable metagenome assembler
@@
 \hypertarget{ref-Langmead:2012jh}{}
-49. Langmead B, Salzberg SL. 2012. Fast gapped-read alignment with
+48. Langmead B, Salzberg SL. 2012. Fast gapped-read alignment with
 Bowtie 2. Nature Methods 9:357--359.
@@
 \hypertarget{ref-Alneberg:2014fc}{}
-50. Alneberg J, Bjarnason BS aacute ri, Bruijn I de, Schirmer M, Quick
+49. Alneberg J, Bjarnason BS aacute ri, Bruijn I de, Schirmer M, Quick
 J, Ijaz UZ, Lahti L, Loman NJ, Andersson AF, Quince C. 2014. Binning
@@
 \hypertarget{ref-veganCommunityEco:xdLliqSE}{}
-51. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
+50. Oksanen J, Blanchet FG, Friendly M, Kindt R, Legendre P, McGlinn D,
 Minchin PR, OHara RB, Simpson GL, Solymos P, Stevens MHH, Szoecs E,
@@
 \hypertarget{ref-caretClassificatio:ux5fU2Litux5f1}{}
-52. Kuhn M. 2016. caret: Classification and Regression Training. CRAN.
+51. Kuhn M. 2016. caret: Classification and Regression Training. CRAN.
 
 \hypertarget{ref-Guidi:2016kf}{}
-53. Guidi L, Chaffron S, Bittner L, Eveillard D, Larhlimi A, Roux S,
+52. Guidi L, Chaffron S, Bittner L, Eveillard D, Larhlimi A, Roux S,
 Darzi Y, Audic S, Berline L, Brum JR, Coelho LP, Espinoza JCI, Malviya
@@
 \hypertarget{ref-Camacho:2009fc}{}
-54. Camacho C, Coulouris G, Avagyan V, Ma N, Papadopoulos J, Bealer K,
+53. Camacho C, Coulouris G, Avagyan V, Ma N, Papadopoulos J, Bealer K,
 Madden TL. 2009. BLAST+: architecture and applications. BMC
@@
 \centering
-\includegraphics[width=0.50000\textwidth]{}
-\caption{\emph{Lysogenic phage relative abundance in disease states.
-Phage OVUs were predicted to be either lytic or lysogenic, and the
-relative abundance of lysogenic phages was quantified and represented as
-a boxplot. No disease groups were statistically
-significant.}\label{replicationstyles}}
-\end{figure}
-
-\newpage
-
-\begin{figure}[htbp]
-\centering
 \includegraphics{}
@@
 \centering
-\includegraphics{}
-\caption{\emph{Community network analysis utilizing predicted
-interactions between bacteria and phage operational genomic units. A)
-Visualization of the community network for our colorectal cancer cohort.
-B) Scatter plot illustrating the correlation between importance (mean
-decrease in accuracy) and the degree of centrality for each OVU. A
-linear regression line was fit to illustrate the correlation (blue)
-which was found to be statistically significantly and weakly correlated
-(p-value = 0.00409, R = 0.176).}\label{network}}
+\includegraphics[width=0.50000\textwidth]{}
+\caption{\emph{Lysogenic phage relative abundance in disease states.
+Phage OVUs were predicted to be either lytic or lysogenic, and the
+relative abundance of lysogenic phages was quantified and represented as
+a boxplot. No disease groups were statistically
+significant.}\label{replicationstyles}}
 \end{figure}
@@
 \includegraphics{}
-\caption{\emph{Final working hypothesis from this study. These panels
-summarize our thoughts on our results and represent interesting future
-directions that we predict will build on the presented work. A) Basic
-model illustrating the connections between the virome, bacterial
+\caption{\emph{Final model and working hypothesis from this study. A)
+Basic model illustrating the connections between the virome, bacterial
 communities, and colorectal cancer. B) Working hypothesis of how the
@@
 following quality control including quality score filtering and human
-decontamination. Dashed line shows rarefaction line in which all samples
-with lower sequence yield were excluded from downstream
-analysis.}\label{qualcontrol}}
+decontamination.}\label{qualcontrol}}
 \end{figure}
@@
 \caption{\emph{Beta-diversity comparing disease states and the study
-negative controls which remained following rarefaction. Differences in
-means between disease group centroids with 95\% confidence intervals
-based on an ANOSIM test with a post hoc multivariate Tukey test.
-Comparisons in which the intervals cross the zero mean difference line
-(dashed line) were not significantly different.}\label{betaogunegative}}
+negative controls. Differences in means between disease group centroids
+with 95\% confidence intervals based on an ANOSIM test with a post hoc
+multivariate Tukey test. Comparisons in which the intervals cross the
+zero mean difference line (dashed line) were not significantly
+different.}\label{betaogunegative}}
 \end{figure}
@@
 
+\newpage
+
+\begin{figure}[htbp]
+\centering
+\includegraphics{}
+\caption{\emph{Community network analysis utilizing predicted
+interactions between bacteria and phage operational genomic units. A)
+Visualization of the community network for our colorectal cancer cohort.
+B) Scatter plot illustrating the correlation between importance (mean
+decrease in accuracy) and the degree of centrality for each OVU. A
+linear regression line was fit to illustrate the correlation (blue)
+which was found to be statistically significantly and weakly correlated
+(p-value = 0.0173, R = 0.14).}\label{network}}
+\end{figure}
+
 \end{document}

### 2018-09-30 revfiles/internalComments.rmd

@@
 
-Rewording was also great.
+Rewording was also great, thanks.
 
@@
 
-Thanks for the feedback. I'll also try to make it sound less dodgy.
+Thanks for the feedback. I also tried to make it sound less dodgy.
 
@@
 
-I agree that it should be self evident but this is something that some of my colleagues, who are not familiar with phages, appreciated hearing.
+I agree that it should be self evident but this is something that some of my colleagues, who are not familiar with phages (e.g. clinicians), have appreciated hearing.
 

### 2018-09-30 bin/IdentifyContigs.sh
Filter with more hits

@@ -24,6 +24,7 @@ RunBlast () {
 		-in "${2}" \
 		-out ./data/tmpid/ReferenceGenomes
 
+	# Use a higher max target seqs for filtering below
 	echo Running tblastx...
 	${BlastPath}tblastx \
     	-query "${1}" \
@@ -31,7 +32,7 @@ RunBlast () {
     	-db ./data/tmpid/ReferenceGenomes \
     	-evalue 1e-25 \
     	-num_threads 8 \
-    	-max_target_seqs 1 \
+    	-max_target_seqs 100 \
     	-outfmt 6
 }
 

### 2018-09-30 bin/IdentifyContigs.sh
And some additional fixes downstream.

@@ -72,7 +72,10 @@ else
 fi
 
 # Add cluster ID to the table
-cut -f 1,2 ./data/tmpid/blastout.tsv | sort | uniq > ./data/tmpid/cutblastout.tsv
+sort -k2,2 -k11,11g -k 12,12Vr ./data/tmpid/blastout.tsv \
+	| awk '!seen[substr($1,0,15)]++' \
+	| cut -f 1,2 \
+	> ./data/tmpid/cutblastout.tsv
 sed 's/\,/\t/' ./data/ContigClustersVirus/clustering_gt1000.csv > ./data/tmpid/clusterids.tsv
 awk -F "\t" 'FNR==NR { a[$1] = $2; next } { for( i in a ) if($1 ~ i) {print a[$1]"\t"$2} }' \
 	./data/tmpid/clusterids.tsv \

### 2018-10-11 doc/manuscript.tex
Doc updates including word doc

@@
 abundances. The viral model performed as well as the bacterial model
-(corrected p-value = 0.4), with the viral and bacterial models achieving
-mean area under the curve (AUC) values of 0.792 and 0.809, respectively
+(corrected p-value = 0.6), with the viral and bacterial models achieving
+mean area under the curve (AUC) values of 0.768 and 0.775, respectively
 \textbf{(Figure \ref{predmodel} A - B)}. To evaluate the ability of both
@@
 model did not yield a statistically significant performance improvement
-beyond the viral (corrected p-value = 0.4) and bacterial (corrected
-p-value = 0.08) models, yielding an AUC of 0.768 \textbf{(Figure
+beyond the viral (corrected p-value = 0.08) and bacterial (corrected
+p-value = 0.1) models, yielding an AUC of 0.807 \textbf{(Figure
 \ref{predmodel} A - B)}.
@@
 profiles from bacterial metagenomic shotgun sequencing data. This
-bacterial model performed worse than the other models (mean AUC = 0.474)
+bacterial model performed worse than the other models (mean AUC = 0.458)
 \textbf{(Figure \ref{predmodel} A - B)}. To determine the cause of the
@@
 identified beyond their broad phage identification (i.e.
-``unclassified''). Many of the important viruses, that were confirmed to
-not have genomic similarity to known bacterial genomes, were
-unidentifiable (denoted ``unknown''). This is common in viromes across
-habitats; studies have reported as much as 95\% of virus sequences
-belonging to unknown genomic units (14, 28--30). When the bacterial and
-viral community signatures were combined, both bacterial and viral
-organisms drove the community association with cancer \textbf{(Figure
-\ref{predmodel} E)}.
+``unclassified''). Viruses, which were confirmed to not have genomic
+similarity to known bacterial genomes, were unidentifiable (denoted
+``unknown''). This is common in viromes across habitats; studies have
+reported as much as 95\% of virus sequences belonging to unknown genomic
+units (14, 28--30). When the bacterial and viral community signatures
+were combined, both bacterial and viral organisms drove the community
+association with cancer \textbf{(Figure \ref{predmodel} E)}.
 
@@
 states, we performed a three-class random forest model including all
-disease states. The 16S rRNA gene model yielded a mean AUC of 0.765 and
+disease states. The 16S rRNA gene model yielded a mean AUC of 0.784 and
 outperformed the viral community model, which yielded a mean AUC of
-0.658 (p-value \textless{} 0.001, \textbf{Figure \ref{threewaymodel}
+0.654 (p-value \textless{} 0.001, \textbf{Figure \ref{threewaymodel}
 A-C}). The microbes important for the healthy versus cancer and healthy

### 2018-10-11 bin/ConnectionImportanceCorr.R
Code updates

@@ -18,6 +18,8 @@ library("cowplot")
 library("vegan")
 library("extrafont")
 loadfonts()
+library("parallel")
+
 
 ###################
 # Set Subroutines #
@@ -140,6 +142,16 @@ alphacent$sampleID <- rownames(alphacent)
 input <- read.delim("./data/VirusClusteredContigAbund.tsv", header=TRUE, sep="\t")
 datadisease <- read.delim("./data/metadata/MasterMeta.tsv", header=FALSE, sep="\t")[,c(2,30,22)]
 
+# Remove the bacteria-not-virus clusters
+removaltable <- read.delim("./data/contigclustersidentity/BacteriaNotVirus.tsv", header = FALSE, sep = "\t")
+removaltable$V1 <- gsub("^", "Cluster_", removaltable$V1, perl = TRUE)
+# Clean input
+input <- input[!c(input$V1 %in% removaltable$V1),]
+# Remove the nodes from the network
+removenodevector <- data.frame(V1 = gsub("Cluster_", "Phage_", removaltable$V1, perl = TRUE))
+edgeout <- edgeout[!c(edgeout$from %in% removenodevector$V1),]
+
+
 # Rarefy input table
 minimumsubsample <- 1000000
 inputcast <- dcast(input, V1 ~ V2)
@@ -184,7 +196,7 @@ absmissingid <- abssubset[,-which(names(abssubset) %in% c("V22"))]
 outmodel <- caretmodel(absmissingid)
 outmodel
 
-avgimportance <- lapply(c(1:25), function(i) GetAverageImportance(absmissingid, i))
+avgimportance <- mclapply(c(1:25), mc.cores = 5, function(i) GetAverageImportance(absmissingid, i))
 avgimportance <- ldply(avgimportance, data.frame)
 avgimportance$sampleID <- gsub("Cluster", "Phage", avgimportance$categories)
 davg <- ddply(avgimportance, c("sampleID"), summarize, mean = mean(Overall))

### 2018-10-11 bin/ContigLengthDist.R

@@ -39,6 +39,10 @@ colnames(clusters) <- c("ContigID", "Cluster")
 
 mergeddf <- merge(input, clusters, by = "ContigID")
 
+# Remove the bacteria-not-virus clusters
+removaltable <- read.delim("./data/contigclustersidentity/BacteriaNotVirus.tsv", header = FALSE, sep = "\t")
+mergeddf <- mergeddf[!c(mergeddf$Cluster %in% removaltable$V1),]
+
 lengthscatter <- ggplot(mergeddf, aes(x = Cluster, y = Length)) +
     theme_classic() +
     geom_point() +

### 2018-10-11 bin/ContigLengthHits.R

@@ -57,6 +57,10 @@ mergedclusters <- merge(mergedabund, clusters, by = "ContigID")
 
 clusterstats <- ddply(mergedclusters, "Cluster", summarize, sumcount = sum(sum), avglength = mean(Length))
 
+# Remove the bacteria-not-virus clusters
+removaltable <- read.delim("./data/contigclustersidentity/BacteriaNotVirus.tsv", header = FALSE, sep = "\t")
+clusterstats <- clusterstats[!c(clusterstats$Cluster %in% removaltable$V1),]
+
 virusclusterplot <- ggplot(clusterstats, aes(x=avglength, y=sumcount)) +
     theme_classic() +
     geom_point() +

### 2018-10-11 bin/diversity-beta.R

@@ -13,10 +13,13 @@ library("optparse")
 library("ggplot2")
 library("plotROC")
 library("reshape2")
-library("wesanderson")
+library("viridis")
 library("vegan")
 library("plyr")
 library("cowplot")
+library("tidyr")
+library("dplyr")
+library("wesanderson")
 
 option_list <- list(
   make_option(c("-i", "--input"),
@@ -65,6 +68,28 @@ input <- read.delim("./data/VirusClusteredContigAbund.tsv", header=TRUE, sep="\t
 # Get the metadata table
 datadisease <- read.delim("./data/metadata/MasterMeta.tsv", header=FALSE, sep="\t")[,c(2,30,22)]
 
+# Remove the bacteria-not-virus clusters
+removaltable <- read.delim("./data/contigclustersidentity/BacteriaNotVirus.tsv", header = FALSE, sep = "\t")
+removaltable$V1 <- gsub("^", "Cluster_", removaltable$V1, perl = TRUE)
+# Clean input
+input <- input[!c(input$V1 %in% removaltable$V1),]
+
+# Get corrected aligned sequence count for each
+totalcounts <- input %>%
+    group_by(V2) %>%
+    summarize(total = sum(sum)) %>%
+    as.data.frame()
+
+# Write this to a file
+write.table(
+    file = "./data/ProjectSeqDepth.tsv",
+    x = totalcounts,
+    sep = "\t",
+    quote = FALSE,
+    row.names = FALSE,
+    col.names = TRUE
+    )
+
 # Format for vegan
 inputcast <- dcast(input, V2 ~ V1, value.var = "sum")
 inputcast[is.na(inputcast)] <- 0
@@ -78,8 +103,8 @@ castsum <- rowSums(inputcast)
 
 distlist <- lapply(c(1:25), function(i) {
     print(i)
-    inputcast <- rrarefy(inputcast, sample=100000)
-    inputcast <- inputcast[c(rowSums(inputcast) %in% 100000),]
+    inputcast <- rrarefy(inputcast, sample=1000000)
+    inputcast <- inputcast[c(rowSums(inputcast) %in% 1000000),]
     outdist <- vegdist(inputcast, method="bray", diag = TRUE, upper = TRUE)
     return(outdist)
 })
@@ -88,8 +113,8 @@ castdist <- Reduce("+", distlist) / length(distlist)
 
 distlist2 <- lapply(c(1:25), function(i) {
[…]

### 2018-10-11 bin/phagebacteriacorr.R

@@ -70,6 +70,13 @@ virustax <- read.delim("./data/contigclustersidentity/clustax.tsv", header = FAL
 taxonomy$Taxonomy <- sub(".+\\;(.+)\\(\\d+\\)\\;$", "\\1", taxonomy$Taxonomy, perl=TRUE)
 virustax$V1 <- sub("^", "Cluster_", virustax$V1, perl = TRUE)
 
+# Remove the bacteria-not-virus clusters
+removaltable <- read.delim("./data/contigclustersidentity/BacteriaNotVirus.tsv", header = FALSE, sep = "\t")
+removaltable$V1 <- gsub("^", "Cluster_", removaltable$V1, perl = TRUE)
+# Clean input
+input <- input[!c(input$V1 %in% removaltable$V1),]
+virustax <- virustax[!c(virustax$V1 %in% removaltable$V1),]
+
 # Rarefy input table
 minimumsubsample <- 1000000
 inputcast <- dcast(input, V1 ~ V2)

### 2018-10-11 bin/predictProgression.R

@@ -102,15 +102,15 @@ plotimportance <- function(x, iterationcount = 10, topcount = 10, corecount = 5)
 	importaverage <- merge(avgimportancedf, import, by = "categories")
 
 	write("Getting Virus Tax", stderr())
-	virustax <- virustax[,c(1,3,7)]
+	virustax <- virustax[,c(1,3,6)]
 
 	importaverage <- merge(importaverage, virustax, by.x = "categories", by.y = "V1", all = TRUE)
-	importaverage$V7 <- as.character(importaverage$V7)
+	importaverage$V6 <- as.character(importaverage$V6)
 	importaverage <- importaverage[!c(importaverage$Overall %in% NA),]
 	importaverage[is.na(importaverage)] <- "Unknown"
 	importaverage <- importaverage[order(importaverage$mean, decreasing = FALSE),]
 	importaverage$categories <- factor(importaverage$categories, levels = unique(importaverage$categories))
-	importaverage$V7 <- factor(importaverage$V7)
+	importaverage$V6 <- factor(importaverage$V6)
 	
 	binlength <- c(1:topcount) + 0.5
 
@@ -134,7 +134,7 @@ plotimportance <- function(x, iterationcount = 10, topcount = 10, corecount = 5)
 	  	labels=parse(
 	  		text = paste0(
 	  			"italic('",
-	  			dfplot[c(0:(topcount - 1))*iterationcount+1,"V7"],
+	  			dfplot[c(0:(topcount - 1))*iterationcount+1,"V6"],
 	  			"')~",
 	  			paste(
 	  				" (",
@@ -149,8 +149,6 @@ plotimportance <- function(x, iterationcount = 10, topcount = 10, corecount = 5)
 
 	  return(importanceplot)
 }
-
-hvai <- plotimportance(absmissingid, iterationcount = 3)
 
 pbi <- function(x, iterationcount = 10, topcount = 10) {
 	avgimportance <- lapply(c(1:iterationcount), function(i) GetAverageImportance(x, i))

### 2018-10-11 bin/predictThreeClass.R

@@ -65,15 +65,15 @@ plotimportance <- function(x, iterationcount = 25, topcount = 10, corecount = 5)
 	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall), stder = sd(Overall)/sqrt(length(Overall)))
 	importaverage <- merge(avgimportancedf, import, by = "categories")
 
-	virustaxdf <- virustax[,c(1,3,7)]
+	virustaxdf <- virustax[,c(1,3,6)]
 
 	importaverage <- merge(importaverage, virustaxdf, by.x = "categories", by.y = "V1", all = TRUE)
-	importaverage$V7 <- as.character(importaverage$V7)
+	importaverage$V6 <- as.character(importaverage$V6)
 	importaverage <- importaverage[!c(importaverage$Overall %in% NA),]
 	importaverage[is.na(importaverage)] <- "Unknown"
 	importaverage <- importaverage[order(importaverage$mean, decreasing = FALSE),]
 	importaverage$categories <- factor(importaverage$categories, levels = importaverage$categories)
-	importaverage$V7 <- factor(importaverage$V7, levels = importaverage$V7)
+	importaverage$V6 <- factor(importaverage$V6, levels = importaverage$V6)
 	
 	binlength <- c(1:topcount) + 0.5
 
@@ -113,7 +113,7 @@ plotimportance <- function(x, iterationcount = 25, topcount = 10, corecount = 5)
 			labels=parse(
 				text = paste0(
 					"italic('",
-					dfplot[c(0:(topcount - 1))*iterationcount+1,"V7"],
+					dfplot[c(0:(topcount - 1))*iterationcount+1,"V6"],
 					"')~",
 					paste(
 						" (",
@@ -236,6 +236,13 @@ virustax <- read.delim("./data/contigclustersidentity/clustax.tsv", header = FAL
 virustax$V1 <- sub("^", "Cluster_", virustax$V1, perl = TRUE)
 virustax <- unique(virustax)
 
+# Remove the bacteria-not-virus clusters
+removaltable <- read.delim("./data/contigclustersidentity/BacteriaNotVirus.tsv", header = FALSE, sep = "\t")
+removaltable$V1 <- gsub("^", "Cluster_", removaltable$V1, perl = TRUE)
+# Clean input
[…]

### 2018-10-11 bin/predictVirome.R

@@ -99,15 +99,15 @@ plotimportance <- function(x, iterationcount = 25, topcount = 10, corecount = 5)
 	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall))
 	importaverage <- merge(avgimportancedf, import, by = "categories")
 
-	virustax <- virustax[,c(1,3,7)]
+	virustax <- virustax[,c(1,3,6)]
 
 	importaverage <- merge(importaverage, virustax, by.x = "categories", by.y = "V1", all = TRUE)
-	importaverage$V7 <- as.character(importaverage$V7)
+	importaverage$V6 <- as.character(importaverage$V6)
 	importaverage <- importaverage[!c(importaverage$Overall %in% NA),]
 	importaverage[is.na(importaverage)] <- "Unknown"
 	importaverage <- importaverage[order(importaverage$mean, decreasing = FALSE),]
 	importaverage$categories <- factor(importaverage$categories, levels = importaverage$categories)
-	importaverage$V7 <- factor(importaverage$V7, levels = importaverage$V7)
+	importaverage$V6 <- factor(importaverage$V6, levels = importaverage$V6)
 	
 	binlength <- c(1:topcount) + 0.5
 
@@ -145,7 +145,7 @@ plotimportance <- function(x, iterationcount = 25, topcount = 10, corecount = 5)
 			labels = parse(
 				text = paste0(
 					"italic('",
-					dfplot[c(0:(topcount - 1))*iterationcount+1,"V7"],
+					dfplot[c(0:(topcount - 1))*iterationcount+1,"V6"],
 					"')~",
 					paste(
 						" (",
@@ -213,7 +213,7 @@ pcomb <- function(x, iterationcount = 25, topcount = 10, corecount = 5) {
 	avgimportancedf <- ldply(avgimportance, data.frame)
 	import <- ddply(avgimportancedf, c("categories"), summarize, mean = mean(Overall))
 	importaverage <- merge(avgimportancedf, import, by = "categories")
-	coltax <- virustax[,c(1, 3, 7)]
+	coltax <- virustax[,c(1, 3, 6)]
 	colnames(coltax) <- c("OTU", "Size", "Taxonomy")
 	mtax <- rbind(taxonomy, coltax)
[…]

### 2018-10-11 bin/repsty.R

@@ -20,6 +20,13 @@ input <- read.delim("./data/VirusClusteredContigAbund.tsv", header=TRUE, sep="\t
 tempteratelist <- read.delim("./data/repcycle/repclust.tsv", header = FALSE, sep = "\t")
 datadisease <- read.delim("./data/metadata/MasterMeta.tsv", header=FALSE, sep="\t")[,c(2,30,22)]
 
+# Remove the bacteria-not-virus clusters
+removaltable <- read.delim("./data/contigclustersidentity/BacteriaNotVirus.tsv", header = FALSE, sep = "\t")
+removaltable$V1 <- gsub("^", "Cluster_", removaltable$V1, perl = TRUE)
+# Clean input
+input <- input[!c(input$V1 %in% removaltable$V1),]
+
+
 # Rarefy input table
 minimumsubsample <- 1000000
 inputcast <- dcast(input, V1 ~ V2)

### 2018-10-11 bin/IdentifyContigs.sh
Update shell

@@ -73,7 +73,7 @@ fi
 
 # Add cluster ID to the table
 sort -k2,2 -k11,11g -k 12,12Vr ./data/tmpid/blastout.tsv \
-	| awk '!seen[substr($1,0,15)]++' \
+	| awk '!seen[substr($1,0,35)]++' \
 	| cut -f 1,2 \
 	> ./data/tmpid/cutblastout.tsv
 sed 's/\,/\t/' ./data/ContigClustersVirus/clustering_gt1000.csv > ./data/tmpid/clusterids.tsv
@@ -111,10 +111,13 @@ for i in $(seq 1 8); do
 	awk -v col=$i '{ if($col) print $col; else print "Unclassified" }' ./data/tmpid/taxawnum.tsv > ./data/tmpcter/${i}.tsv
 done
 paste ./data/tmpcter/* > ./data/tmpid/fullcols.tsv
+sort ./data/tmpid/fullcols.tsv | uniq > ./data/tmpid/fullcols_filtered.tsv
 
-awk -F "\t" 'FNR==NR { a[$3] = $1"\t"$2; next } { for( i in a ) if($1 ~ i) {print a[$1]"\t"$0} }' \
-	./data/tmpid/ctax.tsv  \
-	./data/tmpid/fullcols.tsv \
-	> ${outputfile}
+# Get together the taxonomy reference and the contig IDs
+awk -F "\t" 'FNR==NR { a[$1] = $0; next } { {print $1"\t"$2"\t"a[$3]} }' \
+	./data/tmpid/fullcols_filtered.tsv \
+	./data/tmpid/ctax.tsv \
+	| cut -f -2,4- \
+	> ./data/contigclustersidentity/clustax.tsv
 
 
